---
title: "Chapter 2: R Fundamentals and Data Structures"
---

# R Fundamentals and Data Structures

## 2.1 Data Types in R: The Building Blocks

Every value in R has a data type. Understanding these types is essential because R's functions behave differently depending on the type of data they receive. In clinical research, we work primarily with six data types.

### Numeric (Real Numbers)

Numeric values are decimal numbers used for measurements, lab values, and continuous outcomes.

```{r}
#| label: numeric-data-type
#| eval: false

# Define a numeric value (systolic blood pressure)
systolic_bp <- 145.3

# Verify it is numeric
class(systolic_bp)
# [1] "numeric"

# R stores this as a floating-point number (double precision)
# It can be used in arithmetic operations
diastolic_bp <- 92.5
mean_arterial_pressure <- (systolic_bp + 2 * diastolic_bp) / 3
print(mean_arterial_pressure)
# [1] 110.1
```

In clinical research, numeric data represents:
- **Lab values**: Hemoglobin (g/dL), white blood cells (K/uL), glucose (mg/dL)
- **Vital signs**: Blood pressure (mmHg), heart rate (bpm), temperature (°C or °F)
- **Efficacy measures**: Tumor size (cm), FEV1 (liters), ejection fraction (%)
- **Pharmacokinetic parameters**: AUC, Cmax, clearance

::: {.callout-tip}
**Numeric Precision**: R uses double-precision floating-point arithmetic, which can introduce small rounding errors. In clinical data, these are negligible (differences at 16 decimal places), but when comparing floating-point numbers for exact equality, use `all.equal()` instead of `==`.
:::

### Integer

Integers are whole numbers without decimal places. They occupy less memory than numeric values but are less commonly needed in clinical analysis.

```{r}
#| label: integer-data-type
#| eval: false

# Define an integer explicitly using the L suffix
subject_id <- 101L
visit_number <- 3L
number_of_ae <- 2L

class(subject_id)
# [1] "integer"

# Count data is naturally integer:
# - Number of adverse events per subject
# - Number of protocol violations
# - Number of hospitalizations

# If you forget the L suffix, R treats it as numeric (not integer)
subject_id_numeric <- 101
class(subject_id_numeric)
# [1] "numeric"
```

### Character (Text Strings)

Character data represents text: names, identifiers, categorical values, and descriptive information.

```{r}
#| label: character-data-type
#| eval: false

# Character strings use double or single quotes
subject_id <- "001-101"
gender <- "Female"
adverse_event <- "Grade 3 Rash"

class(subject_id)
# [1] "character"

# In clinical data, character variables include:
# - Subject IDs: "001-101", "001-102"
# - Treatment groups: "Active", "Placebo"
# - Event terms: "Nausea", "Headache", "Elevated ALT"
# - Adverse event severity: "Mild", "Moderate", "Severe"
# - Lab units: "g/dL", "K/uL", "%"

# Character vectors can be concatenated
patient_label <- paste("Subject", subject_id, "is", gender)
print(patient_label)
# [1] "Subject 001-101 is Female"
```

### Logical (Boolean)

Logical values are TRUE or FALSE. They result from comparisons and conditional tests.

```{r}
#| label: logical-data-type
#| eval: false

# Logical values from comparisons
glucose <- 165  # mg/dL
is_hyperglycemic <- glucose > 125
print(is_hyperglycemic)
# [1] TRUE

# In clinical research, logical flags indicate:
# - Abnormal result: is_abnormal = TRUE/FALSE
# - Safety flag: has_ae = TRUE/FALSE
# - Protocol violation: violated_inclusion = TRUE/FALSE
# - Baseline status: at_baseline = TRUE/FALSE

# Logical arithmetic (TRUE = 1, FALSE = 0)
num_adverse_events <- 3
ae_flag <- num_adverse_events > 0
print(ae_flag)
# [1] TRUE

# Sum of logical vector counts TRUE values
safety_flags <- c(TRUE, FALSE, TRUE, FALSE, TRUE)
count_flags <- sum(safety_flags)
print(count_flags)
# [1] 3
```

### Factor (Categorical Data with Levels)

Factors are categorical variables where levels have a specific order or meaning. This is crucial for clinical data.

```{r}
#| label: factor-data-type
#| eval: false

# Define a factor: treatment group with two levels
treatment <- factor(c("Active", "Placebo", "Active", "Placebo"),
                    levels = c("Placebo", "Active"))

class(treatment)
# [1] "factor"

# Display the factor
print(treatment)
# [1] Active  Placebo Active  Placebo
# Levels: Placebo Active

# Note: The levels are ordered (Placebo first, Active second)
# This ordering matters for statistical models and visualizations

# Clinical factors include:
# - Treatment group: Levels = c("Control", "Low Dose", "High Dose")
# - Adverse event severity: Levels = c("Mild", "Moderate", "Severe", "Life-threatening")
# - Visit number: Levels = c("Screening", "Baseline", "Week 4", "Week 8", "Week 12")
# - Gender: Levels = c("Male", "Female")

# When you analyze factors in statistics, the reference level matters
# Example: A logistic regression with treatment as a factor
# will compare each level against the reference (first level)

# Ordered factors (for ordinal data like severity)
severity <- ordered(c("Mild", "Moderate", "Severe"),
                    levels = c("Mild", "Moderate", "Severe"))

print(severity)
# [1] Mild     Moderate Severe
# Levels: Mild < Moderate < Severe
```

::: {.callout-important}
**Factors vs. Characters**: If a variable is categorical (e.g., treatment group, gender, adverse event grade), use `factor()`, not character. Factors enable proper statistical modeling and prevent accidental misspellings (R will warn if you try to assign a level that doesn't exist).
:::

### Date and Date-Time

Dates are specialized data types that handle temporal data correctly.

```{r}
#| label: date-data-type
#| eval: false

# Create a date using as.Date() with a standard format
screening_date <- as.Date("2024-03-15", format = "%Y-%m-%d")
baseline_date <- as.Date("2024-03-22")

class(screening_date)
# [1] "Date"

# Dates allow temporal arithmetic
days_to_baseline <- baseline_date - screening_date
print(days_to_baseline)
# Time difference of 7 days

# The lubridate package provides convenient date functions
library(lubridate)

# Parse dates flexibly
enrollment_date <- ymd("2024-03-22")  # Year-Month-Day
birth_date <- mdy("01/15/1970")        # Month-Day-Year
lab_datetime <- dmy_hm("22/03/2024 14:30")  # Day-Month-Year Hour:Minute

# Extract components
age_at_enrollment <- year(enrollment_date) - year(birth_date)
print(age_at_enrollment)
# [1] 54
```

In clinical research, dates are essential for:
- **Trial timeline**: Screening, baseline, treatment, follow-up dates
- **Adverse event reporting**: Date of onset, resolution, reporting
- **Medication history**: Start/stop dates for concomitant medications
- **Lab test dates**: Comparing results across visits

## 2.2 Data Structures: Vectors, Matrices, Lists, and Data Frames

### Vectors: The Foundation

A vector is a sequence of values of the same type. Vectors are the fundamental building block of R.

```{r}
#| label: vectors-intro
#| eval: false

# Numeric vector: hemoglobin values for 5 subjects
hemoglobin <- c(13.5, 12.8, 14.2, 11.9, 13.1)
print(hemoglobin)
# [1] 13.5 12.8 14.2 11.9 13.1

# Character vector: treatment groups
treatment_groups <- c("Active", "Placebo", "Active", "Placebo", "Active")
print(treatment_groups)
# [1] "Active"  "Placebo" "Active"  "Placebo" "Active"

# Logical vector: abnormality flags
is_abnormal <- c(TRUE, TRUE, FALSE, TRUE, FALSE)
print(is_abnormal)
# [1]  TRUE  TRUE FALSE  TRUE FALSE

# Access individual elements using [index]
first_hemoglobin <- hemoglobin[1]
print(first_hemoglobin)
# [1] 13.5

# Access multiple elements
first_three <- hemoglobin[1:3]
print(first_three)
# [1] 13.5 12.8 14.2

# Conditional indexing: get values > 13
high_hemoglobin <- hemoglobin[hemoglobin > 13]
print(high_hemoglobin)
# [1] 13.5 14.2 13.1
```

### Named Vectors

Named vectors associate each element with a meaningful label.

```{r}
#| label: named-vectors
#| eval: false

# Create a vector of baseline vital signs
vitals <- c(
  systolic_bp = 145,
  diastolic_bp = 92,
  heart_rate = 78,
  temperature = 36.8
)

print(vitals)
# systolic_bp diastolic_bp   heart_rate temperature
#         145           92           78        36.8

# Access by name
hr <- vitals["heart_rate"]
print(hr)
# heart_rate
#         78

# In clinical data, named vectors represent:
# - Subject-level baseline measurements
# - Protocol-specified reference values
# - Adverse event counts by type
```

### Matrices: Two-Dimensional Arrays

A matrix is a two-dimensional collection of values (all the same type).

```{r}
#| label: matrices
#| eval: false

# Create a matrix: 5 subjects, 3 lab measurements
lab_values <- matrix(
  c(13.5, 12.8, 14.2, 11.9, 13.1,  # Hemoglobin (g/dL)
    7.2, 6.1, 7.5, 5.8, 6.9,        # WBC (K/uL)
    95, 88, 102, 85, 98),           # Glucose (mg/dL)
  nrow = 5,
  ncol = 3,
  byrow = FALSE
)

# Add informative row and column names
rownames(lab_values) <- paste0("Subject_", 1:5)
colnames(lab_values) <- c("Hemoglobin", "WBC", "Glucose")

print(lab_values)
#         Hemoglobin WBC Glucose
# Subject_1       13.5 7.2      95
# Subject_2       12.8 6.1      88
# ...

# Access elements
hgb_subject_2 <- lab_values[2, "Hemoglobin"]
print(hgb_subject_2)
# [1] 12.8

# Extract a column
hemoglobin_values <- lab_values[, "Hemoglobin"]
print(hemoglobin_values)
# Subject_1 Subject_2 Subject_3 Subject_4 Subject_5
#      13.5      12.8      14.2      11.9      13.1
```

::: {.callout-note}
**Matrices vs. Data Frames**: Matrices are homogeneous (all elements same type) and are useful for mathematical operations. Data frames (next section) are heterogeneous (mixed types) and are the standard for clinical datasets. In practice, you'll use data frames far more often than matrices.
:::

### Lists: Flexible Collections

A list can contain elements of different types and structures. Lists are extremely flexible.

```{r}
#| label: lists
#| eval: false

# Create a list with diverse content
subject_data <- list(
  id = "001-101",
  age = 62,
  gender = "Female",
  baseline_labs = c(hemoglobin = 13.5, wbc = 7.2, glucose = 95),
  adverse_events = c("Nausea", "Headache"),
  completed_trial = TRUE
)

print(subject_data)
# $id
# [1] "001-101"
#
# $age
# [1] 62
#
# $gender
# [1] "Female"
#
# $baseline_labs
# hemoglobin       wbc    glucose
#       13.5       7.2       95.0
#
# $adverse_events
# [1] "Nausea"   "Headache"
#
# $completed_trial
# [1] TRUE

# Access list elements using $ or [[]]
subject_id <- subject_data$id
print(subject_id)
# [1] "001-101"

# Access by index
age <- subject_data[[2]]
print(age)
# [1] 62

# Lists are used for:
# - Storing mixed-type data for a single subject
# - Returning multiple outputs from a function
# - Nested data structures (list of lists)
```

### Data Frames: The Standard for Tabular Data

A data frame is a list of vectors (columns) of equal length. This is the standard structure for clinical datasets.

```{r}
#| label: data-frames
#| eval: false

# Create a data frame: 5 subjects with demographics and baseline labs
clinical_data <- data.frame(
  USUBJID = c("001-101", "001-102", "001-103", "001-104", "001-105"),
  AGE = c(62, 32, 39, 56, 31),
  SEX = factor(c("F", "M", "F", "F", "M"), levels = c("F", "M")),
  TREATMENT = factor(c("Active", "Placebo", "Active", "Placebo", "Active")),
  BASELINE_HGB = c(13.5, 12.8, 14.2, 11.9, 13.1),
  BASELINE_GLU = c(95, 88, 102, 85, 98)
)

print(clinical_data)
#    USUBJID AGE SEX TREATMENT BASELINE_HGB BASELINE_GLU
# 1 001-101  62   F    Active         13.5           95
# 2 001-102  32   M   Placebo         12.8           88
# 3 001-103  39   F    Active         14.2          102
# 4 001-104  56   F   Placebo         11.9           85
# 5 001-105  31   M    Active         13.1           98

# Check dimensions
dim(clinical_data)
# [1] 5 6

# Access columns
age_vector <- clinical_data$AGE
print(age_vector)
# [1] 62 32 39 56 31

# Or by index
treatment_vector <- clinical_data[, 4]
print(treatment_vector)
# [1] Active  Placebo Active  Placebo Active
# Levels: Active Placebo

# Access rows
subject_102_data <- clinical_data[2, ]
print(subject_102_data)
#    USUBJID AGE SEX TREATMENT BASELINE_HGB BASELINE_GLU
# 2 001-102  32   M   Placebo         12.8           88

# Access a specific cell
age_of_subject_103 <- clinical_data[3, "AGE"]
print(age_of_subject_103)
# [1] 39
```

### Tibbles: The Modern Data Frame

Tibbles (from the {tibble} package, part of tidyverse) are enhanced data frames with better printing and behavior.

```{r}
#| label: tibbles
#| eval: false

library(tibble)

# Create a tibble
clinical_tbl <- tibble(
  USUBJID = c("001-101", "001-102", "001-103", "001-104", "001-105"),
  AGE = c(62, 32, 39, 56, 31),
  SEX = factor(c("F", "M", "F", "F", "M")),
  TREATMENT = factor(c("Active", "Placebo", "Active", "Placebo", "Active")),
  BASELINE_HGB = c(13.5, 12.8, 14.2, 11.9, 13.1)
)

# Printing a tibble shows:
# - Column types (e.g., <chr>, <dbl>, <fct>)
# - Only as many rows as fit on screen
# - Only as many columns as fit on screen
print(clinical_tbl)

# # A tibble: 5 × 5
#   USUBJID   AGE SEX   TREATMENT BASELINE_HGB
#   <chr>   <dbl> <fct> <fct>            <dbl>
# 1 001-101    62 F     Active            13.5
# 2 001-102    32 M     Placebo           12.8
# ...

# Tibbles are stricter about subsetting (which prevents accidental bugs)
# They're the default output of many tidyverse functions
```

## 2.3 Operators: Arithmetic, Comparison, and Logical

Operators allow us to perform calculations, comparisons, and logical tests on data.

### Arithmetic Operators

```{r}
#| label: arithmetic-operators
#| eval: false

# Addition
total_dose <- 100 + 50
print(total_dose)
# [1] 150

# Subtraction
weight_change <- 85 - 82
print(weight_change)
# [1] 3

# Multiplication
total_dose <- 10 * 5  # 10 subjects, 5 mg each
print(total_dose)
# [1] 50

# Division
average_dose <- 50 / 5
print(average_dose)
# [1] 10

# Exponentiation
squared <- 5 ^ 2
print(squared)
# [1] 25

# Modulo (remainder)
remainder <- 17 %% 5
print(remainder)
# [1] 2

# Integer division
quotient <- 17 %/% 5
print(quotient)
# [1] 3

# These operators work on vectors element-wise
weights <- c(70, 85, 92, 78, 81)
weights_in_pounds <- weights * 2.20462
print(weights_in_pounds)
# [1] 154.32 187.39 202.86 171.96 178.57
```

### Comparison Operators

Comparison operators return logical (TRUE/FALSE) values. They're essential for filtering clinical data.

```{r}
#| label: comparison-operators
#| eval: false

# Equal to
glucose <- 125
is_equal_threshold <- glucose == 125
print(is_equal_threshold)
# [1] TRUE

# Not equal to
is_hyperglycemic <- glucose != 100
print(is_hyperglycemic)
# [1] TRUE

# Less than
is_low <- glucose < 70
print(is_low)
# [1] FALSE

# Greater than
is_elevated <- glucose > 100
print(is_elevated)
# [1] TRUE

# Less than or equal to
is_acceptable <- glucose <= 140
print(is_acceptable)
# [1] TRUE

# Greater than or equal to
is_high_normal <- glucose >= 100
print(is_high_normal)
# [1] TRUE

# Clinical application: flag abnormal labs
hemoglobin_values <- c(13.5, 12.8, 14.2, 11.9, 13.1)
is_abnormal <- hemoglobin_values < 12 | hemoglobin_values > 14
print(is_abnormal)
# [1] FALSE FALSE  TRUE  TRUE FALSE
```

### Logical Operators

Logical operators combine logical conditions.

```{r}
#| label: logical-operators
#| eval: false

# AND (&): Both conditions must be TRUE
age <- 45
glucose <- 125
is_middle_aged <- age > 40
is_hyperglycemic <- glucose > 100
is_target_population <- is_middle_aged & is_hyperglycemic
print(is_target_population)
# [1] TRUE

# OR (|): At least one condition must be TRUE
has_hypertension <- FALSE
has_diabetes <- TRUE
requires_medication <- has_hypertension | has_diabetes
print(requires_medication)
# [1] TRUE

# NOT (!): Inverts the condition
on_active_treatment <- TRUE
on_placebo <- !on_active_treatment
print(on_placebo)
# [1] FALSE

# Complex conditions: Clinical inclusion criteria
age <- 35
bmi <- 28
glucose <- 110

meets_criteria <- (age >= 18 & age <= 65) &
                  (bmi >= 20 & bmi < 40) &
                  (glucose >= 100 & glucose <= 200)

print(meets_criteria)
# [1] TRUE
```

## 2.4 Control Flow: Decisions and Loops

### if/else Statements

```{r}
#| label: if-else-statements
#| eval: false

# Simple if statement
hemoglobin <- 12.5

if (hemoglobin < 12) {
  status <- "Anemic"
} else if (hemoglobin < 13.5) {
  status <- "Low-normal"
} else {
  status <- "Normal"
}

print(status)
# [1] "Low-normal"

# Clinical application: Flag abnormal lab values based on CTCAE criteria
wbc <- 2.5  # White blood cell count in K/uL

if (wbc >= 3) {
  grade <- 0
} else if (wbc >= 2) {
  grade <- 1
} else if (wbc >= 1) {
  grade <- 2
} else if (wbc > 0.5) {
  grade <- 3
} else {
  grade <- 4
}

print(paste("WBC Grade:", grade))
# [1] "WBC Grade: 1"

# Vectorized if/else: ifelse() for entire vectors
hemoglobin_values <- c(13.5, 12.8, 14.2, 11.9, 13.1)
abnormality <- ifelse(hemoglobin_values < 12 | hemoglobin_values > 14,
                      "Abnormal", "Normal")
print(abnormality)
# [1] "Normal"  "Normal"  "Abnormal" "Abnormal" "Normal"
```

### for Loops

For loops iterate a fixed number of times, processing each element.

```{r}
#| label: for-loops
#| eval: false

# Simple loop: Calculate BMI for each subject
heights_m <- c(1.75, 1.68, 1.82, 1.70, 1.65)  # meters
weights_kg <- c(70, 65, 82, 75, 60)

bmi_values <- numeric(length = 5)  # Create empty vector

for (i in 1:5) {
  bmi_values[i] <- weights_kg[i] / (heights_m[i] ^ 2)
}

print(bmi_values)
# [1] 22.86 23.03 24.68 25.95 22.04

# Loop over a vector directly
subject_ids <- c("001-101", "001-102", "001-103")

for (id in subject_ids) {
  cat("Processing subject", id, "\n")
}
# Processing subject 001-101
# Processing subject 001-102
# Processing subject 001-103

# Clinical application: Process multiple domains
domains <- c("DM", "AE", "LB")

for (domain in domains) {
  filename <- paste0(domain, ".csv")
  cat("Would load:", filename, "\n")
}
# Would load: DM.csv
# Would load: AE.csv
# Would load: LB.csv
```

::: {.callout-tip}
**Vectorization > Loops**: In R, vectorized operations are much faster than loops. When possible, use built-in functions that work on entire vectors rather than explicit loops. We'll see this in detail in later chapters with dplyr.
:::

### while Loops

While loops continue as long as a condition is TRUE.

```{r}
#| label: while-loops
#| eval: false

# Simple while loop
count <- 0
while (count < 5) {
  cat("Count:", count, "\n")
  count <- count + 1
}
# Count: 0
# Count: 1
# Count: 2
# Count: 3
# Count: 4

# Clinical application: Iterate until convergence
# (simplified example of iterative algorithm)
parameter <- 10
iteration <- 0
tolerance <- 0.01

while (abs(parameter - 9.5) > tolerance & iteration < 100) {
  parameter <- parameter - (parameter - 9.5) / 10
  iteration <- iteration + 1
}

cat("Converged to", parameter, "after", iteration, "iterations\n")
# Converged to 9.5 after 29 iterations
```

## 2.5 Functions: Creating Reusable Code

Functions encapsulate code logic for reuse. In clinical research, functions standardize calculations across datasets.

### Creating a Function

```{r}
#| label: function-creation
#| eval: false

# Function to calculate BMI
calculate_bmi <- function(weight_kg, height_m) {
  # Input validation
  if (weight_kg <= 0 | height_m <= 0) {
    stop("Weight and height must be positive values")
  }

  # Calculate BMI
  bmi <- weight_kg / (height_m ^ 2)

  # Return the result
  return(bmi)
}

# Use the function
subject_1_bmi <- calculate_bmi(weight_kg = 70, height_m = 1.75)
print(subject_1_bmi)
# [1] 22.86

# Function with multiple subjects
weights <- c(70, 65, 82, 75, 60)
heights <- c(1.75, 1.68, 1.82, 1.70, 1.65)
bmi_all <- mapply(calculate_bmi, weights, heights)
print(bmi_all)
# [1] 22.86 23.03 24.68 25.95 22.04
```

### Functions with Multiple Return Values

```{r}
#| label: function-multiple-returns
#| eval: false

# Function to classify lab abnormality by CTCAE
classify_wbc_grade <- function(wbc_value) {
  # WBC reference: 3.0-11.0 K/uL
  # CTCAE grades: 0=normal, 1=mild, 2=moderate, 3=severe, 4=life-threatening

  if (wbc_value >= 3.0) {
    grade <- 0
    severity <- "Normal"
  } else if (wbc_value >= 2.0) {
    grade <- 1
    severity <- "Mild"
  } else if (wbc_value >= 1.0) {
    grade <- 2
    severity <- "Moderate"
  } else if (wbc_value > 0.5) {
    grade <- 3
    severity <- "Severe"
  } else {
    grade <- 4
    severity <- "Life-threatening"
  }

  # Return a list with multiple values
  return(list(
    grade = grade,
    severity = severity,
    value = wbc_value
  ))
}

# Use the function
result <- classify_wbc_grade(2.1)
print(result$grade)
# [1] 1
print(result$severity)
# [1] "Mild"
```

### Function Scope

Variables defined inside a function are local (exist only within that function). This prevents namespace pollution.

```{r}
#| label: function-scope
#| eval: false

# Define a global variable
patient_age <- 50

calculate_dose <- function(weight_kg) {
  # Local variable: only exists within this function
  dose <- weight_kg * 10

  # Can access global variables
  cat("Patient age:", patient_age, "\n")

  return(dose)
}

dose_for_70kg <- calculate_dose(70)
print(dose_for_70kg)
# Patient age: 50
# [1] 700

# Local variable 'dose' doesn't exist outside the function
print(exists("dose"))
# [1] FALSE

# But function name and global variable still exist
print(exists("calculate_dose"))
# [1] TRUE
print(exists("patient_age"))
# [1] TRUE
```

## 2.6 The apply Family: Efficient Iteration

The apply family of functions offers vectorized alternatives to loops.

```{r}
#| label: apply-family
#| eval: false

# Create a sample clinical dataset (matrix form)
lab_matrix <- matrix(
  c(13.5, 12.8, 14.2, 11.9, 13.1,  # Hemoglobin
    7.2, 6.1, 7.5, 5.8, 6.9,        # WBC
    95, 88, 102, 85, 98),           # Glucose
  nrow = 5,
  ncol = 3,
  byrow = FALSE,
  dimnames = list(paste0("S", 1:5), c("HGB", "WBC", "GLU"))
)

# apply() on rows (margin = 1): Calculate mean for each subject
mean_per_subject <- apply(lab_matrix, 1, mean)
print(mean_per_subject)
# S1        S2        S3        S4        S5
# 38.57     35.30     41.23     33.93     39.33

# apply() on columns (margin = 2): Calculate mean for each lab
mean_per_lab <- apply(lab_matrix, 2, mean)
print(mean_per_lab)
# HGB       WBC       GLU
# 13.10     6.70      93.60

# lapply(): Apply function to list elements, returns a list
lab_list <- list(
  hemoglobin = c(13.5, 12.8, 14.2),
  wbc = c(7.2, 6.1, 7.5),
  glucose = c(95, 88, 102)
)

means_list <- lapply(lab_list, mean)
print(means_list)
# $hemoglobin
# [1] 13.50
#
# $wbc
# [1] 6.93
#
# $glucose
# [1] 95

# sapply(): Simplify results to a vector
means_vector <- sapply(lab_list, mean)
print(means_vector)
# hemoglobin       wbc    glucose
#       13.50      6.93      95.00
```

## 2.7 Creating a Complete Clinical Patient Record

Let's integrate everything: data types, structures, operators, and functions.

```{r}
#| label: complete-patient-record
#| eval: false

library(lubridate)

# Create a function to build a patient record
create_patient_record <- function(subject_id, age, sex, treatment,
                                  baseline_weight, baseline_height,
                                  baseline_glucose) {

  # Calculate baseline BMI
  baseline_bmi <- baseline_weight / (baseline_height ^ 2)

  # Determine gender code
  sex_code <- ifelse(sex == "Female", "F", "M")

  # Create a patient record (list)
  patient <- list(
    demographics = data.frame(
      USUBJID = subject_id,
      AGE = age,
      SEX = factor(sex_code, levels = c("F", "M")),
      TREATMENT = factor(treatment, levels = c("Placebo", "Active")),
      stringsAsFactors = FALSE
    ),
    baseline_vitals = data.frame(
      WEIGHT_KG = baseline_weight,
      HEIGHT_M = baseline_height,
      BMI = baseline_bmi,
      GLUCOSE_MGDL = baseline_glucose
    ),
    status = list(
      enrolled = TRUE,
      enrollment_date = Sys.Date(),
      completed = NA
    )
  )

  return(patient)
}

# Create records for 3 subjects
subjects <- list()

subjects[[1]] <- create_patient_record(
  subject_id = "001-101",
  age = 62,
  sex = "Female",
  treatment = "Active",
  baseline_weight = 70,
  baseline_height = 1.65,
  baseline_glucose = 95
)

subjects[[2]] <- create_patient_record(
  subject_id = "001-102",
  age = 45,
  sex = "Male",
  treatment = "Placebo",
  baseline_weight = 85,
  baseline_height = 1.78,
  baseline_glucose = 102
)

subjects[[3]] <- create_patient_record(
  subject_id = "001-103",
  age = 55,
  sex = "Female",
  treatment = "Active",
  baseline_weight = 68,
  baseline_height = 1.62,
  baseline_glucose = 98
)

# Access data from the first subject
print(subjects[[1]]$demographics)
#   USUBJID AGE SEX TREATMENT
# 1 001-101  62   F    Active

print(subjects[[1]]$baseline_vitals)
#   WEIGHT_KG HEIGHT_M      BMI GLUCOSE_MGDL
# 1        70     1.65 25.67761            95

# Extract all treatment assignments
treatments <- sapply(subjects, function(x) x$demographics$TREATMENT)
print(treatments)
# [1] Active  Placebo Active
```

## 2.8 Best Practices for Clinical R Code

1. **Use meaningful variable names**: `baseline_hemoglobin`, not `x` or `hgb1`
2. **Validate inputs**: Check that ages are positive, weights are reasonable
3. **Use factors for categorical data**: Treatment, gender, adverse event grades
4. **Handle missing data explicitly**: `is.na()` before calculations
5. **Comment your code**: Especially CDISC variable names and clinical interpretations
6. **Use functions for repeated operations**: Standardizes calculations across datasets
7. **Test your code**: Run on known examples before production use

---

## Exercises

### Beginner Exercises

**Exercise 2.1: Working with Data Types**

Create the following variables and verify their types:
- A numeric vector of five systolic blood pressure values (e.g., 145, 152, 138, 148, 141)
- A character vector of subject IDs (e.g., "001-101", "001-102", etc.)
- A logical vector indicating whether each subject is on active treatment (TRUE/FALSE)
- A factor representing adverse event severity (Mild, Moderate, Severe)

Use `class()` to verify each variable's type.

**Exercise 2.2: Creating and Manipulating a Clinical Data Frame**

Create a data frame with 10 subjects containing:
- USUBJID: Subject IDs from "001-101" to "001-110"
- AGE: Ages ranging from 30 to 70
- WEIGHT_KG: Weights from 50 to 95 kg
- TREATMENT: Randomly assign "Active" or "Placebo" (Hint: use `sample()`)

Calculate:
a) The mean age of all subjects
b) The mean weight of subjects on active treatment
c) How many subjects are over 50 years old

**Exercise 2.3: Using Logical Operators**

Using the data frame from Exercise 2.2:
- Create a logical vector indicating subjects who are BOTH over 50 years old AND on active treatment
- Create a logical vector indicating subjects who are EITHER over 65 years old OR weigh more than 85 kg
- Count how many subjects meet each criterion

### Advanced Exercises

**Exercise 2.4: Creating a Clinical Utility Function**

Write a function `classify_bmi()` that:
- Takes weight (kg) and height (m) as inputs
- Calculates BMI
- Classifies BMI as: "Underweight" (< 18.5), "Normal" (18.5–24.9), "Overweight" (25–29.9), or "Obese" (≥ 30)
- Returns a named list with bmi value and classification

Test the function with at least 5 different weight/height combinations.

**Exercise 2.5: Simulating Trial Enrollment**

Create a function `simulate_trial_enrollment()` that:
- Generates a cohort of 100 subjects with random age (18–75), weight (50–110 kg), and height (1.50–1.95 m)
- Randomly assigns treatment (Active/Placebo, 1:1 ratio)
- Calculates BMI for all subjects
- Returns a data frame with USUBJID, AGE, TREATMENT, WEIGHT_KG, HEIGHT_M, BMI
- Provides summary statistics (mean age, mean BMI by treatment, etc.)

---
