
# Data Import and Export in Clinical Research

## 3.1 Understanding Clinical Data File Formats

Clinical data arrives in multiple formats, each with specific advantages and requirements. Understanding these formats and how to work with them is essential for any clinical analyst.

### CSV: The Universal Format

CSV (Comma-Separated Values) is plain text: simple, portable, and human-readable. Most clinical data systems can export to CSV.

```
USUBJID,AGE,SEX,TREATMENT,BASELINE_HGB
001-101,62,F,Active,13.5
001-102,45,M,Placebo,12.8
001-103,55,F,Active,14.2
```

Advantages:
- Platform-independent (works on Windows, Mac, Linux)
- Readable by any text editor
- Easy to version control with Git
- Small file sizes

Disadvantages:
- Loses data types (everything is text initially)
- No support for multiple sheets
- Comma conflicts in text fields can cause parsing issues

### Excel: The Industry Standard

Excel files (.xlsx, .xls) are ubiquitous in clinical research. While not ideal for reproducible analysis, they're often unavoidable.

```
├─ Sheet1: Demographics
│  ├─ USUBJID
│  ├─ AGE
│  ├─ SEX
│  ├─ TREATMENT
│  └─ BASELINE_HGB
├─ Sheet2: Lab Results
└─ Sheet3: Adverse Events
```

Advantages:
- Multiple sheets in one file
- Preserves some formatting
- Familiar to non-programmers
- Can include formulas (though importing them is problematic)

Disadvantages:
- Proprietary format (XML-based, but complex)
- Formula preservation issues
- Formatting can create parsing problems
- Less transparent than plain text

### SAS Formats: The Regulatory Standard

SAS data files (.sas7bdat) are the pharmaceutical industry standard. If submitting to the FDA, you'll likely encounter them.

**SDTM domains** are typically provided in SAS format:
- dm.sas7bdat (Demographics)
- ae.sas7bdat (Adverse Events)
- lb.sas7bdat (Laboratory)
- vs.sas7bdat (Vital Signs)

### SAS Transport (XPT) Format

XPT files (.xpt) are the FDA-mandated format for data submissions. They're text-based and transportable across systems, but compressed.

::: {.callout-important}
**XPT Format in Regulatory Submissions**: The FDA requires clinical trial data in XPT (SAS Transport format) v5, which has specific formatting requirements:
- Variable names limited to 8 characters
- No special characters
- Specific metadata structure
- Can be read and written by R but requires care
:::

## 3.2 Reading CSV Files

### Using Base R: read.csv()

```{r}
#| label: read-csv-base
#| eval: false

# Read a CSV file
dm_data <- read.csv("dm.csv")

# Display the first few rows
head(dm_data)

# Check structure
str(dm_data)

# Common parameters:
# - sep: Field separator (default ",", can be "\t" for tab, "|" for pipe)
# - header: TRUE if first row contains column names (default TRUE)
# - na.strings: How missing values are represented (default "NA")
# - stringsAsFactors: Should character columns be factors? (default FALSE in modern R)

# Example: Tab-separated file with missing values as empty strings
vs_data <- read.csv("vs.txt",
                    sep = "\t",
                    na.strings = "")

# Example: Pipe-separated file with comments
cm_data <- read.csv("cm.txt",
                    sep = "|",
                    comment.char = "#")
```

### Using readr: read_csv()

The {readr} package (part of tidyverse) provides faster, more robust CSV reading.

```{r}
#| label: read-csv-readr
#| eval: false

library(readr)

# Read with readr (returns a tibble)
dm_data <- read_csv("dm.csv")

# readr automatically detects column types
# and shows a message like:
# ── Column specification ────────────────────────────
# cols(
#   USUBJID = col_character(),
#   AGE = col_double(),
#   SEX = col_character(),
#   TREATMENT = col_character(),
#   BASELINE_HGB = col_double()
# )

# Explicit column specification
ae_data <- read_csv(
  "ae.csv",
  col_types = cols(
    USUBJID = col_character(),
    AETERM = col_character(),
    AESEV = col_factor(levels = c("Mild", "Moderate", "Severe")),
    AEREL = col_factor(levels = c("Unrelated", "Remote", "Possible", "Probable", "Definite")),
    AESTDTC = col_date(format = "%Y-%m-%d")
  )
)

# Skip the first few rows if they contain metadata
lb_data <- read_csv("lb.csv", skip = 3)

# Read only specific columns
dm_subset <- read_csv("dm.csv",
                      col_select = c("USUBJID", "AGE", "SEX"))

# Handle problematic files
# Show problems encountered during parsing
problems(dm_data)
```

### Practical CSV Reading Workflow

```{r}
#| label: csv-workflow
#| eval: false

library(readr)

# Step 1: Read with automatic detection
raw_data <- read_csv("clinical_data.csv")

# Step 2: Check what was parsed
str(raw_data)

# Step 3: Review the first few rows
head(raw_data, n = 10)

# Step 4: Check for parsing problems
if (nrow(problems(raw_data)) > 0) {
  print("Parsing problems detected:")
  print(problems(raw_data))
  # Re-read with corrected col_types if needed
  raw_data <- read_csv("clinical_data.csv",
                       col_types = cols(
                         # Specify corrections here
                       ))
}

# Step 5: Verify structure matches expectations
expected_cols <- c("USUBJID", "AGE", "SEX", "TREATMENT", "BASELINE_HGB")
missing_cols <- setdiff(expected_cols, names(raw_data))
if (length(missing_cols) > 0) {
  warning(paste("Missing columns:", paste(missing_cols, collapse = ", ")))
}
```

## 3.3 Reading Excel Files

### Using readxl: read_excel()

The {readxl} package provides robust Excel reading without external dependencies.

```{r}
#| label: read-excel
#| eval: false

library(readxl)

# Read a single sheet (first sheet by default)
dm_data <- read_excel("clinical_data.xlsx")

# Read a specific sheet by name
ae_data <- read_excel("clinical_data.xlsx", sheet = "Adverse_Events")

# Read a specific sheet by number
lb_data <- read_excel("clinical_data.xlsx", sheet = 3)

# See all sheet names
sheet_names <- excel_sheets("clinical_data.xlsx")
print(sheet_names)
# [1] "Demographics" "Adverse_Events" "Laboratory" "Vital_Signs"

# Read with specific column types
vs_data <- read_excel(
  "clinical_data.xlsx",
  sheet = "Vital_Signs",
  col_types = c("text", "numeric", "numeric", "numeric", "date")
)

# Skip rows (e.g., if there's a title row)
dm_data <- read_excel("clinical_data.xlsx",
                      skip = 2)

# Read only specific rows and columns
ae_subset <- read_excel("clinical_data.xlsx",
                        sheet = "Adverse_Events",
                        range = "A1:D100")

# Specify missing value representations
cm_data <- read_excel("clinical_data.xlsx",
                      sheet = "Concomitant_Meds",
                      na = c("", "NA", "N/A", "."))
```

### Reading Multiple Sheets at Once

Often clinical Excel files contain multiple domains. Reading them all into a list is efficient:

```{r}
#| label: read-excel-multiple-sheets
#| eval: false

library(readxl)
library(purrr)

# Get sheet names from Excel file
sheets <- excel_sheets("clinical_trial.xlsx")

# Read all sheets into a list
all_data <- map(sheets, ~read_excel("clinical_trial.xlsx", sheet = .x))

# Name the list elements
names(all_data) <- sheets

# Access specific domains
dm <- all_data$Demographics
ae <- all_data$`Adverse_Events`
lb <- all_data$Laboratory

# Or using the $ operator directly
dm <- all_data[["Demographics"]]
```

## 3.4 Reading SAS Data Files

### Using haven: read_sas()

The {haven} package reads SAS, SPSS, and Stata files. It preserves SAS metadata like variable labels and formats.

```{r}
#| label: read-sas
#| eval: false

library(haven)

# Read a SAS dataset
dm_data <- read_sas("dm.sas7bdat")

# SAS labels are preserved as "label" attribute
attributes(dm_data$AGE)$label
# [1] "Age (years)"

# View labels with a helper function
label_list <- sapply(dm_data, function(x) attr(x, "label"))
print(label_list)
# USUBJID: "Unique Subject ID"
# AGE: "Age (years)"
# SEX: "Sex"
# ...

# Use multiple SAS files from same study
dm <- read_sas("sdtm/dm.sas7bdat")
ae <- read_sas("sdtm/ae.sas7bdat")
lb <- read_sas("sdtm/lb.sas7bdat")
vs <- read_sas("sdtm/vs.sas7bdat")
ex <- read_sas("sdtm/ex.sas7bdat")
```

### Understanding SAS Variable Formats

SAS uses "formats" to control how values are displayed. Haven preserves these in attributes.

```{r}
#| label: sas-formats
#| eval: false

library(haven)

# Read SAS data
dm <- read_sas("dm.sas7bdat")

# Check format information
attr(dm$SEX, "format.sas")
# [1] "$SEX."

# SAS value labels (stored as attributes)
attr(dm$SEX, "labels")
# F       M
# Female Male

# When converting to R factor, use these labels
dm$SEX <- factor(dm$SEX,
                 labels = attr(dm$SEX, "labels"))

# The as_factor() function from haven does this automatically
dm$SEX <- as_factor(dm$SEX)

# Check the result
print(dm$SEX)
# [1] Female Male Female Male Male ...
# Levels: Female Male
```

### Practical SAS Import Workflow

```{r}
#| label: sas-workflow
#| eval: false

library(haven)
library(tidyverse)

# Function to import and standardize SDTM domains
import_sdtm <- function(domain_code, folder = "sdtm") {
  filename <- paste0(folder, "/", tolower(domain_code), ".sas7bdat")

  if (!file.exists(filename)) {
    stop(paste("File not found:", filename))
  }

  # Read the SAS file
  data <- read_sas(filename)

  # Convert SAS variable labels to a comment
  var_labels <- sapply(data, function(x) attr(x, "label"))
  attr(data, "variable_labels") <- var_labels

  # Convert factors with SAS formats
  data <- data %>%
    mutate(across(where(is.character), ~as_factor(.) %||% .))

  return(data)
}

# Load all SDTM domains
dm <- import_sdtm("DM")
ae <- import_sdtm("AE")
lb <- import_sdtm("LB")
vs <- import_sdtm("VS")
ex <- import_sdtm("EX")

# Verify structure
str(dm)
```

## 3.5 Reading SAS Transport (XPT) Files

### Using haven: read_xpt()

XPT (SAS Transport v5) is the FDA-mandated format for data submission.

```{r}
#| label: read-xpt
#| eval: false

library(haven)

# Read an XPT file
dm <- read_xpt("dm.xpt")

# XPT files follow CDISC standards strictly
# Check column names (limited to 8 characters in SDTM)
names(dm)
# [1] "STUDYID" "DOMAIN" "USUBJID" "SUBJID" "RFSTDTC" ...

# Read multiple XPT files
xpt_files <- c("dm.xpt", "ae.xpt", "lb.xpt", "vs.xpt")

domains <- lapply(xpt_files, read_xpt)
names(domains) <- c("DM", "AE", "LB", "VS")

# Access individual domains
dm <- domains$DM
ae <- domains$AE
```

::: {.callout-tip}
**XPT File Validation**: XPT files from FDA submissions must be valid and compliant. R can read them, but when writing XPT files, use specialized tools like {xportr} from the pharmaverse for full compliance.
:::

## 3.6 Reading Data from Pharmaversesdtm Package

The {pharmaversesdtm} package provides simulated clinical data in SDTM format, perfect for learning and testing.

### Loading Datasets

```{r}
#| label: pharmaversesdtm-load

library(pharmaversesdtm)
library(tidyverse)

# See available datasets
data(package = "pharmaversesdtm")

# Load specific domains
data("dm")   # Demographics
data("ae")   # Adverse Events
data("lb")   # Laboratory
data("vs")   # Vital Signs
data("ex")   # Exposure
data("cm")   # Concomitant Medications
data("ds")   # Disposition

# These are immediately available
head(dm)
head(ae)

# Check the structure of each domain
str(lb)
str(vs)
```

### Understanding Pharmaversesdtm Structure

```{r}
#| label: pharmaversesdtm-structure

library(pharmaversesdtm)
library(dplyr)

# Load a domain
data("lb")

# Examine structure
glimpse(lb)

# Get a summary of the LB (Laboratory) domain
lb %>%
  summarise(
    n_records = n(),
    n_subjects = n_distinct(USUBJID),
    n_test_types = n_distinct(LBTEST),
    n_visits = n_distinct(VISIT),
    date_range = paste(min(LBDTC), "to", max(LBDTC))
  )

# View unique lab test types
unique_tests <- lb %>%
  distinct(LBTEST, LBCAT) %>%
  arrange(LBCAT, LBTEST)

print(unique_tests)
```

### Using Pharmaversesdtm for ADaM Development

The pharmaversesdtm data can be used directly with the {admiral} package to create analysis datasets:

```{r}
#| label: pharmaversesdtm-adam

library(pharmaversesdtm)
library(tidyverse)

# Load SDTM domains
data("dm")
data("lb")

# Create a simple ADLB (laboratory analysis dataset)
# using admiral functions

adlb <- lb %>%
  # Join with demographics for subject-level variables
  left_join(dm %>% select(USUBJID, ARM, ACTARMCD),
            by = "USUBJID") %>%
  # Calculate change from baseline
  group_by(USUBJID, LBTEST) %>%
  arrange(LBDTC) %>%
  mutate(
    LBBLFL = if_else(row_number() == 1, "Y", "N"),
    BASELINE = if_else(LBBLFL == "Y", LBSTRESN, NA_real_)
  ) %>%
  ungroup() %>%
  fill(BASELINE, .direction = "down") %>%
  mutate(
    CHG = if_else(!is.na(BASELINE), LBSTRESN - BASELINE, NA_real_)
  ) %>%
  select(USUBJID, LBTEST, LBDTC, LBSTRESN, BASELINE, CHG, ARM, everything())

head(adlb)
```

## 3.7 Writing Data Files

### Exporting to CSV

```{r}
#| label: write-csv
#| eval: false

# Write to CSV with base R
write.csv(dm_data, file = "dm_output.csv", row.names = FALSE)

# Write to CSV with readr (recommended)
library(readr)
write_csv(dm_data, file = "dm_output.csv")

# Write with custom separator
write.table(vs_data, file = "vs_output.txt",
            sep = "\t",
            row.names = FALSE)

# Write with custom NA representation
write_csv(ae_data, file = "ae_output.csv", na = "")

# Write specific columns
write_csv(dm_data %>% select(USUBJID, AGE, SEX),
          file = "dm_subset.csv")
```

### Exporting to Excel

```{r}
#| label: write-excel
#| eval: false

library(writexl)

# Write a single sheet
write_xlsx(dm_data, path = "output.xlsx")

# Write multiple sheets
write_xlsx(
  list(
    Demographics = dm_data,
    `Adverse Events` = ae_data,
    Laboratory = lb_data,
    `Vital Signs` = vs_data
  ),
  path = "clinical_data.xlsx"
)

# Note: writexl does NOT support formatting or formulas
# For complex Excel workbooks, consider openxlsx package
```

### Exporting to SAS Format

```{r}
#| label: write-sas
#| eval: false

library(haven)

# Write to SAS dataset (.sas7bdat)
write_sas(dm_data, path = "dm_output.sas7bdat")

# Add variable labels before writing
dm_data <- dm_data %>%
  set_variable_labels(
    USUBJID = "Unique Subject ID",
    AGE = "Age (years)",
    SEX = "Sex",
    TREATMENT = "Assigned Treatment"
  )

write_sas(dm_data, path = "dm_labeled.sas7bdat")
```

### Exporting to XPT (SAS Transport)

::: {.callout-important}
**FDA XPT Compliance**: Writing XPT files for FDA submission requires strict adherence to standards. Use the {xportr} package from pharmaverse for full compliance, including:
- 8-character variable name truncation
- Proper variable ordering
- Standard metadata
- Label truncation to 40 characters
:::

```{r}
#| label: write-xpt
#| eval: false

library(haven)

# Basic XPT write (careful with CDISC compliance)
write_xpt(dm_data, path = "dm.xpt")

# For FDA submissions, use xportr from pharmaverse
# install.packages("remotes")
# remotes::install_github("pharmaverse/xportr")

library(xportr)

# Create a specification of required variables
var_spec <- tibble(
  Variable = c("USUBJID", "AGE", "SEX"),
  Type = c("text", "numeric", "text"),
  `Variable Label` = c("Unique Subject ID", "Age (years)", "Sex"),
  Length = c(8, 8, 2)
)

# Write with full compliance checking
dm_xpt <- dm_data %>%
  xportr_type(var_spec) %>%
  xportr_label(var_spec) %>%
  xportr_order(var_spec)

write_xpt(dm_xpt, path = "dm.xpt")
```

## 3.8 Handling Large Clinical Datasets

### Data Size Considerations

Clinical datasets can be large:
- A single SDTM LB (laboratory) domain from a 1000-subject trial with 50 labs per subject per visit across 10 visits = 500,000 rows
- Full text files with imaging data can be gigabytes

### Using data.table for Performance

For very large datasets, {data.table} is faster than {dplyr}:

```{r}
#| label: data-table-import
#| eval: false

library(data.table)

# Read large CSV with data.table
# Much faster than read.csv() or read_csv()
lb_large <- fread("lb_large.csv")

# Efficient subsetting
# Filter to specific lab tests
hematology <- lb_large[LBCAT == "HEMATOLOGY"]

# Group by calculations
summary_by_test <- lb_large[, .(
  n = .N,
  mean_value = mean(LBSTRESN, na.rm = TRUE),
  sd_value = sd(LBSTRESN, na.rm = TRUE)
), by = LBTEST]

print(summary_by_test)

# Write efficiently
fwrite(lb_large, file = "lb_output.csv")
```

### Chunked Reading for Memory Efficiency

```{r}
#| label: chunked-reading
#| eval: false

library(readr)

# Read a large file in chunks
process_large_file <- function(filename, chunk_size = 10000) {
  # Create a connection
  f <- file(filename, "r")

  # Read chunks
  chunk_num <- 0
  while (TRUE) {
    # Read chunk_size lines
    chunk <- read_csv(f, n_max = chunk_size, progress = FALSE)

    if (nrow(chunk) == 0) break  # End of file

    chunk_num <- chunk_num + 1
    cat("Processing chunk", chunk_num, "with", nrow(chunk), "rows\n")

    # Process this chunk (example: write to output)
    # your_analysis_function(chunk)
  }

  close(f)
}
```

## 3.9 Complete Data Import Workflow for a Clinical Trial

Let's integrate everything into a realistic clinical data workflow:

```{r}
#| label: complete-import-workflow
#| eval: false

library(readr)
library(haven)
library(tidyverse)

# Define study parameters
study_id <- "EUSTC1"
data_folder <- "./data/raw"

# Function to load and validate SDTM domains
load_sdtm_domains <- function(study_id, data_folder) {

  # Define expected domains
  domains <- list(
    DM = "Demographics",
    AE = "Adverse Events",
    LB = "Laboratory",
    VS = "Vital Signs",
    EX = "Exposure"
  )

  # Load each domain
  sdtm <- list()

  for (domain_code in names(domains)) {
    # Try reading as SAS first
    sas_file <- file.path(data_folder, paste0(tolower(domain_code), ".sas7bdat"))

    if (file.exists(sas_file)) {
      cat("Loading", domain_code, "from SAS...\n")
      sdtm[[domain_code]] <- read_sas(sas_file)
    } else {
      # Try CSV as fallback
      csv_file <- file.path(data_folder, paste0(tolower(domain_code), ".csv"))
      if (file.exists(csv_file)) {
        cat("Loading", domain_code, "from CSV...\n")
        sdtm[[domain_code]] <- read_csv(csv_file)
      } else {
        warning(paste("Could not find data for domain", domain_code))
      }
    }
  }

  return(sdtm)
}

# Load the data
sdtm_data <- load_sdtm_domains(study_id, data_folder)

# Extract individual domains
dm <- sdtm_data$DM
ae <- sdtm_data$AE
lb <- sdtm_data$LB
vs <- sdtm_data$VS
ex <- sdtm_data$EX

# Perform validation checks
cat("=== Data Quality Summary ===\n")
cat("DM (Demographics):", nrow(dm), "records for", n_distinct(dm$USUBJID), "subjects\n")
cat("AE (Adverse Events):", nrow(ae), "records\n")
cat("LB (Laboratory):", nrow(lb), "records\n")
cat("VS (Vital Signs):", nrow(vs), "records\n")
cat("EX (Exposure):", nrow(ex), "records\n")

# Check for missing data
cat("\n=== Missing Data Check ===\n")
missing_by_domain <- list(
  DM = colSums(is.na(dm)),
  AE = colSums(is.na(ae)),
  LB = colSums(is.na(lb))
)

print(missing_by_domain)

# Export cleaned data in multiple formats
output_folder <- "./data/processed"
dir.create(output_folder, showWarnings = FALSE)

# Export to CSV (for transparency)
write_csv(dm, file.path(output_folder, "dm.csv"))
write_csv(ae, file.path(output_folder, "ae.csv"))
write_csv(lb, file.path(output_folder, "lb.csv"))

# Export to XPT (for regulatory submission)
write_xpt(dm, file.path(output_folder, "dm.xpt"))
write_xpt(ae, file.path(output_folder, "ae.xpt"))
write_xpt(lb, file.path(output_folder, "lb.xpt"))

cat("\nData import and export complete.\n")
cat("Output saved to:", output_folder, "\n")
```

## 3.10 Best Practices for Clinical Data Import/Export

1. **Always validate after import**: Check dimensions, data types, missing values
2. **Use consistent naming**: Follow CDISC conventions (8-character variable names for submission)
3. **Preserve metadata**: Keep variable labels and formats when possible
4. **Version your raw data**: Never overwrite original files; keep dated backups
5. **Document transformations**: Record which files were loaded, when, and any transformations applied
6. **Use reproducible scripts**: Write R scripts that can reload data identically
7. **Test on sample data first**: Before processing 500,000 rows, test on a subset
8. **Check for encoding issues**: Clinical data may come from different locales; watch for encoding problems
9. **Validate against source systems**: Compare record counts and key statistics with the source system
10. **Archive processed data**: Keep exported datasets for audit trails and reproducibility

---

## Exercises

### Beginner Exercises

**Exercise 3.1: Reading and Comparing File Formats**

Using the {pharmaversesdtm} package:
a) Load the `dm` (Demographics) domain
b) Export it to three formats: CSV, Excel, and SAS (.sas7bdat)
c) Read each file back into R
d) Verify that the dimensions are identical for all three formats
e) Check if any data types have changed

**Exercise 3.2: Handling Missing Data During Import**

Create a CSV file with intentional missing values:
```
USUBJID,AGE,SEX,TREATMENT,BASELINE_HGB
001-101,62,F,Active,13.5
001-102,,M,Placebo,12.8
001-103,55,F,,14.2
001-104,56,,Active,
```

a) Read this file using `read_csv()` with default settings
b) Identify which cells have missing values
c) Re-read the file specifying that missing values should also be represented as "."
d) Check how the data has changed

**Exercise 3.3: Multi-Sheet Excel Import**

Create an Excel file with three sheets (DM, AE, LB) containing sample clinical data:
a) Read all three sheets into a single list
b) Access each sheet individually from the list
c) Verify that column names are identical across sheets where expected
d) Export the list to three separate CSV files

### Advanced Exercises

**Exercise 3.4: Batch Import of SDTM Domains**

Write a function `import_all_sdtm_domains()` that:
a) Takes a folder path as input
b) Automatically detects all SAS files in that folder
c) Reads each file using the domain code from the filename (e.g., dm.sas7bdat → DM)
d) Returns a named list of all domains found
e) Provides a summary report: number of subjects per domain, record counts, date ranges

**Exercise 3.5: Data Export Pipeline**

Using data from pharmaversesdtm:
a) Load the DM and LB domains
b) Create a merged analysis dataset combining demographics with baseline lab values
c) Export to four formats: CSV, Excel (single file, multiple sheets), SAS, and XPT
d) Verify that all files can be read back correctly
e) Document (in code comments) what information is preserved/lost in each format

---
