# Data Manipulation with dplyr: Filtering, Selecting, Mutating, Grouping, and Summarising Clinical Data

## Introduction

Once data has been imported and cleaned, the next essential step is manipulation — transforming raw datasets into analysis-ready structures. The `dplyr` package provides a consistent, readable grammar for data manipulation built around a small set of core verbs. Each verb does one thing well, and they compose naturally through the pipe operator (`|>` or `%>%`).

This chapter covers `dplyr`'s core verbs in the context of clinical trial data:

- `filter()` — Keep rows that meet conditions (patient selection criteria)
- `select()` — Keep or drop columns (variable selection)
- `mutate()` — Create or transform variables (derived endpoints)
- `summarise()` + `group_by()` — Compute summaries by group (treatment arm summaries)
- `arrange()` — Sort data
- `join` functions — Combine datasets (merging ADSL with ADAE, etc.)
- `across()` — Apply operations to multiple columns at once

---

## 1. Setup and Data

```{r}
#| label: setup-04a
#| message: false
#| warning: false

library(tidyverse)
library(pharmaversesdtm)

# Load key SDTM domains
data("dm")    # Demographics
data("ae")    # Adverse Events
data("lb")    # Laboratory
data("vs")    # Vital Signs
data("ex")    # Exposure

cat("Loaded SDTM domains:\n")
cat("  DM:", nrow(dm), "rows x", ncol(dm), "cols\n")
cat("  AE:", nrow(ae), "rows x", ncol(ae), "cols\n")
cat("  LB:", nrow(lb), "rows x", ncol(lb), "cols\n")
cat("  VS:", nrow(vs), "rows x", ncol(vs), "cols\n")
```

---

## 2. filter() — Selecting Rows

`filter()` keeps rows where a logical condition is `TRUE`. In clinical data, this maps directly to subject selection criteria, population flags, and analysis windows.

### 2.1 Basic Filtering

```{r}
#| label: filter-basics

# Select the safety population: all subjects who received at least one dose
# (ACTARM not equal to "Screen Failure")
safety_pop <- dm |>
  filter(ACTARM != "Screen Failure", !is.na(ACTARM))

cat("Safety population:", nrow(safety_pop), "subjects\n")

# Select treatment-emergent adverse events only
# AESTDTC >= EXSTDTC (AE started on or after first dose)
# We use a simple proxy: AESER == "Y" for serious AEs
serious_ae <- ae |>
  filter(AESER == "Y")

cat("Serious adverse events:", nrow(serious_ae), "\n")

# Select subjects aged >= 65 (elderly sub-group)
elderly <- dm |>
  filter(AGE >= 65)
cat("Subjects aged ≥ 65:", nrow(elderly), "\n")
```

### 2.2 Compound Conditions

```{r}
#| label: filter-compound

# AND: female subjects in the active arm
female_active <- dm |>
  filter(SEX == "F", ARM != "Placebo")
cat("Female active-arm subjects:", nrow(female_active), "\n")

# OR: subjects with severe AEs
# (pharmaversesdtm uses AESEV; real studies often also carry AETOXGR for toxicity grade)
severe_ae <- ae |>
  filter(AESEV == "SEVERE")
cat("Severe AEs:", nrow(severe_ae), "\n")

# NOT: exclude screen failures and not-treated subjects
analysable <- dm |>
  filter(!ARM %in% c("Screen Failure", "Not Assigned"))
cat("Analysable subjects:", nrow(analysable), "\n")
```

### 2.3 Filtering with Helper Functions

```{r}
#| label: filter-helpers

# between() — numeric range
lb_normal_range <- lb |>
  filter(LBTESTCD == "ALT", between(LBSTRESN, 0, 56))
cat("ALT values in normal range (0–56):", nrow(lb_normal_range), "\n")

# str_detect() — pattern matching on character columns
ae_infect <- ae |>
  filter(str_detect(AEDECOD, regex("infection|infect", ignore_case = TRUE)))
cat("Infection-related AEs:", nrow(ae_infect), "\n")

# is.na() — missing value filter
complete_sbp <- vs |>
  filter(VSTESTCD == "SYSBP", !is.na(VSSTRESN))
cat("Non-missing SBP records:", nrow(complete_sbp), "\n")
```

---

## 3. select() — Choosing Columns

In clinical datasets, domains can have 50+ columns. `select()` lets you work with only the variables you need.

```{r}
#| label: select-basics

# Select key demographic variables
dm_subset <- dm |>
  select(USUBJID, SUBJID, ARM, ACTARM, AGE, SEX, RACE, COUNTRY)

cat("DM columns selected:", names(dm_subset), "\n")

# Select columns by pattern
ae_key <- ae |>
  select(USUBJID, starts_with("AE"), -ends_with("DTC"))  # AE vars, no dates

cat("AE columns:", ncol(ae_key), "\n")

# Drop specific columns
lb_no_dates <- lb |>
  select(-contains("DTC"), -contains("DTM"))

cat("LB without date columns:", ncol(lb_no_dates), "cols (was", ncol(lb), ")\n")

# Rename during select
dm_clean <- dm |>
  select(
    subject_id   = USUBJID,
    arm          = ARM,
    age_years    = AGE,
    sex          = SEX
  )
head(dm_clean, 3)
```

---

## 4. mutate() — Creating and Transforming Variables

`mutate()` adds new columns or transforms existing ones — the primary tool for creating derived endpoints.

### 4.1 Numeric Transformations

```{r}
#| label: mutate-numeric

# Create age groups (a common demographic stratification variable)
dm_with_agegroup <- dm |>
  mutate(
    age_group = case_when(
      AGE < 18            ~ "< 18",
      between(AGE, 18, 44) ~ "18–44",
      between(AGE, 45, 64) ~ "45–64",
      AGE >= 65           ~ "≥ 65",
      TRUE                ~ NA_character_
    )
  ) |>
  select(USUBJID, AGE, age_group)

table(dm_with_agegroup$age_group, useNA = "ifany")
```

```{r}
#| label: mutate-lab-flags

# Flag abnormal laboratory values based on reference ranges
lb_flagged <- lb |>
  filter(!is.na(LBSTRESN), !is.na(LBSTNRLO), !is.na(LBSTNRHI)) |>
  mutate(
    below_low  = LBSTRESN < LBSTNRLO,
    above_high = LBSTRESN > LBSTNRHI,
    abnormal   = below_low | above_high,
    # Relative deviation from reference range midpoint
    ref_midpoint = (LBSTNRLO + LBSTNRHI) / 2,
    pct_from_mid = 100 * (LBSTRESN - ref_midpoint) / ref_midpoint
  )

cat("Proportion of abnormal lab results:\n")
lb_flagged |>
  group_by(LBTESTCD) |>
  summarise(
    n          = n(),
    pct_abnorm = round(100 * mean(abnormal), 1)
  ) |>
  arrange(desc(pct_abnorm)) |>
  head(10) |>
  print()
```

### 4.2 String Operations

```{r}
#| label: mutate-strings

# Standardise AE preferred terms to title case
ae_clean <- ae |>
  mutate(
    AEDECOD_std = str_to_title(AEDECOD),
    AEBODSYS_std = str_to_title(AEBODSYS)
  ) |>
  select(USUBJID, AEDECOD, AEDECOD_std, AEBODSYS_std)

head(ae_clean, 5)
```

### 4.3 Conditional Variables with case_when()

```{r}
#| label: mutate-case-when

# Classify AE severity for reporting
ae_severity <- ae |>
  mutate(
    severity_label = case_when(
      AESEV == "MILD"     ~ "Mild",
      AESEV == "MODERATE" ~ "Moderate",
      AESEV == "SEVERE"   ~ "Severe",
      AESER == "Y"        ~ "Serious",
      TRUE                ~ "Unknown"
    ),
    is_related = case_when(
      # pharmaversesdtm encodes AEREL as "Y" (related) / "N" (not related)
      AEREL == "Y" ~ TRUE,
      AEREL == "N" ~ FALSE,
      TRUE         ~ NA
    )
  )

cat("AE severity distribution:\n")
print(table(ae_severity$severity_label, useNA = "ifany"))
```

---

## 5. group_by() and summarise() — Grouped Summaries

The combination of `group_by()` + `summarise()` produces clinical summary tables — the building blocks of Table 14.1 (subject disposition) and Table 14.3 (adverse events) in a clinical study report.

```{r}
#| label: group-summarise

# Summary statistics by treatment arm
dm_summary <- dm |>
  filter(!is.na(ARM), ARM != "Screen Failure") |>
  group_by(ARM) |>
  summarise(
    N           = n(),
    Age_mean    = round(mean(AGE, na.rm = TRUE), 1),
    Age_sd      = round(sd(AGE,  na.rm = TRUE), 1),
    Age_median  = median(AGE, na.rm = TRUE),
    Pct_female  = round(100 * mean(SEX == "F", na.rm = TRUE), 1),
    .groups     = "drop"
  ) |>
  mutate(Age_summary = paste0(Age_mean, " (", Age_sd, ")"))

dm_summary |>
  select(ARM, N, Age_summary, Age_median, Pct_female) |>
  print()
```

```{r}
#| label: ae-frequency-table

# AE incidence table: frequency of each preferred term by arm
ae_incidence <- ae |>
  left_join(dm |> select(USUBJID, ARM), by = "USUBJID") |>
  filter(!is.na(ARM), ARM != "Screen Failure") |>
  group_by(ARM, AEDECOD) |>
  summarise(
    n_events   = n(),
    n_subjects = n_distinct(USUBJID),
    .groups    = "drop"
  ) |>
  left_join(
    dm |> filter(!is.na(ARM), ARM != "Screen Failure") |>
      count(ARM, name = "N_arm"),
    by = "ARM"
  ) |>
  mutate(pct = round(100 * n_subjects / N_arm, 1)) |>
  arrange(ARM, desc(n_subjects))

cat("Top 10 AEs (all arms combined):\n")
ae_incidence |>
  group_by(AEDECOD) |>
  summarise(total_subjects = sum(n_subjects), .groups = "drop") |>
  arrange(desc(total_subjects)) |>
  head(10) |>
  print()
```

---

## 6. arrange() — Sorting Data

```{r}
#| label: arrange-examples

# Sort vital signs by subject then visit then test
vs_sorted <- vs |>
  arrange(USUBJID, VISITNUM, VSTESTCD)

head(vs_sorted |> select(USUBJID, VISITNUM, VSTESTCD, VSSTRESN), 8)

# Sort AEs by onset date descending (most recent first)
ae_by_date <- ae |>
  filter(!is.na(AESTDTC)) |>
  arrange(desc(AESTDTC))

head(ae_by_date |> select(USUBJID, AEDECOD, AESTDTC, AESER), 5)
```

---

## 7. Joining Datasets

Combining datasets (DM + AE, ADSL + ADAE) is fundamental to clinical data analysis. `dplyr` provides clear, explicit join functions.

```{r}
#| label: joins

# Left join: add demographic info to every AE record
ae_dm <- ae |>
  left_join(
    dm |> select(USUBJID, ARM, AGE, SEX),
    by = "USUBJID"
  )

cat("AE with demographics:", nrow(ae_dm), "rows\n")
cat("AEs without matched subject:", sum(is.na(ae_dm$ARM)), "\n")

# Inner join: only subjects who have both AEs AND lab data
# (subjects with at least one record in both domains)
ae_subjects <- distinct(ae, USUBJID)
lb_subjects <- distinct(lb, USUBJID)
both_domains <- inner_join(ae_subjects, lb_subjects, by = "USUBJID")
cat("Subjects with both AE and LB data:", nrow(both_domains), "\n")

# Anti-join: subjects in DM but NOT in AE (no AEs reported)
no_ae_subjects <- dm |>
  anti_join(ae, by = "USUBJID")
cat("Subjects with no AEs:", nrow(no_ae_subjects), "\n")
```

---

## 8. across() — Applying Functions to Multiple Columns

`across()` eliminates repetitive code when the same operation needs to be applied to several columns — useful for standardising lab values or converting multiple date columns.

```{r}
#| label: across-examples

# Compute mean and SD for all numeric vital sign measurements per subject
vs_numeric_summary <- vs |>
  filter(!is.na(VSSTRESN)) |>
  group_by(USUBJID, VSTESTCD) |>
  summarise(
    across(VSSTRESN, list(mean = mean, sd = sd, n = \(x) sum(!is.na(x)))),
    .groups = "drop"
  )

head(vs_numeric_summary, 6)

# Convert multiple character columns to upper case
ae_upper <- ae |>
  mutate(across(c(AEBODSYS, AEDECOD, AESEV), str_to_upper))

cat("Standardised AE terms (first row):\n")
ae_upper |> select(AEBODSYS, AEDECOD, AESEV) |> head(2) |> print()
```

---

## 9. Practical Workflow: Building an Analysis Dataset

Combining all verbs into a complete data-preparation pipeline:

```{r}
#| label: full-pipeline

# Build a subject-level safety dataset (ADSL-like)
adsl_safety <- dm |>
  # Keep randomised subjects only
  filter(!ARM %in% c("Screen Failure", "Not Assigned"), !is.na(ARM)) |>
  # Select and rename key variables
  select(
    USUBJID, SUBJID,
    ARM, ACTARM,
    AGE, SEX, RACE, COUNTRY
  ) |>
  # Add derived variables
  mutate(
    age_group = case_when(
      AGE < 65  ~ "< 65",
      AGE >= 65 ~ "≥ 65"
    ),
    female_flag = (SEX == "F")
  ) |>
  # Join AE summary
  left_join(
    ae |>
      group_by(USUBJID) |>
      summarise(
        n_ae       = n(),
        any_serious = any(AESER == "Y", na.rm = TRUE),
        .groups    = "drop"
      ),
    by = "USUBJID"
  ) |>
  # Replace NA (no AEs) with 0 / FALSE
  mutate(
    n_ae        = replace_na(n_ae, 0L),
    any_serious = replace_na(any_serious, FALSE)
  ) |>
  arrange(USUBJID)

cat("Analysis dataset dimensions:", nrow(adsl_safety), "×", ncol(adsl_safety), "\n")
cat("Subjects with any SAE:", sum(adsl_safety$any_serious), "\n")
cat("Mean AEs per subject:", round(mean(adsl_safety$n_ae), 2), "\n")

glimpse(adsl_safety)
```

---

## Summary

`dplyr`'s six core verbs — `filter()`, `select()`, `mutate()`, `summarise()`, `group_by()`, and `arrange()` — together with the join functions, cover the vast majority of clinical data manipulation tasks:

- **`filter()`** implements inclusion/exclusion criteria and analysis population flags
- **`select()`** shapes datasets to contain only the needed variables
- **`mutate()`** creates derived endpoints, group flags, and normalised values
- **`group_by()` + `summarise()`** produces the subject-level and group-level summary tables that appear in clinical study reports
- **`arrange()`** orders records for review, QC, and listing creation
- **`join()`** functions merge across SDTM domains and ADaM datasets
- **`across()`** applies the same transformation to many variables at once, eliminating repetitive code

> **Best practice**: Write manipulation code as a single pipe chain from raw data to analysis-ready dataset. This makes the data lineage transparent, reproducible, and easy to audit — all essential for regulatory submissions.
