
# Data Wrangling with tidyr: Reshaping, Pivoting, Nesting, and Structuring Clinical Data

## Introduction

Data wrangling — transforming data between wide and long formats, handling nested structures, and resolving shape mismatches — is a daily task in clinical data programming. Clinical trial data is particularly challenging because:

- SDTM stores lab and vital sign data in **long** format (one row per test per visit)
- Statistical analysis often requires **wide** format (one row per subject, one column per timepoint)
- ADaM datasets require specific structures that differ from SDTM
- Regulatory outputs (listings, tables) demand yet another shape

The `tidyr` package provides the tools to reshape data efficiently and consistently.

This chapter covers:

- `pivot_longer()` and `pivot_wider()` — reshaping between long and wide
- `separate()` and `unite()` — splitting and combining columns
- `complete()` — filling in missing combinations
- `nest()` and `unnest()` — working with list-columns and grouped models
- `fill()` and `replace_na()` — handling structural missing values
- Real SDTM-to-ADaM reshaping workflows

---

## 1. Setup

```{r}
#| label: setup-04b
#| message: false
#| warning: false

library(tidyverse)
library(pharmaversesdtm)

data("dm")
data("lb")
data("vs")
data("ae")

cat("Domains loaded.\n")
```

---

## 2. Tidy Data Principles

Before reshaping, it helps to understand what "tidy" means:

1. Every **variable** has its own column
2. Every **observation** has its own row
3. Every **value** has its own cell

Most clinical data issues arise from violations of these rules. For example, a lab dataset with `BASELINE`, `WEEK4`, `WEEK8`, `WEEK12` as separate columns violates rule 2 — the four columns represent four observations of the same variable.

```{r}
#| label: tidy-concept

# SDTM LB is already tidy (long): one row per subject × test × visit
lb |>
  select(USUBJID, LBTESTCD, VISITNUM, LBSTRESN) |>
  filter(LBTESTCD == "ALT", USUBJID == unique(USUBJID)[1]) |>
  head(8)
```

---

## 3. pivot_longer() — Wide to Long

`pivot_longer()` stacks multiple columns into key–value pairs. Use it when column names represent values of a variable (e.g., `visit_1`, `visit_2`, `visit_4`).

### 3.1 Clinical Example: Summary Statistics to Long Format

```{r}
#| label: pivot-longer-basic

# Create a wide summary table (e.g., from a reporting tool)
dm_wide <- dm |>
  filter(!is.na(ARM), ARM != "Screen Failure") |>
  group_by(ARM) |>
  summarise(
    N           = n(),
    Mean_age    = round(mean(AGE, na.rm = TRUE), 1),
    SD_age      = round(sd(AGE,  na.rm = TRUE), 1),
    Pct_female  = round(100 * mean(SEX == "F"), 1),
    .groups     = "drop"
  )

print(dm_wide)

# Pivot to long for a reporting-friendly layout
dm_long <- dm_wide |>
  pivot_longer(
    cols      = -ARM,
    names_to  = "Statistic",
    values_to = "Value"
  )

print(dm_long)
```

### 3.2 SDTM Vital Signs: Multiple Parameters to Long

```{r}
#| label: pivot-longer-vs

# Wide vital signs (hypothetical pre-processed table)
vs_wide_demo <- vs |>
  filter(VSTESTCD %in% c("SYSBP", "DIABP", "PULSE"),
         VISITNUM %in% c(1, 2, 3)) |>
  select(USUBJID, VISITNUM, VSTESTCD, VSSTRESN) |>
  pivot_wider(names_from = VSTESTCD, values_from = VSSTRESN) |>
  head(20)

cat("Wide VS (first 5 rows):\n")
print(head(vs_wide_demo, 5))

# Now pivot back to long (with multiple value columns)
vs_long_back <- vs_wide_demo |>
  pivot_longer(
    cols      = any_of(c("SYSBP", "DIABP", "PULSE")),
    names_to  = "Parameter",
    values_to = "Value",
    values_drop_na = TRUE
  )

cat("\nLong VS (first 8 rows):\n")
print(head(vs_long_back, 8))
```

---

## 4. pivot_wider() — Long to Wide

`pivot_wider()` spreads key–value pairs across multiple columns. Critical for:

- Creating change-from-baseline (CFB) datasets where baseline and post-baseline are side by side
- Building subject-level datasets with one column per timepoint
- Constructing cross-tabulations for regulatory tables

### 4.1 Lab Data: One Column per Visit

```{r}
#| label: pivot-wider-lab

# Create an ALT profile table: one column per visit
alt_profile <- lb |>
  filter(LBTESTCD == "ALT",
         !is.na(LBSTRESN),
         VISITNUM %in% c(1, 2, 3, 4, 5)) |>
  select(USUBJID, VISITNUM, LBSTRESN) |>
  pivot_wider(
    names_from   = VISITNUM,
    values_from  = LBSTRESN,
    names_prefix = "Visit_"
  )

cat("ALT by visit (wide format — first 6 subjects):\n")
print(head(alt_profile, 6))
cat("Dimensions:", nrow(alt_profile), "×", ncol(alt_profile), "\n")
```

### 4.2 Computing Change from Baseline

```{r}
#| label: cfb-calculation

# Compute change from baseline (CFB) and percent CFB for SBP
sbp_cfb <- vs |>
  filter(VSTESTCD == "SYSBP", !is.na(VSSTRESN)) |>
  select(USUBJID, VISITNUM, VSSTRESN) |>
  # Create BASELINE and POST-BASELINE columns
  mutate(
    timepoint = if_else(VISITNUM == 1, "Baseline", paste0("Visit_", VISITNUM))
  ) |>
  pivot_wider(
    id_cols      = USUBJID,
    names_from   = timepoint,
    values_from  = VSSTRESN,
    values_fn    = mean   # average if multiple baseline readings
  ) |>
  # Compute CFB for each post-baseline visit
  mutate(
    across(
      starts_with("Visit_"),
      list(
        CFB    = \(x) x - Baseline,
        PCTCFB = \(x) 100 * (x - Baseline) / Baseline
      ),
      .names = "{.col}_{.fn}"
    )
  )

cat("SBP with CFB (first 3 subjects):\n")
sbp_cfb |>
  select(USUBJID, Baseline, starts_with("Visit_2"), starts_with("Visit_3")) |>
  head(3) |>
  print()
```

---

## 5. separate() and unite() — Splitting and Combining Columns

SDTM date-time columns (ISO 8601: `2023-04-15T09:30:00`) are stored as strings and often need splitting. Conversely, separate pieces of information sometimes need joining.

### 5.1 separate() — Split a Column

```{r}
#| label: separate-dates

# Split SDTM ISO date into date and time components
lb_dates <- lb |>
  filter(!is.na(LBDTC)) |>
  select(USUBJID, LBTESTCD, LBDTC) |>
  separate(
    col  = LBDTC,
    into = c("LBDT", "LBTM"),
    sep  = "T",
    fill = "right",    # fill right if no time component
    remove = FALSE
  ) |>
  separate(
    col  = LBDT,
    into = c("LBYR", "LBMO", "LBDY"),
    sep  = "-",
    fill = "right"
  )

head(lb_dates |> select(USUBJID, LBDTC, LBYR, LBMO, LBDY, LBTM), 5)
```

### 5.2 unite() — Combine Columns

```{r}
#| label: unite-example

# Create a unique visit-test identifier for pivot keys
lb_key <- lb |>
  select(USUBJID, LBTESTCD, VISITNUM, LBSTRESN) |>
  unite(
    col    = "visit_test",
    VISITNUM, LBTESTCD,
    sep    = "_V",
    remove = FALSE
  ) |>
  mutate(visit_test = paste0("V", visit_test))

head(lb_key |> select(USUBJID, VISITNUM, LBTESTCD, visit_test, LBSTRESN), 5)
```

---

## 6. complete() — Making Implicit Missing Values Explicit

`complete()` ensures that all combinations of variables are present in the dataset, inserting `NA` for missing combinations. This is essential for creating LOCF (Last Observation Carried Forward) datasets and for identifying subjects missing planned assessments.

```{r}
#| label: complete-example

# Identify which subjects are missing planned VS assessments
# Planned: all subjects should have SYSBP at visits 1, 2, 3, 4, 5
subjects <- dm |>
  filter(!is.na(ARM), ARM != "Screen Failure") |>
  pull(USUBJID)

planned_visits <- 1:5

sbp_observed <- vs |>
  filter(VSTESTCD == "SYSBP",
         USUBJID %in% subjects,
         VISITNUM %in% planned_visits) |>
  select(USUBJID, VISITNUM, VSSTRESN)

# complete() fills in missing USUBJID × VISITNUM combinations
sbp_complete <- sbp_observed |>
  complete(
    USUBJID  = subjects,
    VISITNUM = planned_visits
  )

n_missing <- sum(is.na(sbp_complete$VSSTRESN))
cat("Missing planned SBP assessments:", n_missing, "\n")
cat("Out of:", nrow(sbp_complete), "planned assessments\n")
cat("Missingness rate:", round(100 * n_missing / nrow(sbp_complete), 1), "%\n")
```

---

## 7. fill() — Carrying Values Forward or Backward

`fill()` propagates non-missing values across rows — implementing LOCF and BOCF imputation strategies.

```{r}
#| label: fill-locf

# LOCF: Last Observation Carried Forward for SBP
sbp_locf <- sbp_complete |>
  arrange(USUBJID, VISITNUM) |>
  group_by(USUBJID) |>
  fill(VSSTRESN, .direction = "down") |>   # LOCF
  ungroup() |>
  rename(SBP_locf = VSSTRESN)

# BOCF: Baseline Observation Carried Backward
sbp_bocf <- sbp_complete |>
  arrange(USUBJID, VISITNUM) |>
  group_by(USUBJID) |>
  fill(VSSTRESN, .direction = "up") |>     # BOCF
  ungroup() |>
  rename(SBP_bocf = VSSTRESN)

cat("Remaining missing after LOCF:", sum(is.na(sbp_locf$SBP_locf)), "\n")
cat("Remaining missing after BOCF:", sum(is.na(sbp_bocf$SBP_bocf)), "\n")
```

::: {.callout-warning}
**Regulatory caution**: LOCF and BOCF are simple imputation approaches that were common in older trials but are no longer considered acceptable as the primary approach for handling missing data in regulatory submissions. Modern guidance (ICH E9 R1) recommends reference-based imputation or mixed models for repeated measures (MMRM). LOCF is useful for sensitivity analyses only.
:::

---

## 8. nest() and unnest() — List-Columns and Grouped Models

`nest()` stores sub-datasets as list-column elements, enabling per-group modelling workflows — fitting a separate model for each lab test, subgroup, or country.

### 8.1 Nested Linear Models by Lab Test

```{r}
#| label: nest-models

library(broom)

# Fit a linear time trend for each lab test
lab_trends <- lb |>
  filter(!is.na(LBSTRESN), !is.na(VISITNUM)) |>
  group_by(LBTESTCD) |>
  filter(n() >= 30) |>  # require at least 30 records
  nest() |>
  mutate(
    model  = map(data, \(d) lm(LBSTRESN ~ VISITNUM, data = d)),
    tidy   = map(model, broom::tidy),
    glance = map(model, broom::glance)
  )

# Extract slope (time trend) for each test
lab_slopes <- lab_trends |>
  unnest(tidy) |>
  filter(term == "VISITNUM") |>
  select(LBTESTCD, estimate, std.error, p.value) |>
  arrange(p.value) |>
  head(10)

cat("Lab tests with significant time trends (top 10):\n")
print(lab_slopes)
```

### 8.2 Unnesting for Reporting

```{r}
#| label: unnest-example

# Unnest model summaries for a clean R² table
lab_fit <- lab_trends |>
  unnest(glance) |>
  select(LBTESTCD, r.squared, adj.r.squared, p.value) |>
  arrange(desc(r.squared)) |>
  head(8)

cat("Model fit by lab test:\n")
print(lab_fit)
```

---

## 9. Full Workflow: SDTM Lab Data → ADaM ADLB-like Structure

Combining all `tidyr` and `dplyr` tools into a single ADLB-style preparation:

```{r}
#| label: adlb-workflow

# Build an ADLB-like dataset for ALT and AST
adlb <- lb |>
  filter(
    LBTESTCD %in% c("ALT", "AST"),
    !is.na(LBSTRESN),
    VISITNUM %in% 1:6
  ) |>
  select(USUBJID, LBTESTCD, VISITNUM, LBSTRESN, LBSTNRHI) |>
  # Add treatment arm from DM
  left_join(
    dm |> select(USUBJID, ARM) |>
      filter(!is.na(ARM), ARM != "Screen Failure"),
    by = "USUBJID"
  ) |>
  filter(!is.na(ARM)) |>
  # Compute ULN ratio
  mutate(
    ULN_ratio  = round(LBSTRESN / LBSTNRHI, 3),
    ULN_flag   = ULN_ratio > 3,   # >3× ULN is a common hepatotoxicity flag
    visit_label = paste0("Visit ", VISITNUM)
  ) |>
  # Pivot: one column per visit for each test
  arrange(USUBJID, LBTESTCD, VISITNUM)

cat("ADLB-like dataset:\n")
glimpse(adlb)

cat("\nSubjects with ALT >3× ULN at any visit:\n")
adlb |>
  filter(LBTESTCD == "ALT", ULN_flag) |>
  distinct(USUBJID, ARM) |>
  count(ARM)
```

---

## Summary

`tidyr` and `dplyr` together form the backbone of clinical data wrangling in R:

- **`pivot_longer()`** converts wide formats (e.g., one column per visit) into tidy long format required by most statistical models
- **`pivot_wider()`** produces wide formats needed for CFB calculations, listings, and regulatory tables
- **`separate()` / `unite()`** handle the messy ISO date strings and compound identifiers ubiquitous in SDTM
- **`complete()`** makes implicit missingness explicit — essential for identifying protocol deviations and unplanned missing assessments
- **`fill()`** implements LOCF/BOCF imputation (as sensitivity analysis only)
- **`nest()` / `unnest()`** enable scalable per-group modelling — running the same analysis across all lab parameters or subgroups with a single pipeline

> **Regulatory note**: Any reshaping or imputation applied to the data must be documented in the Statistical Analysis Plan and the dataset specification. The transformation logic should be captured in production-quality, version-controlled R code that can be reviewed and reproduced by statisticians, programmers, and regulators.
