
# Chapter 6: Exploratory Data Analysis (EDA): Understanding Your Data Before Analysis

## Learning Objectives

By the end of this chapter, you will:
- Understand why "look before you model" is critical in clinical research
- Compute and interpret summary statistics (mean, median, SD, IQR, quantiles)
- Assess distributional assumptions using histograms, Q-Q plots, and normality tests
- Detect and handle outliers using multiple methods
- Perform correlation analysis (Pearson, Spearman)
- Create publication-ready summary tables with gtsummary
- Apply EDA to real CDISC-formatted clinical data

## 6.1 Why Exploratory Data Analysis Matters

### "Look Before You Model"

In 2003, statistician Anscombe created four datasets, each with identical summary statistics:
- Mean of X: 9.0
- Mean of Y: 7.5
- Correlation: 0.82
- Regression line: y = 3 + 0.5x

Yet when plotted, the datasets told four completely different stories: one linear, one quadratic, one non-linear with one outlier, and one with a single influencer point.

**The lesson**: Summary statistics can hide the truth. Always visualize and explore your data first.

### Clinical Context: A Hypertension Trial

Imagine analyzing a 12-week randomized trial of 300 patients on three antihypertensive drugs. You plan a simple t-test to compare systolic BP reduction between arms.

Before running any statistical test, EDA should answer:

1. **Are the data normally distributed?** (If not, maybe use non-parametric tests)
2. **Are there outliers?** (A single patient with BP = 40 mmHg would skew results)
3. **Are baseline characteristics balanced?** (Randomization should achieve this, but verify)
4. **What's the missing data pattern?** (If 40% of Week 12 data is missing, results may be biased)
5. **Are assumptions met?** (Equal variances, independence of observations)

Missing just one of these checks could invalidate your entire analysis.

::: {.callout-important}
## FDA Guidance on Data Quality
The FDA's guidance on biostatistical principles emphasizes exploratory analysis to identify data quality issues *before* confirmatory analysis. This is not optional—it's expected in submissions.
:::

---

## 6.2 Summary Statistics

### Measures of Central Tendency and Spread

Let's define the statistics we'll compute:

**Central Tendency:**
- **Mean**: $$\bar{x} = \frac{1}{n}\sum_{i=1}^{n} x_i$$
- **Median**: Middle value when data are sorted
- **Mode**: Most frequently occurring value

**Spread:**
- **Standard Deviation**: $$SD = \sqrt{\frac{1}{n-1}\sum_{i=1}^{n}(x_i - \bar{x})^2}$$
- **Variance**: $$Var = SD^2$$
- **Interquartile Range**: IQR = Q3 - Q1 (25th to 75th percentile)
- **Range**: Max - Min

### When to Use Each Statistic

| Statistic | Use When | Not When |
|-----------|----------|----------|
| Mean | Data are normally distributed | Outliers present or skewed data |
| Median | Data are skewed or non-normal | You need a parametric test |
| SD | Data are normal, comparing to reference | You're describing a non-normal distribution |
| IQR | Data are skewed or have outliers | You want to compare to parametric tests |

---

## 6.3 Distribution Assessment

### Why Check Normality?

Many statistical tests (t-tests, ANOVA, linear regression) assume normality. If violated:
- Test results may be unreliable
- P-values may be incorrect
- Confidence intervals may be inaccurate

### Visual Assessment: Histogram

```{r}
# Load required packages
library(pharmaversesdtm)
library(dplyr)
library(tidyr)
library(ggplot2)

# Load vital signs data
data(vs)

# Extract systolic blood pressure at baseline
sbp_baseline <- vs %>%
  filter(VSTESTCD == "SYSBP", VISITNUM == 1) %>%
  pull(VSSTRESN) %>%
  as.numeric()

# Histogram with normal curve overlay
hist(sbp_baseline,
     main = "Distribution of Baseline Systolic Blood Pressure",
     xlab = "Systolic BP (mmHg)",
     ylab = "Frequency",
     breaks = 15,
     prob = TRUE)

# Overlay normal distribution curve
x <- seq(min(sbp_baseline, na.rm = TRUE),
         max(sbp_baseline, na.rm = TRUE),
         length = 100)
normal_curve <- dnorm(x, mean = mean(sbp_baseline, na.rm = TRUE),
                      sd = sd(sbp_baseline, na.rm = TRUE))
lines(x, normal_curve, col = "red", lwd = 2)
```

**Interpretation**: If the histogram roughly matches the red normal curve, data are approximately normal. Asymmetry (skewness) suggests non-normality.

### Q-Q Plot: Quantile-Quantile

The Q-Q plot compares observed quantiles to theoretical normal quantiles:

```{r}
# Q-Q plot
qqnorm(sbp_baseline,
       main = "Q-Q Plot: Systolic BP")
qqline(sbp_baseline, col = "red", lwd = 2)
```

**Interpretation**:
- Points closely follow the red line → approximately normal
- Points deviate from the line → non-normal
- "S" curve → data are skewed
- Extreme deviations at tails → heavy tails (outliers likely)

### Shapiro-Wilk Normality Test

The **Shapiro-Wilk test** tests the null hypothesis: "Data are normally distributed."

$$H_0: \text{Data come from a normal distribution}$$
$$H_A: \text{Data do not come from a normal distribution}$$

```{r}
# Shapiro-Wilk test
test_result <- shapiro.test(sbp_baseline)

cat("Shapiro-Wilk Test for Normality\n")
cat("Test Statistic W =", round(test_result$statistic, 4), "\n")
cat("P-value =", round(test_result$p.value, 4), "\n")

if (test_result$p.value < 0.05) {
  cat("Conclusion: Data are NOT normally distributed (p < 0.05)\n")
} else {
  cat("Conclusion: Data may be normally distributed (p >= 0.05)\n")
}
```

::: {.callout-tip}
## Sample Size and Normality
With large samples (n > 100), even trivial deviations from normality may give p < 0.05. Always combine the test with visual inspection. A histogram might show the data are "close enough" to normal for practical purposes.
:::

---

## 6.4 Outlier Detection

### Method 1: IQR Method (Tukey Fences)

The **Tukey fence** method flags outliers as:
- **Lower fence** = Q1 - 1.5 × IQR
- **Upper fence** = Q3 + 1.5 × IQR
- Any point beyond these fences is a potential outlier

```{r}
# Calculate quartiles and IQR
Q1 <- quantile(sbp_baseline, 0.25, na.rm = TRUE)
Q3 <- quantile(sbp_baseline, 0.75, na.rm = TRUE)
IQR <- Q3 - Q1

lower_fence <- Q1 - 1.5 * IQR
upper_fence <- Q3 + 1.5 * IQR

cat("Q1 (25th percentile) =", Q1, "\n")
cat("Q3 (75th percentile) =", Q3, "\n")
cat("IQR =", IQR, "\n")
cat("Lower Fence =", lower_fence, "\n")
cat("Upper Fence =", upper_fence, "\n\n")

# Identify outliers
outliers_iqr <- sbp_baseline[sbp_baseline < lower_fence | sbp_baseline > upper_fence]
cat("Number of outliers:", length(outliers_iqr), "\n")
cat("Outlier values:", outliers_iqr, "\n")
```

### Method 2: Z-Score Method

The **Z-score** measures how many standard deviations a point is from the mean:

$$Z = \frac{x_i - \bar{x}}{SD}$$

Points with |Z| > 3 are often considered outliers; |Z| > 2.5 are flagged for review.

```{r}
# Calculate Z-scores
sbp_df <- data.frame(SBP = sbp_baseline)
sbp_df <- sbp_df %>%
  mutate(
    Z_score = (SBP - mean(SBP, na.rm = TRUE)) / sd(SBP, na.rm = TRUE),
    Outlier_Z3 = abs(Z_score) > 3,
    Outlier_Z25 = abs(Z_score) > 2.5
  )

# Display outliers
outliers_z <- sbp_df %>%
  filter(Outlier_Z3 | Outlier_Z25) %>%
  arrange(desc(abs(Z_score)))

cat("Outliers by Z-score (|Z| > 2.5):\n")
print(outliers_z)

cat("\nNumber of extreme outliers (|Z| > 3):", sum(sbp_df$Outlier_Z3, na.rm = TRUE), "\n")
```

### Boxplot Visualization

```{r}
# Boxplot shows quartiles and outliers visually
boxplot(sbp_baseline,
        main = "Systolic BP: Boxplot with Outliers",
        ylab = "SBP (mmHg)")

# Add mean point
points(1, mean(sbp_baseline, na.rm = TRUE), pch = 4, col = "red", cex = 2)
legend("topright", "Mean", pch = 4, col = "red")
```

::: {.callout-warning}
## Outliers Require Investigation, Not Deletion
Never automatically delete outliers. First investigate:
- **Data entry error?** Verify in source documents
- **Measurement error?** Was the device calibrated?
- **Real biological extreme?** Some patients do have very low or high BP
- **Eligibility violation?** Does this patient meet criteria?

Only after investigation should you decide to exclude or transform.
:::

---

## 6.5 Correlation Analysis

### Pearson Correlation (Normal Data)

For normally distributed data, **Pearson's r** measures linear association:

$$r = \frac{\sum_{i=1}^{n}(x_i - \bar{x})(y_i - \bar{y})}{\sqrt{\sum_{i=1}^{n}(x_i - \bar{x})^2 \sum_{i=1}^{n}(y_i - \bar{y})^2}}$$

Where: r ranges from -1 (perfect negative) to +1 (perfect positive); r = 0 (no linear relationship)

```{r}
# Load lab data
data(lb)

# Extract baseline lab values (first non-NA observation per subject × test)
# SDTM lb does not carry LBBLFL; we derive the baseline as the earliest visit
baseline_labs <- lb %>%
  filter(!is.na(LBSTRESN)) %>%
  group_by(USUBJID, LBTESTCD) %>%
  slice_min(VISITNUM, n = 1, with_ties = FALSE) %>%
  ungroup() %>%
  select(USUBJID, LBTESTCD, LBSTRESN) %>%
  pivot_wider(
    names_from  = LBTESTCD,
    values_from = LBSTRESN,
    values_fn   = mean
  ) %>%
  # Use tests available in pharmaversesdtm: ALB, CREAT, ALT
  select(USUBJID, any_of(c("ALB", "CREAT", "ALT", "AST")))

# Compute Pearson correlations (pairwise — robust to sparse data)
lab_mat <- baseline_labs %>% select(-USUBJID) %>% select(where(is.numeric))

if (ncol(lab_mat) >= 2 && nrow(lab_mat) >= 5) {
  cor_matrix <- cor(lab_mat, use = "pairwise.complete.obs", method = "pearson")
  cat("Pearson Correlation Matrix (Baseline Labs)\n")
  print(round(cor_matrix, 3))

  # P-value for the first available pair
  col_pairs <- combn(colnames(lab_mat), 2)
  c1 <- col_pairs[1, 1]; c2 <- col_pairs[2, 1]
  ct <- cor.test(lab_mat[[c1]], lab_mat[[c2]], method = "pearson")
  cat(sprintf("\nPearson p-values (%s vs %s): r = %.3f, p = %.4f\n",
              c1, c2, ct$estimate, ct$p.value))
} else {
  cat("Insufficient data for correlation analysis.\n")
}
```

### Spearman Correlation (Non-Normal Data)

When data are non-normal or you're measuring monotonic (not necessarily linear) association, use **Spearman's rank correlation**:

$$\rho = 1 - \frac{6 \sum_{i=1}^{n} d_i^2}{n(n^2 - 1)}$$

Where: $d_i$ = difference in ranks for each pair

```{r}
# Spearman correlation (rank-based, robust to non-normality)
lab_mat2 <- baseline_labs %>% select(-USUBJID) %>% select(where(is.numeric))
spearman_matrix <- cor(lab_mat2, use = "pairwise.complete.obs", method = "spearman")

cat("\nSpearman Correlation Matrix (Baseline Labs)\n")
print(round(spearman_matrix, 3))

# Pearson vs. Spearman comparison
cat("\nComparison:\n")
if (exists("cor_matrix") &&
    "ALB" %in% rownames(cor_matrix) && "CREAT" %in% rownames(cor_matrix) &&
    "ALB" %in% rownames(spearman_matrix) && "CREAT" %in% rownames(spearman_matrix)) {
  cat("Pearson r (ALB vs CREAT) =",   round(cor_matrix["ALB",     "CREAT"], 3), "\n")
  cat("Spearman rho (ALB vs CREAT) =", round(spearman_matrix["ALB","CREAT"], 3), "\n")
} else {
  cat("Correlation comparison: see matrices above\n")
}
```

---

## 6.6 Cross-Tabulation and Categorical Analysis

### Table Function

```{r}
# Load demographics
data(dm)

# Simple 1-way frequency table
treatment_table <- table(dm$ARM)
cat("Treatment Arm Distribution:\n")
print(treatment_table)

# 2-way crosstab: Treatment by Sex
crosstab <- table(dm$ARM, dm$SEX)
cat("\nTreatment Arm by Sex:\n")
print(crosstab)

# Add margins (row and column totals)
addmargins(crosstab)
```

### Proportions and Percentages

```{r}
# Row proportions (% within each treatment arm)
row_prop <- prop.table(crosstab, margin = 1) * 100
cat("\nPercentage by Sex within each Arm:\n")
print(round(row_prop, 1))

# Column proportions (% within each sex)
col_prop <- prop.table(crosstab, margin = 2) * 100
cat("\nPercentage by Arm within each Sex:\n")
print(round(col_prop, 1))
```

---

## 6.7 Publication-Ready Tables with gtsummary

The **gtsummary** package creates FDA-submission-quality tables with minimal code.

```{r}
# Install if needed: install.packages("gtsummary")
library(gtsummary)

# Load data
data(dm)
data(vs)
data(lb)

# Merge demographics with baseline vitals and labs
baseline_data <- dm %>%
  left_join(
    vs %>%
      filter(VSTESTCD == "SYSBP", VISITNUM == 1) %>%
      select(USUBJID, VSSTRESN) %>%
      mutate(VSSTRESN = as.numeric(VSSTRESN)) %>%
      rename(SBP_Baseline = VSSTRESN),
    by = "USUBJID"
  ) %>%
  left_join(
    lb %>%
      filter(LBTESTCD == "CREAT", !is.na(LBSTRESN)) %>%
      group_by(USUBJID) %>%
      slice_min(VISITNUM, n = 1, with_ties = FALSE) %>%
      ungroup() %>%
      select(USUBJID, LBSTRESN) %>%
      rename(CREATININE = LBSTRESN),
    by = "USUBJID"
  ) %>%
  mutate(
    # Convert age to categorical for table
    AGE_GR = case_when(
      AGE < 40 ~ "<40",
      AGE >= 40 & AGE < 65 ~ "40-64",
      AGE >= 65 ~ ">=65"
    )
  ) %>%
  select(ARM, AGE, AGE_GR, SEX, RACE, SBP_Baseline, CREATININE)

# Create baseline characteristics table
baseline_table <- baseline_data %>%
  tbl_summary(
    by = ARM,  # Stratify by treatment arm
    statistic = list(
      all_continuous() ~ "{mean} ({sd})",  # Mean (SD) for continuous
      all_categorical() ~ "{n} ({p}%)"  # N (%) for categorical
    ),
    digits = list(
      all_continuous() ~ 1,
      all_categorical() ~ 0
    ),
    label = list(
      AGE ~ "Age (years)",
      SEX ~ "Sex",
      RACE ~ "Race",
      SBP_Baseline ~ "Baseline Systolic BP (mmHg)",
      CREATININE ~ "Baseline Creatinine (mg/dL)"
    )
  ) %>%
  # Add overall column
  add_overall() %>%
  # Add p-values for comparison
  add_p() %>%
  # Modify header
  modify_header(label ~ "**Characteristic**") %>%
  # Convert to standard table format for display
  as_gt()

print(baseline_table)
```

### Customized Summary Statistics

```{r}
# Example: Summary by treatment arm with custom statistics
baseline_data %>%
  select(ARM, AGE, SBP_Baseline, CREATININE) %>%
  tbl_summary(
    by = ARM,
    statistic = list(
      AGE ~ "{median} ({p25}, {p75})",  # Median and IQR
      SBP_Baseline ~ "{mean} ± {sd}",   # Mean ± SD
      CREATININE ~ "{min}-{max}"        # Range
    ),
    missing = "no"  # Don't show missing counts
  ) %>%
  add_overall() %>%
  as_gt()
```

---

## 6.8 Complete EDA Workflow: Hypertension Trial

Here's a comprehensive EDA for our hypertension trial example:

```{r}
# ============================================
# COMPLETE EDA WORKFLOW
# Hypertension Trial - Baseline Analysis
# ============================================

library(pharmaversesdtm)
library(dplyr)
library(tidyr)
library(gtsummary)

# Load SDTM domains
data(dm)
data(vs)
data(lb)
data(ae)

# ========== STEP 1: DEMOGRAPHIC SUMMARY ==========
cat("=== DEMOGRAPHIC SUMMARY ===\n\n")

demo_summary <- dm %>%
  select(ARM, AGE, SEX, RACE, ETHNIC) %>%
  tbl_summary(
    by = ARM,
    statistic = list(
      AGE ~ "{mean} ({sd})",
      all_categorical() ~ "{n} ({p}%)"
    ),
    digits = list(AGE ~ 1, everything() ~ 0),
    missing = "no"
  ) %>%
  add_overall() %>%
  add_p()

print(demo_summary)

# ========== STEP 2: BASELINE VITALS ==========
cat("\n=== BASELINE VITALS BY ARM ===\n\n")

baseline_vitals <- vs %>%
  filter(VISITNUM == 1, VSTESTCD %in% c("SYSBP", "DIABP", "PULSE"),
         !is.na(VSSTRESN)) %>%
  select(USUBJID, VSTESTCD, VSSTRESN) %>%
  mutate(VSSTRESN = as.numeric(VSSTRESN)) %>%
  # Average if multiple records per subject × parameter (avoids list columns)
  group_by(USUBJID, VSTESTCD) %>%
  summarise(VSSTRESN = mean(VSSTRESN, na.rm = TRUE), .groups = "drop") %>%
  pivot_wider(
    names_from  = VSTESTCD,
    values_from = VSSTRESN
  ) %>%
  left_join(dm %>% select(USUBJID, ARM), by = "USUBJID") %>%
  select(ARM, any_of(c("SYSBP", "DIABP", "PULSE")))

baseline_vitals %>%
  tbl_summary(
    by = ARM,
    statistic = list(
      all_continuous() ~ "{mean} ({sd}) [{min}-{max}]"
    ),
    digits = list(all_continuous() ~ 1),
    label = list(
      SYSBP ~ "Systolic BP (mmHg)",
      DIABP ~ "Diastolic BP (mmHg)",
      PULSE ~ "Heart Rate (bpm)"
    ),
    missing = "no"
  ) %>%
  add_overall() %>%
  add_p()

# ========== STEP 3: DISTRIBUTION CHECKS ==========
cat("\n=== NORMALITY TESTS (SHAPIRO-WILK) ===\n\n")

sbp_by_arm <- vs %>%
  filter(VISITNUM == 1, VSTESTCD == "SYSBP") %>%
  select(USUBJID, VSSTRESN) %>%
  mutate(VSSTRESN = as.numeric(VSSTRESN)) %>%
  left_join(dm %>% select(USUBJID, ARM), by = "USUBJID")

for (arm in unique(sbp_by_arm$ARM)) {
  sbp_values <- sbp_by_arm %>%
    filter(ARM == arm) %>%
    pull(VSSTRESN)

  test <- shapiro.test(sbp_values)
  cat(sprintf("%s: W = %.4f, p-value = %.4f\n",
              arm, test$statistic, test$p.value))
}

# ========== STEP 4: OUTLIER DETECTION ==========
cat("\n=== OUTLIER DETECTION (IQR METHOD) ===\n\n")

sbp_all <- vs %>%
  filter(VISITNUM == 1, VSTESTCD == "SYSBP") %>%
  pull(VSSTRESN) %>%
  as.numeric()

Q1 <- quantile(sbp_all, 0.25, na.rm = TRUE)
Q3 <- quantile(sbp_all, 0.75, na.rm = TRUE)
IQR_val <- Q3 - Q1
lower_fence <- Q1 - 1.5 * IQR_val
upper_fence <- Q3 + 1.5 * IQR_val

outlier_count <- sum(sbp_all < lower_fence | sbp_all > upper_fence, na.rm = TRUE)
cat(sprintf("Fences: [%.1f, %.1f]\n", lower_fence, upper_fence))
cat(sprintf("Number of outliers: %d\n", outlier_count))

# ========== STEP 5: BASELINE LAB CORRELATIONS ==========
cat("\n=== BASELINE LAB CORRELATIONS ===\n\n")

# Select a handful of well-populated tests to ensure enough complete pairs
common_tests <- c("ALT", "AST", "CREAT", "ALB", "BILI")

baseline_labs_corr <- lb %>%
  filter(!is.na(LBSTRESN), LBTESTCD %in% common_tests) %>%
  group_by(USUBJID, LBTESTCD) %>%
  slice_min(VISITNUM, n = 1, with_ties = FALSE) %>%
  ungroup() %>%
  select(USUBJID, LBTESTCD, LBSTRESN) %>%
  pivot_wider(
    names_from  = LBTESTCD,
    values_from = LBSTRESN,
    values_fn   = mean
  ) %>%
  select(-USUBJID) %>%
  select(where(is.numeric))

# pairwise.complete.obs uses all available pairs — robust to sparse data
if (ncol(baseline_labs_corr) >= 2) {
  cor_matrix <- cor(baseline_labs_corr, use = "pairwise.complete.obs", method = "pearson")
  cat("Pearson Correlation Matrix (key liver/renal labs):\n")
  print(round(cor_matrix, 2))
} else {
  cat("Insufficient lab columns for correlation analysis.\n")
}

# ========== STEP 6: ADVERSE EVENTS OVERVIEW ==========
cat("\n=== ADVERSE EVENTS OVERVIEW ===\n\n")

ae_summary <- ae %>%
  left_join(dm %>% select(USUBJID, ARM), by = "USUBJID") %>%
  group_by(ARM) %>%
  summarise(
    Total_AE = n(),
    Subjects_with_AE = n_distinct(USUBJID),
    .groups = 'drop'
  ) %>%
  mutate(AE_per_Subject = round(Total_AE / Subjects_with_AE, 2))

print(ae_summary)

cat("\n=== MOST COMMON ADVERSE EVENTS ===\n\n")
ae %>%
  left_join(dm %>% select(USUBJID, ARM), by = "USUBJID") %>%
  group_by(ARM, AETERM) %>%
  summarise(
    N = n(),
    .groups = 'drop'
  ) %>%
  arrange(ARM, desc(N)) %>%
  head(15) %>%
  print()

cat("\n=== EDA COMPLETE ===\n")
```

---

## 6.9 Key Decisions from EDA

Before moving to formal analysis, EDA should inform your analytical approach:

| EDA Finding | Implication |
|-------------|-------------|
| Data non-normal, outliers present | Use non-parametric tests (Mann-Whitney, Kruskal-Wallis) |
| Baseline characteristics unbalanced | Consider stratified analysis or ANCOVA |
| High missing data (>20%) | Use multiple imputation or sensitivity analysis |
| Extreme outliers present | Investigate and possibly exclude; document reasoning |
| Variables highly correlated (r > 0.9) | May have multicollinearity; consider dropping one |

---

## Summary: Why EDA is Non-Negotiable

- **Detects data quality issues** before they invalidate analysis
- **Informs methodological choices** (parametric vs. non-parametric)
- **Builds credibility** with regulators (shows you understand your data)
- **Prevents analytical errors** (outliers, missing data patterns)
- **Produces publication tables** (baseline characteristics, summary statistics)

In clinical trials, EDA is not a "nice-to-have"—it's essential due diligence.

---

## Exercises

### Beginner Exercises

**6.1 Summary Statistics**

Using the pharmaversesdtm::vs dataset, filter to VSTESTCD == "SYSBP" and VISITNUM == 1 (baseline), then calculate:
1. Mean
2. Median
3. Standard Deviation
4. IQR
5. Range (min to max)

**Answer template:**
```{r}
data(vs)
sbp <- vs %>%
  filter(VSTESTCD == "SYSBP", VISITNUM == 1) %>%
  pull(VSSTRESN) %>%
  as.numeric()

mean(sbp, na.rm = TRUE)     # Mean
median(sbp, na.rm = TRUE)   # Median
sd(sbp, na.rm = TRUE)       # SD
IQR(sbp, na.rm = TRUE)      # IQR
range(sbp, na.rm = TRUE)    # Range
```

**6.2 Outlier Detection**

For the same baseline SBP data:
1. Calculate the Tukey fences (lower and upper)
2. Identify how many values fall outside these fences
3. What are the outlier values?

**Hint**: Q1 - 1.5*IQR and Q3 + 1.5*IQR

**6.3 Normality Assessment**

1. Create a histogram of baseline SBP
2. Create a Q-Q plot
3. Run Shapiro-Wilk test
4. Based on all three, is the data approximately normal? Why or why not?

---

### Advanced Exercises

**6.4 Baseline Characteristics Table**

Using pharmaversesdtm data:
1. Merge dm, vs (baseline SYSBP), and lb (baseline CREATININE)
2. Create a gtsummary table stratified by ARM showing:
   - Age (mean ± SD)
   - Sex (N %)
   - Race (N %)
   - Baseline SBP (mean ± SD)
   - Baseline Creatinine (mean ± SD)
3. Add an overall column and p-values for comparison
4. Format for publication

**Expected output**: A professional baseline characteristics table suitable for a regulatory submission.

**6.5 Correlation Analysis and Missing Data Investigation**

Using pharmaversesdtm::lb:
1. Create a matrix of baseline lab values (filter to LBBLFL == "Y")
2. Calculate both Pearson and Spearman correlations for all numeric labs
3. Identify which pairs are highly correlated (|r| > 0.7)
4. Investigate the missing data pattern:
   - What percentage of baseline labs are missing for each test?
   - Are missings correlated with treatment arm?
5. Recommend which labs to include in analysis based on data completeness

---

## Key Takeaways

- **EDA precedes all formal analysis**: No exceptions
- **Look for violations of assumptions**: Normality, outliers, missing data
- **Use multiple methods**: Visual (histogram, boxplot, Q-Q) and statistical (Shapiro-Wilk)
- **gtsummary creates publication tables** with minimal effort
- **Document all findings** for regulatory compliance
- **Outliers need investigation, not deletion**: Understand before deciding

