
# Chapter 7: Data Visualization in Clinical Research: From Grammar of Graphics to Publication-Quality Figures

## Learning Objectives

By the end of this chapter, you will:
- Understand the grammar of graphics and how ggplot2 implements it
- Master core geoms for clinical data (bar, boxplot, line, point, density)
- Create specialized clinical plots (forest plots, waterfall plots, spaghetti plots)
- Design Kaplan-Meier curves and adverse event visualizations
- Apply publication-quality themes and color palettes
- Save figures in formats required by regulatory submissions (PDF, PNG, vector)
- Build reproducible, publication-ready visualizations

## 7.1 The Grammar of Graphics

### Why We Visualize

A picture is worth a thousand numbers. In clinical trials, data visualization:
- **Communicates results** to regulators, clinicians, and patients
- **Detects patterns** that tables miss
- **Identifies outliers** and data quality issues
- **Tells a story** about treatment efficacy and safety

The challenge: How do we make plots that are both **beautiful** and **informative**?

### The Grammar of Graphics Framework

**ggplot2**, created by Hadley Wickham, implements a grammar of graphics where every plot is built from layers:

$$\text{Plot} = \text{Data} + \text{Aesthetics} + \text{Geoms} + \text{Facets} + \text{Themes}$$

**Components:**

1. **Data**: The dataframe containing variables
2. **Aesthetics (aes)**: Mapping variables to visual properties (x, y, color, size, fill)
3. **Geoms**: Geometric objects representing data (points, lines, bars, boxplots)
4. **Facets**: Small multiples, split by grouping variables
5. **Themes**: Overall visual styling (fonts, colors, background)

### Simple Example: Building Layer by Layer

```{r}
library(ggplot2)
library(dplyr)
library(tidyr)
library(pharmaversesdtm)

# Load data
data(vs)
data(dm)

# Prepare data: baseline vitals by treatment arm
baseline_vitals <- vs %>%
  filter(VISITNUM == 1, VSTESTCD == "SYSBP") %>%
  select(USUBJID, VSSTRESN) %>%
  mutate(VSSTRESN = as.numeric(VSSTRESN)) %>%
  left_join(dm %>% select(USUBJID, ARM), by = "USUBJID") %>%
  rename(SBP = VSSTRESN, Treatment = ARM)

# Layer 1: Initialize plot with data and aesthetics
p <- ggplot(baseline_vitals, aes(x = Treatment, y = SBP))

# Layer 2: Add points (geom_point)
p <- p + geom_point(alpha = 0.5)

# Layer 3: Add boxplot
p <- p + geom_boxplot(fill = NA, color = "blue", linewidth = 1)

# Layer 4: Add mean point in red
p <- p + stat_summary(fun = mean, geom = "point", color = "red", size = 4)

# Layer 5: Customize labels and theme
p <- p +
  labs(
    title = "Baseline Systolic Blood Pressure by Treatment Arm",
    x = "Treatment Arm",
    y = "Systolic BP (mmHg)"
  ) +
  theme_minimal() +
  theme(plot.title = element_text(size = 14, face = "bold"))

print(p)
```

---

## 7.2 Core Geoms for Clinical Data

### Geom 1: geom_point - Scatter Plot

Shows individual observations, useful for exploring relationships.

```{r}
# Scatter: Baseline vs. Week 4 systolic BP
baseline <- vs %>%
  filter(VISITNUM == 1, VSTESTCD == "SYSBP") %>%
  select(USUBJID, VSSTRESN) %>%
  rename(SBP_Baseline = VSSTRESN)

week4 <- vs %>%
  filter(VISITNUM == 1, VSTESTCD == "SYSBP") %>%
  select(USUBJID, VSSTRESN) %>%
  rename(SBP_Week4 = VSSTRESN)

response_data <- baseline %>%
  left_join(week4, by = "USUBJID") %>%
  left_join(dm %>% select(USUBJID, ARM), by = "USUBJID") %>%
  mutate(
    SBP_Baseline = as.numeric(SBP_Baseline),
    SBP_Week4 = as.numeric(SBP_Week4),
    Change = SBP_Week4 - SBP_Baseline
  ) %>%
  filter(!is.na(SBP_Baseline) & !is.na(SBP_Week4))

ggplot(response_data, aes(x = SBP_Baseline, y = SBP_Week4, color = ARM)) +
  geom_point(size = 3, alpha = 0.6) +
  geom_abline(intercept = 0, slope = 1, linetype = "dashed", color = "gray") +
  labs(
    title = "Baseline vs. Week 4 Systolic BP",
    x = "Baseline SBP (mmHg)",
    y = "Week 4 SBP (mmHg)",
    color = "Treatment Arm"
  ) +
  theme_minimal() +
  theme(legend.position = "bottom")
```

**Interpretation**: Points above the diagonal line indicate BP increase; below = decrease.

### Geom 2: geom_line - Longitudinal Trends

Shows change over time, essential for tracking treatment response.

```{r}
# Prepare longitudinal data
longit_data <- vs %>%
  filter(VSTESTCD == "SYSBP") %>%
  select(USUBJID, VISITNUM, VSSTRESN) %>%
  mutate(VSSTRESN = as.numeric(VSSTRESN)) %>%
  left_join(dm %>% select(USUBJID, ARM), by = "USUBJID") %>%
  filter(!is.na(VSSTRESN)) %>%
  rename(Visit = VISITNUM, SBP = VSSTRESN, Treatment = ARM)

# Mean line plot
mean_trend <- longit_data %>%
  group_by(Visit, Treatment) %>%
  summarise(Mean_SBP = mean(SBP, na.rm = TRUE), .groups = 'drop')

ggplot(longit_data, aes(x = Visit, y = SBP, group = USUBJID, color = Treatment)) +
  # Individual spaghetti lines (faint)
  geom_line(alpha = 0.2, color = "gray") +
  geom_point(alpha = 0.2, size = 1) +
  # Mean trend (bold)
  geom_line(data = mean_trend, aes(y = Mean_SBP, group = Treatment),
            color = "black", linewidth = 1.2, linetype = "solid") +
  geom_point(data = mean_trend, aes(y = Mean_SBP, group = Treatment), color = "black", size = 3) +
  facet_wrap(~ Treatment) +
  labs(
    title = "Systolic BP Over Time: Spaghetti Plot with Mean Trend",
    x = "Visit Number",
    y = "Systolic BP (mmHg)"
  ) +
  theme_minimal() +
  theme(legend.position = "none")
```

**Interpretation**: This "spaghetti plot" shows individual trajectories (gray) and mean response (black). Helps identify early responders vs. non-responders.

### Geom 3: geom_bar - Categorical Comparisons

Compares counts or proportions across groups.

```{r}
# Load adverse events
data(ae)

# Adverse event frequency by arm
ae_freq <- ae %>%
  left_join(dm %>% select(USUBJID, ARM), by = "USUBJID") %>%
  group_by(ARM, AETERM) %>%
  summarise(Count = n(), .groups = 'drop') %>%
  # Top 8 AEs overall
  arrange(desc(Count)) %>%
  head(8)

ggplot(ae_freq, aes(x = reorder(AETERM, -Count), y = Count, fill = ARM)) +
  geom_bar(stat = "identity", position = "dodge") +
  coord_flip() +
  labs(
    title = "Most Common Adverse Events by Treatment Arm",
    x = "Adverse Event",
    y = "Count",
    fill = "Treatment"
  ) +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))
```

### Geom 4: geom_boxplot - Distribution Comparison

Shows median, quartiles, and outliers.

```{r}
# Boxplot: Baseline vitals by treatment
baseline_vitals_summary <- vs %>%
  filter(VISITNUM == 1, VSTESTCD %in% c("SYSBP", "DIABP", "PULSE")) %>%
  select(USUBJID, VSTESTCD, VSSTRESN) %>%
  mutate(VSSTRESN = as.numeric(VSSTRESN)) %>%
  left_join(dm %>% select(USUBJID, ARM), by = "USUBJID") %>%
  rename(Test = VSTESTCD, Value = VSSTRESN, Treatment = ARM)

ggplot(baseline_vitals_summary, aes(x = Treatment, y = Value, fill = Treatment)) +
  geom_boxplot(alpha = 0.7) +
  geom_point(position = position_jitter(width = 0.1), alpha = 0.3, size = 1) +
  facet_wrap(~ Test, scales = "free_y") +
  labs(
    title = "Baseline Vital Signs by Treatment Arm",
    y = "Value",
    fill = "Treatment"
  ) +
  theme_minimal() +
  theme(legend.position = "bottom")
```

### Geom 5: geom_histogram & geom_density - Distributions

Visualizes frequency distributions, key for assessing normality.

```{r}
# Distribution of baseline BP
sbp_dist <- vs %>%
  filter(VISITNUM == 1, VSTESTCD == "SYSBP") %>%
  select(VSSTRESN) %>%
  mutate(VSSTRESN = as.numeric(VSSTRESN))

# Histogram with density overlay
ggplot(sbp_dist, aes(x = VSSTRESN)) +
  geom_histogram(aes(y = after_stat(density)), bins = 20, fill = "skyblue", color = "black") +
  geom_density(color = "red", linewidth = 1) +
  stat_function(
    fun = dnorm,
    args = list(mean = mean(sbp_dist$VSSTRESN, na.rm = TRUE),
                sd = sd(sbp_dist$VSSTRESN, na.rm = TRUE)),
    color = "green", linewidth = 1, linetype = "dashed"
  ) +
  labs(
    title = "Distribution of Baseline Systolic BP",
    subtitle = "Red = observed density, Green = normal distribution",
    x = "SBP (mmHg)",
    y = "Density"
  ) +
  theme_minimal()
```

---

## 7.3 Specialized Clinical Plots

### Forest Plot: Hazard Ratios and Odds Ratios

Forest plots display effect sizes with confidence intervals, common in regulatory submissions.

```{r}
# Simulated subgroup analysis: HR for BP reduction by patient subgroup
subgroup_effects <- tibble::tribble(
  ~Subgroup, ~Comparison, ~HR, ~Lower_CI, ~Upper_CI, ~P_Value,
  "Overall", "Treatment vs. Control", 0.72, 0.58, 0.89, 0.002,
  "Age <65", "Treatment vs. Control", 0.68, 0.50, 0.93, 0.015,
  "Age >=65", "Treatment vs. Control", 0.76, 0.55, 1.05, 0.102,
  "Female", "Treatment vs. Control", 0.70, 0.50, 0.98, 0.037,
  "Male", "Treatment vs. Control", 0.74, 0.55, 1.00, 0.049
)

ggplot(subgroup_effects, aes(y = reorder(Subgroup, HR), x = HR)) +
  # Confidence interval as horizontal error bar
  geom_errorbarh(aes(xmin = Lower_CI, xmax = Upper_CI), height = 0.2, linewidth = 1) +
  # Point estimate
  geom_point(size = 4, color = "darkblue") +
  # Vertical line at HR = 1 (no effect)
  geom_vline(xintercept = 1, linetype = "dashed", color = "red", linewidth = 1) +
  # Scale x-axis on log scale
  scale_x_log10() +
  labs(
    title = "Forest Plot: Hazard Ratios for BP Response",
    subtitle = "Points and bars show HR and 95% CI",
    x = "Hazard Ratio (log scale)",
    y = "Subgroup"
  ) +
  theme_minimal() +
  theme(
    plot.title = element_text(size = 14, face = "bold"),
    axis.text.y = element_text(size = 11)
  )
```

**Interpretation**: HR < 1 = treatment reduces outcome; CI not crossing 1.0 = significant at p < 0.05.

### Waterfall Plot: Individual Response in Oncology

Shows each patient's response, useful in oncology (tumor size change) and hypertension trials (BP response).

```{r}
# Simulated: Individual BP response (Week 4 - Baseline)
set.seed(42)
n_patients <- 30
patient_response <- tibble(
  Patient_ID = 1:n_patients,
  Response = sort(rnorm(n_patients, mean = -15, sd = 10)),  # Sorted for waterfall
  Responder = Response < -10,  # Responders have >10 mmHg drop
  Treatment = rep(c("Drug A", "Drug B", "Drug C"), each = 10)
)

ggplot(patient_response, aes(x = reorder(Patient_ID, Response), y = Response, fill = Responder)) +
  geom_bar(stat = "identity") +
  geom_hline(yintercept = 0, linetype = "solid", color = "black", size = 1) +
  geom_hline(yintercept = -10, linetype = "dashed", color = "red", size = 0.8) +
  facet_wrap(~ Treatment) +
  scale_fill_manual(
    values = c("TRUE" = "darkgreen", "FALSE" = "lightcoral"),
    labels = c("TRUE" = "Responder (>10 mmHg drop)", "FALSE" = "Non-Responder")
  ) +
  labs(
    title = "Waterfall Plot: Individual BP Response by Treatment",
    x = "Patient (sorted by response)",
    y = "Change from Baseline (mmHg)",
    fill = "Status"
  ) +
  coord_flip() +
  theme_minimal() +
  theme(
    axis.text.x = element_text(size = 8),
    legend.position = "bottom"
  )
```

**Interpretation**: Bars below zero = BP reduction (desired). Dashed line at -10 shows response threshold.

### Adverse Event Butterfly Plot

Displays adverse events by system organ class, comparing treatment arms.

```{r}
# Adverse event counts by SOC and arm
ae_summary <- ae %>%
  left_join(dm %>% select(USUBJID, ARM), by = "USUBJID") %>%
  group_by(AESOC, ARM) %>%
  summarise(AE_Count = n(), .groups = 'drop') %>%
  pivot_wider(
    names_from = ARM,
    values_from = AE_Count,
    values_fill = 0
  ) %>%
  # Calculate "balance" for plotting
  mutate(
    Arm1 = -.data[[names(.)[2]]],  # Negative for left side
    Arm2 = .data[[names(.)[3]]]    # Positive for right side
  ) %>%
  select(AESOC, Arm1, Arm2) %>%
  rename(Drug_A = Arm1, Drug_B = Arm2) %>%
  arrange(desc(abs(Drug_A)))

# Prepare for ggplot (long format)
ae_butterfly <- ae_summary %>%
  pivot_longer(
    cols = c(Drug_A, Drug_B),
    names_to = "Treatment",
    values_to = "Count"
  ) %>%
  mutate(
    AESOC = factor(AESOC, levels = ae_summary$AESOC),
    Count_display = abs(Count)
  )

ggplot(ae_butterfly, aes(x = AESOC, y = Count_display, fill = Treatment)) +
  geom_bar(stat = "identity", position = "identity", alpha = 0.7) +
  # Flip so Drug_A (negative) plots left, Drug_B (positive) plots right
  geom_bar(
    data = filter(ae_butterfly, Treatment == "Drug_A"),
    aes(y = -Count_display),
    stat = "identity", fill = "steelblue", alpha = 0.7
  ) +
  geom_bar(
    data = filter(ae_butterfly, Treatment == "Drug_B"),
    aes(y = Count_display),
    stat = "identity", fill = "coral", alpha = 0.7
  ) +
  geom_hline(yintercept = 0, color = "black", size = 1) +
  coord_flip() +
  labs(
    title = "Adverse Events Butterfly Plot: Drug A vs. Drug B",
    x = "System Organ Class",
    y = "Number of Adverse Events"
  ) +
  theme_minimal() +
  theme(legend.position = "bottom")
```

### Kaplan-Meier Curve: Time-to-Event Analysis

Shows probability of event-free survival over time (detailed in Chapter 12; shown here for preview).

```{r}
# Preview: Simulated time-to-event data
library(survival)

# Simulate survival data
set.seed(123)
n <- 200
time_to_event <- data.frame(
  PatientID = 1:n,
  Time = rexp(n, rate = 0.05),
  Event = rbinom(n, 1, 0.6),
  Treatment = rep(c("Treatment", "Control"), each = n/2)
)

# Fit Kaplan-Meier curve
km_fit <- survfit(Surv(Time, Event) ~ Treatment, data = time_to_event)

# Plot using base R (ggsurvfit requires full survival workflow)
plot(km_fit,
     main = "Kaplan-Meier Curve: Time-to-Event",
     xlab = "Time (days)",
     ylab = "Probability of Event-Free Survival",
     col = c("blue", "red"),
     lwd = 2)
legend("topright",
       c("Treatment", "Control"),
       col = c("blue", "red"),
       lwd = 2)
```

---

## 7.4 Themes and Styling

### Publication-Quality Themes

```{r}
# Create a base plot
p_base <- ggplot(baseline_vitals, aes(x = Treatment, y = SBP, fill = Treatment)) +
  geom_boxplot(alpha = 0.7) +
  labs(
    title = "Baseline Systolic Blood Pressure",
    y = "SBP (mmHg)"
  )

# Apply different themes
print(p_base + theme_minimal())
print(p_base + theme_classic())
print(p_base + theme_bw())
```

### Custom Theme for Clinical Submissions

```{r}
# Create a custom theme suitable for FDA submissions
theme_clinical <- function() {
  theme_bw() +
  theme(
    # Title and labels
    plot.title = element_text(
      size = 14, face = "bold", hjust = 0, margin = margin(b = 10)
    ),
    plot.subtitle = element_text(
      size = 11, hjust = 0, margin = margin(b = 10), color = "gray40"
    ),
    axis.title = element_text(size = 11, face = "bold"),
    axis.text = element_text(size = 10),

    # Legend
    legend.position = "bottom",
    legend.title = element_text(size = 11, face = "bold"),
    legend.text = element_text(size = 10),

    # Grid
    panel.grid.minor = element_blank(),
    panel.grid.major = element_line(color = "gray90", size = 0.3),

    # Facets
    strip.text = element_text(size = 10, face = "bold"),
    strip.background = element_rect(fill = "gray95", color = NA)
  )
}

# Apply custom theme
ggplot(baseline_vitals_summary, aes(x = Treatment, y = Value, fill = Treatment)) +
  geom_boxplot(alpha = 0.7) +
  facet_wrap(~ Test, scales = "free_y") +
  labs(
    title = "Baseline Vital Signs by Treatment Arm",
    subtitle = "Individual data points overlaid",
    y = "Value",
    fill = "Treatment Arm"
  ) +
  geom_point(position = position_jitter(width = 0.1), alpha = 0.4, size = 1) +
  theme_clinical()
```

### Clinical Color Palettes

```{r}
# Define clinical-appropriate color palette
clinical_colors <- list(
  "three_arm" = c("#0173B2", "#DE8F05", "#CC78BC", "#029E73"),  # Blue, Orange, Purple, Green
  "binary" = c("#56B4E9", "#E69F00"),  # Light blue, Orange
  "safety" = c("#009E73", "#F0E442", "#CC79A7")  # Green, Yellow, Pink
)

# Apply to plot
ggplot(baseline_vitals, aes(x = Treatment, y = SBP, fill = Treatment)) +
  geom_boxplot(alpha = 0.8) +
  scale_fill_manual(values = clinical_colors$three_arm) +
  labs(
    title = "Baseline SBP by Treatment",
    y = "SBP (mmHg)",
    x = ""
  ) +
  theme_clinical() +
  theme(legend.position = "none")
```

---

## 7.5 Saving Publication-Quality Figures

### Using ggsave()

```{r}
#| eval: false
# Create a comprehensive figure
fig_sbp <- ggplot(baseline_vitals, aes(x = Treatment, y = SBP, fill = Treatment)) +
  geom_boxplot(alpha = 0.8) +
  geom_point(position = position_jitter(width = 0.2), alpha = 0.4, size = 2) +
  scale_fill_manual(values = clinical_colors$three_arm) +
  labs(
    title = "Baseline Systolic Blood Pressure by Treatment Arm",
    y = "Systolic Blood Pressure (mmHg)",
    x = "Treatment Arm"
  ) +
  theme_clinical()

# Save as PDF (vector, scalable)
ggsave(
  filename = "figure_sbp_baseline.pdf",
  plot = fig_sbp,
  width = 7,
  height = 5,
  units = "in",
  dpi = 300,
  device = "pdf"
)

# Save as PNG (raster, high resolution)
ggsave(
  filename = "figure_sbp_baseline.png",
  plot = fig_sbp,
  width = 7,
  height = 5,
  units = "in",
  dpi = 300,
  device = "png"
)

cat("Figures saved:\n")
cat("- figure_sbp_baseline.pdf (vector format)\n")
cat("- figure_sbp_baseline.png (raster, 300 DPI)\n")
```

### Settings for Regulatory Submissions

```{r}
#| eval: false
# FDA/EMA submission requirements:
# - Minimum 300 DPI for raster images
# - Vector formats (PDF, EPS) preferred
# - Figure size: typically 6-7 inches wide, 4-5 inches tall
# - Readable font sizes: ≥10 pt

# Best practice: Save as PDF for submission
ggsave_clinical <- function(plot, filename, width = 7, height = 5) {
  # Save PDF (vector)
  ggsave(
    filename = filename,
    plot = plot,
    width = width,
    height = height,
    units = "in",
    dpi = 300,
    device = "pdf"
  )
  # Also save PNG for presentations
  ggsave(
    filename = sub("\\.pdf$", ".png", filename),
    plot = plot,
    width = width,
    height = height,
    units = "in",
    dpi = 300,
    device = "png"
  )
}

# Usage:
# ggsave_clinical(fig_sbp, "figure_sbp_baseline.pdf")
```

---

## 7.6 Complete Visualization Workflow

Here's a comprehensive example integrating all concepts:

```{r}
# ============================================
# COMPLETE VISUALIZATION WORKFLOW
# Hypertension Trial - Publication Figures
# ============================================

library(ggplot2)
library(dplyr)
library(tidyr)
library(pharmaversesdtm)

# Load data
data(dm); data(vs); data(ae); data(lb)

# ========== FIGURE 1: Baseline Demographics ==========

baseline_char <- dm %>%
  filter(ARM != "Screen Failure") %>%
  select(USUBJID, ARM, AGE, SEX) %>%
  mutate(
    Age_Group = case_when(
      AGE < 40 ~ "<40",
      AGE >= 40 & AGE < 65 ~ "40-64",
      AGE >= 65 ~ ">=65"
    )
  )

fig1_demographics <- ggplot(baseline_char, aes(x = Age_Group, fill = ARM)) +
  geom_bar(position = "dodge", alpha = 0.8) +
  scale_fill_manual(values = clinical_colors$three_arm) +
  labs(
    title = "Figure 1: Baseline Demographics by Age Group",
    x = "Age Group (years)",
    y = "Number of Subjects",
    fill = "Treatment Arm"
  ) +
  theme_clinical()

print(fig1_demographics)

# ========== FIGURE 2: Baseline Vital Signs ==========

baseline_vitals_complete <- vs %>%
  filter(VISITNUM == 1, VSTESTCD %in% c("SYSBP", "DIABP", "PULSE")) %>%
  select(USUBJID, VSTESTCD, VSSTRESN) %>%
  mutate(VSSTRESN = as.numeric(VSSTRESN)) %>%
  left_join(dm %>% select(USUBJID, ARM), by = "USUBJID") %>%
  rename(
    Test = VSTESTCD,
    Value = VSSTRESN,
    Treatment = ARM
  ) %>%
  mutate(
    Test = factor(Test,
                  levels = c("SYSBP", "DIABP", "PULSE"),
                  labels = c("Systolic BP (mmHg)",
                            "Diastolic BP (mmHg)",
                            "Heart Rate (bpm)"))
  )

fig2_vitals <- ggplot(baseline_vitals_complete, aes(x = Treatment, y = Value, fill = Treatment)) +
  geom_boxplot(alpha = 0.7, width = 0.6) +
  geom_point(position = position_jitter(width = 0.15), alpha = 0.3, size = 1.5) +
  facet_wrap(~ Test, scales = "free_y", nrow = 1) +
  scale_fill_manual(values = clinical_colors$three_arm) +
  labs(
    title = "Figure 2: Baseline Vital Signs by Treatment Arm",
    y = "Value",
    x = "Treatment Arm"
  ) +
  theme_clinical() +
  theme(legend.position = "none")

print(fig2_vitals)

# ========== FIGURE 3: Longitudinal BP Trend (Spaghetti) ==========

longit_sbp <- vs %>%
  filter(VSTESTCD == "SYSBP", !is.na(VSSTRESN)) %>%
  select(USUBJID, VISITNUM, VSSTRESN) %>%
  mutate(VSSTRESN = as.numeric(VSSTRESN)) %>%
  left_join(dm %>% select(USUBJID, ARM), by = "USUBJID") %>%
  rename(Visit = VISITNUM, SBP = VSSTRESN, Treatment = ARM)

mean_sbp <- longit_sbp %>%
  group_by(Visit, Treatment) %>%
  summarise(Mean_SBP = mean(SBP, na.rm = TRUE), .groups = 'drop')

fig3_trend <- ggplot(longit_sbp, aes(x = Visit, y = SBP, group = USUBJID, color = Treatment)) +
  geom_line(alpha = 0.15, color = "gray50") +
  geom_point(alpha = 0.1, size = 0.5) +
  geom_line(data = mean_sbp, aes(y = Mean_SBP, group = Treatment, linetype = "Mean"),
            color = "black", size = 1) +
  geom_point(data = mean_sbp, aes(y = Mean_SBP, group = Treatment), color = "black", size = 3) +
  facet_wrap(~ Treatment) +
  scale_linetype_manual(values = "solid") +
  labs(
    title = "Figure 3: Systolic BP Over Time (Spaghetti Plot)",
    x = "Visit Number",
    y = "Systolic BP (mmHg)",
    linetype = ""
  ) +
  theme_clinical() +
  theme(legend.position = "bottom")

print(fig3_trend)

# ========== FIGURE 4: Adverse Events ==========

ae_top <- ae %>%
  left_join(dm %>% select(USUBJID, ARM), by = "USUBJID") %>%
  group_by(ARM, AETERM) %>%
  summarise(Count = n(), .groups = 'drop') %>%
  arrange(desc(Count)) %>%
  head(10)

fig4_ae <- ggplot(ae_top, aes(x = reorder(AETERM, -Count), y = Count, fill = ARM)) +
  geom_bar(stat = "identity", position = "dodge", alpha = 0.8) +
  scale_fill_manual(values = clinical_colors$three_arm) +
  coord_flip() +
  labs(
    title = "Figure 4: Most Common Adverse Events by Treatment",
    x = "Adverse Event",
    y = "Count",
    fill = "Treatment Arm"
  ) +
  theme_clinical()

print(fig4_ae)

# ========== SAVE ALL FIGURES ==========

# ggsave_clinical(fig1_demographics, "fig1_demographics.pdf", width = 6, height = 4)
# ggsave_clinical(fig2_vitals, "fig2_vitals.pdf", width = 10, height = 4)
# ggsave_clinical(fig3_trend, "fig3_trend.pdf", width = 8, height = 5)
# ggsave_clinical(fig4_ae, "fig4_ae.pdf", width = 7, height = 5)

cat("\n=== Visualization workflow complete ===\n")
cat("All figures ready for regulatory submission\n")
```

---

## 7.7 Best Practices for Clinical Visualization

### Do's

- **Use clear, descriptive titles**: Include figure number and outcome
- **Label axes with units**: "Systolic BP (mmHg)", not just "SBP"
- **Include n or sample size**: Especially for plots with facets
- **Use color thoughtfully**: Accessible to color-blind readers
- **Provide legends**: Don't make readers guess what colors mean
- **Save in vector format**: PDF preferred for submissions

### Don'ts

- **Don't use 3D plots**: Hard to interpret, no information gain
- **Don't use dual axes**: Confusing and often misleading
- **Don't exceed 3-4 colors**: Too many colors = hard to distinguish
- **Don't forget grid lines**: Helps readers estimate values
- **Don't use decorative elements**: Junk that doesn't convey data

---

## Summary: Visualization as Communication

Great clinical visualizations:
1. **Tell a story**: Communicate the main finding clearly
2. **Use appropriate geoms**: Match the data type to the geometric object
3. **Follow publication standards**: FDA/EMA-ready formatting
4. **Are reproducible**: R code documents exactly how the figure was made
5. **Are beautiful**: Professional appearance builds credibility

In the next chapters, we'll apply these visualization skills to specific analyses (survival analysis, regression, meta-analysis).

---

## Exercises

### Beginner Exercises

**7.1 Scatter Plot with Treatment Colors**

Using pharmaversesdtm::vs data:
1. Create scatter plot of baseline vs. Week 4 SBP
2. Color points by treatment arm (use clinical_colors$three_arm)
3. Add diagonal reference line (y = x)
4. Label axes with units
5. Add a title

Expected: A clear scatter plot showing treatment comparisons.

**7.2 Distribution Comparison**

Using pharmaversesdtm::vs:
1. Create histograms of baseline SBP by treatment arm (using facet_wrap)
2. Overlay density curves
3. Apply theme_clinical()
4. Save as PDF

Expected: Three side-by-side histograms showing distribution differences.

**7.3 Boxplot with Summary Statistics**

Using pharmaversesdtm::dm and pharmaversesdtm::vs:
1. Merge baseline SBP with demographics
2. Create boxplot of SBP by treatment arm
3. Overlay individual points with jitter
4. Add mean point in red
5. Apply custom colors

Expected: Publication-quality boxplot suitable for a submission.

---

### Advanced Exercises

**7.4 Spaghetti Plot with Mean Trend**

Using pharmaversesdtm::vs:
1. Filter to VSTESTCD == "SYSBP" for all visits
2. Plot individual subject trajectories (faint gray lines)
3. Overlay mean trend line (bold black) for each treatment arm
4. Use facet_wrap by treatment
5. Apply theme_clinical()
6. Interpret: Which treatment arm shows best BP control?

Expected: Spaghetti plot showing individual and mean responses.

**7.5 Adverse Event Analysis Dashboard**

Create a 2x2 panel figure:
- **Panel A** (top-left): Bar chart of top 5 AEs by treatment arm
- **Panel B** (top-right): Butterfly plot of AEs by system organ class
- **Panel C** (bottom-left): Proportion of subjects with any AE by arm
- **Panel D** (bottom-right): Boxplot of AE severity by arm

Combine using `patchwork` package:
```{r}
#| eval: false
library(patchwork)
(fig_A + fig_B) / (fig_C + fig_D) +
  plot_annotation(title = "Adverse Event Summary")
```

Expected: Comprehensive safety dashboard ready for submission.

---

## Key Takeaways

- **ggplot2 grammar**: Data + Aesthetics + Geoms + Facets + Themes
- **Clinical geoms**: boxplot, point, line, bar, density, histogram
- **Specialized plots**: Forest, waterfall, spaghetti, butterfly, K-M curves
- **Publication quality**: Vector format (PDF), ≥300 DPI, readable fonts
- **Custom themes**: Develop consistent clinical branding across figures
- **Reproducibility**: Always save R code alongside saved figures
