# Bayesian Methods in Clinical Trials


## Introduction

For over a century, clinical trials have been analyzed using **frequentist statistics**—computing p-values, confidence intervals, and power, all conditional on the null hypothesis being true. But frequentist inference cannot directly answer the question clinicians really want to know: *Given my data, what is the probability that my drug actually works?*

**Bayesian methods** answer this question directly. They combine prior knowledge with observed data to compute the **posterior probability** that a hypothesis is true. This chapter introduces Bayesian thinking, explains Bayes' theorem, covers conjugate models, introduces MCMC for complex problems, and demonstrates how to implement Bayesian clinical trial analyses in R.

## 18.1 Frequentist vs. Bayesian Philosophy

### The Fundamental Difference

**Frequentist inference:**
- Parameters are fixed but unknown
- Data are random; imagine repeated sampling under the same conditions
- P-value = Probability of seeing data *as or more extreme* than observed, *if the null hypothesis were true*
- Cannot say "probability the drug works" — only "probability of data given null"

**Bayesian inference:**
- Parameters are random (follow a probability distribution)
- Data are fixed (what we observe)
- Posterior = Probability of the parameter *given the observed data*
- Can directly answer: "What is P(drug is effective | my data)?"

### A Concrete Example

A clinical trial shows a new pain medication reduces pain score by 3 points vs. placebo. A frequentist might report: "p = 0.04; we reject the null hypothesis at the 0.05 level."

A Bayesian might report: "Given our prior knowledge about similar drugs and the data from this trial, there is an 87% probability the true pain reduction is greater than 2 points (clinically meaningful)."

## 18.2 Bayes' Theorem

### The Foundation

Bayes' theorem expresses how to update beliefs in light of new evidence:

$$P(\theta | \text{data}) = \frac{P(\text{data} | \theta) \cdot P(\theta)}{P(\text{data})}$$

Breaking this down:

- **$P(\theta | \text{data})$** = **Posterior**: Probability of the parameter (e.g., treatment effect) given the observed data. This is what we want.
- **$P(\text{data} | \theta)$** = **Likelihood**: Probability of observing the data under a specific parameter value. This comes from the statistical model (e.g., binomial, normal).
- **$P(\theta)$** = **Prior**: Probability distribution for the parameter *before* seeing the data. This encodes prior knowledge or assumptions.
- **$P(\text{data})$** = **Marginal likelihood** or **evidence**: Probability of the data averaging over all possible parameter values. This is a normalizing constant.

### Simplified Form

Often written as:

$$\text{Posterior} \propto \text{Likelihood} \times \text{Prior}$$

The proportionality symbol (∝) reminds us that the posterior is proportional to likelihood times prior, up to a normalizing constant.

## 18.3 Prior Distributions

### Three Types of Priors

**1. Informative Priors**
- Incorporate substantial prior knowledge
- Example: Beta(10, 2) for a treatment response rate, based on historical data showing typically high response
- Advantages: More efficient (smaller sample size), reflects real knowledge
- Disadvantages: Subjective, can be controversial

**2. Weakly Informative Priors**
- Provide minimal constraint but prevent extreme values
- Example: Normal(0, 10) for a treatment effect (mean 0, SD 10)
- Advantages: Balance between objectivity and efficiency
- Disadvantages: Still involves some choice

**3. Non-Informative (Flat) Priors**
- Assign equal probability to all parameter values
- Example: Uniform(0, 1) for a response rate
- Advantages: Aim for "objectivity" (letting data dominate)
- Disadvantages: Actually still impose assumptions; can lead to improper posteriors

### The Beta Distribution for Proportions

For binary outcomes (response = yes/no), the **Beta distribution** is natural:

$$\theta \sim \text{Beta}(\alpha, \beta)$$

where $\alpha$ and $\beta$ are shape parameters:
- **Mean**: $\frac{\alpha}{\alpha + \beta}$
- **Variance**: $\frac{\alpha \beta}{(\alpha+\beta)^2(\alpha+\beta+1)}$

**Interpretation:**
- $\alpha$ = number of "successes" (pseudo-observations)
- $\beta$ = number of "failures" (pseudo-observations)

**Examples:**
- Beta(1, 1): Uniform prior (non-informative)
- Beta(5, 5): Centered at 0.5, moderately informative
- Beta(10, 2): High response rate expected (mean = 10/12 ≈ 0.83)
- Beta(0.5, 0.5): U-shaped, bimodal prior (rare)

## 18.4 Conjugate Priors: Beta-Binomial Model

### The Conjugate Family

When the prior and likelihood are **conjugate**, the posterior has a closed-form solution (no MCMC needed). The **Beta-Binomial** model is the simplest:

- **Prior**: $\theta \sim \text{Beta}(\alpha, \beta)$
- **Likelihood**: $X \sim \text{Binomial}(n, \theta)$ where $X$ = number of successes
- **Posterior**: $\theta | X \sim \text{Beta}(\alpha + X, \beta + n - X)$

### Why This Is Elegant

The posterior is also Beta with updated shape parameters:
- New successes: $\alpha + X$
- New failures: $\beta + n - X$

Updating is intuitive: add observed successes to $\alpha$, add observed failures to $\beta$.

### Interpretation of Posterior

Once we have the posterior Beta distribution, we can compute:
- **Posterior mean**: $\frac{\alpha + X}{\alpha + \beta + n}$ (estimates the true response rate)
- **Posterior credible interval**: Quantiles of the Beta distribution (e.g., 2.5th and 97.5th percentiles)
- **Probability of clinical benefit**: $P(\theta > \theta_0 | X)$ where $\theta_0$ is a threshold (e.g., 0.50 response rate)

## 18.5 The Normal-Normal Model

For continuous outcomes, when both the prior and likelihood are normal:

- **Prior**: $\mu \sim \text{Normal}(\mu_0, \tau_0^2)$
- **Likelihood**: $X_i \sim \text{Normal}(\mu, \sigma^2)$ (known variance $\sigma^2$)
- **Posterior**: $\mu | X \sim \text{Normal}(\mu_1, \tau_1^2)$

where:
$$\mu_1 = \frac{\mu_0 / \tau_0^2 + \sum X_i / \sigma^2}{1/\tau_0^2 + n/\sigma^2}, \quad \tau_1^{-2} = \tau_0^{-2} + \frac{n}{\sigma^2}$$

The posterior mean is a weighted average of the prior mean and sample mean, with weights reflecting precision (inverse variance).

## 18.6 Markov Chain Monte Carlo (MCMC)

### When Conjugacy Fails

Many realistic clinical trial models (hierarchical models, regression with multiple covariates, mixture models) do not have conjugate priors. For these, we use **Markov Chain Monte Carlo (MCMC)**.

### Conceptual Overview

MCMC generates a sequence of posterior samples by:
1. Starting at an initial parameter value
2. Proposing a new value
3. Accepting or rejecting based on the ratio of posterior probabilities
4. Repeating thousands of times to build a representative sample from the posterior

The sequence forms a "chain" whose long-run empirical distribution approximates the true posterior.

### Popular MCMC Algorithms

- **Metropolis-Hastings**: Symmetric proposal distribution
- **Gibbs Sampling**: Special case for models with conjugate conditionals
- **Hamiltonian MC (HMC)**: Uses gradient information; very efficient but complex
- **No-U-Turn Sampler (NUTS)**: Adaptive HMC; implemented in Stan and rstanarm

## 18.7 Convergence Diagnostics

After running MCMC, we must verify that the chain has converged to the posterior.

### Trace Plots

A trace plot shows the parameter value at each iteration. Good traces look like "white noise" (no trends or long excursions).

### The Gelman-Rubin Statistic (R-hat)

Run multiple MCMC chains (e.g., 4) in parallel. The R-hat compares within-chain and between-chain variance:

$$\hat{R} = \sqrt{\frac{\text{Var}_{\text{between}}}{\text{Var}_{\text{within}}}}$$

- $\hat{R} \approx 1.00$ = chains have converged (chains indistinguishable)
- $\hat{R} > 1.05$ = convergence questionable; run longer
- $\hat{R} > 1.10$ = convergence problematic; reassess model or priors

## 18.8 Credible Intervals vs. Confidence Intervals

- **Credible Interval (Bayesian)**: Range of parameter values containing 95% of the posterior probability mass. Interpretation: "Given the data, there is a 95% probability the true parameter lies in this interval."

- **Confidence Interval (Frequentist)**: Constructed so that 95% of hypothetical repeated samples would produce an interval containing the true parameter. Interpretation: "If we repeated the study many times, 95% of such intervals would contain the true value."

Credible intervals are more intuitive for clinicians but require specification of prior.

## 18.9 Practical Analysis in R

### Example 1: Phase II Oncology Trial (Beta-Binomial Analysis)

A Phase II trial of a new immunotherapy evaluates response rate. Historical data suggest prior response rates in this cancer type range from 20% to 50%.

```{r}
library(tidyverse)

# Data: 12 out of 30 patients responded
n_patients <- 30
n_responses <- 12

# Prior: Beta(2, 8) — represents belief that response rate is around 20-30%
# (mean = 2/(2+8) = 0.20; concentrated on lower rates)
alpha_prior <- 2
beta_prior <- 8

# Posterior: Beta(2 + 12, 8 + 30 - 12) = Beta(14, 26)
alpha_post <- alpha_prior + n_responses
beta_post <- beta_prior + (n_patients - n_responses)

cat("=== Phase II Oncology Trial: Bayesian Response Rate Analysis ===\n\n")
cat("Prior: Beta(", alpha_prior, ",", beta_prior, ")\n")
cat("Prior mean:", round(alpha_prior / (alpha_prior + beta_prior), 3), "\n\n")

cat("Data: ", n_responses, " responses out of ", n_patients, " patients\n\n")

cat("Posterior: Beta(", alpha_post, ",", beta_post, ")\n")

# Posterior mean (best point estimate)
post_mean <- alpha_post / (alpha_post + beta_post)
cat("Posterior mean:", round(post_mean, 3), "\n")

# Posterior standard deviation
post_sd <- sqrt((alpha_post * beta_post) / ((alpha_post + beta_post)^2 * (alpha_post + beta_post + 1)))
cat("Posterior SD:", round(post_sd, 3), "\n\n")

# Posterior credible interval (95%)
ci_lower <- qbeta(0.025, alpha_post, beta_post)
ci_upper <- qbeta(0.975, alpha_post, beta_post)

cat("95% Credible Interval: [", round(ci_lower, 3), ", ", round(ci_upper, 3), "]\n\n")

# Probability of clinically relevant response (e.g., >30%)
prob_clinical_benefit <- 1 - pbeta(0.30, alpha_post, beta_post)
cat("P(response rate > 30% | data):", round(prob_clinical_benefit, 3), "\n")
```

### Visualizing Prior, Likelihood, and Posterior

```{r}
# Create a range of theta values
theta <- seq(0, 1, by = 0.01)

# Prior density
prior_density <- dbeta(theta, alpha_prior, beta_prior)

# Likelihood (proportional to Binomial pmf)
# L(theta) ∝ theta^12 * (1-theta)^18
likelihood_density <- theta^n_responses * (1 - theta)^(n_patients - n_responses)
# Normalize for plotting
likelihood_density <- likelihood_density / max(likelihood_density) * max(prior_density)

# Posterior density
posterior_density <- dbeta(theta, alpha_post, beta_post)

# Create data frame for plotting
density_df <- data.frame(
  theta = rep(theta, 3),
  density = c(prior_density, likelihood_density, posterior_density),
  component = rep(c("Prior", "Likelihood (scaled)", "Posterior"), each = length(theta))
)

ggplot(density_df, aes(x = theta, y = density, fill = component, color = component)) +
  geom_area(alpha = 0.3, outline.type = "full") +
  geom_line(size = 1) +
  scale_color_manual(values = c("Prior" = "blue", "Likelihood (scaled)" = "orange", "Posterior" = "red")) +
  scale_fill_manual(values = c("Prior" = "blue", "Likelihood (scaled)" = "orange", "Posterior" = "red")) +
  annotate("vline", xintercept = post_mean, linetype = "dashed", color = "red", size = 1) +
  annotate("text", x = post_mean + 0.05, y = max(posterior_density) * 0.9,
           label = paste("Posterior Mean =", round(post_mean, 3)), size = 4) +
  labs(
    title = "Bayesian Analysis: Prior × Likelihood → Posterior",
    subtitle = "Phase II Oncology Trial Response Rate",
    x = "Response Rate (θ)", y = "Density",
    caption = "The posterior (red) represents the updated belief after observing 12/30 responses."
  ) +
  theme_minimal() +
  theme(
    plot.title = element_text(face = "bold", size = 13),
    legend.position = "topright"
  )

# Plot just the posterior with credible interval
posterior_plot_df <- data.frame(theta = theta, density = posterior_density)

ggplot(posterior_plot_df, aes(x = theta, y = density)) +
  geom_area(fill = "steelblue", alpha = 0.4) +
  geom_line(color = "steelblue", size = 1) +
  geom_vline(xintercept = ci_lower, linetype = "dashed", color = "red", size = 1) +
  geom_vline(xintercept = ci_upper, linetype = "dashed", color = "red", size = 1) +
  geom_vline(xintercept = post_mean, linetype = "solid", color = "darkblue", size = 1) +
  annotate("text", x = post_mean, y = max(posterior_density) * 0.85,
           label = paste("Mean:", round(post_mean, 3)), size = 4, color = "darkblue") +
  annotate("text", x = ci_lower, y = max(posterior_density) * 0.70,
           label = round(ci_lower, 3), size = 3.5, color = "red") +
  annotate("text", x = ci_upper, y = max(posterior_density) * 0.70,
           label = round(ci_upper, 3), size = 3.5, color = "red") +
  labs(
    title = "Posterior Distribution of Response Rate",
    subtitle = paste("95% Credible Interval: [", round(ci_lower, 3), ", ", round(ci_upper, 3), "]"),
    x = "Response Rate (θ)", y = "Posterior Density"
  ) +
  theme_minimal() +
  theme(plot.title = element_text(face = "bold", size = 13))
```

### Example 2: Bayesian Logistic Regression with rstanarm

For more complex models (e.g., multiple covariates, hierarchical structures), we use MCMC. The `rstanarm` package provides a user-friendly interface to Stan.

```{r}
#| eval: false
if (requireNamespace("rstanarm", quietly = TRUE)) {
  library(rstanarm)
}

# Create a simulated clinical trial dataset
# Outcome: response (1) vs. no response (0)
# Predictor 1: treatment group (0 = control, 1 = intervention)
# Predictor 2: baseline severity score (0-100)

set.seed(42)

n_total <- 200
treatment <- rep(c(0, 1), each = n_total / 2)
baseline_severity <- rnorm(n_total, mean = 50, sd = 15)

# True model: log-odds = -2 + 1.5 * treatment - 0.02 * baseline_severity
true_intercept <- -2
true_treatment_effect <- 1.5
true_severity_effect <- -0.02

linear_predictor <- true_intercept + true_treatment_effect * treatment + true_severity_effect * baseline_severity
response_prob <- plogis(linear_predictor)  # inverse logit
response <- rbinom(n_total, size = 1, prob = response_prob)

clinical_data <- data.frame(
  response = response,
  treatment = factor(treatment, labels = c("Control", "Intervention")),
  baseline_severity = baseline_severity
)

# Fit Bayesian logistic regression using rstanarm
# Priors:
# - intercept: Normal(0, 2.5) — weakly informative
# - coefficients: Normal(0, 1) — centered at 0, weak effect
# - uses QR decomposition for stability

bayesian_logistic <- stan_glm(
  response ~ treatment + baseline_severity,
  data = clinical_data,
  family = binomial(link = "logit"),
  prior = normal(location = 0, scale = 1),
  prior_intercept = normal(location = 0, scale = 2.5),
  chains = 4,
  iter = 2000,  # 2000 iterations per chain
  warmup = 1000,  # first 1000 are burn-in
  refresh = 0  # suppress iteration printing
)

# Summarize results
cat("=== Bayesian Logistic Regression Results ===\n")
summary(bayesian_logistic)
```

### Extracting and Visualizing Posterior Samples

```{r}
#| eval: false
# Extract posterior samples for treatment effect
posterior_samples <- as.data.frame(bayesian_logistic)

# Transform log-odds to odds ratio
posterior_odds_ratio <- exp(posterior_samples$treatmentIntervention)

cat("Posterior Summary for Treatment Effect (Odds Ratio):\n")
cat("  Mean:", round(mean(posterior_odds_ratio), 3), "\n")
cat("  Median:", round(median(posterior_odds_ratio), 3), "\n")
cat("  95% Credible Interval: [",
    round(quantile(posterior_odds_ratio, 0.025), 3), ", ",
    round(quantile(posterior_odds_ratio, 0.975), 3), "]\n\n")

# Probability that treatment is beneficial (OR > 1)
prob_beneficial <- mean(posterior_odds_ratio > 1)
cat("P(OR > 1 | data) = P(treatment beneficial) =", round(prob_beneficial, 3), "\n\n")

# Visualize posterior distribution
posterior_or_df <- data.frame(odds_ratio = posterior_odds_ratio)

ggplot(posterior_or_df, aes(x = odds_ratio)) +
  geom_histogram(bins = 40, fill = "steelblue", alpha = 0.7, color = "white") +
  geom_vline(xintercept = 1, linetype = "dashed", color = "red", size = 1) +
  geom_vline(xintercept = mean(posterior_odds_ratio), linetype = "solid", color = "darkblue", size = 1) +
  annotate("text", x = mean(posterior_odds_ratio) + 0.15, y = max(table(round(posterior_odds_ratio, 2)))/2,
           label = paste("Mean OR =", round(mean(posterior_odds_ratio), 3)), size = 4) +
  labs(
    title = "Posterior Distribution of Treatment Odds Ratio",
    subtitle = "Bayesian Logistic Regression: Intervention vs. Control",
    x = "Odds Ratio (Intervention / Control)",
    y = "Count (MCMC Samples)"
  ) +
  theme_minimal() +
  theme(plot.title = element_text(face = "bold", size = 13))
```

### Convergence Diagnostics

```{r}
#| eval: false
# Check R-hat (Gelman-Rubin statistic)
cat("R-hat values (should be < 1.05 for convergence):\n")
print(bayesian_logistic$diagnostics)

# Trace plots for visual inspection
plot(bayesian_logistic, ask = FALSE)
```

## 18.10 Bayesian Clinical Trial Designs

### Interim Analyses and Predictive Probability

Bayesian designs enable **adaptive trials** with interim analyses. At an interim look:

1. Compute posterior probability of efficacy based on observed data
2. Compute **predictive probability**: P(efficacy at final analysis | current data)
3. Possible decisions: stop for efficacy, stop for futility, continue enrollment

### Example: Predictive Probability Stopping Rule

Stop for efficacy if P(future response rate > 40% | current data) > 90%.

```{r}
# Simple illustration: predict response rate at final analysis
# Current data: 8 responses out of 20 patients
# Assume 40 more patients will be enrolled

n_current <- 20
r_current <- 8
n_future <- 40

# Posterior after current data (conjugate Beta update)
alpha_current <- 1 + r_current
beta_current <- 1 + (n_current - r_current)

# Prior for future patients is the current posterior
# Predictive distribution: what response rates could we observe in future patients?

# Simulate: draw 10,000 posterior samples of theta, then for each, simulate future responses
set.seed(42)

posterior_theta_samples <- rbeta(10000, alpha_current, beta_current)
simulated_future_responses <- rbinom(10000, size = n_future, prob = posterior_theta_samples)

# Final response rate if we continue and see median future responses
median_future <- median(simulated_future_responses)
total_future_responses <- r_current + simulated_future_responses
total_patients <- n_current + n_future
final_response_rates <- total_future_responses / total_patients

# Predictive probability of >40% response at final analysis
pred_prob_efficacy <- mean(final_response_rates > 0.40)

cat("Interim Analysis Predictive Probability:\n")
cat("  Current data: ", r_current, "/", n_current, " responses\n")
cat("  Projected: ", median_future, " additional responses in next ", n_future, " patients\n")
cat("  Projected final response rate: ", round(median(final_response_rates), 3), "\n")
cat("  P(final response rate > 40% | current data, continued enrollment) =", round(pred_prob_efficacy, 3), "\n\n")

if (isTRUE(pred_prob_efficacy > 0.90)) {
  cat("DECISION: Stop for efficacy (predictive probability > 90%)\n")
} else if (isTRUE(pred_prob_efficacy < 0.10)) {
  cat("DECISION: Stop for futility (predictive probability < 10%)\n")
} else {
  cat("DECISION: Continue enrollment\n")
}
```

## 18.11 Exercises

### Beginner Exercises

**Exercise 18.1:** A historical study of a pain medication found a response rate of 55% (95% CI: 45%–65%). Based on this, specify a Beta prior distribution for the response rate in a new trial. (Hint: match the mean and approximate the 95% CI range.)

**Exercise 18.2:** In a new Phase II trial of a cancer drug, 7 out of 20 patients respond. Using a uniform prior Beta(1,1), compute:
- Posterior mean response rate
- 95% credible interval
- Probability that response rate exceeds 25%

**Exercise 18.3:** Explain in plain language why the posterior is a weighted average of the prior and the data in the Normal-Normal model. Which weight (prior vs. data) dominates when the prior variance is very large?

### Advanced Exercises

**Exercise 18.4:** Using the `bayesian_logistic` model from Example 2, extract the posterior samples for the baseline_severity effect. Compute:
- Posterior mean and 95% credible interval for the log-odds ratio
- Posterior probability that increasing baseline severity by 10 points *decreases* the log-odds of response by at least 0.15
- Create a trace plot and verify convergence (R-hat < 1.05)

**Exercise 18.5:** Design a Bayesian adaptive trial for a Phase IIb antihypertensive drug:
- Primary endpoint: systolic blood pressure reduction at 12 weeks
- Interim analysis at 50 patients, final analysis at 100 patients
- Efficacy stopping rule: If P(mean reduction > 10 mmHg | interim data) > 85%, stop and declare efficacy
- Futility stopping rule: If P(mean reduction > 10 mmHg | interim data) < 20%, stop for futility

Propose informative priors based on historical trials and sketch a decision tree for the adaptive design.

