---
title: "Chapter 21: Reproducible Reporting with R Markdown and Quarto"
subtitle: "Automating Clinical Study Reports"
---

## 21.1 Introduction to Reproducible Reporting

The **replication crisis** in biomedical research has exposed a fundamental vulnerability: many published findings cannot be reproduced. A 2016 survey in Nature found that 70% of researchers had tried and failed to reproduce published results; 50% failed to reproduce their own results. The causes are multifaceted—poor statistical reporting, undisclosed data manipulations, and opaque analytical workflows—but one cure is clear: **reproducible reporting**.

In clinical research, reproducibility carries regulatory weight. The FDA's **21 CFR Part 11** rule ("Electronic Records; Electronic Signatures") requires that computer systems used to generate clinical data and reports maintain an **audit trail**—a permanent record of who did what and when. Reproducible reports, version-controlled through git and executed via scripts, provide exactly this traceability.

**Reproducible reporting** means:
1. Raw data → Code → Results (no manual copy-pasting of numbers into tables).
2. All analytical choices are documented and version-controlled.
3. Any reviewer can re-run the entire pipeline and obtain identical results.
4. The report is a "living document" that updates when data or code changes.

This chapter focuses on two tools: **R Markdown** and **Quarto**. While R Markdown has dominated clinical reporting, Quarto is its modern successor—faster, more flexible, and increasingly the standard in biostatistics and regulatory circles.

## 21.2 R Markdown vs. Quarto: Key Differences

### 21.2.1 What They Have in Common

Both R Markdown and Quarto are **literate programming** environments that weave code and narrative text:

- YAML header for metadata (title, author, date).
- Markdown text for documentation.
- Code chunks (R, Python, Julia, etc.) executed during rendering.
- Automatic generation of tables, figures, and inline values.
- Multiple output formats (HTML, PDF, Word, etc.).

### 21.2.2 Key Differences

| Feature | R Markdown | Quarto |
|---------|-----------|--------|
| **Underlying Engine** | knitr (R only) | Pandoc + knitr/jupyter (R, Python, Julia) |
| **Development** | RStudio (stable) | Posit + community (active development) |
| **Execution Speed** | Slower (knitr parsing) | Faster (Pandoc architecture) |
| **Cross-References** | Manual numbering | Native `@fig-`, `@tbl-` syntax |
| **Multi-file Projects** | bookdown (add-on) | _quarto.yml (native) |
| **Parameter Rendering** | Limited (params:) | Robust parameterized reports |
| **Code Reuse** | Difficult | _quarto.yml `include:` directives |
| **Future Roadmap** | Limited updates | Active development (2024+) |

**Why Quarto is the future**: Quarto is Posit's (formerly RStudio) strategic direction. While R Markdown will remain supported, Quarto offers superior architecture for complex, multi-file regulatory reports. Major pharmaceutical companies (Roche, Novartis) are adopting Quarto for CSR generation.

### 21.2.3 Recommendation

For **new projects** (2024+), use **Quarto**. For legacy reports using R Markdown, migration is straightforward (minor YAML syntax adjustments).

## 21.3 Quarto Document Anatomy

A Quarto document comprises three sections:

### 21.3.1 YAML Header

``` yaml
---
title: "Safety Report: Treatment Arm A"
subtitle: "Phase III Clinical Trial XYZ-201"
author:
  - "John Smith (Biostatistics)"
  - "Jane Doe (Data Management)"
date: today
format:
  html:
    toc: true
    toc-depth: 3
    number-sections: true
  docx:
    reference-doc: template.docx
params:
  treatment_arm: "A"
  analysis_date: !r Sys.Date()
---

```

Key YAML options:
- `title`, `subtitle`, `author`, `date`: Document metadata.
- `format`: Output format(s) and options.
- `params`: Parameters for parameterized reports (discussed in 21.4).
- `execute`: Global code execution options (see 21.3.2).

### 21.3.2 Quarto Markdown Text

Regular Markdown text with embedded cross-references:

```markdown
## Safety Summary

As shown in @fig-ae-trend, adverse events increased over time in Treatment A.
Table @tbl-ae-summary presents a detailed breakdown.

### Serious Adverse Events

Six serious adverse events (SAEs) were reported: @ref_saes_count.
```

### 21.3.3 Code Chunks

Code chunks are delimited by triple backticks with language and options:

```{r}
#| label: load-data
#| eval: true
#| echo: false
#| warning: false
#| message: false

library(dplyr)
library(pharmaversesdtm)

data("ae")
data("dm")

# Load data
sae_count <- ae %>%
  filter(AESER == "Y") %>%
  distinct(USUBJID) %>%
  nrow()
```

## 21.4 Code Chunk Options

Quarto chunk options are specified in YAML-like comments (using `#|`):

| Option | Purpose | Example |
|--------|---------|---------|
| `label` | Chunk identifier (for cross-references) | `#\| label: create-ae-table` |
| `eval` | Execute chunk? (Y/N) | `#\| eval: true` |
| `echo` | Display code in output? (Y/N) | `#\| echo: false` |
| `warning` | Display warnings? (Y/N) | `#\| warning: false` |
| `message` | Display messages? (Y/N) | `#\| message: false` |
| `fig.width`, `fig.height` | Figure dimensions (inches) | `#\| fig.width: 8` |
| `fig.cap` | Figure caption | `#\| fig.cap: "Figure 1: AE Trends"` |
| `cache` | Cache chunk results? (Y/N) | `#\| cache: true` |
| `include` | Include output in document? (Y/N) | `#\| include: false` |
| `error` | Continue on error? (Y/N) | `#\| error: false` |

### 21.4.1 Global Execution Options

Set defaults in the YAML header:

```yaml
execute:
  echo: false
  warning: false
  message: false
  cache: false
```

## 21.5 Parameterized Reports

A **parameterized report** is a Quarto document that accepts input parameters, allowing you to generate multiple versions automatically without changing code.

### 21.5.1 Define Parameters in YAML

```yaml
---
title: "Safety Report"
params:
  treatment_arm: "A"
  analysis_date: !r Sys.Date()
  include_serious_ae: true
---
```

### 21.5.2 Reference Parameters in Code

Use `params$variable_name` to access parameters:

```{r}
#| label: filter-by-arm

library(dplyr)

# Filter to the selected treatment arm
data_filtered <- ae %>%
  filter(ARM == params$treatment_arm)

# Get analysis date
analysis_date <- params$analysis_date

# Report
cat("Analysis Date:", analysis_date, "\n")
cat("Treatment Arm:", params$treatment_arm, "\n")
cat("N subjects with AE:", n_distinct(data_filtered$SUBJID), "\n")
```

### 21.5.3 Render with Different Parameters

Render a report with parameters via command line or R:

```bash
# Bash: Render for Treatment A
quarto render safety_report.qmd \
  -P treatment_arm:"A" \
  -P include_serious_ae:true \
  -o "safety_report_arm_a.html"

# Render for Treatment B
quarto render safety_report.qmd \
  -P treatment_arm:"B" \
  -P include_serious_ae:true \
  -o "safety_report_arm_b.html"
```

Or in R:

```{r}
library(quarto)

# Render for each treatment arm
for (arm in c("A", "B", "Control")) {
  quarto_render(
    input = "safety_report.qmd",
    execute_params = list(treatment_arm = arm),
    output_file = paste0("safety_report_arm_", arm, ".html")
  )
}
```

### 21.5.4 Example: Parameterized Safety Report Template

Create a file `safety_report_template.qmd`:

```{r}
#| eval: false
#| echo: true

---
title: "Safety Report: Treatment {{params$treatment_arm}}"
params:
  treatment_arm: "A"
  data_cutoff: !r Sys.Date()
  min_aesev: "MODERATE"
format:
  html:
    toc: true
  docx:
    reference-doc: style_template.docx
---

## Executive Summary

This report presents a safety analysis for **`r params$treatment_arm`**
as of **`r format(params$data_cutoff, '%B %d, %Y')`**.

## Adverse Events by Severity

```{r}
#| label: ae-by-severity
#| echo: false

library(dplyr)
library(gtsummary)
library(pharmaversesdtm)

data("ae")
data("dm")

# Filter to severity threshold
ae_filtered <- ae %>%
  filter(
    AESEV %in% c(params$min_aesev, "SEVERE")
  )

# Create summary table
ae_table <- ae_filtered %>%
  select(SUBJID, AETERM, AESEV, AEONSET = AESTDTC) %>%
  distinct() %>%
  arrange(AEONSET) %>%
  head(20)  # Top 20 for display

knitr::kable(ae_table, caption = "Adverse Events (Top 20)")
```

## Demographics

```{r}
#| label: demographics-table
#| echo: false

# Get N for the selected arm
n_arm <- dm %>%
  filter(ARM == params$treatment_arm) %>%
  nrow()

cat("Total subjects in", params$treatment_arm, ":", n_arm, "\n")
```

## Recommendations

- [ ] Further investigation of [EVENT_NAME] recommended
- [ ] Safety monitoring plan review scheduled
- [ ] DSMB notification completed

``` r
*Report generated: `r Sys.time()`*
```

## 21.6 Cross-References in Quarto

Quarto supports native cross-references using the syntax `@fig-label`, `@tbl-label`, `@sec-label`, and `@eq-label`.

### 21.6.1 Figure Cross-References

```{r}
#| label: fig-ae-trend
#| fig.cap: "Adverse Event Incidence Over Time"
#| echo: false

library(ggplot2)

# Simulate AE trend data
ae_trend <- data.frame(
  week = rep(1:12, 3),
  arm = rep(c("Treatment A", "Treatment B", "Control"), each = 12),
  ae_incidence = c(
    seq(0.05, 0.25, length.out = 12),
    seq(0.04, 0.15, length.out = 12),
    seq(0.03, 0.12, length.out = 12)
  )
)

ggplot(ae_trend, aes(x = week, y = ae_incidence, color = arm)) +
  geom_line(size = 1) +
  geom_point(size = 2) +
  labs(
    x = "Week",
    y = "Cumulative AE Incidence",
    color = "Treatment Arm"
  ) +
  theme_minimal()
```

Now reference it: "As shown in @fig-ae-trend, AE incidence increased over time."

### 21.6.2 Table Cross-References

```{r}
#| label: tbl-ae-summary
#| echo: false

library(gtsummary)
library(kableExtra)

ae_summary_table <- mtcars %>%  # Placeholder; replace with actual AE data
  head(5) %>%
  knitr::kable(caption = "Adverse Events Summary")

ae_summary_table
```

Reference: "See @tbl-ae-summary for details."

### 21.6.3 Section Cross-References

```markdown
## Background {#sec-background}

See @sec-background for an overview of the trial design.
```

## 21.7 Inline R Code

Insert computed values directly into narrative text using `` `r ` `` backticks:

```markdown
A total of `r nrow(dm)` subjects were randomized across three treatment arms.
The mean age was `r round(mean(dm$AGE), 1)` years (SD = `r round(sd(dm$AGE), 1)`).
```

Renders as: "A total of 353 subjects were randomized across three treatment arms. The mean age was 45.7 years (SD = 12.2)."

This approach ensures that all reported numbers are **programmatically generated** from data, preventing transcription errors.

## 21.8 Quarto Books and Multi-File Projects

For large clinical study reports (hundreds of pages), organize content across multiple files using Quarto projects.

### 21.8.1 Project Structure

```
clinical_study_report/
├── _quarto.yml
├── index.qmd
├── chapters/
│   ├── 01_background.qmd
│   ├── 02_methods.qmd
│   ├── 03_results.qmd
│   ├── 04_safety.qmd
│   └── 05_conclusions.qmd
├── appendices/
│   ├── a_demographics_tables.qmd
│   └── b_ae_listings.qmd
└── data/
    ├── adsl.rds
    ├── adae.rds
    └── adlb.rds
```

### 21.8.2 Configuration with _quarto.yml

Create `_quarto.yml` to control project settings:

```yaml
project:
  type: book
  output-dir: _book

book:
  title: "Clinical Study Report: Trial XYZ-201"
  author:
    - "Clinical Development Team"
    - "Biostatistics"
  date: today

  chapters:
    - index.qmd
    - chapters/01_background.qmd
    - chapters/02_methods.qmd
    - chapters/03_results.qmd
    - chapters/04_safety.qmd
    - chapters/05_conclusions.qmd
    - appendices/a_demographics_tables.qmd
    - appendices/b_ae_listings.qmd

format:
  html:
    theme: cosmo
    toc: true
    number-sections: true
  pdf:
    toc: true
    number-sections: true
  docx:
    reference-doc: csr_template.docx
    number-sections: true

execute:
  echo: false
  warning: false
  message: false
```

### 21.8.3 Render the Entire Book

From the project root, run:

```bash
quarto render
```

This generates HTML, PDF, and Word versions of the entire CSR in the `_book/` directory.

## 21.9 Automation Pipelines with targets

The **targets** package (by Will Landau) is an R-native pipeline tool that tracks dependencies and only re-executes code when inputs change—essential for large analyses.

### 21.9.1 Basic targets Workflow

Create `_targets.R` in your project root:

```{r}
#| eval: false
#| echo: true

library(targets)
library(tarchetypes)

# Define pipeline
list(
  # Load raw data
  tar_target(
    name = raw_data,
    command = {
      # Download or load from server
      read.csv("data/raw_adsl.csv")
    }
  ),

  # Clean and preprocess
  tar_target(
    name = clean_data,
    command = {
      raw_data %>%
        filter(AVAL >= 0) %>%
        mutate(AVAL_log = log(AVAL + 1))
    },
    dependson = raw_data
  ),

  # Run analysis
  tar_target(
    name = efficacy_analysis,
    command = {
      fit <- lm(AVAL ~ ARM + ABASE + AGE, data = clean_data)
      broom::tidy(fit)
    },
    dependson = clean_data
  ),

  # Generate report
  tar_quarto(
    name = report,
    path = "clinical_report.qmd",
    execute_params = list(
      analysis_results = efficacy_analysis,
      data_timestamp = Sys.time()
    )
  )
)
```

### 21.9.2 Execute the Pipeline

```r
# Run pipeline
targets::tar_make()

# Check status
targets::tar_visnetwork()

# Load results
results <- targets::tar_read(efficacy_analysis)
```

The **targets** package ensures that:
- Raw data changes trigger all downstream reprocessing.
- Reports only re-render if their inputs change.
- A visual dependency graph tracks all relationships.

## 21.10 Version Control Integration: git + RStudio

### 21.10.1 Initialize a git Repository

```bash
cd ~/clinical_study_report
git init
git add .
git commit -m "Initial commit: CSR template"
```

### 21.10.2 .gitignore for Sensitive Data

Create `.gitignore` to exclude sensitive files:

```
# Data files
data/*.csv
data/*.rds
*.xlsx

# Rendered outputs
_book/
*.html
*.pdf
*.docx

# System files
.Rhistory
.Rdata
.DS_Store
.Rproj.user/

# Credentials (NEVER commit!)
.env
credentials.txt
api_keys.txt
```

### 21.10.3 Collaborative Workflow

```bash
# Feature branch for new safety analysis
git checkout -b feature/sae-analysis

# Make changes, commit
git add chapters/04_safety.qmd
git commit -m "Add serious AE subgroup analysis"

# Push to remote
git push -u origin feature/sae-analysis

# Create pull request for peer review
# (GitHub/GitLab web interface)
```

### 21.10.4 Audit Trail Benefits

Once merged to main, git history provides:
- WHO made each change (author metadata).
- WHEN the change was made (timestamp).
- WHAT was changed (diff).
- WHY it was changed (commit message).

This satisfies 21 CFR Part 11 audit trail requirements.

## 21.11 Example: Complete Parameterized Safety Report

Here's a comprehensive safety report template (`safety_report_full.qmd`) that demonstrates all concepts:

```{r}
#| eval: false
#| echo: true

---
title: "Safety Report Template"
subtitle: "Phase III Clinical Trial XYZ-201"
author:
  - "Biostatistics Team"
date: today
params:
  treatment_arm: "Treatment A"
  data_cutoff: !r Sys.Date()
  include_serious_ae: true

format:
  html:
    toc: true
    toc-depth: 3
    number-sections: true
    theme: flatly
  docx:
    reference-doc: csr_style_template.docx
    number-sections: true

execute:
  echo: false
  warning: false
  message: false
  cache: false
---

## {#sec-intro}

This report summarizes safety data for **`r params$treatment_arm`**
as of **`r format(params$data_cutoff, '%B %d, %Y')`**.

```{r}
#| label: load-libraries

library(dplyr)
library(gtsummary)
library(ggplot2)
library(flextable)
library(pharmaversesdtm)

# Load data
data("dm")
data("ae")
data("adsl")
```

## Study Population {#sec-population}

```{r}
#| label: population-summary

# Filter to treatment arm
dm_arm <- dm %>%
  filter(ARM == params$treatment_arm)

n_subjects <- nrow(dm_arm)
mean_age <- round(mean(dm_arm$AGE), 1)
sd_age <- round(sd(dm_arm$AGE), 1)
n_female <- sum(dm_arm$SEX == "F")
pct_female <- round(100 * n_female / n_subjects, 1)
```

A total of `r n_subjects` subjects were enrolled in `r params$treatment_arm`.
The population was `r pct_female`% female with a mean age of `r mean_age` ± `r sd_age` years.

### Demographics Table {#sec-demo-table}

```{r}
#| label: tbl-demographics
#| tbl-cap: "Baseline Demographics"

dm %>%
  filter(ARM == params$treatment_arm) %>%
  select(AGE, SEX, RACE, ETHNIC) %>%
  tbl_summary(
    statistic = list(
      all_continuous() ~ "{mean} ({sd})",
      all_categorical() ~ "{n} ({p}%)"
    ),
    label = list(
      AGE ~ "Age (years)",
      SEX ~ "Sex",
      RACE ~ "Race",
      ETHNIC ~ "Ethnicity"
    )
  ) %>%
  modify_caption("Table 1: Baseline Demographics")
```

## Adverse Events {#sec-ae}

```{r}
#| label: ae-summary-data

ae_arm <- ae %>%
  filter(ARM == params$treatment_arm)

n_ae_subjects <- n_distinct(ae_arm$SUBJID)
total_ae_events <- nrow(ae_arm)
ae_incidence_pct <- round(100 * n_ae_subjects / n_subjects, 1)

# SAEs if requested
sae_count <- 0
if (params$include_serious_ae) {
  sae_count <- ae_arm %>%
    filter(AESER == "Y") %>%
    distinct(SUBJID) %>%
    nrow()
}
```

A total of `r total_ae_events` adverse events were reported in `r n_ae_subjects`
(`r ae_incidence_pct`%) subjects.

`if (params$include_serious_ae) {`
  - **Serious AEs**: `r sae_count` subjects had at least one serious adverse event.
`}`

### AE Table by SOC {#sec-ae-table}

```{r}
#| label: tbl-ae-soc
#| tbl-cap: "Adverse Events by System Organ Class"

ae_summary <- ae_arm %>%
  distinct(SUBJID, AESOC, AETERM) %>%
  group_by(AESOC) %>%
  summarise(
    n = n_distinct(SUBJID),
    .groups = 'drop'
  ) %>%
  arrange(desc(n)) %>%
  mutate(pct = round(100 * n / n_subjects, 1)) %>%
  rename(
    `System Organ Class` = AESOC,
    `N Subjects` = n,
    `%` = pct
  )

flextable(ae_summary) %>%
  theme_vanilla() %>%
  autofit() %>%
  add_header_row(
    values = c(
      paste("Adverse Events by SOC –", params$treatment_arm),
      "", ""
    )
  )
```

### AE Trend Figure {#sec-ae-trend}

```{r}
#| label: fig-ae-cumulative
#| fig.cap: "Cumulative AE Incidence Over Time"

ae_timeline <- ae_arm %>%
  arrange(AESTDTC) %>%
  mutate(
    cumulative_subjects = cumsum(!duplicated(SUBJID))
  ) %>%
  select(AESTDTC, cumulative_subjects) %>%
  distinct() %>%
  mutate(cumulative_pct = 100 * cumulative_subjects / n_subjects)

ggplot(ae_timeline, aes(x = AESTDTC, y = cumulative_pct)) +
  geom_step(color = "darkred", size = 1) +
  geom_point(color = "darkred", size = 2) +
  labs(
    x = "Date",
    y = "Cumulative AE Incidence (%)",
    title = paste("Cumulative Adverse Events –", params$treatment_arm)
  ) +
  theme_minimal() +
  ylim(0, 100)
```

As shown in @fig-ae-cumulative, AE incidence accumulated over the study period,
reaching `r tail(ae_timeline$cumulative_pct, 1)`% by the data cutoff.

## Serious Adverse Events Listing {#sec-sae-listing}

`if (params$include_serious_ae) {`

```{r}
#| label: sae-listing-data

sae_listing <- ae_arm %>%
  filter(AESER == "Y") %>%
  select(SUBJID, AESOC, AETERM, AESEV, AESTDTC, AEENDTC) %>%
  arrange(AESTDTC)
```

```{r}
#| label: tbl-sae-listing
#| tbl-cap: "Listing of Serious Adverse Events"

flextable(sae_listing) %>%
  set_header_labels(
    SUBJID = "Subject",
    AESOC = "SOC",
    AETERM = "Preferred Term",
    AESEV = "Severity",
    AESTDTC = "Onset Date",
    AEENDTC = "End Date"
  ) %>%
  theme_vanilla() %>%
  autofit()
```

`}`

## Conclusions

- Safety profile is consistent with previous observations.
- No unexpected safety signals identified.
- Adverse events were generally manageable and reversible.

---

**Report Generated**: `r Sys.time()`
**Analysis Date**: `r params$data_cutoff`
**Treatment Arm**: `r params$treatment_arm`
```

## 21.12 Rendering and Output Options

### 21.12.1 Render from Command Line

```bash
# HTML output
quarto render safety_report_full.qmd -t html

# Word output
quarto render safety_report_full.qmd -t docx

# PDF output (requires TinyTeX or LaTeX)
quarto render safety_report_full.qmd -t pdf

# With parameters
quarto render safety_report_full.qmd \
  -P treatment_arm:"Treatment B" \
  -t html \
  -o "safety_report_arm_b.html"
```

### 21.12.2 Render from R

```r
library(quarto)

quarto_render(
  input = "safety_report_full.qmd",
  execute_params = list(
    treatment_arm = "Treatment A",
    include_serious_ae = TRUE
  ),
  output_format = "all"
)
```

### 21.12.3 Batch Rendering for Multiple Sites/Arms

```r
# Render report for each treatment arm
arms <- c("Treatment A", "Treatment B", "Control")

for (arm in arms) {
  quarto_render(
    input = "safety_report_full.qmd",
    execute_params = list(treatment_arm = arm),
    output_file = paste0("safety_report_", tolower(gsub(" ", "_", arm)), ".html")
  )
}
```

## 21.13 Troubleshooting Common Issues

| Issue | Cause | Solution |
|-------|-------|----------|
| Cross-references not numbering | Missing `number-sections: true` in YAML | Add to format block |
| Parameters not updating | Caching enabled | Set `cache: false` in execute block |
| Figure dimensions wrong | `fig.width` not set in chunk | Add `#\| fig.width: 8` |
| Word output styling lost | Missing `reference-doc:` | Provide template Word file |
| Slow rendering | Data loading in every chunk | Cache data load: `#\| cache: true` |
| PDF errors | LaTeX not installed | `quarto::quarto_install_tinytex()` |

## 21.14 Summary

- **Reproducible reporting** is essential for regulatory compliance (21 CFR Part 11) and scientific integrity.
- **Quarto** is the modern successor to R Markdown, with superior architecture for clinical reports.
- **Parameterized reports** allow automated generation of multiple report variants (by site, arm, data cutoff).
- **Cross-references** (`@fig-`, `@tbl-`) enable automatic numbering and updates.
- **Inline R code** ensures all reported numbers are programmatically generated.
- **Quarto books** organize large reports across multiple files with `_quarto.yml` configuration.
- **targets** pipelines track dependencies and enable efficient re-execution.
- **git** integration provides audit trails and version control for regulatory compliance.

## 21.15 Exercises

**Exercise 21.1**: Convert an existing R Markdown report to Quarto. Update YAML syntax, code chunk options (from `{r, echo=FALSE}` to `#| echo: false`), and test all output formats.

**Exercise 21.2**: Create a parameterized safety report that accepts `treatment_arm` and `data_cutoff` parameters. Render it for three arms (A, B, Control) automatically using a for loop.

**Exercise 21.3**: Build a Quarto book project with three chapters: Background, Methods, and Results. Use `_quarto.yml` to configure multi-file rendering, and test HTML, PDF, and Word outputs.

**Exercise 21.4**: Implement a targets pipeline that:
  - Loads raw ADaM data (simulated or pharmaversesdtm)
  - Performs efficacy analysis
  - Generates a Quarto report with inline results
  - Visualizes the dependency graph with `tar_visnetwork()`

**Exercise 21.5**: Set up git version control for a clinical report project, create a feature branch for a safety analysis, and simulate a pull request review workflow.

---

**Key References**:
- Quarto Official Documentation: https://quarto.org/
- R Markdown: The Definitive Guide: https://bookdown.org/yihui/rmarkdown/
- targets Package: https://books.ropensci.org/targets/
- 21 CFR Part 11 (FDA): https://www.ecfr.gov/
- PHUSE Guidelines on Reproducibility: https://www.phuse.global/
