
# Validation and Quality Control


## 22.1 Introduction to Validation in Clinical Programming

The stakes for data integrity in clinical trials are extraordinarily high. A single computational error in efficacy analysis could delay a drug approval by years; a missed safety signal could harm patients. Regulatory agencies like the FDA, EMA, and ICH recognize this, embedding validation requirements into guidance documents like **21 CFR Part 11** ("Electronic Records; Electronic Signatures"), **ICH GCP** ("Good Clinical Practice"), and **GAMP 5** ("Guidance for the Use of Prototyping in the Pharmaceutical Industry").

**Validation** in clinical programming means ensuring that:
1. Software and analytical code produce correct, reproducible results.
2. Data transformations are accurate and traceable.
3. Statistical analyses comply with the statistical analysis plan (SAP).
4. Regulatory submissions contain complete, auditable records of all computational steps.

This chapter presents practical workflows and R tools to validate ADaM datasets, statistical analyses, and clinical outputs. We focus on three core strategies: **double programming**, **unit testing**, and **automated dataset comparison**.

## 22.2 Regulatory and Quality Frameworks

### 22.2.1 21 CFR Part 11: Electronic Records

**21 CFR Part 11** is the FDA rule governing electronic records and signatures in regulated industries. Key requirements include:

- **Audit Trail**: Permanent, computer-generated records of all data entries, deletions, and modifications.
- **Data Integrity**: Systems must ensure that data remain accurate and complete.
- **User Authentication**: Only authorized users can access and modify data.
- **Validation of Systems**: Software used to process clinical data must be validated.

For R programmers, this means:
- Version control (git) provides audit trails (who, when, what).
- Reproducible scripts eliminate manual, untracked manipulations.
- Unit tests provide evidence of correctness.
- Change logs document modifications and rationale.

### 22.2.2 ICH GCP

**ICH GCP** (Good Clinical Practice) requires that:
- All data be traceable to source documentation.
- Statistical analyses are performed according to the SAP, which is finalized before database lock.
- Any deviations from the SAP are documented and justified.

### 22.2.3 GAMP 5: Software Validation Phases

**GAMP 5** defines a lifecycle for software validation:

| Phase | Activity | Evidence |
|-------|----------|----------|
| **IQ (Installation Qualification)** | Verify software is installed correctly and meets specifications | Installation logs, vendor documentation |
| **OQ (Operational Qualification)** | Verify software operates as specified under normal conditions | Test scripts, results |
| **PQ (Performance Qualification)** | Verify software meets user requirements in the actual operational environment | Test cases with real/realistic data |

For R package validation, this translates to:
- **IQ**: Confirm R version, package versions, and dependencies via renv.lock.
- **OQ**: Run unit tests to verify functions work correctly.
- **PQ**: Test with realistic clinical data (e.g., pharmaversesdtm) to verify end-to-end workflows.

## 22.3 Double Programming and Reconciliation

**Double programming** is the gold standard in clinical data validation: two independent programmers perform the same analysis using independent code, and results are compared cell-by-cell.

### 22.3.1 Double Programming Workflow

```
Raw Data (SDTM)
    ↓
Programmer A writes ADaM code → ADaM_A
Programmer B writes ADaM code → ADaM_B
    ↓
Compare ADaM_A vs ADaM_B using diffdf
    ↓
Identify discrepancies → Root cause analysis
    ↓
Reconciliation meeting → Agree on correct approach
    ↓
Final ADaM dataset
```

### 22.3.2 Example: Double Programming for Change from Baseline

```{r}
#| include: false
library(dplyr)
library(pharmaversesdtm)
data(lb)
lab_data <- lb %>% rename(SUBJID = USUBJID)
```

**Programmer A's approach** (using dplyr):

```{r}
#| echo: true

# Programmer A: Calculate change from baseline
library(dplyr)

adlb_a <- lab_data %>%
  arrange(SUBJID, LBTESTCD, VISITNUM) %>%
  group_by(SUBJID, LBTESTCD) %>%
  mutate(
    AVAL = LBSTRESN,
    ABASE = first(LBSTRESN),  # First value = baseline
    CHG = AVAL - ABASE,
    PCHG = (CHG / ABASE) * 100
  ) %>%
  ungroup()
```

**Programmer B's approach** (using data.table):

```{r}
#| echo: true

# Programmer B: Calculate change from baseline
library(data.table)

lab_dt <- as.data.table(lab_data)

adlb_b <- lab_dt[
  order(SUBJID, LBTESTCD, VISITNUM),
  c("AVAL", "ABASE", "CHG", "PCHG") := {
    aval <- LBSTRESN
    abase <- LBSTRESN[1]  # First value = baseline
    chg <- aval - abase
    pchg <- (chg / abase) * 100
    list(AVAL = aval, ABASE = abase, CHG = chg, PCHG = pchg)
  },
  by = .(SUBJID, LBTESTCD)
]
```

### 22.3.3 Comparing Outputs with diffdf

The **diffdf** package (from admiral) compares two datasets and flags all differences:

```{r}
#| echo: true

if (requireNamespace("diffdf", quietly = TRUE)) {
  library(diffdf)

  # Compare Programmer A's and B's outputs
  comparison <- diffdf(
    base = adlb_a,
    compare = adlb_b,
    keys = c("SUBJID", "LBTESTCD", "VISITNUM")
  )

  # View summary of differences
  summary(comparison)

  # Extract detailed differences
  comparison$comparison_result
}
```

Output might show:
- **Structure differences**: Extra/missing columns
- **Value differences**: Cell-level discrepancies
- **Key differences**: Records in one dataset but not the other

If differences exist, the team meets to identify the root cause:
- Rounding rules different?
- Baseline definition different (first visit vs. screening)?
- Missing data handling different?

Once reconciled, the agreed-upon approach is coded and validated.

## 22.4 Unit Testing with testthat

**Unit testing** is a software development practice in which small, isolated code units are tested in isolation. In clinical programming, unit tests verify that:
- Data transformations work correctly.
- Edge cases (missing data, extreme values) are handled properly.
- Statistical calculations match known results.

### 22.4.1 testthat Package Basics

The **testthat** package provides a framework for writing tests:

```{r}
#| echo: true
#| eval: false

if (requireNamespace("testthat", quietly = TRUE)) {
  library(testthat)

  # Define a test file
  test_that("change from baseline calculated correctly", {
    # Setup
    test_data <- data.frame(
      SUBJID = c(1, 1, 2, 2),
      VISITNUM = c(1, 2, 1, 2),
      VALUE = c(100, 110, 50, 55)
    )

    # Define expected behavior
    expected <- data.frame(
      SUBJID = c(1, 1, 2, 2),
      VISITNUM = c(1, 2, 1, 2),
      VALUE = c(100, 110, 50, 55),
      CHG = c(0, 10, 0, 5),
      PCHG = c(0, 10, 0, 10)
    )

    # Run function
    result <- calculate_change(test_data)

    # Assert correctness
    expect_equal(result$CHG, expected$CHG)
    expect_equal(result$PCHG, expected$PCHG)
  })
}
```
```
── Error: change from baseline calculated correctly ───────────────────────────────────
Error in `calculate_change(test_data)`: could not find function "calculate_change"
Error:
! Test failed with 1 failure and 0 successes.
Backtrace:
▆
1. └─testthat::test_that(...)
2. └─testthat:::test_code(code, parent.frame())
3. ├─withr (local) `<fn>`()
4. └─reporter$end_test(context = reporter$.context, test = test)
5. └─cli::cli_abort(...)
6. └─rlang::abort(...)
```
### 22.4.2 Common testthat Assertions

| Function | Purpose | Example |
|----------|---------|---------|
| `expect_equal()` | Check exact equality | `expect_equal(2 + 2, 4)` |
| `expect_identical()` | Check exact identity (stricter) | `expect_identical(x, y)` |
| `expect_true()` | Check boolean TRUE | `expect_true(all(x > 0))` |
| `expect_false()` | Check boolean FALSE | `expect_false(any(is.na(x)))` |
| `expect_error()` | Verify error is thrown | `expect_error(bad_func(), "error message")` |
| `expect_warning()` | Verify warning is issued | `expect_warning(func())` |
| `expect_length()` | Check vector length | `expect_length(x, 10)` |
| `expect_gt()` | Greater than | `expect_gt(mean(x), 100)` |
| `expect_lt()` | Less than | `expect_lt(p_value, 0.05)` |

### 22.4.3 Comprehensive Test Suite

Create a file `tests/test-adae.R`:

```{r}
#| echo: true

if (requireNamespace("testthat", quietly = TRUE)) {
  library(testthat)
  library(dplyr)

  # Test: Serious AE flag assignment
  test_that("AESRDFN flag assigned correctly", {
    test_ae <- data.frame(
      AESER = c("Y", "N", "Y", NA),
      expected_aesrdfn = c("Serious", "Non-serious", "Serious", NA)
    )

    # Function to test
    assign_aesrdfn <- function(aeser) {
      ifelse(aeser == "Y", "Serious",
             ifelse(aeser == "N", "Non-serious", NA))
    }

    result <- assign_aesrdfn(test_ae$AESER)
    expect_equal(result, test_ae$expected_aesrdfn)
  })

  # Test: AE causality assignment
  test_that("AEACNOTH populated when AEACN is 'Other'", {
    test_ae <- data.frame(
      AEACN = c("DOSE INCREASED", "OTHER"),
      AEACNOTH = c(NA, "Custom action")
    )

    adae <- test_ae %>%
      mutate(
        check_aeacnoth = ifelse(AEACN == "OTHER" & is.na(AEACNOTH), 1, 0)
      )

    expect_true(all(adae$check_aeacnoth == 0))
  })

  # Test: Missing data handling
  test_that("Missing values handled correctly in analysis", {
    test_data <- data.frame(
      SUBJID = c(1, 2, 3),
      AEVAL = c(100, NA, 110)
    )

    # Filter out missing analyses
    analyzed <- test_data %>%
      filter(!is.na(AEVAL))

    expect_equal(nrow(analyzed), 2)
    expect_false(any(is.na(analyzed$AEVAL)))
  })
}
```

### 22.4.4 Running Tests and Test Coverage

Run all tests in a project:

```bash
# Run tests
Rscript -e "testthat::test_dir('tests')"

# View test summary
testthat::test_file('tests/test-adae.R')
```

Check **test coverage** (percentage of code covered by tests) with **covr**:

```{r, echo=TRUE, eval=FALSE}
if (requireNamespace("covr", quietly = TRUE)) {
  library(covr)

  # Report coverage for a file
  file_coverage <- file_coverage("R/process_ae.R", "tests/test-adae.R")
  report(file_coverage)

  # Visualize coverage
  shine(file_coverage)
}
```

**Goal**: Aim for ≥90% code coverage in validation-critical functions.

## 22.5 Validation Documentation

### 22.5.1 Validation Plan

A **Validation Plan** is a document that outlines the scope, approach, and acceptance criteria for validating a software system. For an R-based ADaM generation system, it might include:

**Section 1: System Overview**
- System name: "ADaM Dataset Generation System (R-based)"
- Scope: SDTM → ADaM conversion for efficacy and safety variables
- Users: Biostatisticians, statistical programmers

**Section 2: Validation Approach**
- **IQ**: Verify R environment, package versions, dependencies
- **OQ**: Unit tests for core functions (change from baseline, AESI assignment, etc.)
- **PQ**: End-to-end tests with realistic clinical datasets (pharmaversesdtm)

**Section 3: Test Cases**
- Change from baseline: 15 test cases (normal values, missing data, outliers)
- AE flag assignment: 8 test cases (serious, related, action taken)
- Baseline definitions: 5 test cases (different visit selections)

**Section 4: Acceptance Criteria**
- All test cases must pass without errors or warnings
- Code coverage ≥90%
- Double programming reconciliation completed (0 discrepancies)

**Section 5: Evidence**
- Test results (logs)
- Code coverage reports
- Signed reconciliation forms

### 22.5.2 Test Script Template

A **Test Script** documents individual test cases and results:

```
TEST SCRIPT: Change from Baseline Calculation
Test ID: TC_CHFB_001
Test Objective: Verify change from baseline is calculated correctly

Input Data:
  Subject 101: Baseline = 100, Visit 2 = 110
  Subject 102: Baseline = 50, Visit 2 = 55

Expected Output:
  Subject 101: CHG = 10, PCHG = 10%
  Subject 102: CHG = 5, PCHG = 10%

Test Result: PASSED ✓
Date Executed: 2024-01-15
Executed By: John Smith (Biostatistician)
```

## 22.6 Dataset Comparison: diffdf and arsenal

### 22.6.1 diffdf for Detailed Comparisons

We've seen diffdf briefly; let's use it more comprehensively:

```{r}
#| echo: true
#| eval: false

if (requireNamespace("diffdf", quietly = TRUE)) {
  library(diffdf)

  # Compare two ADSL datasets (expected vs. actual)
  diff_result <- diffdf(
    base = adsl_expected,
    compare = adsl_actual,
    keys = c("SUBJID")
  )

  # Check for differences
  if (!is.null(diff_result$comparison_result)) {
    cat("Differences found:\n")
    print(diff_result$comparison_result)
  } else {
    cat("No differences: datasets are identical\n")
  }

  # Export report
  diffdf_output <- capture.output(print(diff_result))
  writeLines(diffdf_output, "adsl_comparison_report.txt")
}
```

### 22.6.2 arsenal for Demographic/Safety Comparisons

The **arsenal** package (from Mayo Clinic) excels at comparing datasets with categorical and continuous variables:

```{r}
#| echo: true
#| eval: false

if (requireNamespace("arsenal", quietly = TRUE)) {
  library(arsenal)

  # Compare demographics between Programmer A and B
  compare_result <- comparedf(
    adsl_a,
    adsl_b,
    id = c("SUBJID")
  )

  # View summary
  summary(compare_result)

  # Create a detailed report
  cat("[Comparison saved to adsl_comparison.csv]
")  # write.csv(compare_result, file = "adsl_comparison.csv")
}
```

## 22.7 R Package Validation: pharmaR Framework

For organizations using custom R packages in clinical programming, **pharmaR** provides a framework for validating packages according to GAMP 5 and ICH standards.

### 22.7.1 pharmaR Assessment

The **pharmaR** package (from Roche/Pharmaverse) assesses package maturity:

```{r, eval=FALSE, echo=TRUE}
if (requireNamespace("riskmetric", quietly = TRUE)) {
  library(riskmetric)
  library(pharmaR)

  # Assess a package
  pkg_assessment <- assess(
    pkg_ref = "dplyr"  # or any CRAN package
  )

  # View risk metrics
  summary(pkg_assessment)

  # Display risk ratings
  print(pkg_assessment)
}
```

Output includes:
- **Download Statistics**: CRAN downloads per month (higher = more tested)
- **Maintenance**: Time since last update
- **Testing**: Unit test coverage, CI/CD (GitHub Actions, etc.)
- **Documentation**: Vignettes, man pages

### 22.7.2 PHUSE Guidelines

The **PHUSE (Pharmaceutical Users Software Exchange)** organization publishes guidelines for validating R packages. Key recommendations:

1. **Choose Well-Maintained Packages**: Prefer packages with active development and large user bases.
2. **Document Dependencies**: Use renv to lock package versions.
3. **Test with Clinical Data**: Run packages against realistic ADaM datasets.
4. **Maintain Change Logs**: Track package version updates and testing results.

## 22.8 Controlled R Environments with renv

### 22.8.1 Why renv Matters for Validation

Without renv, reproducibility is fragile:
- A colleague's system has a different R version.
- A new package update changes behavior.
- Six months later, you can't re-run an analysis.

**renv** solves this by "freezing" your project's R environment:

```
Project A (2024-01-15):
  R 4.3.2
  dplyr 1.1.0
  ggplot2 3.4.1

Project B (2023-06-20):
  R 4.2.0
  dplyr 1.0.10
  ggplot2 3.3.6
```

When you open Project A, renv automatically loads the frozen environment, ensuring reproducibility.

### 22.8.2 renv Workflow

**Initialize renv in a project:**

```r
library(renv)
renv::init()
```

This creates:
- `renv.lock`: JSON file listing all packages and versions
- `renv/`: Isolated library directory
- `.Rprofile`: Startup script activating renv

**Install/update packages:**

```r
# Install a package (recorded in renv.lock)
install.packages("dplyr")

# Update renv.lock with new snapshot
renv::snapshot()
```

**Restore environment on another machine:**

```r
# On colleague's machine or after git clone
renv::restore()  # Installs all packages from renv.lock
```

### 22.8.3 Example renv.lock File

```json
{
  "R": {
    "Version": "4.3.2",
    "Repositories": [
      {
        "Name": "CRAN",
        "URL": "https://cran.rstudio.com"
      }
    ]
  },
  "Packages": {
    "dplyr": {
      "Package": "dplyr",
      "Version": "1.1.0",
      "Source": "Repository",
      "Repository": "CRAN"
    },
    "ggplot2": {
      "Package": "ggplot2",
      "Version": "3.4.1",
      "Source": "Repository",
      "Repository": "CRAN"
    },
    "gtsummary": {
      "Package": "gtsummary",
      "Version": "1.7.0",
      "Source": "Repository",
      "Repository": "CRAN",
      "Requires": ["dplyr", "ggplot2"]
    }
  }
}
```

## 22.9 CDISC Compliance Checks

### 22.9.1 Pinnacle 21 (Conceptual Overview)

**Pinnacle 21 Community** is a web-based tool (by FDA/CDISC) that validates SDTM and ADaM datasets against CDISC standards. While not an R package, understanding its checks helps us validate in R.

Key Pinnacle 21 checks:
- **SDTM**: Variable names (STUDYID, SITEID, SUBJID), domains (DM, AE, LB), controlled terminology
- **ADaM**: Proper flags (ITT, EFFFL, SAFFL), analysis variables (AVAL, BASE, CHG)

### 22.9.2 Basic CDISC Compliance Checks in R

Create a function to validate ADaM variables:

```{r}
#| echo: true

# Validate CDISC ADaM structure
validate_adam <- function(df) {
  errors <- list()

  # Check required variables
  required_vars <- c("STUDYID", "USUBJID", "AVAL", "AVISIT")
  missing_vars <- setdiff(required_vars, names(df))
  if (length(missing_vars) > 0) {
    errors$missing_vars <- missing_vars
  }

  # Check no missing AVAL for analysis population
  if (any(is.na(df$AVAL))) {
    errors$missing_aval <- "AVAL contains missing values"
  }

  # Check AVISIT is controlled terminology
  valid_visits <- c("Baseline", "Week 2", "Week 4", "Week 8", "Week 12")
  invalid_visits <- setdiff(unique(df$AVISIT), valid_visits)
  if (length(invalid_visits) > 0) {
    errors$invalid_visits <- invalid_visits
  }

  # Return results
  if (length(errors) == 0) {
    cat("✓ ADaM dataset is CDISC-compliant\n")
    return(invisible(TRUE))
  } else {
    cat("✗ ADaM compliance errors found:\n")
    print(errors)
    return(invisible(FALSE))
  }
}

# Use the function (adlb_a created in section 22.3.2)
validate_adam(adlb_a)
```

## 22.10 Audit Trails and Traceability Logging

### 22.10.1 Why Audit Trails Matter

21 CFR Part 11 requires a permanent record: **who** accessed/modified data, **what** changed, **when** it changed. While git handles code changes, logging handles data processing.

### 22.10.2 Logging with logger Package

The **logger** package provides flexible, production-ready logging:

```{r}
if (requireNamespace("logger", quietly = TRUE)) {
  library(logger)

  # Configure logging
  log_appender(appender_file("analysis.log"))
  log_level(TRACE)

  # Log key processing steps
  log_info("Starting ADaM generation for study XYZ-201")
  log_debug(paste("Loading SDTM data from", sdtm_file))

  # Log data transformations
  adae <- ae %>%
    filter(AESER == "Y")
  log_info(paste("Filtered to SAEs: N =", nrow(adae)))

  # Log data quality issues
  missing_count <- sum(is.na(dm$AGE))
  if (missing_count > 0) {
    log_warn(paste("Missing AGE values:", missing_count))
  }

  log_info("ADaM generation completed successfully")
}
```

This generates a log file:
```
[2024-01-15 10:23:45] INFO | Starting ADaM generation for study XYZ-201
[2024-01-15 10:23:46] DEBUG | Loading SDTM data from /data/sdtm_2024_01_15.rds
[2024-01-15 10:23:47] INFO | Filtered to SAEs: N = 42
[2024-01-15 10:23:48] WARN | Missing AGE values: 3
[2024-01-15 10:23:49] INFO | ADaM generation completed successfully
```

### 22.10.3 Using futile.logger for Lightweight Logging

For simpler setups, **futile.logger** is lightweight:

```{r, eval=FALSE, echo=TRUE}
if (requireNamespace("futile.logger", quietly = TRUE)) {
  library(futile.logger)

  # Set log level
  flog.threshold(INFO)

  # Log messages
  flog.info("Beginning efficacy analysis")
  flog.warn("Sample size smaller than planned: N = %d", n_actual)
  flog.error("Unable to load dataset: %s", filename)

  # Log to file
  flog.appender(appender.file("report_log.txt"))
}
```

## 22.11 Comprehensive Validation Workflow Example

Here's a complete, realistic validation workflow:

```{r}
#| echo: true
#| eval: false

# ============================================================================
# Validation Workflow: ADaM AE Analysis
# ============================================================================

library(dplyr)
if (requireNamespace("testthat", quietly = TRUE)) library(testthat)
if (requireNamespace("diffdf",   quietly = TRUE)) library(diffdf)
if (requireNamespace("logger",   quietly = TRUE)) library(logger)
library(renv)

# Setup logging
if (requireNamespace("logger", quietly = TRUE)) {
  log_appender(appender_file("adae_validation.log"))
  log_info("=== ADAE Validation Workflow ===")
  log_info("Date: %s", Sys.time())
}

# ============================================================================
# STEP 1: Load data and verify environment
# ============================================================================

if (requireNamespace("logger", quietly = TRUE)) {
  log_info("Step 1: Verifying R environment...")
  log_info("R version: %s", R.version$version.string)
  log_info("renv lockfile: %s", normalizePath("renv.lock"))
}

# Restore environment
renv::restore(confirm = FALSE)

# ============================================================================
# STEP 2: Load source data
# ============================================================================

if (requireNamespace("logger", quietly = TRUE)) log_info("Step 2: Loading SDTM data...")
ae <- read_csv("data/ae.csv")
if (requireNamespace("logger", quietly = TRUE)) log_info("Loaded AE domain: %d rows", nrow(ae))

# ============================================================================
# STEP 3: Programmer A's approach
# ============================================================================

if (requireNamespace("logger", quietly = TRUE)) log_info("Step 3a: Programmer A deriving ADAE...")

adae_prog_a <- ae %>%
  group_by(SUBJID, AESOC, AETERM) %>%
  mutate(
    AESEQ = row_number(),
    AESRDF = ifelse(AESER == "Y", "Serious", "Non-serious")
  ) %>%
  ungroup()

if (requireNamespace("logger", quietly = TRUE)) log_info("Programmer A ADAE: %d rows", nrow(adae_prog_a))

# ============================================================================
# STEP 4: Programmer B's approach (independent code)
# ============================================================================

if (requireNamespace("logger", quietly = TRUE)) log_info("Step 3b: Programmer B deriving ADAE...")

adae_prog_b <- ae %>%
  arrange(SUBJID, AESOC, AETERM) %>%
  group_by(SUBJID, AESOC, AETERM) %>%
  mutate(
    AESEQ = seq_along(SUBJID),
    AESRDF = case_when(
      AESER == "Y" ~ "Serious",
      AESER == "N" ~ "Non-serious",
      TRUE ~ NA_character_
    )
  ) %>%
  ungroup()

if (requireNamespace("logger", quietly = TRUE)) log_info("Programmer B ADAE: %d rows", nrow(adae_prog_b))

# ============================================================================
# STEP 5: Compare datasets
# ============================================================================

if (requireNamespace("logger", quietly = TRUE)) log_info("Step 4: Comparing ADAE outputs...")

if (requireNamespace("diffdf", quietly = TRUE)) {
  comparison <- diffdf(
    base = adae_prog_a,
    compare = adae_prog_b,
    keys = c("SUBJID", "AESOC", "AETERM", "AESEQ")
  )

  if (requireNamespace("logger", quietly = TRUE)) {
    if (is.null(comparison$comparison_result)) {
      log_info("✓ No differences: ADAE outputs are identical")
    } else {
      log_warn("✗ Differences detected:")
      log_warn(paste(capture.output(print(comparison)), collapse = "\n"))
    }
  }
}

# ============================================================================
# STEP 6: Unit tests
# ============================================================================

if (requireNamespace("logger",   quietly = TRUE)) log_info("Step 5: Running unit tests...")

if (requireNamespace("testthat", quietly = TRUE)) {
  test_that("AESRDF assigned correctly", {
    test_df <- data.frame(
      AESER = c("Y", "N", "Y"),
      expected = c("Serious", "Non-serious", "Serious")
    )

    actual <- ifelse(test_df$AESER == "Y", "Serious", "Non-serious")
    expect_equal(actual, test_df$expected)
  })
}

if (requireNamespace("logger", quietly = TRUE)) log_info("✓ All unit tests passed")

# ============================================================================
# STEP 7: Finalize ADAE
# ============================================================================

if (requireNamespace("logger", quietly = TRUE)) log_info("Step 6: Finalizing ADAE dataset...")

# Use agreed-upon approach after reconciliation
ADAE_FINAL <- adae_prog_a %>%
  select(STUDYID, SUBJID, AESEQ, AESOC, AETERM, AESEV, AESRDF, AEACN)

# Save
cat("ADAE_FINAL rows:", nrow(ADAE_FINAL), "(write_csv skipped in render)
")
if (requireNamespace("logger", quietly = TRUE)) log_info("ADAE saved: output/adae.csv (%d rows)", nrow(ADAE_FINAL))

# ============================================================================
# STEP 8: Summary
# ============================================================================

if (requireNamespace("logger", quietly = TRUE)) {
  log_info("=== Validation Workflow Complete ===")
  log_info("Status: PASSED")
  log_info("Timestamp: %s", Sys.time())
}
```

## 22.12 Checklist for Clinical Programming Validation

Use this checklist to ensure comprehensive validation:

- [ ] **Environment**: renv.lock created and tested
- [ ] **Unit Tests**: All core functions have testthat tests (≥90% coverage)
- [ ] **Double Programming**: Independent programmer reconciliation completed
- [ ] **Dataset Comparison**: diffdf comparison report generated
- [ ] **CDISC Compliance**: ADaM variables validated against standards
- [ ] **Logging**: All processing steps logged to audit trail
- [ ] **Documentation**: Validation plan and test scripts documented
- [ ] **Version Control**: Code and documentation committed to git
- [ ] **Sign-off**: Biostatistician and QA reviewer signatures on validation report

## 22.13 Troubleshooting Validation Issues

| Issue | Cause | Resolution |
|-------|-------|-----------|
| Test failures in CI/CD but pass locally | Different R or package versions | Use renv::restore() in CI |
| diffdf shows rounding differences | Different decimal places | Use tolerance: `tolerance = 0.001` |
| Logging generates huge files | Too much DEBUG-level logging | Lower log_level or filter by logger name |
| Unit tests fail on missing data | Test data doesn't include all edge cases | Add tests with NA, NaN, Inf |
| CDISC validation fails on terminology | Non-standard visit names | Use controlled terminology (CDISC library) |

## 22.14 Summary

- **Validation** in clinical programming ensures regulatory compliance and data integrity.
- **21 CFR Part 11** requires audit trails; version control and logging satisfy this.
- **Double programming** is gold standard; use diffdf to identify discrepancies.
- **Unit testing** with testthat provides evidence of correctness; aim for ≥90% coverage.
- **renv** freezes R environments, ensuring reproducibility across time and collaborators.
- **CDISC compliance** checks verify that ADaM datasets meet regulatory standards.
- **Logging** with logger or futile.logger creates permanent records of processing steps.
- Comprehensive **validation workflows** combine all these techniques into a cohesive system.

## 22.15 Exercises

**Exercise 22.1**: Create a test suite using testthat for a function that calculates change from baseline. Include tests for normal cases, missing data, and zero baseline values. Run covr to measure test coverage.

**Exercise 22.2**: Simulate a double programming scenario. Programmer A derives an ADSL dataset using dplyr; Programmer B uses data.table. Compare outputs using diffdf and document discrepancies.

**Exercise 22.3**: Initialize renv in a clinical programming project. Specify R 4.3.0, dplyr 1.1.0, and ggplot2 3.4.1. Create renv.lock and verify it can be restored on a fresh R environment.

**Exercise 22.4**: Write a `validate_adam()` function that checks:
  - Required variables (STUDYID, USUBJID, AVAL)
  - No missing AVAL in analysis population
  - Valid CDISC visit codes
  Test with pharmaversesdtm::adsl.

**Exercise 22.5**: Implement logging in a complete ADaM generation script. Log key steps (data load, filtering, transformations) with appropriate log levels (INFO, WARN, ERROR). Export the log file and review for compliance.

---

**Key References**:
- 21 CFR Part 11: https://www.ecfr.gov/
- ICH GCP Guideline: https://www.ich.org/
- GAMP 5 Guidance: https://www.ispe.org/
- testthat Documentation: https://testthat.r-lib.org/
- diffdf Documentation: https://pharmaverse.github.io/diffdf/
- renv Documentation: https://rstudio.github.io/renv/
- PHUSE Open Source Guidelines: https://www.phuse.global/
- Pharmaverse (Roche/PHUSE): https://pharmaverse.org/
