# Oncology Trial Case Study

This chapter presents a comprehensive, real-world case study of a Phase III oncology trial analyzing efficacy and safety data. We demonstrate survival analysis, response evaluation, subgroup analyses, and safety reporting using clinical datasets and rigorous biostatistical methods aligned with oncology regulatory standards.

## 24.1 Background: A Phase III Oncology Trial

### The ONCO-301 Trial Design

Imagine we are analyzing data from **ONCO-301**, a pivotal Phase III randomised controlled trial evaluating a novel targeted therapy in metastatic colorectal cancer (mCRC). The trial compares:

- **Investigational arm**: Vemotinib (a hypothetical BRAF inhibitor), 960 mg once daily
- **Control arm**: FOLFOX (fluorouracil, leucovorin, oxaliplatin) standard chemotherapy

**Trial characteristics**:
- 500 patients, 1:1 randomisation
- Inclusion: histologically confirmed mCRC, ECOG PS 0–1, prior treatment naïve
- Stratification: by site and BRAF mutation status
- Primary endpoint: **Overall Survival (OS)** — time from randomisation to death from any cause
- Secondary endpoints: **Progression-Free Survival (PFS)**, **Objective Response Rate (ORR)**

### Clinical Endpoints Explained

::: {.callout-note}
## Primary and Secondary Endpoints in Oncology

**Overall Survival (OS)**: Time from randomisation to death from any cause. The gold standard efficacy endpoint, directly measuring patient benefit. Regulatory agencies prioritise OS in their approval decisions.

**Progression-Free Survival (PFS)**: Time from randomisation to documented disease progression (per RECIST 1.1) or death from any cause, whichever occurs first. A surrogate endpoint reflecting treatment effect on tumour control.

**Objective Response Rate (ORR)**: Percentage of patients achieving Complete Response (CR) or Partial Response (PR) per RECIST 1.1 criteria. An intermediate endpoint assessing tumour shrinkage.
:::

### RECIST 1.1 Response Categories

RECIST 1.1 defines tumour response based on change in target lesion diameter from baseline:

- **Complete Response (CR)**: Disappearance of all target lesions. ≥30% decrease from baseline.
- **Partial Response (PR)**: ≥30% decrease in sum of target lesion diameters.
- **Stable Disease (SD)**: Neither PR nor PD criteria met.
- **Progressive Disease (PD)**: ≥20% increase in sum of target lesion diameters from nadir.

## 24.2 Data Setup and Preparation

### Exploring the Survival Package Colon Dataset

We use the built-in `survival::colon` dataset, originally from a study of adjuvant therapy for colorectal cancer. We map it to our oncology trial context.

```{r}
#| label: load-packages
#| eval: true

library(survival)
library(tidyverse)
library(broom)
library(gt)

# Load the colon dataset
data(colon)

# View structure and first rows
str(colon)
```

```{r}
#| label: explore-colon
#| eval: true

head(colon, 10)
summary(colon)
```

::: {.callout-info}
The `colon` dataset contains 1,858 observations with two rows per subject (one for each event type). Key variables:

- **id**: Subject identifier
- **time**: Follow-up time in days from entry
- **status**: Event indicator (1 = recurrence/progression, 2 = death)
- **etype**: Event type (1 = recurrence, 2 = death)
- **rx**: Treatment (Obs, Lev, Lev+5FU)
- **age**: Age in years
- **sex**: 1 = male, 0 = female
- **obstruct**: Obstruction of colon (1 = yes, 0 = no)
- **perfor**: Perforation (1 = yes, 0 = no)
- **adhere**: Adherence (1 = yes, 0 = no)
- **nodes**: Number of lymph nodes examined
- **node4**: Number of positive lymph nodes (1 = 4+, 0 = <4)
- **differ**: Grade of differentiation (1 = well, 2 = moderate, 3 = poor)
- **extent**: Extent of spread (1 = confined, 2 = submucosa, 3 = muscle, 4 = beyond)
- **surg**: Type of surgery (0 = short, 1 = long)
:::

### Data Restructuring for ONCO-301

```{r}
#| label: prepare-onco-301
#| eval: true

set.seed(42)

# Create OS dataset: select death events (etype==2) and create one row per subject
colon_os <- colon %>%
  filter(etype == 2) %>%
  select(id, time, status, rx, age, sex, obstruct, perfor, adhere,
         nodes, node4, differ, extent, surg) %>%
  distinct(id, .keep_all = TRUE) %>%
  # Map treatments: "Obs" → Control, "Lev+5FU" → Vemotinib
  mutate(
    rx_new = case_when(
      rx == "Obs" ~ "Control",
      rx == "Lev+5FU" ~ "Vemotinib",
      TRUE ~ NA_character_
    )
  ) %>%
  filter(!is.na(rx_new)) %>%
  select(-rx) %>%
  rename(rx = rx_new) %>%
  # Convert to years for interpretation
  mutate(
    time_years = time / 365.25,
    # Create binary event indicator
    os_event = status
  ) %>%
  arrange(id)

# Check sample sizes
table(colon_os$rx)

# Summary of key variables
colon_os %>%
  group_by(rx) %>%
  summarise(
    n = n(),
    `Median Age` = median(age),
    `Median Time (years)` = median(time_years),
    `Deaths (n)` = sum(os_event),
    `Death Rate (%)` = round(100 * mean(os_event), 1)
  ) %>%
  gt()
```

### Simulating RECIST Response Data

```{r}
#| label: simulate-recist
#| eval: true

# Create unique subject dataset for RECIST (one row per subject)
colon_os_unique <- colon_os %>%
  distinct(id, .keep_all = TRUE) %>%
  mutate(subject_id = row_number())

n_subjects <- nrow(colon_os_unique)

# Simulate RECIST response with treatment-dependent distributions
set.seed(42)

recist_data <- colon_os_unique %>%
  select(subject_id, id, rx, age, node4) %>%
  mutate(
    # Simulate best response by treatment
    best_response = {
      resp    <- character(n())
      is_vemo <- rx == "Vemotinib"
      resp[is_vemo]  <- sample(c("CR","PR","SD","PD"), sum(is_vemo),  replace=TRUE,
                                prob=c(0.12, 0.28, 0.35, 0.25))
      resp[!is_vemo] <- sample(c("CR","PR","SD","PD"), sum(!is_vemo), replace=TRUE,
                                prob=c(0.05, 0.15, 0.40, 0.40))
      resp
    },
    change_pct = {
      chg     <- numeric(n())
      is_vemo <- rx == "Vemotinib"
      chg[is_vemo]  <- rnorm(sum(is_vemo),  mean=-18, sd=28)
      chg[!is_vemo] <- rnorm(sum(!is_vemo), mean=-5,  sd=35)
      chg
    },
    # Responder indicator (CR or PR)
    responder = as.integer(best_response %in% c("CR", "PR"))
  )

# Merge with OS data
colon_combined <- colon_os_unique %>%
  left_join(recist_data %>% select(subject_id, best_response, change_pct, responder),
            by = "subject_id")

head(recist_data, 10)
```

## 24.3 Overall Survival Analysis

### Kaplan-Meier Estimator

The Kaplan-Meier (KM) estimator provides a non-parametric estimate of the survival function:

$$S(t) = \prod_{i: t_i \leq t} \left(1 - \frac{d_i}{n_i}\right)$$

where:
- $d_i$ = number of events at time $t_i$
- $n_i$ = number at risk at time $t_i$

### Fitting the KM Estimator

```{r}
#| label: km-os-fit
#| eval: true

# Fit Kaplan-Meier curves by treatment
km_os <- survfit(Surv(time_years, os_event) ~ rx,
                 data = colon_os)

# Print summary
print(km_os)
```

### KM Survival Curves with Confidence Bands

```{r}
#| label: km-os-plot
#| message: false

# KM plot using base survival + ggplot2 (no ggsurvfit dependency)
km_summary <- summary(km_os)
km_df <- data.frame(
  time   = km_summary$time,
  surv   = km_summary$surv,
  lower  = km_summary$lower,
  upper  = km_summary$upper,
  strata = sub("ARM=", "", as.character(km_summary$strata))
)

ggplot(km_df, aes(x = time, y = surv, colour = strata, fill = strata)) +
  geom_step(linewidth = 0.9) +
  geom_ribbon(aes(ymin = lower, ymax = upper), alpha = 0.15, colour = NA) +
  scale_y_continuous(labels = scales::percent, limits = c(0, 1)) +
  scale_colour_manual(values = c("#2C6E9E", "#E07B39")) +
  scale_fill_manual(  values = c("#2C6E9E", "#E07B39")) +
  labs(
    title  = "ONCO-301 Trial: Overall Survival (Intent-to-Treat Population)",
    x      = "Time (years)",
    y      = "Overall Survival Probability",
    colour = "Treatment",
    fill   = "Treatment"
  ) +
  theme_bw(base_size = 12) +
  theme(
    legend.position  = "bottom",
    plot.title       = element_text(size = 13, face = "bold"),
    panel.grid.minor = element_blank()
  )
```

### Log-Rank Test

The log-rank test compares survival distributions between treatment groups, testing:

$$H_0: S_{\text{Vemotinib}}(t) = S_{\text{Control}}(t) \text{ for all } t$$

```{r}
#| label: logrank-os
#| eval: true

# Perform log-rank test
lr_test <- survdiff(Surv(time_years, os_event) ~ rx,
                    data = colon_os)

print(lr_test)

# Extract test statistic and p-value
lr_chisq <- lr_test$chisq
lr_pval <- 1 - pchisq(lr_chisq, df = 1)

cat("\nLog-Rank Test Results:\n")
cat("Chi-square statistic:", round(lr_chisq, 3), "\n")
cat("Degrees of freedom:", 1, "\n")
cat("P-value:", round(lr_pval, 4), "\n")
```

### Median Overall Survival and Confidence Intervals

```{r}
#| label: median-os
#| eval: true

# Extract survival summary by treatment
summary_os <- as_tibble(summary(km_os)$table) %>%
  rownames_to_column("Treatment") %>%
  select(Treatment, `median`, `0.95LCL`, `0.95UCL`)

colnames(summary_os) <- c("Treatment", "Median (years)", "95% CI Lower", "95% CI Upper")

gt(summary_os) %>%
  fmt_number(columns = c("Median (years)", "95% CI Lower", "95% CI Upper"),
             decimals = 2)
```

::: {.callout-success}
## Clinical Interpretation

The log-rank test result indicates whether Vemotinib provides a statistically significant survival benefit compared to FOLFOX. If p < 0.05, we conclude the survival curves differ significantly between treatments. The median OS values quantify the difference clinically: a difference of 1–2 years in median OS often represents substantial clinical benefit in mCRC.
:::

## 24.4 Cox Proportional Hazards Model for Overall Survival

### Fitting the Cox Model

The Cox proportional hazards model relates covariates to the hazard rate:

$$h(t | X) = h_0(t) \exp(\beta_1 X_1 + \beta_2 X_2 + \cdots + \beta_p X_p)$$

The hazard ratio for treatment is $\exp(\beta_{\text{rx}})$, representing the relative risk of death.

```{r}
#| label: cox-os-fit
#| eval: true

# Fit Cox model with treatment and key prognostic factors
cox_os <- coxph(Surv(time_years, os_event) ~ rx + age + sex + obstruct + node4,
                data = colon_os)

# Display model summary
summary(cox_os)
```

### Extracting Hazard Ratios and Confidence Intervals

```{r}
#| label: cox-os-tidy
#| eval: true

# Use broom::tidy for clean extraction
cox_tidy <- tidy(cox_os, exponentiate = TRUE, conf.int = TRUE) %>%
  select(term, estimate, conf.low, conf.high, p.value) %>%
  mutate(
    term = case_when(
      term == "rxVemotinib" ~ "Vemotinib vs Control",
      term == "age" ~ "Age (per year)",
      term == "sex" ~ "Sex (male vs female)",
      term == "obstruct" ~ "Obstruction (yes vs no)",
      term == "node4" ~ "≥4 Positive Nodes (yes vs no)",
      TRUE ~ term
    ),
    across(estimate:p.value, ~round(., 4))
  )

gt(cox_tidy) %>%
  cols_label(
    term = "Variable",
    estimate = "Hazard Ratio",
    conf.low = "95% CI Lower",
    conf.high = "95% CI Upper",
    p.value = "P-value"
  ) %>%
  fmt_number(columns = estimate:p.value, decimals = 4)
```

### Checking Proportional Hazards Assumption

The Cox model assumes proportional hazards: the hazard ratio between groups is constant over time. We test this using Schoenfeld residuals.

```{r}
#| label: cox-ph-test
#| eval: true

# Test proportional hazards assumption
ph_test <- cox.zph(cox_os)
print(ph_test)

# Interpretation: large p-values (>0.05) suggest the PH assumption holds
```

::: {.callout-warning}
The proportional hazards assumption is critical for Cox model validity. If any variable has p < 0.05 in the `cox.zph()` test, the PH assumption is violated and stratified or time-varying coefficient models may be needed. In this example, if the assumption is violated, consider fitting a stratified model: `coxph(Surv(time_years, os_event) ~ rx + age + strata(node4), data = colon_os)`.
:::

### Forest Plot of Hazard Ratios

```{r}
#| label: forest-os-ggplot
#| eval: true

forest_data <- cox_tidy %>%
  mutate(
    term = fct_reorder(term, estimate),
    significant = if_else(p.value < 0.05, "Yes", "No")
  )

forest_plot <- ggplot(forest_data, aes(x = estimate, y = term)) +
  geom_vline(xintercept = 1, linetype = "dashed", colour = "red", linewidth = 0.8) +
  geom_point(aes(colour = significant), size = 4) +
  geom_errorbarh(aes(xmin = conf.low, xmax = conf.high, colour = significant),
                 height = 0.2, linewidth = 1) +
  scale_x_log10() +
  scale_colour_manual(values = c("Yes" = "#e41a1c", "No" = "#377eb8")) +
  labs(
    title = "ONCO-301: Cox Model Hazard Ratios for Overall Survival",
    x = "Hazard Ratio (log scale, <1 = benefit)",
    y = NULL,
    colour = "Significant\n(p<0.05)"
  ) +
  theme_minimal() +
  theme(
    plot.title = element_text(size = 13, face = "bold"),
    panel.grid.major.x = element_line(colour = "grey90"),
    legend.position = "bottom"
  )

forest_plot
```

## 24.5 Progression-Free Survival Analysis

PFS (time to progression or death) is evaluated using the recurrence events from the colon dataset (etype==1).

```{r}
#| label: pfs-data
#| eval: true

# Create PFS dataset: select recurrence events (etype==1)
colon_pfs <- colon %>%
  filter(etype == 1) %>%
  select(id, time, status, rx, age, sex, node4) %>%
  distinct(id, .keep_all = TRUE) %>%
  mutate(
    rx_new = case_when(
      rx == "Obs" ~ "Control",
      rx == "Lev+5FU" ~ "Vemotinib",
      TRUE ~ NA_character_
    )
  ) %>%
  filter(!is.na(rx_new)) %>%
  select(-rx) %>%
  rename(rx = rx_new) %>%
  mutate(
    time_years = time / 365.25,
    pfs_event = status
  )

# Fit KM for PFS
km_pfs <- survfit(Surv(time_years, pfs_event) ~ rx, data = colon_pfs)
print(km_pfs)
```

### PFS Log-Rank Test

```{r}
#| label: logrank-pfs
#| eval: true

lr_pfs <- survdiff(Surv(time_years, pfs_event) ~ rx, data = colon_pfs)
print(lr_pfs)
```

### Cox Model for PFS

```{r}
#| label: cox-pfs
#| eval: true

cox_pfs <- coxph(Surv(time_years, pfs_event) ~ rx + age + sex + node4,
                 data = colon_pfs)

cox_pfs_tidy <- tidy(cox_pfs, exponentiate = TRUE, conf.int = TRUE) %>%
  filter(term == "rxVemotinib") %>%
  mutate(endpoint = "PFS")

cox_os_comp <- tidy(cox_os, exponentiate = TRUE, conf.int = TRUE) %>%
  filter(term == "rxVemotinib") %>%
  mutate(endpoint = "OS")

comparison <- bind_rows(cox_os_comp, cox_pfs_tidy) %>%
  select(endpoint, estimate, conf.low, conf.high, p.value) %>%
  mutate(across(estimate:p.value, ~round(., 4)))

gt(comparison) %>%
  cols_label(
    endpoint = "Endpoint",
    estimate = "HR (Vemotinib)",
    conf.low = "95% CI Lower",
    conf.high = "95% CI Upper",
    p.value = "P-value"
  )
```

::: {.callout-info}
**OS vs PFS**: PFS is typically achieved earlier than OS and may show treatment differences earlier in a trial. However, OS is the definitive efficacy endpoint. Both endpoints should be pre-specified and reported to provide a complete picture of treatment effect.
:::

## 24.6 Objective Response Rate Analysis

### ORR by Treatment Arm

```{r}
#| label: orr-summary
#| eval: true

orr_summary <- recist_data %>%
  group_by(rx) %>%
  summarise(
    n = n(),
    `CR (n)` = sum(best_response == "CR"),
    `PR (n)` = sum(best_response == "PR"),
    `SD (n)` = sum(best_response == "SD"),
    `PD (n)` = sum(best_response == "PD"),
    `Responders (CR+PR)` = sum(responder),
    `ORR (%)` = round(100 * mean(responder), 1),
    `95% CI` = sprintf("%.1f–%.1f",
                       100 * (mean(responder) - 1.96 * sqrt(mean(responder) *
                              (1 - mean(responder)) / n())),
                       100 * (mean(responder) + 1.96 * sqrt(mean(responder) *
                              (1 - mean(responder)) / n())))
  )

gt(orr_summary)
```

### Logistic Regression for ORR

```{r}
#| label: logistic-orr
#| eval: true

# Logistic model for response
glm_orr <- glm(responder ~ rx + age + node4,
               family = binomial(link = "logit"),
               data = recist_data)

# Extract odds ratios
glm_tidy <- tidy(glm_orr, exponentiate = TRUE, conf.int = TRUE) %>%
  filter(term != "(Intercept)") %>%
  select(term, estimate, conf.low, conf.high, p.value) %>%
  mutate(
    term = case_when(
      term == "rxVemotinib" ~ "Vemotinib vs Control",
      term == "age" ~ "Age (per year)",
      term == "node4" ~ "≥4 Positive Nodes",
      TRUE ~ term
    ),
    across(estimate:p.value, ~round(., 4))
  )

gt(glm_tidy) %>%
  cols_label(
    term = "Variable",
    estimate = "Odds Ratio",
    conf.low = "95% CI Lower",
    conf.high = "95% CI Upper",
    p.value = "P-value"
  )
```

### Chi-Square Test for ORR Difference

```{r}
#| label: chisq-orr
#| eval: true

# Create contingency table
ct_orr <- table(recist_data$rx, recist_data$responder)
print(ct_orr)

# Chi-square test
chisq_orr <- chisq.test(ct_orr)
print(chisq_orr)

cat("\nInterpretation: Chi-square statistic =", round(chisq_orr$statistic, 3),
    ", p-value =", round(chisq_orr$p.value, 4), "\n")
```

## 24.7 Waterfall Plot (Tumour Response)

The waterfall plot is a classic oncology visualization showing individual patient responses sorted by percent change from baseline.

```{r}
#| label: waterfall-plot
#| eval: true

# Prepare waterfall data
waterfall_data <- recist_data %>%
  arrange(change_pct) %>%
  mutate(
    patient_order = row_number(),
    response_color = case_when(
      best_response == "CR" ~ "#1b7837",
      best_response == "PR" ~ "#66bd63",
      best_response == "SD" ~ "#fee08b",
      best_response == "PD" ~ "#d73027"
    )
  )

# Create waterfall plot
waterfall_plot <- ggplot(waterfall_data,
                         aes(x = reorder(subject_id, change_pct),
                             y = change_pct,
                             fill = best_response)) +
  geom_col(width = 1) +
  geom_hline(yintercept = -30, linetype = "dashed", colour = "black", linewidth = 0.7) +
  geom_hline(yintercept = 20, linetype = "dashed", colour = "black", linewidth = 0.7) +
  scale_fill_manual(
    values = c("CR" = "#1b7837", "PR" = "#66bd63", "SD" = "#fee08b", "PD" = "#d73027"),
    name = "RECIST Response"
  ) +
  labs(
    title = "ONCO-301: Waterfall Plot of Best Percentage Change in Target Lesion Sum",
    x = "Patient",
    y = "% Change from Baseline"
  ) +
  theme_minimal() +
  theme(
    axis.text.x = element_blank(),
    axis.ticks.x = element_blank(),
    plot.title = element_text(size = 13, face = "bold"),
    legend.position = "bottom"
  ) +
  annotate("text", x = Inf, y = -30, label = "PR threshold (-30%)",
           hjust = 1.1, vjust = -0.3, size = 3) +
  annotate("text", x = Inf, y = 20, label = "PD threshold (+20%)",
           hjust = 1.1, vjust = -0.3, size = 3)

waterfall_plot
```

## 24.8 Subgroup Analysis (Forest Plot)

Subgroup analyses evaluate whether treatment effects are consistent across patient populations.

```{r}
#| label: subgroup-analysis
#| eval: true

# Define subgroups
subgroup_vars <- list(
  age_group = list(var = "age", breaks = c(0, 65, Inf),
                   labels = c("<65 years", "≥65 years")),
  sex_group = list(var = "sex", breaks = c(-0.5, 0.5, 1.5),
                   labels = c("Female", "Male")),
  node_status = list(var = "node4", breaks = c(-0.5, 0.5, 1.5),
                     labels = c("<4 nodes", "≥4 nodes")),
  obstruction = list(var = "obstruct", breaks = c(-0.5, 0.5, 1.5),
                     labels = c("No obstruction", "Obstruction"))
)

# Function to fit Cox model in subgroup
fit_subgroup_cox <- function(data, subgroup_var, subgroup_label) {
  data_sub <- data %>%
    mutate(subgroup = subgroup_label)

  cox_sub <- coxph(Surv(time_years, os_event) ~ rx, data = data_sub)

  tidy(cox_sub, exponentiate = TRUE, conf.int = TRUE) %>%
    filter(term == "rxVemotinib") %>%
    mutate(
      subgroup_var = subgroup_var,
      subgroup_label = subgroup_label,
      n = nrow(data_sub)
    )
}

# Age subgroup
subgroup_results <- bind_rows(
  # <65 years
  colon_os %>%
    filter(age < 65) %>%
    nest() %>%
    mutate(result = map_df(data,
                           ~fit_subgroup_cox(., "age", "<65 years"))) %>%
    pull(result),
  # ≥65 years
  colon_os %>%
    filter(age >= 65) %>%
    nest() %>%
    mutate(result = map_df(data,
                           ~fit_subgroup_cox(., "age", "≥65 years"))) %>%
    pull(result),
  # Female
  colon_os %>%
    filter(sex == 0) %>%
    nest() %>%
    mutate(result = map_df(data,
                           ~fit_subgroup_cox(., "sex", "Female"))) %>%
    pull(result),
  # Male
  colon_os %>%
    filter(sex == 1) %>%
    nest() %>%
    mutate(result = map_df(data,
                           ~fit_subgroup_cox(., "sex", "Male"))) %>%
    pull(result)
) %>%
  select(subgroup_var, subgroup_label, estimate, conf.low, conf.high, p.value, n)

# Print subgroup results
head(subgroup_results, 8)
```

### Subgroup Forest Plot

```{r}
#| label: subgroup-forest
#| eval: true

subgroup_forest <- subgroup_results %>%
  mutate(
    subgroup_label = factor(subgroup_label,
                           levels = unique(subgroup_label)),
    significant = if_else(p.value < 0.05, "p<0.05", "p≥0.05")
  )

forest_sg <- ggplot(subgroup_forest,
                    aes(x = estimate, y = subgroup_label, colour = significant)) +
  geom_vline(xintercept = 1, linetype = "dashed", colour = "red", linewidth = 0.8) +
  geom_point(size = 4) +
  geom_errorbarh(aes(xmin = conf.low, xmax = conf.high),
                 height = 0.2, linewidth = 1) +
  scale_x_log10() +
  scale_colour_manual(values = c("p<0.05" = "#e41a1c", "p≥0.05" = "#377eb8")) +
  facet_wrap(~subgroup_var, scales = "free_y") +
  labs(
    title = "ONCO-301: Subgroup Analysis of Vemotinib Effect on Overall Survival",
    x = "Hazard Ratio (95% CI)",
    y = NULL,
    colour = "Significance"
  ) +
  theme_minimal() +
  theme(
    plot.title = element_text(size = 13, face = "bold"),
    panel.grid.major.x = element_line(colour = "grey90")
  )

forest_sg
```

::: {.callout-success}
## Interpretation of Subgroup Analyses

Consistent hazard ratios across subgroups with overlapping confidence intervals suggest a homogeneous treatment effect—the benefit of Vemotinib does not differ meaningfully by age, sex, or disease extent. This supports broad applicability. Conversely, significant interactions (non-overlapping CIs, p < 0.05) may prompt further investigation and subgroup labelling in the labelling information.
:::

## 24.9 Safety Summary

### Simulated Adverse Event Data

```{r}
#| label: simulate-ae
#| eval: true

set.seed(42)

# Simulate AE data
ae_socs <- c("Gastrointestinal", "Hematologic", "Constitutional",
             "Skin", "Renal", "Cardiac")

ae_data <- expand_grid(
  subject_id = 1:nrow(colon_os_unique),
  soc = ae_socs
) %>%
  left_join(colon_os_unique %>% select(subject_id, rx),
            by = "subject_id") %>%
  mutate(
    # Incidence varies by treatment and SOC
    ae_occurred = case_when(
      rx == "Vemotinib" & soc == "Skin" ~ rbinom(n(), 1, prob = 0.35),
      rx == "Vemotinib" & soc == "Gastrointestinal" ~ rbinom(n(), 1, prob = 0.40),
      rx == "Vemotinib" & soc == "Hematologic" ~ rbinom(n(), 1, prob = 0.25),
      rx == "Control" & soc == "Gastrointestinal" ~ rbinom(n(), 1, prob = 0.55),
      rx == "Control" & soc == "Hematologic" ~ rbinom(n(), 1, prob = 0.45),
      TRUE ~ rbinom(n(), 1, prob = 0.15)
    ),
    # Grade for occurred AEs
    grade = if_else(ae_occurred == 1,
                    sample(1:4, size = n(), replace = TRUE,
                           prob = c(0.30, 0.35, 0.25, 0.10)),
                    NA_integer_)
  ) %>%
  filter(ae_occurred == 1)

# Grade 3-4 AE summary by treatment
ae_summary <- ae_data %>%
  mutate(grade_34 = as.integer(grade >= 3)) %>%
  group_by(rx, soc) %>%
  summarise(
    `Any Grade (n, %)` = paste0(n(), " (",
                               round(100 * n() / nrow(colon_os_unique) * 2, 0), ")"),
    `Grade 3-4 (n, %)` = paste0(sum(grade_34), " (",
                               round(100 * sum(grade_34) / nrow(colon_os_unique) * 2, 0), ")"),
    .groups = "drop"
  ) %>%
  pivot_wider(
    names_from = rx,
    values_from = c(`Any Grade (n, %)`, `Grade 3-4 (n, %)`),
    names_glue = "{rx}_{.value}"
  )

col_labels <- list(soc = "System Organ Class")
for (arm in c("Control", "Vemotinib")) {
  ag_col  <- paste0(arm, "_Any Grade (n, %)")
  g34_col <- paste0(arm, "_Grade 3-4 (n, %)")
  if (ag_col  %in% names(ae_summary)) col_labels[[ag_col]]  <- paste0(arm, ": Any Grade")
  if (g34_col %in% names(ae_summary)) col_labels[[g34_col]] <- paste0(arm, ": Grade 3-4")
}
gt(ae_summary) %>%
  gt::cols_label(.list = col_labels)
```

::: {.callout-info}
Safety reporting follows CTCAE v5.0 (Common Terminology Criteria for Adverse Events) grading: Grade 1 (mild), Grade 2 (moderate), Grade 3–4 (severe/life-threatening). Regulatory submissions require comprehensive AE tables by SOC/PT (System Organ Class / Preferred Term), treatment arm, and grade.
:::

## 24.10 Reporting Findings

### CONSORT-Style Efficacy Reporting

When reporting the primary efficacy result, adopt this structured format:

::: {.callout-note}
**Sample Primary Efficacy Result (CONSORT style)**:

"In the intent-to-treat population, median overall survival was **3.2 years** (95% CI: 2.8–3.5) in the Vemotinib arm vs **2.4 years** (95% CI: 2.1–2.8) in the FOLFOX control arm. Vemotinib reduced the risk of death by 32% (hazard ratio [HR] 0.68, 95% CI: 0.54–0.86; p = 0.001 by log-rank test). The OS benefit was consistent across pre-specified subgroups of age, sex, and disease extent."
:::

### Key Tables for a Clinical Study Report (CSR)

1. **Baseline Characteristics Table**: Patient demographics, disease characteristics, baseline labs by treatment
2. **Primary Efficacy Table**: Primary endpoint results with KM estimates, log-rank test, Cox HR
3. **Secondary Efficacy Table**: PFS, ORR with confidence intervals and p-values
4. **Subgroup Analysis Table**: HR with 95% CI for each subgroup
5. **Adverse Event Summary Table**: By SOC and grade
6. **Patient Disposition Table**: Enrolled, randomised, treated, completed, withdrawn, reasons for discontinuation

```{r}
#| label: baseline-table
#| eval: true

# Baseline characteristics table
baseline_table <- colon_os %>%
  group_by(rx) %>%
  summarise(
    n = n(),
    `Age (median, IQR)` = paste0(median(age), " (",
                                quantile(age, 0.25), "–",
                                quantile(age, 0.75), ")"),
    `Male (n, %)` = paste0(sum(sex == 1), " (",
                          round(100 * mean(sex == 1), 1), ")"),
    `Obstruction (n, %)` = paste0(sum(obstruct == 1), " (",
                                 round(100 * mean(obstruct == 1), 1), ")"),
    `≥4 Positive Nodes (n, %)` = paste0(sum(node4 == 1), " (",
                                       round(100 * mean(node4 == 1), 1), ")")
  ) %>%
  mutate(across(-rx, as.character)) %>%
  pivot_longer(cols = -rx, names_to = "Characteristic", values_to = "Value") %>%
  pivot_wider(names_from = rx, values_from = Value)

gt(baseline_table) %>%
  cols_label(Characteristic = "Baseline Characteristic")
```

## Exercises

### Beginner Exercises

**Exercise 24.1**: Replicate the Kaplan-Meier plot for OS but stratify by both `rx` and `node4` (positive node status). How does node status modify the treatment effect visually?

**Exercise 24.2**: Calculate the ORR for each treatment arm independently using base R (without the recist_data table). Use the formula: ORR = (CR + PR) / n × 100. Then compute the 95% exact binomial confidence interval for each arm using `binom.test()`.

**Exercise 24.3**: Fit a Cox model for PFS including treatment, age, and sex. Extract the hazard ratio for Vemotinib and interpret it clinically. Is the PFS benefit greater or smaller than the OS benefit?

### Advanced Exercises

**Exercise 24.4**: Perform a landmark analysis at 1 year of follow-up. Restrict the OS dataset to patients with at least 1 year of follow-up, then re-fit the Cox model. Compare the treatment effect at the 1-year landmark to the full dataset Cox HR. What does this reveal about the timing of treatment benefit?

**Exercise 24.5**: Conduct a competing risks analysis using the `cmprsk` package. For PFS, treat death without progression as a competing event. Fit a Fine-Gray model (subdistribution hazard) and compare the treatment effect to the standard Cox model. Interpret the differences.

---

## Summary

This chapter demonstrated a complete Phase III oncology trial analysis using real-world concepts:

- **Overall Survival Analysis**: Kaplan-Meier curves, log-rank test, Cox proportional hazards regression
- **Progression-Free Survival**: Secondary efficacy endpoint with similar methods
- **Response Evaluation**: Logistic regression for ORR, chi-square testing
- **Data Visualization**: Waterfall plots for individual responses, forest plots for subgroups
- **Subgroup Analyses**: Investigating treatment heterogeneity across patient populations
- **Safety Reporting**: AE summaries by SOC and grade

All analyses followed regulatory standards (CONSORT, ICH-GCP) and clinical best practices. The methods extend directly to real oncology trials and disease areas.
