# Probability Distributions in Clinical Research

## Introduction

Probability distributions are the mathematical foundation underlying every statistical method used in clinical research. From modelling the number of adverse events in a safety population to computing confidence intervals for a primary efficacy endpoint, understanding distributions is essential for both the design and analysis of clinical trials.

This chapter covers:

- **Discrete distributions**: Binomial and Poisson, used for counts and binary outcomes
- **Continuous distributions**: Normal, t, Chi-square, and F, used for measurements and test statistics
- **Key concepts**: Probability density/mass functions, CDFs, quantiles, expected value, and variance
- **Clinical applications**: Sample size planning, adverse event rate modelling, and simulation
- **Visualisation**: Creating publication-ready distribution plots and diagnostic QQ-plots

---

## 1. Fundamental Concepts

### 1.1 Probability Mass and Density Functions

For a **discrete** random variable $X$, the probability mass function (PMF) gives the probability of each exact outcome:

$$P(X = k) = f(k), \quad \sum_k f(k) = 1$$

For a **continuous** random variable $X$, the probability density function (PDF) satisfies:

$$P(a \leq X \leq b) = \int_a^b f(x)\,dx, \quad \int_{-\infty}^{\infty} f(x)\,dx = 1$$

### 1.2 Cumulative Distribution Function

The CDF gives the probability that $X$ takes a value at most $x$:

$$F(x) = P(X \leq x)$$

### 1.3 Expected Value and Variance

For a discrete distribution: $E[X] = \sum_k k \cdot f(k)$

$$\text{Var}(X) = E[X^2] - (E[X])^2$$

```{r}
# Load packages used throughout this chapter
library(dplyr)
library(ggplot2)
library(tidyr)

# Consistent theme for all plots
theme_clinical <- function() {
  theme_bw(base_size = 12) +
    theme(
      plot.title    = element_text(face = "bold", size = 13),
      plot.subtitle = element_text(colour = "grey40"),
      legend.position = "bottom"
    )
}
```

---

## 2. Discrete Distributions

### 2.1 Binomial Distribution

The **Binomial distribution** models the number of successes in $n$ independent Bernoulli trials each with probability $p$:

$$P(X = k) = \binom{n}{k} p^k (1-p)^{n-k}, \quad k = 0, 1, \ldots, n$$

$$E[X] = np, \qquad \text{Var}(X) = np(1-p)$$

**Clinical example**: In a Phase II oncology trial, historical data suggest that 30 % of patients with a given tumour type respond to standard therapy. If we enrol 20 patients, what is the probability that exactly 8 respond?

```{r}
# P(X = 8) where X ~ Binomial(n=20, p=0.30)
n <- 20
p_response <- 0.30

prob_exactly_8 <- dbinom(8, size = n, prob = p_response)
cat(sprintf("P(X = 8)  = %.4f\n", prob_exactly_8))

# P(X >= 8): probability of 8 or more responders
prob_8_or_more <- pbinom(7, size = n, prob = p_response, lower.tail = FALSE)
cat(sprintf("P(X >= 8) = %.4f\n", prob_8_or_more))

# 95th percentile of the distribution
q95 <- qbinom(0.95, size = n, prob = p_response)
cat(sprintf("95th percentile: %d\n", q95))
```

```{r}
# Generate random binomial samples (simulate 1 000 trials of n=20)
set.seed(42)
sim_responses <- rbinom(1000, size = 20, prob = 0.30)
cat(sprintf("Simulated mean  : %.2f  (theoretical: %.2f)\n",
            mean(sim_responses), 20 * 0.30))
cat(sprintf("Simulated SD    : %.2f  (theoretical: %.2f)\n",
            sd(sim_responses), sqrt(20 * 0.30 * 0.70)))
```

```{r fig.cap="PMF of Binomial(20, 0.30) with the observed response highlighted"}
# Visualise the PMF
binom_df <- data.frame(
  k    = 0:20,
  prob = dbinom(0:20, size = 20, prob = 0.30),
  highlight = ifelse(0:20 == 8, "k = 8", "other")
)

ggplot(binom_df, aes(x = k, y = prob, fill = highlight)) +
  geom_col(colour = "white", width = 0.7) +
  scale_fill_manual(values = c("k = 8" = "#E63946", "other" = "#457B9D"),
                    name = "") +
  labs(
    title    = "Binomial(n = 20, p = 0.30) — Oncology Response Example",
    subtitle = "Probability of each number of responders out of 20 patients",
    x        = "Number of responders (k)",
    y        = "P(X = k)"
  ) +
  theme_clinical()
```

#### 2.1.1 Comparing Response Rates Across Arms

We can overlay Binomial PMFs for different response probability assumptions:

```{r fig.cap="Binomial PMFs under three different response rate assumptions"}
probs_compare <- c(0.20, 0.30, 0.45)
labels_compare <- c("20% (control)", "30% (low dose)", "45% (high dose)")

pmf_data <- lapply(seq_along(probs_compare), function(i) {
  data.frame(
    k     = 0:20,
    prob  = dbinom(0:20, size = 20, prob = probs_compare[i]),
    group = labels_compare[i]
  )
}) |> bind_rows()

ggplot(pmf_data, aes(x = k, y = prob, colour = group)) +
  geom_line(linewidth = 1.1) +
  geom_point(size = 2) +
  scale_colour_brewer(palette = "Set1", name = "Arm") +
  labs(
    title    = "Binomial PMFs — Three Response Rate Scenarios",
    subtitle = "n = 20 patients per arm",
    x        = "Number of responders",
    y        = "Probability"
  ) +
  theme_clinical()
```

### 2.2 Poisson Distribution

The **Poisson distribution** models the number of events occurring in a fixed interval when events happen independently at a constant rate $\lambda$:

$$P(X = k) = \frac{e^{-\lambda} \lambda^k}{k!}, \quad k = 0, 1, 2, \ldots$$

$$E[X] = \lambda, \qquad \text{Var}(X) = \lambda$$

**Clinical example**: Suppose a drug's safety database indicates a serious adverse event (SAE) rate of 2.1 per 100 patient-years. For a 200-patient trial with a mean exposure of 0.8 years, the expected total SAE count is $\lambda = 2.1 \times 200 \times 0.8 / 100 = 3.36$. What is the probability of observing 5 or more SAEs?

```{r}
# Poisson AE rate modelling
lambda_sae <- 2.1 / 100   # events per patient-year
n_patients  <- 200
mean_yrs    <- 0.8

lambda_total <- lambda_sae * n_patients * mean_yrs
cat(sprintf("Expected SAE count: %.2f\n", lambda_total))

prob_5plus <- ppois(4, lambda = lambda_total, lower.tail = FALSE)
cat(sprintf("P(SAEs >= 5) = %.4f\n", prob_5plus))

# Distribution of possible SAE counts
k_range <- 0:15
poisson_pmf <- dpois(k_range, lambda = lambda_total)

# Tabulate
poisson_tbl <- data.frame(
  k           = k_range,
  Probability = round(poisson_pmf, 4),
  Cumulative  = round(ppois(k_range, lambda = lambda_total), 4)
)
print(head(poisson_tbl, 12))
```

```{r fig.cap="Poisson PMF for SAE count modelling (lambda = 3.36)"}
ggplot(data.frame(k = k_range, prob = poisson_pmf),
       aes(x = k, y = prob,
           fill = ifelse(k >= 5, ">=5 SAEs", "<5 SAEs"))) +
  geom_col(colour = "white", width = 0.7) +
  scale_fill_manual(values = c(">=5 SAEs" = "#E63946", "<5 SAEs" = "#2A9D8F"),
                    name = "") +
  labs(
    title    = "Poisson(λ = 3.36) — SAE Count Distribution",
    subtitle = "200 patients × 0.8 years; SAE rate = 2.1 per 100 patient-years",
    x        = "Number of SAEs",
    y        = "P(X = k)"
  ) +
  theme_clinical()
```

#### 2.2.1 Simulating AE Rates with Poisson

```{r}
set.seed(42)

# Simulate per-patient AE counts for three arms
n_per_arm <- 87   # typical phase III arm size

ae_placebo  <- rpois(n_per_arm, lambda = 0.8)
ae_low      <- rpois(n_per_arm, lambda = 1.4)
ae_high     <- rpois(n_per_arm, lambda = 2.0)

ae_sim <- data.frame(
  AE_count = c(ae_placebo, ae_low, ae_high),
  ARM      = rep(c("Placebo", "Xanomeline Low Dose",
                   "Xanomeline High Dose"), each = n_per_arm)
)

# Summary
ae_sim |>
  group_by(ARM) |>
  summarise(
    Mean_AEs = round(mean(AE_count), 2),
    SD_AEs   = round(sd(AE_count), 2),
    Rate_per_100 = round(mean(AE_count) * 100, 1)
  )
```

---

## 3. Continuous Distributions

### 3.1 Normal Distribution

The **Normal (Gaussian) distribution** is central to parametric statistics. Its PDF is:

$$f(x) = \frac{1}{\sigma\sqrt{2\pi}} \exp\!\left(-\frac{(x-\mu)^2}{2\sigma^2}\right), \quad -\infty < x < \infty$$

$$E[X] = \mu, \qquad \text{Var}(X) = \sigma^2$$

```{r}
# Standard Normal: P(-1.96 < Z < 1.96)
prob_95ci <- pnorm(1.96) - pnorm(-1.96)
cat(sprintf("P(-1.96 < Z < 1.96) = %.4f\n", prob_95ci))

# Critical values for common significance levels
alpha_levels <- c(0.10, 0.05, 0.025, 0.01)
z_crits <- qnorm(1 - alpha_levels)
cat("\nCritical values (one-tailed):\n")
for (i in seq_along(alpha_levels)) {
  cat(sprintf("  alpha = %.3f  ->  z = %.4f\n", alpha_levels[i], z_crits[i]))
}
```

```{r}
# Clinical example: Systolic blood pressure in a population
# SBP ~ Normal(mu=125, sigma=15)
mu_sbp    <- 125
sigma_sbp <- 15

# P(SBP > 140): probability of hypertensive range
prob_hyper <- pnorm(140, mean = mu_sbp, sd = sigma_sbp, lower.tail = FALSE)
cat(sprintf("P(SBP > 140) = %.4f  (%.1f%%)\n",
            prob_hyper, prob_hyper * 100))

# 90th percentile SBP
q90_sbp <- qnorm(0.90, mean = mu_sbp, sd = sigma_sbp)
cat(sprintf("90th percentile SBP = %.1f mmHg\n", q90_sbp))
```

```{r fig.cap="Normal distribution for systolic blood pressure with shaded hypertensive region"}
x_sbp <- seq(80, 175, length.out = 500)
df_sbp <- data.frame(x = x_sbp, y = dnorm(x_sbp, mu_sbp, sigma_sbp))

ggplot(df_sbp, aes(x, y)) +
  geom_line(linewidth = 1.2, colour = "#1D3557") +
  geom_area(data = df_sbp |> filter(x >= 140),
            aes(x, y), fill = "#E63946", alpha = 0.4) +
  geom_vline(xintercept = 140, linetype = "dashed", colour = "#E63946") +
  annotate("text", x = 152, y = 0.012,
           label = sprintf("P(SBP>140)\n= %.1f%%", prob_hyper * 100),
           colour = "#E63946", size = 3.5) +
  labs(
    title    = "Normal(μ=125, σ=15) — Systolic Blood Pressure",
    subtitle = "Shaded area: proportion in hypertensive range (>140 mmHg)",
    x        = "SBP (mmHg)",
    y        = "Density"
  ) +
  theme_clinical()
```

### 3.2 Student's t-Distribution

The **t-distribution** with $\nu$ degrees of freedom arises when estimating the population mean from a sample with unknown variance. As $\nu \to \infty$, it converges to the standard Normal.

$$f(t) = \frac{\Gamma\!\left(\frac{\nu+1}{2}\right)}{\sqrt{\nu\pi}\,\Gamma\!\left(\frac{\nu}{2}\right)} \left(1 + \frac{t^2}{\nu}\right)^{-\frac{\nu+1}{2}}$$

```{r fig.cap="t-distributions with various degrees of freedom versus the standard Normal"}
x_t <- seq(-5, 5, length.out = 400)
dfs <- c(1, 3, 10, 30)

t_df <- lapply(dfs, function(df_val) {
  data.frame(x = x_t,
             y = dt(x_t, df = df_val),
             df_label = paste0("t (df=", df_val, ")"))
}) |> bind_rows()

# Add Normal reference
norm_ref <- data.frame(x = x_t,
                        y = dnorm(x_t),
                        df_label = "Normal (reference)")
t_df <- bind_rows(t_df, norm_ref)

ggplot(t_df, aes(x = x, y = y, colour = df_label, linetype = df_label)) +
  geom_line(linewidth = 1.0) +
  scale_colour_manual(values = c("#E63946", "#457B9D", "#2A9D8F",
                                  "#F4A261", "#1D3557"),
                      name = "") +
  scale_linetype_manual(values = c("dashed","solid","solid","solid","solid"),
                         name = "") +
  labs(
    title    = "Student's t-Distribution vs. Standard Normal",
    subtitle = "Heavier tails with fewer degrees of freedom",
    x        = "t", y        = "Density"
  ) +
  coord_cartesian(ylim = c(0, 0.42)) +
  theme_clinical()
```

```{r}
# Two-sided critical value for a t-test with 25 df at alpha=0.05
t_crit_25 <- qt(0.975, df = 25)
cat(sprintf("t-critical (df=25, alpha=0.05, two-sided): %.4f\n", t_crit_25))

# Compare with z-critical
cat(sprintf("z-critical (large sample)               : %.4f\n", qnorm(0.975)))
```

### 3.3 Chi-Square Distribution

The **Chi-square distribution** with $\nu$ degrees of freedom is the distribution of the sum of squares of $\nu$ independent standard Normal variables:

$$f(x) = \frac{x^{\nu/2 - 1} e^{-x/2}}{2^{\nu/2}\,\Gamma(\nu/2)}, \quad x > 0$$

$$E[X] = \nu, \qquad \text{Var}(X) = 2\nu$$

It is used in goodness-of-fit tests, tests of independence in contingency tables, and in constructing confidence intervals for the variance.

```{r}
# Chi-square critical values
dfs_chi <- c(1, 2, 3, 5, 10)
chi_crits <- qchisq(0.95, df = dfs_chi)

cat("Chi-square 95th percentiles:\n")
for (i in seq_along(dfs_chi)) {
  cat(sprintf("  df = %2d  ->  X^2_0.95 = %.3f\n",
              dfs_chi[i], chi_crits[i]))
}
```

```{r fig.cap="Chi-square PDFs for various degrees of freedom"}
x_chi <- seq(0.01, 30, length.out = 500)
dfs_plot <- c(1, 2, 3, 5, 10)

chi_df <- lapply(dfs_plot, function(df_val) {
  data.frame(x = x_chi,
             y = dchisq(x_chi, df = df_val),
             df_label = paste0("df = ", df_val))
}) |> bind_rows()

ggplot(chi_df, aes(x = x, y = y, colour = df_label)) +
  geom_line(linewidth = 1.1) +
  scale_colour_brewer(palette = "Set1", name = "Degrees of freedom") +
  coord_cartesian(ylim = c(0, 0.5)) +
  labs(
    title = "Chi-Square Distribution PDFs",
    x     = expression(chi^2),
    y     = "Density"
  ) +
  theme_clinical()
```

### 3.4 F-Distribution

The **F-distribution** with $d_1$ and $d_2$ degrees of freedom is the ratio of two independent Chi-square variates divided by their respective df. It underpins ANOVA and regression F-tests.

$$F = \frac{\chi^2_{d_1}/d_1}{\chi^2_{d_2}/d_2}$$

$$E[F] = \frac{d_2}{d_2 - 2} \text{ for } d_2 > 2$$

```{r fig.cap="F-distribution PDFs for three degree-of-freedom combinations"}
x_f <- seq(0.01, 6, length.out = 500)
f_params <- list(
  list(d1 = 1,  d2 = 10,  lab = "F(1, 10)"),
  list(d1 = 2,  d2 = 20,  lab = "F(2, 20)"),
  list(d1 = 5,  d2 = 50,  lab = "F(5, 50)")
)

f_data <- lapply(f_params, function(p) {
  data.frame(x = x_f,
             y = df(x_f, df1 = p$d1, df2 = p$d2),
             label = p$lab)
}) |> bind_rows()

ggplot(f_data, aes(x = x, y = y, colour = label)) +
  geom_line(linewidth = 1.1) +
  scale_colour_brewer(palette = "Dark2", name = "") +
  coord_cartesian(ylim = c(0, 1.2)) +
  labs(
    title = "F-Distribution PDFs",
    x     = "F",
    y     = "Density"
  ) +
  theme_clinical()
```

```{r}
# ANOVA F critical value: 3 groups, 90 total observations
f_crit_anova <- qf(0.95, df1 = 2, df2 = 87)
cat(sprintf("F-critical for ANOVA (df1=2, df2=87, alpha=0.05): %.4f\n",
            f_crit_anova))
```

---

## 4. Clinical Applications

### 4.1 Sample Size Calculation via Power Analysis

The `power.t.test()` function in base R solves for any one of: sample size, delta (effect size), SD, power, or significance level.

**Scenario**: We design a parallel-group trial to detect a mean SBP reduction of 8 mmHg (SD = 14 mmHg) relative to placebo, at two-sided $\alpha = 0.05$ and 80 % power.

```{r}
# Two-sided two-sample t-test power analysis
pwr_result <- power.t.test(
  delta = 8,     # clinically meaningful difference in SBP (mmHg)
  sd    = 14,    # population standard deviation
  sig.level = 0.05,
  power     = 0.80,
  type      = "two.sample",
  alternative = "two.sided"
)

print(pwr_result)
cat(sprintf("\nRequired n per arm: %d (rounded up)\n",
            ceiling(pwr_result$n)))
```

```{r}
# Power curve: how does power change with sample size?
n_vals  <- seq(20, 150, by = 5)
pwr_vals <- sapply(n_vals, function(n) {
  power.t.test(n = n, delta = 8, sd = 14,
               sig.level = 0.05, type = "two.sample")$power
})

pwr_curve <- data.frame(n = n_vals, power = pwr_vals)

ggplot(pwr_curve, aes(x = n, y = power)) +
  geom_line(colour = "#1D3557", linewidth = 1.2) +
  geom_hline(yintercept = 0.80, linetype = "dashed",
             colour = "#E63946", linewidth = 0.9) +
  geom_vline(xintercept = ceiling(pwr_result$n),
             linetype = "dotted", colour = "#2A9D8F", linewidth = 0.9) +
  annotate("text", x = ceiling(pwr_result$n) + 7, y = 0.55,
           label = paste0("n = ", ceiling(pwr_result$n)),
           colour = "#2A9D8F", size = 3.5) +
  scale_y_continuous(labels = scales::percent_format()) +
  labs(
    title    = "Power Curve — Two-Sample t-Test",
    subtitle = "delta = 8 mmHg, SD = 14 mmHg, alpha = 0.05 (two-sided)",
    x        = "Sample size per arm",
    y        = "Power"
  ) +
  theme_clinical()
```

### 4.2 Adverse Event Rate Modelling with Poisson

Using the pharmaversesdtm AE dataset, we fit a simple Poisson model by arm.

```{r}
library(pharmaversesdtm)

# Load datasets
data("dm")
data("ae")

# Exclude screen failures; keep randomised subjects only
dm_rand <- dm |>
  filter(!is.na(ARM), ARM != "") |>
  select(USUBJID, ARM, AGE, SEX)

# Count AEs per subject
ae_counts <- ae |>
  group_by(USUBJID) |>
  summarise(n_ae = n(), .groups = "drop")

# Merge with dm: subjects with 0 AEs need a zero count
ae_by_subj <- dm_rand |>
  left_join(ae_counts, by = "USUBJID") |>
  mutate(n_ae = replace_na(n_ae, 0))

head(ae_by_subj)
```

```{r}
# Poisson summary by treatment arm
ae_summary <- ae_by_subj |>
  group_by(ARM) |>
  summarise(
    N          = n(),
    Total_AEs  = sum(n_ae),
    Mean_AEs   = round(mean(n_ae), 2),
    Variance   = round(var(n_ae), 2),
    Rate_per_subj = round(mean(n_ae), 3),
    .groups = "drop"
  )

print(ae_summary)
cat("\nNote: Under Poisson, Mean ≈ Variance (overdispersion if Var >> Mean)\n")
```

```{r fig.cap="Distribution of AE counts per subject by treatment arm"}
ggplot(ae_by_subj, aes(x = n_ae, fill = ARM)) +
  geom_histogram(binwidth = 1, colour = "white", alpha = 0.8) +
  facet_wrap(~ARM, ncol = 1,
             labeller = labeller(ARM = label_wrap_gen(25))) +
  scale_fill_brewer(palette = "Set2", name = "Arm") +
  labs(
    title    = "Adverse Event Count Distribution by Treatment Arm",
    subtitle = "pharmaversesdtm AE dataset",
    x        = "Number of AEs per subject",
    y        = "Count of subjects"
  ) +
  theme_clinical() +
  theme(legend.position = "none")
```

---

## 5. Simulation

### 5.1 Simulating a Complete Phase III Trial

```{r}
set.seed(42)

# Trial parameters
n_arm     <- 100   # subjects per arm
arms      <- c("Placebo", "Xanomeline Low Dose", "Xanomeline High Dose")

# Primary endpoint: change from baseline in a clinical score
# Placebo: mean = -1, SD = 3
# Low dose: mean = -3, SD = 3
# High dose: mean = -5, SD = 3

effects <- c(-1, -3, -5)
sd_val  <- 3

trial_data <- lapply(seq_along(arms), function(i) {
  data.frame(
    ARM    = arms[i],
    CFB    = rnorm(n_arm, mean = effects[i], sd = sd_val),
    AE_cnt = rpois(n_arm, lambda = c(1.0, 1.5, 2.2)[i])
  )
}) |> bind_rows()

# Summarise
trial_data |>
  group_by(ARM) |>
  summarise(
    N        = n(),
    Mean_CFB = round(mean(CFB), 2),
    SD_CFB   = round(sd(CFB), 2),
    Mean_AEs = round(mean(AE_cnt), 2),
    .groups  = "drop"
  )
```

```{r fig.cap="Simulated change-from-baseline distributions across three trial arms"}
ggplot(trial_data, aes(x = CFB, fill = ARM)) +
  geom_density(alpha = 0.55, colour = NA) +
  geom_vline(data = trial_data |>
               group_by(ARM) |>
               summarise(m = mean(CFB), .groups = "drop"),
             aes(xintercept = m, colour = ARM),
             linetype = "dashed", linewidth = 1) +
  scale_fill_brewer(palette = "Set1", name = "Arm") +
  scale_colour_brewer(palette = "Set1", name = "Arm (mean)") +
  labs(
    title    = "Simulated Change From Baseline — Phase III Trial",
    subtitle = "Kernel density estimates; dashed lines = arm means",
    x        = "Change from baseline",
    y        = "Density"
  ) +
  theme_clinical()
```

### 5.2 Monte Carlo Estimation of a p-Value

```{r}
set.seed(42)

# Observed difference in means between High Dose and Placebo
obs_diff <- mean(trial_data$CFB[trial_data$ARM == "Xanomeline High Dose"]) -
            mean(trial_data$CFB[trial_data$ARM == "Placebo"])

# Permutation test: shuffle arm labels 5000 times
placebo_cfb   <- trial_data$CFB[trial_data$ARM == "Placebo"]
high_dose_cfb <- trial_data$CFB[trial_data$ARM == "Xanomeline High Dose"]
combined      <- c(placebo_cfb, high_dose_cfb)
n_perm        <- 5000

perm_diffs <- replicate(n_perm, {
  shuf <- sample(combined)
  mean(shuf[1:n_arm]) - mean(shuf[(n_arm+1):(2*n_arm)])
})

mc_pvalue <- mean(abs(perm_diffs) >= abs(obs_diff))
cat(sprintf("Observed difference: %.3f\n", obs_diff))
cat(sprintf("Monte Carlo p-value : %.4f\n", mc_pvalue))
```

---

## 6. Visualisation and Diagnostics

### 6.1 QQ-Plots for Normality Assessment

A **quantile-quantile (QQ) plot** compares empirical quantiles of a sample against theoretical quantiles of a reference distribution. Points lying close to the diagonal line indicate the data follows the reference distribution.

```{r fig.cap="QQ-plots for age data in pharmaversesdtm (by arm)"}
# Use pharmaversesdtm dm for age data
dm_rand_age <- dm |>
  filter(!is.na(ARM), ARM != "", !is.na(AGE))

ggplot(dm_rand_age, aes(sample = AGE)) +
  stat_qq(colour = "#1D3557", alpha = 0.6, size = 1.5) +
  stat_qq_line(colour = "#E63946", linewidth = 1) +
  facet_wrap(~ARM, labeller = labeller(ARM = label_wrap_gen(25))) +
  labs(
    title    = "Normal QQ-Plot of Subject Age by Treatment Arm",
    subtitle = "pharmaversesdtm dm dataset — randomised subjects",
    x        = "Theoretical Normal quantiles",
    y        = "Sample quantiles (age, years)"
  ) +
  theme_clinical()
```

### 6.2 Overlaying Theoretical and Empirical Distributions

```{r fig.cap="Empirical age distribution versus fitted Normal density"}
age_mu <- mean(dm_rand_age$AGE, na.rm = TRUE)
age_sd <- sd(dm_rand_age$AGE, na.rm = TRUE)

cat(sprintf("Age: mean = %.1f, SD = %.1f\n", age_mu, age_sd))

ggplot(dm_rand_age, aes(x = AGE)) +
  geom_histogram(aes(y = after_stat(density)),
                 bins = 20, fill = "#457B9D", colour = "white", alpha = 0.7) +
  stat_function(fun = dnorm,
                args = list(mean = age_mu, sd = age_sd),
                colour = "#E63946", linewidth = 1.2) +
  labs(
    title    = "Age Distribution with Fitted Normal Overlay",
    subtitle = sprintf("Normal(μ=%.1f, σ=%.1f)", age_mu, age_sd),
    x        = "Age (years)",
    y        = "Density"
  ) +
  theme_clinical()
```

### 6.3 CDF Comparison Across Arms

```{r fig.cap="Empirical CDF of age by treatment arm (ECDF plot)"}
ggplot(dm_rand_age, aes(x = AGE, colour = ARM)) +
  stat_ecdf(linewidth = 1.1) +
  scale_colour_brewer(palette = "Set1",
                      name = "",
                      labels = function(x) stringr::str_wrap(x, 20)) +
  labs(
    title    = "Empirical CDF of Age by Treatment Arm",
    subtitle = "Steep curves indicate concentrated age distributions",
    x        = "Age (years)",
    y        = "F(x) = P(AGE ≤ x)"
  ) +
  theme_clinical()
```

---

## 7. Summary of Key Distribution Properties

```{r}
# Summary table of all distributions covered
dist_summary <- data.frame(
  Distribution = c("Binomial(n,p)", "Poisson(λ)",
                   "Normal(μ,σ²)", "t(ν)",
                   "Chi-square(ν)", "F(d1,d2)"),
  Type = c("Discrete","Discrete",
           "Continuous","Continuous",
           "Continuous","Continuous"),
  Mean = c("np","λ","μ","0 (ν>1)",
           "ν","d2/(d2-2) for d2>2"),
  Variance = c("np(1-p)","λ","σ²",
               "ν/(ν-2) for ν>2",
               "2ν","complex formula"),
  Clinical_Use = c("Response/event rates",
                   "AE counts",
                   "Continuous outcomes",
                   "Small-sample inference",
                   "GOF, variance tests",
                   "ANOVA, regression")
)

knitr::kable(dist_summary,
             caption = "Summary of probability distributions in clinical research",
             col.names = c("Distribution","Type","Mean","Variance","Clinical Use"))
```

---

## Exercises

### Beginner

1. Using `dbinom()`, calculate the probability that fewer than 3 patients out of 15 experience a treatment-related adverse event if the true event rate is 25%.

2. A Poisson process models laboratory abnormalities at a rate of $\lambda = 0.5$ per patient. Use `ppois()` to find the probability that a patient has 2 or more laboratory abnormalities.

3. Given a population with a mean diastolic blood pressure of 82 mmHg and SD of 10 mmHg, use `pnorm()` to find the proportion of patients with DBP above 95 mmHg.

4. Use `qnorm()` to find the SBP value below which 90% of a Normal(125, 15²) population falls.

### Intermediate

5. Use `power.t.test()` to find the required sample size per arm to detect a 5 mmHg difference in SBP (SD = 12 mmHg) with 90% power at a two-sided significance level of 0.05.

6. From the pharmaversesdtm `dm` dataset, filter to randomised subjects and use `ggplot2` to create overlapping density plots of AGE stratified by SEX. Add a `stat_function()` overlay of the best-fit Normal distribution.

7. Simulate 1000 Binomial(n=30, p=0.40) random variables. Plot the sampling distribution of the sample proportion and overlay the Normal approximation. Comment on the quality of the approximation.

8. Load the `ae` dataset from pharmaversesdtm. Merge with `dm` and compute the mean AE count per arm. Fit a Poisson GLM (`glm(..., family = poisson)`) with ARM as the predictor and interpret the rate ratios.

### Advanced

9. **Bootstrap the Binomial parameter**: Simulate 50 binary outcomes (Bernoulli with p = 0.35). Use 2000 bootstrap resamples to construct a 95% percentile bootstrap confidence interval for the response rate. Compare with the exact Clopper-Pearson interval from `binom.test()`.

10. **Power comparison simulation**: For a two-sample t-test comparing means with true delta = 6 and SD = 15, run 5000 simulated trials at n = 40, 60, 80 per arm (using `set.seed(123)`). Estimate the empirical power at each n and compare with `power.t.test()` theoretical values. Display results in a table.

11. **Distribution selection for safety data**: Using the full pharmaversesdtm `ae` dataset merged with `dm`, plot a histogram of AE counts per subject in the Placebo arm. Fit both a Poisson and a Negative Binomial distribution (using the `MASS` package `fitdistr()` function). Compare the log-likelihoods and AIC values, and discuss which model is more appropriate and why.
