# Statistical Inference in Clinical Research

## Introduction

Statistical inference is the process of drawing conclusions about a population from a sample of data. In clinical research, we rarely have access to every patient who could ever receive a treatment — instead, we study a carefully selected sample and use the results to make inferences about the broader patient population.

This chapter covers:

- **Point estimation**: Using sample statistics to estimate population parameters
- **Interval estimation**: Constructing confidence intervals to quantify uncertainty
- **Hypothesis testing**: Formal framework for making decisions under uncertainty
- **p-Values and effect sizes**: Interpreting and communicating statistical results
- **Multiple testing**: Controlling error rates when testing many hypotheses
- **Clinical versus statistical significance**: Connecting numbers to patient outcomes

---

## 1. The Framework of Statistical Inference

### 1.1 Population, Sample, and Parameters

In clinical research, the *population* is the collection of all patients we wish to make statements about (e.g., all adults with moderate-to-severe hypertension). A *sample* is the finite group we actually observe. *Parameters* (Greek letters: $\mu$, $\sigma$, $\pi$) describe population characteristics; *statistics* (Roman letters: $\bar{x}$, $s$, $\hat{p}$) describe sample characteristics.

The fundamental challenge: parameters are unknown and often unknowable. Inference uses sample statistics — and knowledge about their sampling distributions — to make probabilistic statements about parameters.

```{r}
#| label: inference-setup
#| message: false
#| warning: false

library(tidyverse)
library(broom)
library(gt)
set.seed(42)

# Simulate a clinical population
# True population: SBP ~ N(140, 20) for untreated hypertensive patients
pop_size  <- 100000
pop_sbp   <- rnorm(pop_size, mean = 140, sd = 20)
pop_mean  <- mean(pop_sbp)
pop_sd    <- sd(pop_sbp)

cat("True population mean SBP:", round(pop_mean, 2), "mmHg\n")
cat("True population SD:      ", round(pop_sd,  2), "mmHg\n")
```

```{r}
#| label: sampling-demo
#| fig-cap: "Sampling variability: 200 sample means (n=50 each) drawn from the hypertensive population"

# Draw many samples and examine the sampling distribution
n_samples  <- 200
sample_n   <- 50
sample_means <- replicate(n_samples, mean(sample(pop_sbp, sample_n)))

tibble(sample_mean = sample_means) |>
  ggplot(aes(x = sample_mean)) +
  geom_histogram(bins = 30, fill = "#2C6E9E", colour = "white", alpha = 0.85) +
  geom_vline(xintercept = pop_mean, colour = "red", linewidth = 1, linetype = "dashed") +
  annotate("text", x = pop_mean + 0.4, y = 22,
           label = paste0("Population mean\n", round(pop_mean, 1), " mmHg"),
           colour = "red", hjust = 0, size = 3.5) +
  labs(
    x = "Sample Mean SBP (mmHg)",
    y = "Count",
    title = "Sampling Distribution of the Mean",
    subtitle = paste0(n_samples, " samples of n=", sample_n, " from hypertensive population")
  ) +
  theme_bw(base_size = 12)
```

### 1.2 The Standard Error

The *standard error* (SE) measures the spread of a sampling distribution — it quantifies how much a statistic varies from sample to sample:

$$SE_{\bar{x}} = \frac{\sigma}{\sqrt{n}}$$

When $\sigma$ is unknown (the usual case), we substitute the sample standard deviation $s$:

$$\widehat{SE}_{\bar{x}} = \frac{s}{\sqrt{n}}$$

```{r}
#| label: se-demo

# One sample from our clinical population
clinical_sample <- sample(pop_sbp, 50)
x_bar <- mean(clinical_sample)
s     <- sd(clinical_sample)
se    <- s / sqrt(length(clinical_sample))

cat("Sample mean:           ", round(x_bar, 2), "mmHg\n")
cat("Sample SD:             ", round(s, 2),     "mmHg\n")
cat("Standard Error (SE):   ", round(se, 2),    "mmHg\n")
cat("Theoretical SE:        ", round(pop_sd / sqrt(50), 2), "mmHg\n")
```

---

## 2. Point Estimation

### 2.1 Desirable Properties of Estimators

A good estimator is:

- **Unbiased**: $E[\hat{\theta}] = \theta$ (on average, it hits the target)
- **Consistent**: converges to $\theta$ as $n \to \infty$
- **Efficient**: smallest variance among all unbiased estimators

The sample mean $\bar{x}$ is unbiased, consistent, and efficient for the population mean $\mu$ when data are normally distributed.

```{r}
#| label: estimator-bias

# Demonstrate unbiasedness: average of many sample means ≈ population mean
bias <- mean(sample_means) - pop_mean
cat("Mean of 200 sample means:", round(mean(sample_means), 3), "mmHg\n")
cat("True population mean:    ", round(pop_mean,           3), "mmHg\n")
cat("Empirical bias:          ", round(bias,               4), "mmHg (≈ 0)\n")
```

### 2.2 Maximum Likelihood Estimation

Maximum likelihood estimation (MLE) finds parameter values that maximise the likelihood of observing the collected data. For normally distributed data, MLE gives the same result as the method of moments: $\hat{\mu}_{MLE} = \bar{x}$.

```{r}
#| label: mle-illustration
#| fig-cap: "Log-likelihood surface for the mean of a normal distribution (σ known)"

# Likelihood of observing our sample under different mean hypotheses
mu_grid  <- seq(120, 160, by = 0.5)
log_lik  <- sapply(mu_grid, function(mu) {
  sum(dnorm(clinical_sample, mean = mu, sd = pop_sd, log = TRUE))
})

tibble(mu = mu_grid, log_likelihood = log_lik) |>
  ggplot(aes(x = mu, y = log_likelihood)) +
  geom_line(colour = "#2C6E9E", linewidth = 1) +
  geom_vline(xintercept = x_bar, colour = "red",
             linetype = "dashed", linewidth = 0.9) +
  annotate("text", x = x_bar + 0.5, y = min(log_lik) + 5,
           label = paste0("MLE = ", round(x_bar, 1)),
           colour = "red", hjust = 0, size = 3.5) +
  labs(
    x = expression(mu ~ "(Hypothesised Population Mean)"),
    y = "Log-Likelihood",
    title = "Log-Likelihood as a Function of the Mean",
    subtitle = "Peak identifies the maximum likelihood estimate"
  ) +
  theme_bw(base_size = 12)
```

---

## 3. Interval Estimation: Confidence Intervals

### 3.1 The Confidence Interval Concept

A $(1-\alpha) \times 100\%$ confidence interval (CI) is constructed so that, **in repeated sampling**, $(1-\alpha) \times 100\%$ of such intervals contain the true parameter. For $\alpha = 0.05$, we get 95% CIs.

**Critical misconception**: A 95% CI does *not* mean "there is a 95% probability that the true mean lies in this interval." The true mean is fixed (though unknown); it either is or is not in the interval. The 95% refers to the *procedure*, not the interval.

For a single mean with unknown $\sigma$:
$$\bar{x} \pm t_{\alpha/2, \, n-1} \cdot \frac{s}{\sqrt{n}}$$

```{r}
#| label: ci-coverage
#| fig-cap: "95% confidence intervals from 50 repeated samples — approximately 95% contain the true mean (red line)"

set.seed(123)
n_ci    <- 50
ci_data <- tibble(
  sample_id = 1:n_ci,
  mean      = replicate(n_ci, {
    s_tmp <- sample(pop_sbp, 50)
    mean(s_tmp)
  }),
  sd_samp   = replicate(n_ci, {
    s_tmp <- sample(pop_sbp, 50)
    sd(s_tmp)
  })
) |>
  mutate(
    se    = sd_samp / sqrt(50),
    lower = mean - qt(0.975, df = 49) * se,
    upper = mean + qt(0.975, df = 49) * se,
    contains_true = lower <= pop_mean & pop_mean <= upper
  )

coverage <- mean(ci_data$contains_true) * 100
cat("Empirical coverage:", coverage, "%\n")

ci_data |>
  ggplot(aes(x = sample_id, y = mean,
             colour = contains_true, ymin = lower, ymax = upper)) +
  geom_hline(yintercept = pop_mean, colour = "red",
             linewidth = 1, linetype = "dashed") +
  geom_errorbar(width = 0.4, linewidth = 0.6) +
  geom_point(size = 2) +
  scale_colour_manual(
    values = c("FALSE" = "#D32F2F", "TRUE" = "#2C6E9E"),
    labels = c("FALSE" = "Does not contain µ", "TRUE" = "Contains µ")
  ) +
  labs(
    x = "Sample Number",
    y = "Sample Mean SBP (mmHg)",
    colour = NULL,
    title = "95% Confidence Intervals Across 50 Samples",
    subtitle = paste0("Coverage: ", coverage, "% (expected ≈ 95%)")
  ) +
  theme_bw(base_size = 12) +
  theme(legend.position = "bottom")
```

### 3.2 CIs for Common Clinical Estimands

```{r}
#| label: ci-examples

# Generate a small clinical dataset
set.seed(7)
trial_df <- tibble(
  subject_id = paste0("SUBJ-", sprintf("%03d", 1:80)),
  arm        = rep(c("Active", "Placebo"), each = 40),
  sbp_base   = rnorm(80, mean = 145, sd = 15),
  sbp_chg    = c(rnorm(40, mean = -12, sd = 8),   # Active: mean reduction
                 rnorm(40, mean =  -3, sd = 8))    # Placebo: small reduction
)

# CI for a single mean
active_chg <- trial_df |> filter(arm == "Active") |> pull(sbp_chg)
ci_single  <- t.test(active_chg)$conf.int

# CI for difference between two means
tt_result  <- t.test(sbp_chg ~ arm, data = trial_df, var.equal = FALSE)
ci_diff    <- tt_result$conf.int

# CI for a proportion (adverse events)
n_ae       <- 12
n_total    <- 80
ci_prop    <- prop.test(n_ae, n_total)$conf.int

cat("=== Confidence Intervals ===\n\n")
cat("Mean SBP change (Active):\n")
cat("  Estimate:", round(mean(active_chg), 2), "mmHg\n")
cat("  95% CI: [", round(ci_single[1], 2), ",", round(ci_single[2], 2), "]\n\n")

cat("Difference in SBP change (Active – Placebo):\n")
cat("  Estimate:", round(tt_result$estimate[1] - tt_result$estimate[2], 2), "mmHg\n")
cat("  95% CI: [", round(ci_diff[1], 2), ",", round(ci_diff[2], 2), "]\n\n")

cat("AE proportion (12/80):\n")
cat("  Estimate:", round(n_ae / n_total * 100, 1), "%\n")
cat("  95% CI: [", round(ci_prop[1] * 100, 1), "%, ", round(ci_prop[2] * 100, 1), "%]\n")
```

---

## 4. Hypothesis Testing

### 4.1 The Neyman–Pearson Framework

Hypothesis testing provides a formal decision procedure:

1. **State hypotheses**: $H_0$ (null) and $H_1$ (alternative)
2. **Choose a test statistic** whose distribution under $H_0$ is known
3. **Set significance level** $\alpha$ (usually 0.05 for Phase III)
4. **Compute the test statistic** from the data
5. **Reject $H_0$** if the test statistic exceeds the critical value

**Error types**:

| Decision | $H_0$ true | $H_0$ false |
|---|---|---|
| Reject $H_0$ | Type I error ($\alpha$) | Correct (Power = $1-\beta$) |
| Fail to reject $H_0$ | Correct ($1-\alpha$) | Type II error ($\beta$) |

```{r}
#| label: error-types
#| fig-cap: "Type I and Type II errors visualised for a one-sample t-test"

mu0    <- 0       # H0: no change
mu1    <- -5      # H1: 5 mmHg reduction
sigma  <- 8
n_viz  <- 40
se_viz <- sigma / sqrt(n_viz)

x_range <- seq(-15, 10, length.out = 500)
df_null <- tibble(x = x_range, y = dnorm(x, mu0, se_viz), dist = "H0: µ = 0")
df_alt  <- tibble(x = x_range, y = dnorm(x, mu1, se_viz), dist = "H1: µ = -5")

crit_val <- qnorm(0.05, mu0, se_viz)  # one-tailed

bind_rows(df_null, df_alt) |>
  ggplot(aes(x = x, y = y, colour = dist)) +
  geom_line(linewidth = 1.1) +
  # Type I error (alpha)
  geom_area(data = filter(df_null, x <= crit_val),
            aes(x = x, y = y), fill = "#D32F2F", alpha = 0.35, inherit.aes = FALSE) +
  # Type II error (beta)
  geom_area(data = filter(df_alt, x >= crit_val),
            aes(x = x, y = y), fill = "#F57C00", alpha = 0.35, inherit.aes = FALSE) +
  geom_vline(xintercept = crit_val, linetype = "dashed", colour = "black") +
  annotate("text", x = crit_val - 0.3, y = 0.16,
           label = paste0("Critical value\n", round(crit_val, 2)),
           hjust = 1, size = 3) +
  annotate("text", x = -13, y = 0.05, label = "Power\n(1-β)",
           colour = "#1565C0", size = 3.5, fontface = "bold") +
  annotate("text", x = -2.5, y = 0.03, label = "Type I\n(α)",
           colour = "#D32F2F", size = 3.5, fontface = "bold") +
  scale_colour_manual(values = c("H0: µ = 0" = "#757575", "H1: µ = -5" = "#1565C0")) +
  labs(
    x = "Mean SBP Change (mmHg)",
    y = "Density",
    colour = "Distribution",
    title = "Type I Error, Type II Error, and Power",
    subtitle = "One-tailed test, α = 0.05, n = 40"
  ) +
  theme_bw(base_size = 12) +
  theme(legend.position = "bottom")
```

### 4.2 p-Values: Definition and Interpretation

The **p-value** is the probability of observing a test statistic at least as extreme as the one calculated, *assuming $H_0$ is true*.

$$p = P(|T| \geq |t_{obs}| \mid H_0)$$

**Common misinterpretations to avoid**:

- ✗ "p = 0.03 means there is a 3% chance the null is true"
- ✗ "p = 0.03 means there is a 97% chance the alternative is true"
- ✓ "If the null were true, we would observe data this extreme 3% of the time"

```{r}
#| label: pvalue-demo

# Two-sample t-test comparing SBP change between arms
tt_out <- t.test(sbp_chg ~ arm, data = trial_df, var.equal = FALSE)
tidy(tt_out) |>
  select(estimate, estimate1, estimate2, statistic, p.value, conf.low, conf.high) |>
  mutate(across(where(is.numeric), \(x) round(x, 4))) |>
  gt() |>
  tab_header(title = "Two-Sample t-Test: SBP Change Active vs. Placebo") |>
  cols_label(
    estimate  = "Difference",
    estimate1 = "Mean (Active)",
    estimate2 = "Mean (Placebo)",
    statistic = "t-statistic",
    p.value   = "p-value",
    conf.low  = "95% CI Lower",
    conf.high = "95% CI Upper"
  )
```

### 4.3 One-Tailed vs. Two-Tailed Tests

Two-tailed tests are standard in confirmatory clinical trials because they protect against detecting effects in either direction. One-tailed tests may be pre-specified when there is a directional prior (e.g., superiority trials), but require strong scientific justification and regulatory agreement.

```{r}
#| label: tails-comparison

# Compare one- and two-tailed p-values
tt_two <- t.test(sbp_chg ~ arm, data = trial_df,
                 alternative = "two.sided", var.equal = FALSE)
tt_one <- t.test(sbp_chg ~ arm, data = trial_df,
                 alternative = "less", var.equal = FALSE)  # Active < Placebo

cat("Two-tailed p-value:", round(tt_two$p.value, 4), "\n")
cat("One-tailed p-value:", round(tt_one$p.value,  4), "\n")
cat("Note: one-tailed p ≈ two-tailed p / 2 when testing in the correct direction\n")
```

---

## 5. Effect Sizes and Clinical Significance

Statistical significance ($p < 0.05$) and clinical significance are different concepts. A large study can detect trivially small effects as statistically significant; a small study may miss clinically important effects.

### 5.1 Cohen's d

For comparing two means, Cohen's d standardises the difference by the pooled standard deviation:

$$d = \frac{\bar{x}_1 - \bar{x}_2}{s_{pooled}}$$

Benchmarks (Cohen, 1988): small = 0.2, medium = 0.5, large = 0.8.

```{r}
#| label: effect-sizes

# Compute effect size
active_mean  <- mean(trial_df$sbp_chg[trial_df$arm == "Active"])
placebo_mean <- mean(trial_df$sbp_chg[trial_df$arm == "Placebo"])
pooled_sd    <- sqrt(
  ((40 - 1) * var(trial_df$sbp_chg[trial_df$arm == "Active"]) +
   (40 - 1) * var(trial_df$sbp_chg[trial_df$arm == "Placebo"])) / (80 - 2)
)
cohens_d <- (active_mean - placebo_mean) / pooled_sd

cat("Mean change (Active): ",  round(active_mean,  2), "mmHg\n")
cat("Mean change (Placebo):",  round(placebo_mean, 2), "mmHg\n")
cat("Difference:           ", round(active_mean - placebo_mean, 2), "mmHg\n")
cat("Pooled SD:            ",  round(pooled_sd,   2), "mmHg\n")
cat("Cohen's d:            ",  round(cohens_d,    3), "\n")
cat("Interpretation:        ",
    ifelse(abs(cohens_d) < 0.2, "negligible",
    ifelse(abs(cohens_d) < 0.5, "small",
    ifelse(abs(cohens_d) < 0.8, "medium", "large"))), "\n")
```

### 5.2 Reporting Both Statistical and Clinical Significance

In clinical trial reports and publications, always present:

1. **Point estimate** with units (e.g., mean difference in mmHg)
2. **Confidence interval** to convey precision
3. **p-value** for hypothesis test result
4. **Effect size** to convey practical magnitude
5. **Clinical threshold** context (e.g., MCID for patient-reported outcomes)

```{r}
#| label: full-reporting

# Summary table: complete inferential report
tibble(
  Metric    = c("Mean SBP change (Active)", "Mean SBP change (Placebo)",
                "Difference (Active – Placebo)", "95% CI for Difference",
                "p-value (two-tailed)", "Cohen's d",
                "Clinical threshold (MCID ≈ 5 mmHg)"),
  Value     = c(
    paste0(round(active_mean, 1), " mmHg"),
    paste0(round(placebo_mean, 1), " mmHg"),
    paste0(round(active_mean - placebo_mean, 1), " mmHg"),
    paste0("[", round(tt_out$conf.int[1], 1), ", ", round(tt_out$conf.int[2], 1), "] mmHg"),
    formatC(tt_out$p.value, format = "f", digits = 4),
    round(cohens_d, 2),
    "Difference exceeds MCID: YES"
  )
) |>
  gt() |>
  tab_header(title = "Complete Inferential Report: SBP Reduction Trial")
```

---

## 6. Multiple Testing and Error Control

### 6.1 The Multiple Comparisons Problem

When testing $m$ hypotheses simultaneously, each at level $\alpha$, the probability of at least one false positive is:

$$P(\text{at least one Type I error}) = 1 - (1 - \alpha)^m$$

For $m = 20$ tests at $\alpha = 0.05$: $P = 1 - 0.95^{20} \approx 0.64$.

```{r}
#| label: fwer-demo
#| fig-cap: "Family-wise error rate rises rapidly with number of tests"

m_vals <- 1:50
fwer   <- 1 - (1 - 0.05)^m_vals

tibble(m = m_vals, FWER = fwer) |>
  ggplot(aes(x = m, y = FWER)) +
  geom_line(colour = "#D32F2F", linewidth = 1.1) +
  geom_hline(yintercept = 0.05, linetype = "dashed", colour = "#2C6E9E") +
  geom_hline(yintercept = 0.20, linetype = "dotted", colour = "#F57C00") +
  annotate("text", x = 48, y = 0.07, label = "α = 0.05", colour = "#2C6E9E", size = 3) +
  annotate("text", x = 48, y = 0.22, label = "20%",      colour = "#F57C00", size = 3) +
  labs(
    x = "Number of Tests (m)",
    y = "Family-Wise Error Rate (FWER)",
    title = "Multiple Testing Inflates the Type I Error Rate",
    subtitle = "Each individual test at α = 0.05"
  ) +
  theme_bw(base_size = 12)
```

### 6.2 Correction Methods

**Bonferroni correction**: Divide $\alpha$ by $m$. Simple, conservative.
$$\alpha^* = \frac{\alpha}{m}$$

**Holm's step-down**: Less conservative than Bonferroni, controls FWER.

**Benjamini–Hochberg (FDR)**: Controls the expected proportion of false discoveries among rejected hypotheses. Preferred for exploratory analyses.

```{r}
#| label: multiple-testing-corrections

set.seed(99)
# Simulate 20 p-values (5 truly non-null, 15 null)
n_tests    <- 20
true_null  <- c(rep(TRUE, 15), rep(FALSE, 5))
raw_pvals  <- c(
  runif(15, 0.01, 0.99),         # null hypotheses
  c(0.001, 0.004, 0.012, 0.031, 0.048)  # non-null (small p-values)
)

corrections <- tibble(
  test         = paste0("Test-", sprintf("%02d", 1:n_tests)),
  p_raw        = raw_pvals,
  p_bonferroni = p.adjust(raw_pvals, method = "bonferroni"),
  p_holm       = p.adjust(raw_pvals, method = "holm"),
  p_BH         = p.adjust(raw_pvals, method = "BH"),
  true_null    = true_null
) |>
  arrange(p_raw)

# Summary of rejections
tibble(
  Method      = c("No correction", "Bonferroni", "Holm", "Benjamini-Hochberg"),
  Threshold   = c(0.05, 0.05 / n_tests, 0.05, 0.05),
  Rejections  = c(
    sum(raw_pvals < 0.05),
    sum(p.adjust(raw_pvals, "bonferroni") < 0.05),
    sum(p.adjust(raw_pvals, "holm")       < 0.05),
    sum(p.adjust(raw_pvals, "BH")         < 0.05)
  ),
  True_Positives = c(
    sum(!true_null & raw_pvals < 0.05),
    sum(!true_null & p.adjust(raw_pvals, "bonferroni") < 0.05),
    sum(!true_null & p.adjust(raw_pvals, "holm")       < 0.05),
    sum(!true_null & p.adjust(raw_pvals, "BH")         < 0.05)
  )
) |>
  mutate(False_Positives = Rejections - True_Positives) |>
  gt() |>
  tab_header(title = "Comparison of Multiple Testing Correction Methods",
             subtitle = "20 tests: 15 true null, 5 true non-null")
```

---

## 7. Bootstrap Inference

The bootstrap is a resampling method for constructing CIs without distributional assumptions — valuable when normality cannot be assumed or sample size is small.

```{r}
#| label: bootstrap-ci
#| fig-cap: "Bootstrap distribution of the mean SBP change and corresponding 95% CI"

set.seed(42)
boot_data <- trial_df |> filter(arm == "Active") |> pull(sbp_chg)
n_boot    <- 5000
boot_means <- replicate(n_boot, mean(sample(boot_data, replace = TRUE)))

# Percentile CI
ci_boot <- quantile(boot_means, c(0.025, 0.975))

cat("Bootstrap 95% CI (percentile):",
    round(ci_boot[1], 2), "to", round(ci_boot[2], 2), "mmHg\n")
cat("Parametric 95% CI:            ",
    round(t.test(boot_data)$conf.int[1], 2), "to",
    round(t.test(boot_data)$conf.int[2], 2), "mmHg\n")

tibble(boot_mean = boot_means) |>
  ggplot(aes(x = boot_mean)) +
  geom_histogram(bins = 50, fill = "#2C6E9E", colour = "white", alpha = 0.85) +
  geom_vline(xintercept = ci_boot, colour = "red",
             linetype = "dashed", linewidth = 0.9) +
  annotate("text", x = ci_boot[1] - 0.3, y = 350,
           label = paste0(round(ci_boot[1], 1)), colour = "red", hjust = 1, size = 3.5) +
  annotate("text", x = ci_boot[2] + 0.3, y = 350,
           label = paste0(round(ci_boot[2], 1)), colour = "red", hjust = 0, size = 3.5) +
  labs(
    x = "Bootstrap Sample Mean (mmHg)",
    y = "Count",
    title = "Bootstrap Distribution of Mean SBP Change (Active Arm)",
    subtitle = paste0(n_boot, " bootstrap replicates; red dashed lines = 95% CI")
  ) +
  theme_bw(base_size = 12)
```

---

## 8. One-Sample and Two-Sample Inferences: Practical Workflow

```{r}
#| label: inference-workflow

# Complete workflow for a two-arm trial comparison

# Step 1: Check assumptions
cat("=== Normality Check (Shapiro-Wilk) ===\n")
for (arm_name in c("Active", "Placebo")) {
  vals <- trial_df |> filter(arm == arm_name) |> pull(sbp_chg)
  sw   <- shapiro.test(vals)
  cat(arm_name, ": W =", round(sw$statistic, 4),
      ", p =", round(sw$p.value, 4),
      ifelse(sw$p.value > 0.05, " (Normal assumption OK)\n", " (Consider non-parametric)\n"))
}

# Step 2: Variance equality (Levene's / F-test)
cat("\n=== Variance Equality (F-test) ===\n")
ftest <- var.test(sbp_chg ~ arm, data = trial_df)
cat("F =", round(ftest$statistic, 3),
    ", p =", round(ftest$p.value, 4),
    ifelse(ftest$p.value > 0.05,
           " (Equal variances OK — use pooled t-test)\n",
           " (Unequal variances — use Welch t-test)\n"))

# Step 3: Conduct test
cat("\n=== Inferential Test ===\n")
final_test <- t.test(sbp_chg ~ arm, data = trial_df,
                     var.equal = ftest$p.value > 0.05)
cat("t =", round(final_test$statistic, 3),
    ", df =", round(final_test$parameter, 1),
    ", p =", round(final_test$p.value, 4), "\n")
cat("95% CI: [", round(final_test$conf.int[1], 2), ",",
    round(final_test$conf.int[2], 2), "] mmHg\n")
```

---

## Summary

Statistical inference bridges the gap between observed data and generalisable conclusions. The key principles for clinical research are:

- **Standard errors** quantify sampling variability; they shrink as $n$ increases
- **Confidence intervals** convey both the direction and precision of effects; they are preferred over p-values alone for regulatory submissions (ICH E9)
- **Hypothesis testing** controls long-run error rates; the significance level $\alpha = 0.05$ is a convention, not a law
- **p-Values** measure the compatibility of data with the null hypothesis — they are *not* the probability that the null is true
- **Effect sizes** (Cohen's d, relative risk) are essential for interpreting practical importance
- **Multiple testing adjustments** are mandatory whenever more than one hypothesis is tested
- **Bootstrap methods** provide distribution-free inference, particularly useful for small or non-normal samples

> **Regulatory note**: ICH E9 (Statistical Principles for Clinical Trials) and its addendum ICH E9(R1) on estimands provide the gold standard framework for statistical inference in confirmatory clinical trials.
