# Parametric Statistical Tests in Clinical Trials

## Introduction

Parametric tests are the workhorses of clinical trial statistics. They assume that the data follow a known probability distribution — most commonly the normal distribution — and use the parameters of that distribution (mean, variance) to draw inferences. When these assumptions hold, parametric tests are the most powerful tools available for detecting true treatment effects.

This chapter covers the full spectrum of parametric tests used in clinical development:

- **t-tests**: One-sample, two-sample independent, and paired designs
- **ANOVA**: One-way and two-way analysis of variance across treatment arms
- **ANCOVA**: Covariate adjustment for baseline measurements
- **Assumption checking**: Normality tests, homogeneity of variance, and diagnostic plots
- **Post-hoc tests**: Multiple comparison procedures after ANOVA

All examples use the `pharmaversesdtm` dataset, which provides realistic CDISC-formatted clinical trial data.

---

## Setup and Data Preparation

```{r}
#| label: setup
#| message: false
#| warning: false

library(dplyr)
library(ggplot2)
library(car)
library(pharmaversesdtm)

# Load CDISC datasets
data("dm")   # Demographics
data("ae")   # Adverse events
data("vs")   # Vital signs
data("lb")   # Laboratory values

# Clean demographics: retain valid ARM assignments
dm_clean <- dm |>
  filter(!is.na(ARM), ARM != "") |>
  mutate(ARM = factor(ARM,
                      levels = c("Placebo",
                                 "Xanomeline Low Dose",
                                 "Xanomeline High Dose")))

# Preview
cat("Treatment arm counts:\n")
print(table(dm_clean$ARM))
cat("\nAGE summary:\n")
print(summary(dm_clean$AGE))
```

```{r}
#| label: vs-prep

# Prepare vital signs: systolic blood pressure
vs_sysbp <- vs |>
  filter(VSTESTCD == "SYSBP", !is.na(VSSTRESN)) |>
  left_join(dm_clean |> select(USUBJID, ARM, SEX), by = "USUBJID") |>
  filter(!is.na(ARM))

cat("Systolic BP records per visit:\n")
print(table(vs_sysbp$VISITNUM))
cat("\nTotal SYSBP records:", nrow(vs_sysbp), "\n")
```

---

## 1. One-Sample t-Test

### 1.1 Theory

The one-sample t-test asks: **does the population mean differ from a specified value (mu)?**

$$t = \frac{\bar{x} - \mu_0}{s / \sqrt{n}}$$

where $\bar{x}$ is the sample mean, $\mu_0$ is the hypothesised value, $s$ is the sample standard deviation, and $n$ is the sample size. The test statistic follows a $t$-distribution with $n - 1$ degrees of freedom under $H_0$.

**Hypotheses:**

- $H_0: \mu = \mu_0$ (population mean equals the hypothesised value)
- $H_1: \mu \neq \mu_0$ (two-tailed)

**Clinical question:** Is the mean age of participants in this trial different from 65 years — the typical cut-off used for elderly subgroup analyses?

### 1.2 Assumption Check: Normality

```{r}
#| label: onesample-normality
#| fig-cap: "Q-Q plot for participant age — assessing normality before the one-sample t-test"

age_vals <- dm_clean$AGE

# Shapiro-Wilk normality test (best for n < 2000)
sw_age <- shapiro.test(age_vals)
cat("Shapiro-Wilk test for AGE:\n")
cat("  W =", round(sw_age$statistic, 4),
    ", p-value =", round(sw_age$p.value, 4), "\n\n")

# Q-Q plot
ggplot(data.frame(AGE = age_vals), aes(sample = AGE)) +
  stat_qq(colour = "#2C6E9E", size = 2, alpha = 0.7) +
  stat_qq_line(colour = "#E74C3C", linewidth = 1) +
  labs(
    title    = "Normal Q-Q Plot of Participant Age",
    subtitle = paste0("Shapiro-Wilk: W = ", round(sw_age$statistic, 3),
                      ", p = ", round(sw_age$p.value, 3)),
    x = "Theoretical Quantiles",
    y = "Sample Quantiles (Age in Years)"
  ) +
  theme_classic(base_size = 13)
```

### 1.3 Performing the Test

```{r}
#| label: onesample-ttest

set.seed(42)

# One-sample t-test: H0: mu_AGE = 65
result_1samp <- t.test(dm_clean$AGE, mu = 65, alternative = "two.sided")
print(result_1samp)

cat("\n--- Clinical Interpretation ---\n")
cat("Mean age:", round(mean(dm_clean$AGE), 1), "years\n")
cat("95% CI: [", round(result_1samp$conf.int[1], 1), ",",
    round(result_1samp$conf.int[2], 1), "]\n")

if (result_1samp$p.value < 0.05) {
  cat("Conclusion: The mean age significantly DIFFERS from 65 years (p =",
      round(result_1samp$p.value, 4), ").\n")
  cat("The trial population is", ifelse(mean(dm_clean$AGE) > 65, "older", "younger"),
      "than the general elderly threshold.\n")
} else {
  cat("Conclusion: No significant difference from 65 years (p =",
      round(result_1samp$p.value, 4), ").\n")
  cat("The trial population age is consistent with the elderly threshold.\n")
}
```

### 1.4 Visualisation

```{r}
#| label: onesample-plot
#| fig-cap: "Distribution of participant age with hypothesis test reference line (mu = 65)"

ggplot(dm_clean, aes(x = AGE)) +
  geom_histogram(aes(y = after_stat(density)),
                 bins = 20, fill = "#2C6E9E", colour = "white", alpha = 0.8) +
  geom_density(colour = "#1A3E5C", linewidth = 1) +
  geom_vline(xintercept = 65, colour = "#E74C3C",
             linetype = "dashed", linewidth = 1.2) +
  geom_vline(xintercept = mean(dm_clean$AGE), colour = "#27AE60",
             linetype = "solid", linewidth = 1.2) +
  annotate("text", x = 65, y = Inf, label = " H\u2080: \u03bc = 65",
           hjust = 0, vjust = 2, colour = "#E74C3C", size = 4) +
  annotate("text", x = mean(dm_clean$AGE), y = Inf,
           label = paste0(" \u0078\u0305 = ", round(mean(dm_clean$AGE), 1)),
           hjust = 0, vjust = 3.5, colour = "#27AE60", size = 4) +
  labs(
    title    = "One-Sample t-Test: Participant Age vs. 65-Year Threshold",
    subtitle = paste0("t(", result_1samp$parameter, ") = ",
                      round(result_1samp$statistic, 2),
                      ", p = ", round(result_1samp$p.value, 4)),
    x = "Age (years)",
    y = "Density"
  ) +
  theme_classic(base_size = 13)
```

> **Clinical note:** In regulatory submissions, the age distribution of trial participants is routinely compared against predefined age cut-offs (e.g., 65, 75) to ensure the study population adequately represents the intended patient population.

---

## 2. Two-Sample Independent t-Test

### 2.1 Theory

The two-sample independent t-test compares the means of two unrelated groups:

$$t = \frac{\bar{x}_1 - \bar{x}_2}{\sqrt{s_p^2 \left(\frac{1}{n_1} + \frac{1}{n_2}\right)}}$$

where $s_p^2$ is the pooled variance (assuming equal variances). When variances are unequal, Welch's correction adjusts the degrees of freedom.

**Clinical question:** Does the mean age differ between the Placebo and Xanomeline High Dose arms? (Age balance at baseline validates randomisation.)

### 2.2 Filter to Two Arms

```{r}
#| label: twosample-filter

# Filter to exactly two arms for comparison
dm_2arm <- dm_clean |>
  filter(ARM %in% c("Placebo", "Xanomeline High Dose")) |>
  droplevels()

cat("Sample sizes:\n")
print(table(dm_2arm$ARM))
cat("\nGroup means:\n")
print(tapply(dm_2arm$AGE, dm_2arm$ARM, mean, na.rm = TRUE))
```

### 2.3 Assumption Checks

```{r}
#| label: twosample-assumptions
#| fig-cap: "Q-Q plots by treatment arm for two-sample t-test assumption checking"

# Shapiro-Wilk per group
sw_by_arm <- tapply(dm_2arm$AGE, dm_2arm$ARM, shapiro.test)
for (arm in names(sw_by_arm)) {
  cat(arm, ": W =", round(sw_by_arm[[arm]]$statistic, 4),
      ", p =", round(sw_by_arm[[arm]]$p.value, 4), "\n")
}

# Levene's test for equal variances (more robust than Bartlett)
lev_result <- car::leveneTest(AGE ~ ARM, data = dm_2arm)
cat("\nLevene's Test for Homogeneity of Variance:\n")
print(lev_result)

# Q-Q plots by arm
ggplot(dm_2arm, aes(sample = AGE, colour = ARM)) +
  stat_qq(size = 2, alpha = 0.7) +
  stat_qq_line(linewidth = 1) +
  facet_wrap(~ARM) +
  scale_colour_manual(values = c("Placebo" = "#2C6E9E",
                                  "Xanomeline High Dose" = "#E74C3C")) +
  labs(
    title    = "Normal Q-Q Plots by Treatment Arm",
    subtitle = "Assessing normality for the two-sample t-test",
    x = "Theoretical Quantiles",
    y = "Sample Quantiles (Age)"
  ) +
  theme_classic(base_size = 13) +
  theme(legend.position = "none")
```

### 2.4 Performing the Test

```{r}
#| label: twosample-ttest

# Welch's t-test (var.equal = FALSE is safer default)
result_2samp <- t.test(AGE ~ ARM, data = dm_2arm,
                       alternative = "two.sided",
                       var.equal   = FALSE)
print(result_2samp)

# Also show pooled variance version for comparison
result_2samp_eq <- t.test(AGE ~ ARM, data = dm_2arm, var.equal = TRUE)

cat("\n--- Summary Table ---\n")
cat("Test               | t-statistic | df    | p-value\n")
cat("Welch (unequal var)|", round(result_2samp$statistic, 3),
    "        |", round(result_2samp$parameter, 1),
    "|", round(result_2samp$p.value, 4), "\n")
cat("Student (equal var)|", round(result_2samp_eq$statistic, 3),
    "        |", round(result_2samp_eq$parameter, 1),
    "|", round(result_2samp_eq$p.value, 4), "\n")

cat("\n--- Clinical Interpretation ---\n")
means <- tapply(dm_2arm$AGE, dm_2arm$ARM, mean, na.rm = TRUE)
cat("Mean age - Placebo:", round(means["Placebo"], 1), "years\n")
cat("Mean age - Xanomeline High Dose:", round(means["Xanomeline High Dose"], 1), "years\n")
if (result_2samp$p.value > 0.05) {
  cat("Conclusion: No significant age imbalance between arms (p =",
      round(result_2samp$p.value, 4), ").\n")
  cat("Randomisation appears successful — age is balanced at baseline.\n")
}
```

### 2.5 Visualisation

```{r}
#| label: twosample-plot
#| fig-cap: "Box plot comparing age between Placebo and Xanomeline High Dose"

ggplot(dm_2arm, aes(x = ARM, y = AGE, fill = ARM)) +
  geom_boxplot(alpha = 0.7, outlier.shape = 16, outlier.size = 2) +
  geom_jitter(width = 0.15, alpha = 0.4, size = 1.5) +
  scale_fill_manual(values = c("Placebo"               = "#2C6E9E",
                                "Xanomeline High Dose" = "#E74C3C")) +
  labs(
    title    = "Age Distribution by Treatment Arm",
    subtitle = paste0("Two-sample Welch t-test: p = ",
                      round(result_2samp$p.value, 4)),
    x        = "Treatment Arm",
    y        = "Age (years)"
  ) +
  theme_classic(base_size = 13) +
  theme(legend.position = "none")
```

> **Clinical note:** Baseline demographic comparability is assessed in **Table 1** of most clinical study reports. A non-significant t-test supports the conclusion that randomisation adequately balanced age across treatment groups.

---

## 3. Paired t-Test

### 3.1 Theory

The paired t-test compares two related measurements — typically the same subject at two time points:

$$t = \frac{\bar{d}}{s_d / \sqrt{n}}$$

where $\bar{d}$ is the mean of within-subject differences and $s_d$ is their standard deviation. This design removes between-subject variability, yielding greater power than the two-sample test.

**Clinical question:** Does systolic blood pressure (SYSBP) change significantly from screening (VISITNUM = 1) to the first follow-up visit (VISITNUM = 2)?

### 3.2 Data Preparation

```{r}
#| label: paired-prep

# Extract baseline (VISITNUM=1) and first follow-up (VISITNUM=2) SYSBP
vs_paired <- vs_sysbp |>
  filter(VISITNUM %in% c(1, 2)) |>
  select(USUBJID, VISITNUM, VSSTRESN) |>
  group_by(USUBJID, VISITNUM) |>
  summarise(VSSTRESN = mean(VSSTRESN, na.rm = TRUE), .groups = "drop") |>
  tidyr::pivot_wider(names_from  = VISITNUM,
                     values_from = VSSTRESN,
                     names_prefix = "VISIT") |>
  filter(!is.na(VISIT1), !is.na(VISIT2)) |>
  mutate(diff = VISIT2 - VISIT1)

cat("Subjects with both VISITNUM 1 and 2:", nrow(vs_paired), "\n")
cat("Mean difference (Visit2 - Visit1):", round(mean(vs_paired$diff), 2), "mmHg\n")
cat("SD of differences:", round(sd(vs_paired$diff), 2), "mmHg\n")
```

### 3.3 Performing the Test

```{r}
#| label: paired-ttest

# Shapiro-Wilk on differences
sw_diff <- shapiro.test(vs_paired$diff)
cat("Shapiro-Wilk on differences: W =", round(sw_diff$statistic, 4),
    ", p =", round(sw_diff$p.value, 4), "\n\n")

# Paired t-test
result_paired <- t.test(vs_paired$VISIT2, vs_paired$VISIT1,
                         paired = TRUE, alternative = "two.sided")
print(result_paired)

cat("\n--- Clinical Interpretation ---\n")
cat("Mean SBP at Visit 1:", round(mean(vs_paired$VISIT1), 1), "mmHg\n")
cat("Mean SBP at Visit 2:", round(mean(vs_paired$VISIT2), 1), "mmHg\n")
cat("Mean change:", round(mean(vs_paired$diff), 1), "mmHg\n")
cat("95% CI for change: [", round(result_paired$conf.int[1], 1), ",",
    round(result_paired$conf.int[2], 1), "] mmHg\n")
if (result_paired$p.value < 0.05) {
  cat("Conclusion: Statistically significant change in SBP between visits (p =",
      round(result_paired$p.value, 4), ").\n")
} else {
  cat("Conclusion: No statistically significant change in SBP between visits (p =",
      round(result_paired$p.value, 4), ").\n")
}
```

### 3.4 Visualisation

```{r}
#| label: paired-plot
#| fig-cap: "Paired SBP measurements at Visit 1 and Visit 2 with change distribution"

# Spaghetti-style paired plot (sample for clarity)
set.seed(42)
vs_long <- vs_paired |>
  slice_sample(n = min(60, nrow(vs_paired))) |>
  tidyr::pivot_longer(cols = c(VISIT1, VISIT2),
                      names_to  = "Visit",
                      values_to = "SBP") |>
  mutate(Visit = case_match(Visit,
                            "VISIT1" ~ "Visit 1\n(Baseline)",
                            "VISIT2" ~ "Visit 2\n(Follow-up)"))

p1 <- ggplot(vs_long, aes(x = Visit, y = SBP, group = USUBJID)) +
  geom_line(alpha = 0.3, colour = "#2C6E9E") +
  geom_point(alpha = 0.5, size = 1.5, colour = "#2C6E9E") +
  stat_summary(aes(group = 1), fun = mean, geom = "line",
               colour = "#E74C3C", linewidth = 1.5, linetype = "dashed") +
  stat_summary(aes(group = 1), fun = mean, geom = "point",
               colour = "#E74C3C", size = 4) +
  labs(
    title    = "Paired SBP: Baseline to Follow-up",
    subtitle = paste0("Paired t-test: p = ", round(result_paired$p.value, 4)),
    x        = NULL,
    y        = "Systolic BP (mmHg)"
  ) +
  theme_classic(base_size = 13)

# Distribution of differences
p2 <- ggplot(vs_paired, aes(x = diff)) +
  geom_histogram(bins = 20, fill = "#27AE60", colour = "white", alpha = 0.8) +
  geom_vline(xintercept = 0, colour = "#E74C3C",
             linetype = "dashed", linewidth = 1) +
  geom_vline(xintercept = mean(vs_paired$diff), colour = "#1A3E5C",
             linewidth = 1.2) +
  labs(
    title    = "Distribution of Within-Subject SBP Change",
    subtitle = paste0("Mean change = ", round(mean(vs_paired$diff), 1), " mmHg"),
    x        = "SBP Change (Visit 2 - Visit 1, mmHg)",
    y        = "Count"
  ) +
  theme_classic(base_size = 13)

print(p1)
print(p2)
```

> **Clinical note:** Paired designs are standard for within-patient efficacy analyses (e.g., change from baseline). The paired t-test is the parametric equivalent of the matched-pairs analysis used in crossover trials and pre-post studies.

---

## 4. One-Way ANOVA

### 4.1 Theory

One-way ANOVA extends the two-sample t-test to $k \geq 3$ groups. It partitions total variability into:

$$\text{SS}_{\text{Total}} = \text{SS}_{\text{Between}} + \text{SS}_{\text{Within}}$$

The F-statistic:

$$F = \frac{\text{MS}_{\text{Between}}}{\text{MS}_{\text{Within}}} = \frac{\text{SS}_{\text{Between}} / (k-1)}{\text{SS}_{\text{Within}} / (N-k)}$$

**Clinical question:** Does mean systolic blood pressure differ across the three treatment arms?

### 4.2 Assumption Checks

```{r}
#| label: anova1-assumptions
#| fig-cap: "Diagnostic plots for one-way ANOVA assumption checking"

# Use Visit 1 (baseline) SBP across all 3 arms
vs_anova <- vs_sysbp |>
  filter(VISITNUM == 1, !is.na(VSSTRESN)) |>
  select(USUBJID, ARM, VSSTRESN)

cat("Sample sizes by ARM:\n")
print(table(vs_anova$ARM))

# Shapiro-Wilk per group
cat("\nShapiro-Wilk normality tests by ARM:\n")
by(vs_anova$VSSTRESN, vs_anova$ARM, function(x) {
  sw <- shapiro.test(x)
  cat("  W =", round(sw$statistic, 4), ", p =", round(sw$p.value, 4), "\n")
})

# Levene's test for homogeneity of variance
lev_anova <- car::leveneTest(VSSTRESN ~ ARM, data = vs_anova)
cat("\nLevene's Test:\n")
print(lev_anova)

# Q-Q plots per group
ggplot(vs_anova, aes(sample = VSSTRESN, colour = ARM)) +
  stat_qq(size = 2, alpha = 0.7) +
  stat_qq_line(linewidth = 0.8) +
  facet_wrap(~ARM, ncol = 3) +
  scale_colour_manual(values = c(
    "Placebo"               = "#2C6E9E",
    "Xanomeline Low Dose"  = "#27AE60",
    "Xanomeline High Dose" = "#E74C3C"
  )) +
  labs(
    title    = "Q-Q Plots of Baseline SBP by Treatment Arm",
    x = "Theoretical Quantiles",
    y = "Sample Quantiles (SBP, mmHg)"
  ) +
  theme_classic(base_size = 12) +
  theme(legend.position = "none")
```

### 4.3 Performing One-Way ANOVA

```{r}
#| label: anova1-test

# Fit one-way ANOVA
model_anova1 <- aov(VSSTRESN ~ ARM, data = vs_anova)
anova_summary <- summary(model_anova1)
print(anova_summary)

cat("\n--- Clinical Interpretation ---\n")
group_means <- tapply(vs_anova$VSSTRESN, vs_anova$ARM, mean, na.rm = TRUE)
for (arm in names(group_means)) {
  cat(arm, ": mean SBP =", round(group_means[arm], 1), "mmHg\n")
}

f_val <- anova_summary[[1]]$`F value`[1]
p_val <- anova_summary[[1]]$`Pr(>F)`[1]
cat("\nF-statistic:", round(f_val, 3), ", p-value:", round(p_val, 4), "\n")

if (p_val < 0.05) {
  cat("Conclusion: Significant difference in baseline SBP across arms (p =",
      round(p_val, 4), ").\n")
  cat("Post-hoc tests are needed to identify which pairs differ.\n")
} else {
  cat("Conclusion: No significant difference in SBP across arms (p =",
      round(p_val, 4), ").\n")
  cat("Baseline SBP is balanced across treatment groups — consistent with good randomisation.\n")
}
```

### 4.4 Visualisation

```{r}
#| label: anova1-plot
#| fig-cap: "Violin plot of baseline SBP across treatment arms"

ggplot(vs_anova, aes(x = ARM, y = VSSTRESN, fill = ARM)) +
  geom_violin(alpha = 0.6, trim = FALSE) +
  geom_boxplot(width = 0.15, fill = "white", alpha = 0.8, outlier.size = 1.5) +
  scale_fill_manual(values = c(
    "Placebo"               = "#2C6E9E",
    "Xanomeline Low Dose"  = "#27AE60",
    "Xanomeline High Dose" = "#E74C3C"
  )) +
  labs(
    title    = "Baseline Systolic BP by Treatment Arm",
    subtitle = paste0("One-Way ANOVA: F = ", round(f_val, 2),
                      ", p = ", round(p_val, 4)),
    x        = "Treatment Arm",
    y        = "Systolic BP (mmHg)"
  ) +
  theme_classic(base_size = 13) +
  theme(
    legend.position  = "none",
    axis.text.x      = element_text(size = 10)
  )
```

---

## 5. Two-Way ANOVA: Treatment × Sex Interaction

### 5.1 Theory

Two-way ANOVA simultaneously tests the effects of two categorical factors and their interaction:

$$y_{ijk} = \mu + \alpha_i + \beta_j + (\alpha\beta)_{ij} + \epsilon_{ijk}$$

where $\alpha_i$ = ARM effect, $\beta_j$ = SEX effect, $(\alpha\beta)_{ij}$ = interaction. A significant interaction means the effect of treatment differs by sex (or vice versa) — critical for subgroup analyses in regulatory submissions.

**Clinical question:** Does the treatment arm effect on SBP differ by sex?

### 5.2 Fitting the Model

```{r}
#| label: anova2-fit

vs_anova2 <- vs_anova |>
  left_join(dm_clean |> select(USUBJID, SEX), by = "USUBJID") |>
  filter(!is.na(SEX))

cat("Cell counts (ARM x SEX):\n")
print(table(vs_anova2$ARM, vs_anova2$SEX))

# Two-way ANOVA with interaction
model_anova2 <- aov(VSSTRESN ~ ARM * SEX, data = vs_anova2)
print(summary(model_anova2))

cat("\n--- Clinical Interpretation ---\n")
anova2_table <- summary(model_anova2)[[1]]
p_arm        <- anova2_table["ARM",          "Pr(>F)"]
p_sex        <- anova2_table["SEX",          "Pr(>F)"]
p_interaction <- anova2_table["ARM:SEX",     "Pr(>F)"]

cat("ARM main effect: p =", round(p_arm, 4), "\n")
cat("SEX main effect: p =", round(p_sex, 4), "\n")
cat("ARM:SEX interaction: p =", round(p_interaction, 4), "\n\n")

if (p_interaction < 0.05) {
  cat("Significant interaction detected. The treatment effect on SBP differs by sex.\n")
  cat("Subgroup analyses by sex are warranted.\n")
} else {
  cat("No significant treatment-by-sex interaction (p =", round(p_interaction, 4), ").\n")
  cat("The treatment effect is consistent across sexes — supports a pooled analysis.\n")
}
```

### 5.3 Interaction Plot

```{r}
#| label: anova2-plot
#| fig-cap: "Interaction plot: mean SBP by treatment arm and sex"

cell_means <- vs_anova2 |>
  group_by(ARM, SEX) |>
  summarise(
    mean_sbp = mean(VSSTRESN, na.rm = TRUE),
    se_sbp   = sd(VSSTRESN, na.rm = TRUE) / sqrt(n()),
    .groups  = "drop"
  )

ggplot(cell_means, aes(x = ARM, y = mean_sbp,
                        colour = SEX, group = SEX)) +
  geom_line(linewidth = 1.2) +
  geom_point(size = 4) +
  geom_errorbar(aes(ymin = mean_sbp - 1.96 * se_sbp,
                    ymax = mean_sbp + 1.96 * se_sbp),
                width = 0.15, linewidth = 0.8) +
  scale_colour_manual(values = c("F" = "#E74C3C", "M" = "#2C6E9E"),
                      labels  = c("F" = "Female", "M" = "Male")) +
  labs(
    title    = "Treatment × Sex Interaction: Baseline SBP",
    subtitle = paste0("Interaction p = ", round(p_interaction, 4)),
    x        = "Treatment Arm",
    y        = "Mean Systolic BP \u00b1 95% CI (mmHg)",
    colour   = "Sex"
  ) +
  theme_classic(base_size = 13) +
  theme(axis.text.x = element_text(size = 10))
```

> **Clinical note:** Interaction tests are routinely included in subgroup analyses for regulatory submissions (ICH E9 guideline). A non-significant treatment-by-sex interaction provides support for reporting a pooled primary result.

---

## 6. ANCOVA: Adjusting for Baseline

### 6.1 Theory

ANCOVA (Analysis of Covariance) combines ANOVA with linear regression. It removes the variability explained by a covariate (e.g., baseline measurement) before testing for group differences:

$$y_{ij} = \mu + \tau_i + \beta(x_{ij} - \bar{x}) + \epsilon_{ij}$$

where $\tau_i$ is the treatment effect and $\beta$ is the regression coefficient for the baseline covariate $x$.

**Benefits in clinical trials:**
- Increased statistical power (reduced residual variance)
- Control for baseline imbalances even after randomisation
- Required by ICH E9(R1) for many primary efficacy analyses

**Clinical question:** Does treatment arm affect SBP at Visit 2 after adjusting for baseline (Visit 1) SBP?

### 6.2 Fitting ANCOVA

```{r}
#| label: ancova-fit

# Build dataset: baseline (V1) and post-treatment (V2) SBP with ARM
vs_ancova <- vs_sysbp |>
  filter(VISITNUM %in% c(1, 2)) |>
  select(USUBJID, VISITNUM, VSSTRESN, ARM) |>
  group_by(USUBJID, VISITNUM, ARM) |>
  summarise(VSSTRESN = mean(VSSTRESN, na.rm = TRUE), .groups = "drop") |>
  tidyr::pivot_wider(names_from  = VISITNUM,
                     values_from = VSSTRESN,
                     names_prefix = "V") |>
  filter(!is.na(V1), !is.na(V2))

cat("ANCOVA dataset: n =", nrow(vs_ancova), "\n")
cat("Missingness check: V1 missing =", sum(is.na(vs_ancova$V1)),
    ", V2 missing =", sum(is.na(vs_ancova$V2)), "\n\n")

# ANCOVA model: post-treatment SBP ~ ARM + baseline SBP
model_ancova <- lm(V2 ~ ARM + V1, data = vs_ancova)
ancova_anova <- car::Anova(model_ancova, type = "III")
print(ancova_anova)

cat("\n--- Coefficient Table ---\n")
print(summary(model_ancova)$coefficients)

cat("\n--- Clinical Interpretation ---\n")
p_arm_ancova <- ancova_anova["ARM", "Pr(>F)"]
cat("Treatment ARM effect after adjusting for baseline SBP:\n")
cat("  p-value =", round(p_arm_ancova, 4), "\n")
cat("  R-squared =", round(summary(model_ancova)$r.squared, 3), "\n")
cat("  Baseline SBP coefficient:", round(coef(model_ancova)["V1"], 3), "\n\n")

if (p_arm_ancova < 0.05) {
  cat("Conclusion: Significant treatment effect on SBP after baseline adjustment.\n")
} else {
  cat("Conclusion: No significant treatment effect after baseline adjustment (p =",
      round(p_arm_ancova, 4), ").\n")
}
```

### 6.3 ANCOVA Visualisation

```{r}
#| label: ancova-plot
#| fig-cap: "ANCOVA plot: post-treatment SBP vs baseline SBP, with regression lines by arm"

ggplot(vs_ancova, aes(x = V1, y = V2, colour = ARM)) +
  geom_point(alpha = 0.5, size = 2) +
  geom_smooth(method = "lm", se = TRUE, linewidth = 1.2, alpha = 0.15) +
  scale_colour_manual(values = c(
    "Placebo"               = "#2C6E9E",
    "Xanomeline Low Dose"  = "#27AE60",
    "Xanomeline High Dose" = "#E74C3C"
  )) +
  geom_abline(intercept = 0, slope = 1,
              linetype = "dashed", colour = "grey50", linewidth = 0.8) +
  labs(
    title    = "ANCOVA: Visit 2 SBP Adjusted for Baseline (Visit 1)",
    subtitle = paste0("ARM effect p = ", round(p_arm_ancova, 4),
                      " after adjusting for baseline"),
    x        = "Baseline SBP — Visit 1 (mmHg)",
    y        = "Post-Treatment SBP — Visit 2 (mmHg)",
    colour   = "Treatment Arm"
  ) +
  theme_classic(base_size = 13) +
  theme(legend.position = "bottom")
```

---

## 7. Assumption Checking: A Systematic Approach

### 7.1 ANOVA Residual Diagnostics

```{r}
#| label: residual-diagnostics
#| fig-cap: "Four-panel residual diagnostic plot for one-way ANOVA"
#| fig-width: 9
#| fig-height: 7

par(mfrow = c(2, 2))
plot(model_anova1, which = 1:4)
par(mfrow = c(1, 1))
```

### 7.2 Interpreting Diagnostic Plots

```{r}
#| label: diagnostics-interpretation

# Extract residuals
resids <- residuals(model_anova1)

# Shapiro-Wilk on residuals
sw_resid <- shapiro.test(resids)
cat("Shapiro-Wilk on ANOVA residuals:\n")
cat("  W =", round(sw_resid$statistic, 4), ", p =", round(sw_resid$p.value, 4), "\n\n")

cat("Diagnostic Plot Guide:\n")
cat("1. Residuals vs Fitted: Should show no pattern — tests independence\n")
cat("2. Normal Q-Q:          Points near the line — tests normality of residuals\n")
cat("3. Scale-Location:      Horizontal band — tests homoscedasticity\n")
cat("4. Cook's Distance:     No point > 0.5 — identifies influential observations\n\n")

cat("Cook's Distance: max =", round(max(cooks.distance(model_anova1)), 4), "\n")
n_influential <- sum(cooks.distance(model_anova1) > 4 / nrow(vs_anova))
cat("Observations with Cook's D > 4/n:", n_influential, "\n")
```

### 7.3 Testing All Three Key Assumptions

```{r}
#| label: all-assumptions

cat("=========================================\n")
cat("ASSUMPTION CHECKING SUMMARY\n")
cat("=========================================\n\n")

# 1. Normality
cat("1. NORMALITY (Shapiro-Wilk per group):\n")
sw_results <- tapply(vs_anova$VSSTRESN, vs_anova$ARM, function(x) {
  if (length(x) >= 3) shapiro.test(x) else list(statistic = NA, p.value = NA)
})
for (nm in names(sw_results)) {
  cat("  ", nm, ": p =", round(sw_results[[nm]]$p.value, 4), "\n")
}

# 2. Homogeneity of variance
lev <- car::leveneTest(VSSTRESN ~ ARM, data = vs_anova)
cat("\n2. HOMOGENEITY OF VARIANCE (Levene's Test):\n")
cat("   F =", round(lev[1, "F value"], 3),
    ", p =", round(lev[1, "Pr(>F)"], 4), "\n")

# 3. Independence — structural assumption
cat("\n3. INDEPENDENCE:\n")
cat("   Assumed by CDISC ADSL/ADVS design (one record per subject per visit).\n")
cat("   Parallel group design with randomisation ensures independence.\n")

cat("\nDecision matrix:\n")
cat("Normality OK?  Variance OK?  Recommended test\n")
cat("YES            YES           One-way ANOVA (current)\n")
cat("YES            NO            Welch's ANOVA (oneway.test)\n")
cat("NO             -             Kruskal-Wallis (see Chapter BS4)\n")
```

---

## 8. Post-Hoc Tests: TukeyHSD

### 8.1 Why Post-Hoc Tests?

When ANOVA rejects $H_0$, we know at least one group differs — but not *which* pairs. Post-hoc procedures make all pairwise comparisons while controlling the family-wise error rate (FWER).

**Tukey's Honestly Significant Difference (HSD):** Controls FWER at $\alpha$ for all $\binom{k}{2}$ pairwise comparisons. Appropriate when comparing all pairs with equal sample sizes.

### 8.2 Tukey HSD with Significant ANOVA

```{r}
#| label: posthoc-tukey

# For illustration, use AGE across all 3 arms (re-fit ANOVA)
model_age_anova <- aov(AGE ~ ARM, data = dm_clean)
cat("ANOVA for AGE across all 3 arms:\n")
print(summary(model_age_anova))

# TukeyHSD
tukey_result <- TukeyHSD(model_age_anova)
cat("\nTukey HSD pairwise comparisons:\n")
print(tukey_result)
```

### 8.3 Visualise Tukey Intervals

```{r}
#| label: posthoc-plot
#| fig-cap: "Tukey HSD confidence intervals for all pairwise age comparisons"

tukey_df <- as.data.frame(tukey_result$ARM) |>
  tibble::rownames_to_column("comparison") |>
  rename(diff = diff, lower = lwr, upper = upr, p_adj = `p adj`)

ggplot(tukey_df, aes(x = comparison, y = diff)) +
  geom_hline(yintercept = 0, linetype = "dashed",
             colour = "#E74C3C", linewidth = 1) +
  geom_point(size = 4, colour = "#2C6E9E") +
  geom_errorbar(aes(ymin = lower, ymax = upper),
                width = 0.25, linewidth = 1.2, colour = "#2C6E9E") +
  geom_text(aes(label = paste0("p = ", round(p_adj, 3))),
            vjust = -1.2, size = 3.5) +
  coord_flip() +
  labs(
    title    = "Tukey HSD: Pairwise Comparisons of Age",
    subtitle = "95% family-wise confidence level",
    x        = "Comparison",
    y        = "Difference in Mean Age (years)"
  ) +
  theme_classic(base_size = 13)
```

### 8.4 Other Post-Hoc Options

```{r}
#| label: posthoc-alternatives

# Bonferroni correction — more conservative
pairwise_result <- pairwise.t.test(dm_clean$AGE, dm_clean$ARM,
                                    p.adjust.method = "bonferroni")
cat("Pairwise t-tests with Bonferroni correction:\n")
print(pairwise_result)

# Holm method — step-down, uniformly better than Bonferroni
pairwise_holm <- pairwise.t.test(dm_clean$AGE, dm_clean$ARM,
                                   p.adjust.method = "holm")
cat("\nPairwise t-tests with Holm correction:\n")
print(pairwise_holm)

cat("\nPost-hoc method selection guide:\n")
cat("TukeyHSD:   All pairwise comparisons, equal n, controls FWER\n")
cat("Bonferroni: Conservative, any comparison structure\n")
cat("Holm:       Less conservative than Bonferroni, recommended default\n")
cat("Dunnett:    Compare each treatment to control only (most powerful)\n")
```

> **Clinical note:** ICH E9 guidance recommends pre-specifying the multiple comparison procedure in the Statistical Analysis Plan (SAP). Dunnett's test is often preferred in dose-response trials where each dose is compared only to placebo, as it is more powerful than Tukey HSD.

---

## 9. Welch's ANOVA: When Variances Are Unequal

```{r}
#| label: welch-anova

# Welch's ANOVA — robust to heteroscedasticity
welch_anova <- oneway.test(VSSTRESN ~ ARM, data = vs_anova, var.equal = FALSE)
cat("Welch's ANOVA (heteroscedasticity-robust):\n")
print(welch_anova)

# Compare with standard ANOVA
std_anova <- oneway.test(VSSTRESN ~ ARM, data = vs_anova, var.equal = TRUE)
cat("\nStandard ANOVA (assumes equal variance):\n")
print(std_anova)

cat("\nComparison:\n")
cat("Standard ANOVA F =", round(std_anova$statistic, 3),
    ", p =", round(std_anova$p.value, 4), "\n")
cat("Welch's ANOVA  F =", round(welch_anova$statistic, 3),
    ", p =", round(welch_anova$p.value, 4), "\n")
cat("\nIf Levene's p < 0.05, prefer Welch's ANOVA.\n")
```

---

## 10. Reporting Parametric Results: A Template

```{r}
#| label: reporting-template

cat("=========================================\n")
cat("STATISTICAL RESULTS REPORTING TEMPLATE\n")
cat("(Suitable for Clinical Study Report)\n")
cat("=========================================\n\n")

# Pull key values
n_total  <- nrow(dm_clean)
mean_age <- round(mean(dm_clean$AGE), 1)
sd_age   <- round(sd(dm_clean$AGE),   1)
t_stat   <- round(result_1samp$statistic, 2)
df_val   <- result_1samp$parameter
ci_lo    <- round(result_1samp$conf.int[1], 1)
ci_hi    <- round(result_1samp$conf.int[2], 1)
p1       <- round(result_1samp$p.value, 4)

cat("One-Sample t-Test:\n")
cat("  The mean age of participants (n =", n_total, ") was",
    mean_age, "(SD =", sd_age, ") years.\n")
cat("  A one-sample t-test showed that the mean age was significantly\n")
cat("  different from 65 years: t(", df_val, ") =", t_stat,
    ", 95% CI [", ci_lo, ",", ci_hi, "],",
    "p =", p1, ".\n\n")

cat("Two-Sample t-Test:\n")
cat("  Age was compared between Placebo and Xanomeline High Dose arms.\n")
cat("  A Welch two-sample t-test indicated no significant imbalance:\n")
cat("  t(", round(result_2samp$parameter, 1), ") =",
    round(result_2samp$statistic, 2),
    ", p =", round(result_2samp$p.value, 4), ".\n\n")

cat("ANOVA:\n")
cat("  One-way ANOVA of baseline SBP across three arms:\n")
cat("  F(", anova_summary[[1]]$Df[1], ",",
    anova_summary[[1]]$Df[2], ") =",
    round(anova_summary[[1]]$`F value`[1], 2),
    ", p =", round(anova_summary[[1]]$`Pr(>F)`[1], 4), ".\n")
```

---

## Summary

| Test | Use Case | Key Assumption | R Function |
|------|----------|---------------|------------|
| One-sample t-test | Mean vs. known value | Normality | `t.test(x, mu=)` |
| Two-sample t-test | Two group means | Normality, (equal var) | `t.test(y ~ grp)` |
| Paired t-test | Before-after / matched | Differences normal | `t.test(x, y, paired=TRUE)` |
| One-way ANOVA | ≥3 group means | Normality, equal var | `aov(y ~ grp)` |
| Two-way ANOVA | 2 factors + interaction | Normality, equal var | `aov(y ~ A * B)` |
| ANCOVA | Groups + covariate | Linearity, normality | `lm(y ~ grp + cov)` |
| Welch's ANOVA | ≥3 groups, unequal var | Normality | `oneway.test(var.equal=FALSE)` |
| TukeyHSD | Post-hoc all pairs | ANOVA significant | `TukeyHSD(aov_obj)` |

---

## Exercises

::: {.callout-note icon="false"}
## Exercise 1: One-Sample t-Test on BMI

Using the `dm` dataset (if `BMIBL` is available) or weight-related variables from `vs`:

1. Extract body weight (`VSTESTCD == "WEIGHT"`) from the `vs` dataset at baseline.
2. Test whether the mean weight differs from 70 kg using a one-sample t-test.
3. Check the normality assumption using Shapiro-Wilk and a Q-Q plot.
4. Write two sentences interpreting the result as you would in a clinical study report.
:::

::: {.callout-note icon="false"}
## Exercise 2: Two-Sample t-Test

1. Filter `dm_clean` to Placebo and Xanomeline **Low** Dose arms only.
2. Compare mean `AGE` between these two arms using a Welch t-test.
3. First check whether variances are equal using Levene's test.
4. Create a publication-ready box plot with the p-value annotated.
:::

::: {.callout-note icon="false"}
## Exercise 3: One-Way ANOVA with Post-Hoc

1. Extract diastolic BP (`VSTESTCD == "DIABP"`) at VISITNUM == 1.
2. Run a one-way ANOVA comparing diastolic BP across all three ARM groups.
3. Check the four standard ANOVA assumptions.
4. If ANOVA is significant, run Tukey HSD and visualise the confidence intervals.
5. Which pairs of arms differ significantly?
:::

::: {.callout-note icon="false"}
## Exercise 4: ANCOVA Analysis

1. Use `VSTESTCD == "DIABP"` at VISITNUM 1 (baseline) and VISITNUM 3 (post-treatment).
2. Fit an ANCOVA model: `V3 ~ ARM + V1`.
3. Interpret the ARM coefficient and the baseline covariate coefficient.
4. Compare the p-value for ARM from plain ANOVA vs ANCOVA — which has more power? Why?
:::

::: {.callout-note icon="false"}
## Exercise 5: Two-Way ANOVA

1. Use baseline SYSBP data across all three arms.
2. Fit a two-way ANOVA: `VSSTRESN ~ ARM * SEX`.
3. Interpret the interaction term.
4. Create an interaction plot showing mean SBP by ARM and SEX with 95% CI error bars.
5. Discuss the clinical implications if the interaction were statistically significant.
:::

::: {.callout-note icon="false"}
## Challenge Exercise: Complete Parametric Analysis Pipeline

Write a single R function `parametric_pipeline(data, outcome_var, group_var, baseline_var = NULL)` that:

1. Checks normality (Shapiro-Wilk) for each group.
2. Checks variance equality (Levene's test).
3. Selects and runs the appropriate test (t-test, Welch t-test, ANOVA, Welch ANOVA, or ANCOVA).
4. Runs post-hoc tests if ANOVA is significant.
5. Returns a list with test name, statistic, p-value, and a ggplot visualisation.

Test your function on at least two different variables from the `vs` dataset.
:::
