# Non-Parametric Tests in Clinical Research

## Introduction

Non-parametric (distribution-free) tests make fewer assumptions about the underlying data distribution. They are particularly valuable in clinical research when:

- Sample sizes are small (n < 30) and normality cannot be verified
- Data are ordinal (e.g., pain scores, quality-of-life ratings)
- Distributions are heavily skewed or contain outliers
- Outcomes are censored counts or time-to-event with small samples

This chapter covers:

- **Wilcoxon signed-rank test**: Paired or one-sample comparisons
- **Mann–Whitney U test**: Two independent groups
- **Kruskal–Wallis test**: Multiple independent groups
- **Friedman test**: Repeated measures across groups
- **Spearman correlation**: Monotonic association between variables
- **Permutation tests**: Exact non-parametric inference
- **Choosing between parametric and non-parametric approaches**

---

## 1. When to Use Non-Parametric Tests

```{r}
#| label: np-setup
#| message: false
#| warning: false

library(tidyverse)
library(broom)
library(gt)
set.seed(42)

# Generate clinical dataset: pain score (0-10, ordinal/skewed)
n_per_arm <- 30
pain_data <- tibble(
  subject_id = paste0("PT-", sprintf("%03d", 1:(n_per_arm * 2))),
  arm        = rep(c("Active", "Placebo"), each = n_per_arm),
  pain_base  = c(round(rbeta(n_per_arm, 3, 1.5) * 10),
                 round(rbeta(n_per_arm, 3, 1.5) * 10)),
  pain_week8 = c(pmax(0, round(rbeta(n_per_arm, 1.5, 3) * 10)),   # Active: lower
                 pmax(0, round(rbeta(n_per_arm, 2.5, 2) * 10)))    # Placebo: moderate
) |>
  mutate(pain_chg = pain_week8 - pain_base)

# Check normality
cat("=== Normality Assessment ===\n")
for (arm_name in c("Active", "Placebo")) {
  vals <- pain_data |> filter(arm == arm_name) |> pull(pain_chg)
  sw   <- shapiro.test(vals)
  cat(arm_name, ": W =", round(sw$statistic, 4),
      ", p =", round(sw$p.value, 4), "\n")
}
```

```{r}
#| label: distribution-check
#| fig-cap: "Pain score changes: departure from normality motivates non-parametric tests"

pain_data |>
  ggplot(aes(x = pain_chg, fill = arm)) +
  geom_histogram(bins = 15, alpha = 0.75, colour = "white") +
  facet_wrap(~ arm, ncol = 1) +
  scale_fill_manual(values = c("Active" = "#2C6E9E", "Placebo" = "#E07B39")) +
  labs(
    x = "Change in Pain Score (Week 8 – Baseline)",
    y = "Count",
    fill = "Treatment",
    title = "Distribution of Pain Score Changes by Treatment Arm",
    subtitle = "Skewed distributions suggest non-parametric methods"
  ) +
  theme_bw(base_size = 12) +
  theme(legend.position = "none")
```

---

## 2. Wilcoxon Signed-Rank Test

### 2.1 Concept and Hypotheses

The Wilcoxon signed-rank test evaluates whether the median of paired differences (or a one-sample median) equals a hypothesised value. It replaces the paired t-test when the normality assumption fails.

**Procedure**:
1. Compute pairwise differences $d_i = x_{i,\text{post}} - x_{i,\text{pre}}$
2. Rank the absolute values $|d_i|$ (excluding zeros)
3. Assign signs to ranks based on the sign of $d_i$
4. Test statistic: $W^+ = \sum_{d_i > 0} \text{rank}(|d_i|)$

**Hypotheses** (two-tailed): $H_0: \tilde{d} = 0$ vs $H_1: \tilde{d} \neq 0$

```{r}
#| label: signed-rank-test

# Paired comparison: pain score baseline vs. week 8 (Active arm only)
active_pain <- pain_data |> filter(arm == "Active")

# Parametric paired t-test for comparison
t_paired <- t.test(active_pain$pain_week8, active_pain$pain_base, paired = TRUE)

# Wilcoxon signed-rank test
w_signed <- wilcox.test(active_pain$pain_week8, active_pain$pain_base,
                        paired = TRUE, conf.int = TRUE)

cat("=== Paired Comparison: Pain Score (Active Arm) ===\n\n")
cat("Parametric paired t-test:\n")
cat("  t =", round(t_paired$statistic, 3),
    ", p =", round(t_paired$p.value, 4), "\n")
cat("  95% CI:", round(t_paired$conf.int[1], 2), "to",
    round(t_paired$conf.int[2], 2), "\n\n")

cat("Wilcoxon signed-rank test:\n")
cat("  V =", w_signed$statistic,
    ", p =", round(w_signed$p.value, 4), "\n")
cat("  95% CI (pseudomedian):", round(w_signed$conf.int[1], 2),
    "to", round(w_signed$conf.int[2], 2), "\n")
```

### 2.2 Visualising Paired Data

```{r}
#| label: paired-plot
#| fig-cap: "Paired pain scores before and after treatment (Active arm) with connecting lines"

active_pain |>
  pivot_longer(c(pain_base, pain_week8),
               names_to = "timepoint",
               values_to = "score") |>
  mutate(timepoint = case_match(timepoint,
                               "pain_base"  ~ "Baseline",
                               "pain_week8" ~ "Week 8")) |>
  ggplot(aes(x = timepoint, y = score, group = subject_id)) +
  geom_line(alpha = 0.35, colour = "#2C6E9E") +
  geom_point(aes(colour = timepoint), size = 2.5, alpha = 0.7) +
  stat_summary(aes(group = 1), fun = median, geom = "line",
               linewidth = 1.5, colour = "#D32F2F") +
  stat_summary(aes(group = 1), fun = median, geom = "point",
               size = 4, colour = "#D32F2F", shape = 18) +
  scale_colour_manual(values = c("Baseline" = "#757575", "Week 8" = "#2C6E9E")) +
  labs(
    x = "Timepoint",
    y = "Pain Score (0–10)",
    colour = NULL,
    title = "Individual and Median Pain Scores (Active Arm)",
    subtitle = "Red line = median trajectory"
  ) +
  theme_bw(base_size = 12) +
  theme(legend.position = "bottom")
```

---

## 3. Mann–Whitney U Test (Wilcoxon Rank-Sum Test)

### 3.1 Concept

The Mann–Whitney U test (equivalent to the Wilcoxon rank-sum test) compares the distribution of outcomes between two independent groups. It tests whether one group tends to have higher values than the other.

**Hypotheses**: $H_0$: The two populations are identically distributed
$H_1$: One population tends to produce larger values

The test statistic $U$ counts the number of times an observation from group 1 exceeds an observation from group 2.

```{r}
#| label: mann-whitney

# Compare pain changes between Active and Placebo
mw_result <- wilcox.test(pain_chg ~ arm, data = pain_data,
                         conf.int = TRUE, conf.level = 0.95)

# For comparison, parametric test
t_indep <- t.test(pain_chg ~ arm, data = pain_data, var.equal = FALSE)

cat("=== Pain Change: Active vs. Placebo ===\n\n")
cat("Medians:\n")
pain_data |>
  group_by(arm) |>
  summarise(median = median(pain_chg), IQR = IQR(pain_chg)) |>
  print()

cat("\nMann-Whitney U test:\n")
cat("  W =", mw_result$statistic,
    ", p =", round(mw_result$p.value, 4), "\n")
cat("  Hodges-Lehmann estimate:", round(mw_result$estimate, 2), "\n")
cat("  95% CI:", round(mw_result$conf.int[1], 2), "to",
    round(mw_result$conf.int[2], 2), "\n\n")

cat("Two-sample t-test (for comparison):\n")
cat("  t =", round(t_indep$statistic, 3),
    ", p =", round(t_indep$p.value, 4), "\n")
```

### 3.2 The Probability of Superiority

A useful effect size for the Mann–Whitney test is the **common language effect size** (CLES), also called the probability of superiority:

$$P(\text{Active} < \text{Placebo}) = \frac{U}{n_1 \cdot n_2}$$

```{r}
#| label: cles

active_vals  <- pain_data |> filter(arm == "Active")  |> pull(pain_chg)
placebo_vals <- pain_data |> filter(arm == "Placebo") |> pull(pain_chg)

# CLES: P(Active < Placebo) for pain (lower = better)
u_stat <- mw_result$statistic
cles   <- as.numeric(u_stat) / (length(active_vals) * length(placebo_vals))

cat("Probability that Active patient has lower pain change than Placebo patient:\n")
cat("  CLES =", round(cles, 3), "\n")
cat("  Interpretation: In", round(cles * 100, 1),
    "% of Active–Placebo pairs, the Active patient had more pain reduction\n")
```

```{r}
#| label: boxplot-comparison
#| fig-cap: "Pain score changes by treatment arm with median and IQR"

pain_data |>
  ggplot(aes(x = arm, y = pain_chg, fill = arm)) +
  geom_boxplot(alpha = 0.75, width = 0.5, outlier.shape = 21) +
  geom_jitter(width = 0.15, alpha = 0.5, size = 1.5) +
  scale_fill_manual(values = c("Active" = "#2C6E9E", "Placebo" = "#E07B39")) +
  annotate("text", x = 1.5, y = max(pain_data$pain_chg) * 0.95,
           label = paste0("Mann-Whitney p = ", round(mw_result$p.value, 3)),
           size = 3.5, fontface = "italic") +
  labs(
    x = "Treatment Arm",
    y = "Change in Pain Score",
    fill = NULL,
    title = "Pain Score Changes: Active vs. Placebo",
    subtitle = "Box shows median and IQR; points show individual values"
  ) +
  theme_bw(base_size = 12) +
  theme(legend.position = "none")
```

---

## 4. Kruskal–Wallis Test

### 4.1 Extension to Multiple Groups

The Kruskal–Wallis test extends the Mann–Whitney approach to three or more independent groups. It tests whether the population medians are equal.

$$H = \frac{12}{N(N+1)} \sum_{i=1}^{k} \frac{R_i^2}{n_i} - 3(N+1)$$

where $R_i$ is the sum of ranks in group $i$, $n_i$ is the group size, and $N$ is the total sample size. Under $H_0$, $H$ follows a $\chi^2$ distribution with $k-1$ df.

```{r}
#| label: kruskal-wallis

# Three-group comparison: dose-response study
set.seed(55)
dose_data <- tibble(
  subject_id = paste0("DPT-", sprintf("%03d", 1:75)),
  dose       = rep(c("Placebo", "Low Dose", "High Dose"), each = 25),
  response   = c(
    round(rbeta(25, 2.5, 2) * 10),        # Placebo
    round(rbeta(25, 2, 2.5) * 10),        # Low dose
    round(rbeta(25, 1.5, 3.5) * 10)       # High dose
  )
)

# Order factor
dose_data <- dose_data |>
  mutate(dose = factor(dose, levels = c("Placebo", "Low Dose", "High Dose")))

kw_result <- kruskal.test(response ~ dose, data = dose_data)

cat("Kruskal-Wallis Test: Response by Dose Group\n")
cat("  H =", round(kw_result$statistic, 3),
    ", df =", kw_result$parameter,
    ", p =", round(kw_result$p.value, 4), "\n\n")

dose_data |>
  group_by(dose) |>
  summarise(n = n(), median = median(response), IQR = IQR(response)) |>
  gt() |>
  tab_header(title = "Summary Statistics by Dose Group")
```

### 4.2 Post-Hoc Pairwise Comparisons (Dunn's Test)

```{r}
#| label: dunns-test

# Pairwise Wilcoxon tests with Bonferroni correction
pairwise_results <- pairwise.wilcox.test(
  x       = dose_data$response,
  g       = dose_data$dose,
  p.adjust.method = "bonferroni"
)

cat("Pairwise comparisons (Wilcoxon, Bonferroni-adjusted):\n")
print(round(pairwise_results$p.value, 4))
```

```{r}
#| label: dose-response-plot
#| fig-cap: "Dose-response: response scores by dose group with Kruskal-Wallis p-value"

dose_data |>
  ggplot(aes(x = dose, y = response, fill = dose)) +
  geom_violin(alpha = 0.6, trim = FALSE) +
  geom_boxplot(width = 0.15, fill = "white", alpha = 0.9) +
  scale_fill_manual(values = c(
    "Placebo"   = "#757575",
    "Low Dose"  = "#2C6E9E",
    "High Dose" = "#1A3E5C"
  )) +
  annotate("text", x = 2, y = max(dose_data$response),
           label = paste0("KW p = ", round(kw_result$p.value, 4)),
           size = 3.5, fontface = "italic") +
  labs(
    x = "Dose Group",
    y = "Response Score",
    fill = NULL,
    title = "Dose-Response: Violin + Box Plot",
    subtitle = "Kruskal-Wallis test across three groups"
  ) +
  theme_bw(base_size = 12) +
  theme(legend.position = "none")
```

---

## 5. Friedman Test: Repeated Measures

The Friedman test is the non-parametric equivalent of repeated-measures ANOVA. It is used when the same subjects are measured under $k$ conditions.

```{r}
#| label: friedman-test

# Repeated measures: pain scores at 3 timepoints (within subject)
set.seed(77)
n_subj <- 30
long_pain <- tibble(
  subject_id = rep(paste0("PT-", sprintf("%03d", 1:n_subj)), times = 3),
  timepoint  = rep(c("Baseline", "Week 4", "Week 8"), each = n_subj),
  pain_score = c(
    round(rbeta(n_subj, 3, 1.5) * 10),         # Baseline
    round(rbeta(n_subj, 2.5, 2) * 10),          # Week 4
    round(rbeta(n_subj, 1.5, 3.5) * 10)         # Week 8
  )
) |>
  mutate(timepoint = factor(timepoint,
                            levels = c("Baseline", "Week 4", "Week 8")))

# Wide format for Friedman
wide_pain <- long_pain |>
  pivot_wider(names_from = timepoint, values_from = pain_score)

friedman_result <- friedman.test(
  y = as.matrix(wide_pain[, c("Baseline", "Week 4", "Week 8")])
)

cat("Friedman Test: Pain Scores at 3 Timepoints\n")
cat("  χ² =", round(friedman_result$statistic, 3),
    ", df =", friedman_result$parameter,
    ", p =", round(friedman_result$p.value, 4), "\n")

# Descriptive statistics
long_pain |>
  group_by(timepoint) |>
  summarise(
    n      = n(),
    Median = median(pain_score),
    IQR    = IQR(pain_score),
    Mean   = round(mean(pain_score), 1)
  ) |>
  gt() |>
  tab_header(title = "Pain Scores by Timepoint")
```

---

## 6. Spearman Rank Correlation

Spearman's $\rho$ measures the strength of a monotonic (not necessarily linear) relationship between two variables. It is the correlation of the ranked values:

$$r_s = 1 - \frac{6 \sum d_i^2}{n(n^2 - 1)}$$

where $d_i$ is the difference in ranks for observation $i$.

```{r}
#| label: spearman-correlation
#| fig-cap: "Spearman vs. Pearson correlation: Spearman captures monotonic but non-linear relationships"

set.seed(101)
n_corr <- 60
cor_data <- tibble(
  age         = runif(n_corr, 40, 80),
  pain_score  = 2 + 0.05 * age + rnorm(n_corr, 0, 1.5),   # monotonic but noisy
  crp         = exp(0.03 * age + rnorm(n_corr, 0, 0.5))     # non-linear (log-normal)
)

r_pearson  <- cor(cor_data$age, cor_data$crp, method = "pearson")
r_spearman <- cor(cor_data$age, cor_data$crp, method = "spearman")
cor_test   <- cor.test(cor_data$age, cor_data$crp, method = "spearman")

cat("Pearson r:   ", round(r_pearson,  3), "\n")
cat("Spearman rho:", round(r_spearman, 3),
    " (p =", round(cor_test$p.value, 4), ")\n")

cor_data |>
  ggplot(aes(x = age, y = crp)) +
  geom_point(colour = "#2C6E9E", alpha = 0.7, size = 2) +
  geom_smooth(method = "lm", colour = "red",   linetype = "dashed",
              se = FALSE, linewidth = 0.9) +
  geom_smooth(method = "loess", colour = "#E07B39",
              se = FALSE, linewidth = 0.9) +
  annotate("text", x = 42, y = max(cor_data$crp) * 0.95,
           label = paste0("Pearson r = ", round(r_pearson, 2),
                          "\nSpearman ρ = ", round(r_spearman, 2)),
           hjust = 0, size = 3.5) +
  labs(
    x = "Age (years)",
    y = "CRP (mg/L)",
    title = "Age vs. CRP: Spearman Captures the Monotonic Relationship",
    subtitle = "Red dashed = linear fit; orange = LOESS"
  ) +
  theme_bw(base_size = 12)
```

---

## 7. Permutation Tests

Permutation tests provide exact non-parametric p-values by computing the test statistic under all (or a random sample of) permutations of group labels.

```{r}
#| label: permutation-test
#| fig-cap: "Permutation distribution of mean difference; observed value shown in red"

set.seed(123)
obs_diff   <- mean(active_vals) - mean(placebo_vals)
all_values <- c(active_vals, placebo_vals)
n1         <- length(active_vals)
n_perm     <- 5000

perm_diffs <- replicate(n_perm, {
  shuffled <- sample(all_values)
  mean(shuffled[1:n1]) - mean(shuffled[(n1 + 1):length(all_values)])
})

perm_p <- mean(abs(perm_diffs) >= abs(obs_diff))

cat("Observed mean difference:", round(obs_diff, 3), "\n")
cat("Permutation p-value:      ", round(perm_p, 4), "\n")
cat("Mann-Whitney p-value:     ", round(mw_result$p.value, 4), "\n")

tibble(perm_diff = perm_diffs) |>
  ggplot(aes(x = perm_diff)) +
  geom_histogram(bins = 50, fill = "#757575", colour = "white", alpha = 0.8) +
  geom_vline(xintercept =  obs_diff, colour = "red",
             linewidth = 1.1, linetype = "dashed") +
  geom_vline(xintercept = -obs_diff, colour = "red",
             linewidth = 1.1, linetype = "dashed") +
  annotate("text", x = obs_diff + 0.2, y = 400,
           label = paste0("Observed\n", round(obs_diff, 2)),
           colour = "red", hjust = 0, size = 3.5) +
  labs(
    x = "Permutation Mean Difference",
    y = "Count",
    title = "Permutation Distribution of Mean Difference",
    subtitle = paste0(n_perm, " permutations; p = ", round(perm_p, 4))
  ) +
  theme_bw(base_size = 12)
```

---

## 8. Choosing Between Parametric and Non-Parametric Tests

```{r}
#| label: test-comparison-table

tibble(
  Situation = c(
    "One-sample, test median",
    "Paired samples",
    "Two independent groups",
    "3+ independent groups",
    "Repeated measures",
    "Correlation (monotonic)",
    "Correlation (linear)"
  ),
  Parametric = c(
    "One-sample t-test",
    "Paired t-test",
    "Independent t-test",
    "One-way ANOVA",
    "Repeated-measures ANOVA",
    "Pearson r",
    "Pearson r"
  ),
  Non_Parametric = c(
    "Wilcoxon signed-rank",
    "Wilcoxon signed-rank",
    "Mann–Whitney U",
    "Kruskal–Wallis",
    "Friedman test",
    "Spearman ρ",
    "Kendall τ"
  ),
  Use_NP_When = c(
    "n < 20, skewed, ordinal",
    "n < 20, outliers, ordinal",
    "n < 30, non-normal, ordinal",
    "n < 30 per group, non-normal",
    "Non-normal, ordinal outcomes",
    "Non-linear relationship",
    "Large outliers present"
  )
) |>
  gt() |>
  tab_header(
    title = "Parametric vs. Non-Parametric Test Selection Guide",
    subtitle = "Clinical research reference"
  ) |>
  cols_label(
    Non_Parametric = "Non-Parametric",
    Use_NP_When    = "Prefer NP When"
  )
```

---

## 9. Practical Example: Patient-Reported Outcome Comparison

```{r}
#| label: pro-example

# PRO data: EQ-5D utility score (0-1, ordinal-like, bounded)
set.seed(88)
pro_data <- tibble(
  subject_id = paste0("SUBJ-", sprintf("%03d", 1:100)),
  arm        = rep(c("Active", "Placebo"), each = 50),
  eq5d_base  = round(rbeta(100, 3, 2), 2),
  eq5d_w12   = c(
    pmin(1, round(rbeta(50, 4, 1.5) + 0.05, 2)),  # Active: improvement
    round(rbeta(50, 3, 2.5), 2)                    # Placebo: minimal change
  )
) |>
  mutate(eq5d_chg = eq5d_w12 - eq5d_base)

# Test
mw_pro <- wilcox.test(eq5d_chg ~ arm, data = pro_data, conf.int = TRUE)

cat("=== EQ-5D Utility Change: Active vs. Placebo ===\n\n")
pro_data |>
  group_by(arm) |>
  summarise(
    n      = n(),
    Median = round(median(eq5d_chg), 3),
    IQR    = round(IQR(eq5d_chg), 3),
    Mean   = round(mean(eq5d_chg), 3)
  ) |>
  print()

cat("\nMann-Whitney U test:\n")
cat("  W =", mw_pro$statistic, ", p =", round(mw_pro$p.value, 4), "\n")
cat("  HL estimate (median difference):", round(mw_pro$estimate, 3), "\n")
cat("  95% CI:", round(mw_pro$conf.int[1], 3), "to",
    round(mw_pro$conf.int[2], 3), "\n")
```

---

## Summary

Non-parametric tests are indispensable tools in the clinical researcher's toolkit:

- **Use them** when data are ordinal, heavily skewed, or when sample sizes are too small to verify normality
- **Wilcoxon signed-rank** replaces the paired t-test; **Mann–Whitney U** replaces the independent t-test
- **Kruskal–Wallis** handles three or more independent groups; follow with pairwise Wilcoxon post-hoc tests
- **Friedman test** handles repeated measures when normality fails
- **Spearman ρ** captures monotonic (non-linear) associations; appropriate for ordinal variables
- **Permutation tests** provide exact p-values with no distributional assumptions
- Non-parametric tests lose some power relative to parametric tests when normality holds — but in clinical research with small or ordinal data, the extra robustness is worthwhile

> **Regulatory note**: For PRO endpoints scored on ordinal scales (e.g., Likert scales, VAS), regulatory guidance often requires pre-specification of whether parametric or non-parametric analysis will be used in the SAP, with justification based on distributional assumptions.
