
# Regression Analysis in Clinical Trials {#sec-regression}

Regression models are among the most widely used statistical tools in clinical research. They allow us to quantify relationships between a response variable and one or more predictors, adjust for confounders, and make predictions. This chapter covers linear regression, logistic regression, and Poisson regression, all illustrated with the `pharmaversesdtm` clinical datasets.

---

## Setup and Data Loading

```{r setup}
library(dplyr)
library(ggplot2)
library(tidyr)
library(pharmaversesdtm)

data("dm")
data("ae")
data("vs")
data("lb")

# Clean ARM variable throughout
dm_clean <- dm |>
  filter(!is.na(ARM), ARM != "", ARM != "Screen Failure") |>
  mutate(ARM = factor(ARM,
                      levels = c("Placebo",
                                 "Xanomeline Low Dose",
                                 "Xanomeline High Dose")))

cat("Subjects in dm_clean:", nrow(dm_clean), "\n")
cat("ARM distribution:\n")
print(table(dm_clean$ARM))
```

---

## 1. Simple Linear Regression {#sec-simple-lm}

We predict systolic blood pressure (SBP) from subject age. SBP is stored in the `vs` dataset where `VSTESTCD == "SYSBP"`.

### 1.1 Preparing the SBP Dataset

```{r sbp-data}
library(dplyr)

# Extract baseline SBP values (VISIT == "BASELINE" or first available)
sbp_data <- vs |>
  filter(VSTESTCD == "SYSBP", !is.na(VSSTRESN)) |>
  group_by(USUBJID) |>
  slice_min(order_by = VISITNUM, n = 1, with_ties = FALSE) |>
  ungroup() |>
  select(USUBJID, SBP = VSSTRESN)

# Join to demographics
analysis_df <- dm_clean |>
  inner_join(sbp_data, by = "USUBJID") |>
  filter(!is.na(AGE), !is.na(SBP)) |>
  mutate(SEX = factor(SEX),
         RACE = factor(RACE))

cat("Analytic dataset rows:", nrow(analysis_df), "\n")
cat("SBP range:", range(analysis_df$SBP), "\n")
cat("AGE range:", range(analysis_df$AGE), "\n")
```

### 1.2 Fitting the Simple Linear Model

```{r simple-lm}
model_simple <- lm(SBP ~ AGE, data = analysis_df)
summary(model_simple)
```

**Interpretation:** The slope coefficient for AGE represents the expected change in SBP (mmHg) for each additional year of age. A positive coefficient indicates SBP tends to rise with age, consistent with known physiology.

### 1.3 Visualising the Relationship

```{r simple-lm-plot}
library(ggplot2)

ggplot(analysis_df, aes(x = AGE, y = SBP)) +
  geom_point(alpha = 0.5, colour = "#2166ac") +
  geom_smooth(method = "lm", se = TRUE, colour = "#d6604d") +
  labs(title = "Systolic Blood Pressure vs Age",
       subtitle = "Simple linear regression with 95% confidence band",
       x = "Age (years)",
       y = "Systolic Blood Pressure (mmHg)") +
  theme_bw(base_size = 13)
```

---

## 2. Multiple Linear Regression {#sec-multiple-lm}

Adding SEX, RACE, and treatment ARM as additional predictors provides an adjusted estimate of the age effect and allows us to test whether these variables are independently associated with SBP.

### 2.1 Fitting the Multiple Regression Model

```{r multiple-lm}
model_multi <- lm(SBP ~ AGE + SEX + RACE + ARM, data = analysis_df)
summary(model_multi)
```

### 2.2 Extracting Coefficients and Confidence Intervals

```{r multi-coef}
coef_tbl <- as.data.frame(confint(model_multi))
coef_tbl$Estimate <- coef(model_multi)
coef_tbl <- coef_tbl[, c("Estimate", "2.5 %", "97.5 %")]
names(coef_tbl) <- c("Estimate", "Lower_95", "Upper_95")
print(round(coef_tbl, 3))
```

### 2.3 Comparing Models with ANOVA

```{r anova-compare}
anova(model_simple, model_multi)
```

A significant F-test indicates that adding the extra predictors improves model fit beyond what chance would produce.

---

## 3. Regression Diagnostics {#sec-diagnostics}

Before drawing conclusions, we must verify that the four key assumptions of linear regression hold:

1. **Linearity** – the relationship is approximately linear.
2. **Independence** – residuals are not correlated.
3. **Homoscedasticity** – residuals have constant variance.
4. **Normality** – residuals are approximately normally distributed.

### 3.1 Residual Plots

```{r diagnostic-plots, fig.height=8, fig.width=9}
par(mfrow = c(2, 2))
plot(model_multi, which = 1:4)
par(mfrow = c(1, 1))
```

- **Residuals vs Fitted**: checks linearity and homoscedasticity — a random scatter around zero is ideal.
- **Normal Q-Q**: checks normality — points should lie close to the diagonal.
- **Scale-Location**: another view of homoscedasticity.
- **Cook's Distance / Residuals vs Leverage**: identifies influential observations.

### 3.2 Cook's Distance — Identifying Influential Points

```{r cooks}
library(dplyr)

cooks_d <- cooks.distance(model_multi)
influential_threshold <- 4 / nrow(analysis_df)

influential_df <- data.frame(
  USUBJID = analysis_df$USUBJID,
  CooksD   = cooks_d
) |>
  filter(CooksD > influential_threshold) |>
  arrange(desc(CooksD))

cat("Number of potentially influential observations:", nrow(influential_df), "\n")
if (nrow(influential_df) > 0) print(head(influential_df, 10))
```

### 3.3 Cook's Distance Plot

```{r cooks-plot}
library(ggplot2)

data.frame(
  Index  = seq_along(cooks_d),
  CooksD = cooks_d
) |>
  ggplot(aes(x = Index, y = CooksD)) +
  geom_segment(aes(xend = Index, yend = 0), colour = "grey60") +
  geom_point(aes(colour = CooksD > influential_threshold), size = 2) +
  scale_colour_manual(values = c("FALSE" = "#2166ac", "TRUE" = "#d6604d"),
                      name = "Influential") +
  geom_hline(yintercept = influential_threshold, linetype = "dashed",
             colour = "#d6604d") +
  labs(title = "Cook's Distance — Multiple Regression Model",
       x = "Observation Index",
       y = "Cook's Distance") +
  theme_bw(base_size = 13)
```

---

## 4. Model Selection {#sec-model-selection}

### 4.1 AIC Comparison

AIC (Akaike Information Criterion) penalises model complexity. Lower AIC indicates a better balance between fit and parsimony.

```{r aic}
models <- list(
  "Age only"       = lm(SBP ~ AGE, data = analysis_df),
  "Age + Sex"      = lm(SBP ~ AGE + SEX, data = analysis_df),
  "Age + Sex + Race" = lm(SBP ~ AGE + SEX + RACE, data = analysis_df),
  "Full model"     = model_multi
)

aic_tbl <- data.frame(
  Model = names(models),
  AIC   = sapply(models, AIC),
  BIC   = sapply(models, BIC),
  R2    = sapply(models, \(m) summary(m)$r.squared)
)
aic_tbl <- aic_tbl[order(aic_tbl$AIC), ]
aic_tbl[, -1] <- round(aic_tbl[, -1], 3)
print(aic_tbl)
```

### 4.2 Stepwise Variable Selection

```{r stepwise}
# Start from full model; step() uses AIC by default
full_model <- lm(SBP ~ AGE + SEX + RACE + ARM, data = analysis_df)
step_model  <- step(full_model, direction = "both", trace = 0)
cat("Selected model formula:\n")
print(formula(step_model))
summary(step_model)$coefficients |> round(4)
```

---

## 5. Polynomial Regression — Quadratic Age Effect {#sec-polynomial}

A quadratic term allows for a curved (non-linear) relationship between age and SBP.

```{r poly-lm}
model_poly <- lm(SBP ~ poly(AGE, 2, raw = TRUE), data = analysis_df)
summary(model_poly)
```

### 5.1 Comparing Linear vs Quadratic Fits

```{r poly-compare}
anova(model_simple, model_poly)
```

```{r poly-plot}
library(ggplot2)

age_seq <- data.frame(AGE = seq(min(analysis_df$AGE),
                                 max(analysis_df$AGE),
                                 length.out = 100))

age_seq$Linear    <- predict(model_simple, newdata = age_seq)
age_seq$Quadratic <- predict(model_poly,   newdata = age_seq)

pred_long <- tidyr::pivot_longer(age_seq, cols = c("Linear", "Quadratic"),
                                  names_to = "Model", values_to = "SBP")

ggplot() +
  geom_point(data = analysis_df, aes(x = AGE, y = SBP),
             alpha = 0.35, colour = "grey50") +
  geom_line(data = pred_long, aes(x = AGE, y = SBP, colour = Model),
            linewidth = 1.2) +
  scale_colour_manual(values = c("Linear" = "#2166ac",
                                  "Quadratic" = "#d6604d")) +
  labs(title = "Linear vs Quadratic Age Effect on SBP",
       x = "Age (years)", y = "Systolic BP (mmHg)") +
  theme_bw(base_size = 13)
```

---

## 6. Logistic Regression — Predicting AE Occurrence {#sec-logistic}

Logistic regression models the probability of a binary outcome. Here, we ask: which baseline characteristics predict whether a subject experienced any adverse event?

### 6.1 Creating the Binary Outcome

```{r ae-binary}
library(dplyr)

# Subjects with at least one AE
ae_flag <- ae |>
  filter(!is.na(USUBJID)) |>
  distinct(USUBJID) |>
  mutate(ANY_AE = 1L)

# Join to analysis dataset
logit_df <- dm_clean |>
  left_join(ae_flag, by = "USUBJID") |>
  mutate(ANY_AE = if_else(is.na(ANY_AE), 0L, ANY_AE)) |>
  filter(!is.na(AGE)) |>
  mutate(SEX  = factor(SEX),
         RACE = factor(RACE),
         ANY_AE = factor(ANY_AE, levels = c(0, 1),
                         labels = c("No AE", "AE")))

cat("AE occurrence:\n")
print(table(logit_df$ANY_AE, logit_df$ARM))
```

### 6.2 Fitting the Logistic Regression Model

```{r logit-fit}
model_logit <- glm(ANY_AE ~ AGE + SEX + ARM,
                   data   = logit_df,
                   family = binomial(link = "logit"))
summary(model_logit)
```

### 6.3 Odds Ratios and Confidence Intervals

Exponentiate the log-odds coefficients to obtain odds ratios (OR). An OR > 1 means increased odds of AE occurrence.

```{r odds-ratios}
or_tbl <- data.frame(
  OR       = exp(coef(model_logit)),
  Lower_95 = exp(confint.default(model_logit)[, 1]),
  Upper_95 = exp(confint.default(model_logit)[, 2])
)
print(round(or_tbl, 3))
```

### 6.4 Forest Plot of Odds Ratios

```{r or-forest-plot}
library(ggplot2)

or_plot_df <- or_tbl[-1, ]   # drop intercept
or_plot_df$Term <- rownames(or_plot_df)

ggplot(or_plot_df, aes(x = OR, y = reorder(Term, OR))) +
  geom_vline(xintercept = 1, linetype = "dashed", colour = "grey50") +
  geom_errorbarh(aes(xmin = Lower_95, xmax = Upper_95),
                 height = 0.25, colour = "#2166ac") +
  geom_point(size = 3, colour = "#d6604d") +
  labs(title = "Odds Ratios for AE Occurrence",
       x = "Odds Ratio (95% CI)",
       y = NULL) +
  theme_bw(base_size = 13)
```

---

## 7. Model Evaluation {#sec-evaluation}

### 7.1 Confusion Matrix

We use a 0.5 probability threshold to classify predictions.

```{r confusion-matrix}
pred_probs <- predict(model_logit, newdata = logit_df, type = "response")
pred_class <- factor(if_else(pred_probs >= 0.5, "AE", "No AE"),
                     levels = c("No AE", "AE"))
actual     <- factor(as.character(logit_df$ANY_AE), levels = c("No AE", "AE"))

conf_mat <- table(
  Predicted = pred_class,
  Actual    = actual
)
print(conf_mat)

# Derived metrics (with safe fallbacks for zero cells)
tp <- conf_mat["AE", "AE"]
tn <- conf_mat["No AE", "No AE"]
fp <- conf_mat["AE", "No AE"]
fn <- conf_mat["No AE", "AE"]

sensitivity <- if ((tp + fn) > 0) tp / (tp + fn) else NA_real_
specificity  <- if ((tn + fp) > 0) tn / (tn + fp) else NA_real_
accuracy     <- (tp + tn) / sum(conf_mat)

cat(sprintf("\nSensitivity : %.3f\n", sensitivity))
cat(sprintf("Specificity : %.3f\n", specificity))
cat(sprintf("Accuracy    : %.3f\n", accuracy))
```

### 7.2 ROC Curve Concept

The ROC curve plots sensitivity vs (1 - specificity) across all classification thresholds. Here we build it manually without additional packages.

```{r roc-manual}
library(dplyr)
library(ggplot2)

roc_df <- data.frame(
  prob   = pred_probs,
  actual = logit_df$ANY_AE == "AE"
) |>
  arrange(desc(prob))

thresholds <- sort(unique(roc_df$prob), decreasing = TRUE)

roc_points <- lapply(thresholds, function(thr) {
  pred_pos <- roc_df$prob >= thr
  tp_i  <- sum( pred_pos &  roc_df$actual)
  fp_i  <- sum( pred_pos & !roc_df$actual)
  fn_i  <- sum(!pred_pos &  roc_df$actual)
  tn_i  <- sum(!pred_pos & !roc_df$actual)
  data.frame(
    TPR = tp_i / (tp_i + fn_i),
    FPR = fp_i / (fp_i + tn_i)
  )
}) |>
  dplyr::bind_rows()

# Add (0,0) and (1,1)
roc_curve <- rbind(data.frame(TPR = 0, FPR = 0),
                    roc_points,
                    data.frame(TPR = 1, FPR = 1))

ggplot(roc_curve, aes(x = FPR, y = TPR)) +
  geom_line(colour = "#2166ac", linewidth = 1.1) +
  geom_abline(slope = 1, intercept = 0, linetype = "dashed",
              colour = "grey60") +
  labs(title = "ROC Curve — Logistic Regression (AE Occurrence)",
       x = "1 - Specificity (False Positive Rate)",
       y = "Sensitivity (True Positive Rate)") +
  coord_equal() +
  theme_bw(base_size = 13)
```

---

## 8. Poisson Regression — Count of AEs per Subject {#sec-poisson}

When the outcome is a count (e.g., number of AEs), Poisson regression is appropriate.

### 8.1 Building the Count Outcome

```{r ae-count}
library(dplyr)

ae_counts <- ae |>
  filter(!is.na(USUBJID)) |>
  group_by(USUBJID) |>
  summarise(N_AE = n(), .groups = "drop")

poisson_df <- dm_clean |>
  left_join(ae_counts, by = "USUBJID") |>
  mutate(N_AE = if_else(is.na(N_AE), 0L, as.integer(N_AE)),
         SEX  = factor(SEX),
         RACE = factor(RACE)) |>
  filter(!is.na(AGE))

cat("AE count summary:\n")
print(summary(poisson_df$N_AE))
```

### 8.2 Fitting the Poisson Model

```{r poisson-fit}
model_poisson <- glm(N_AE ~ AGE + SEX + ARM,
                     data   = poisson_df,
                     family = poisson(link = "log"))
summary(model_poisson)
```

### 8.3 Incidence Rate Ratios

Exponentiating Poisson coefficients gives Incidence Rate Ratios (IRR).

```{r irr}
irr_tbl <- data.frame(
  IRR      = exp(coef(model_poisson)),
  Lower_95 = exp(confint.default(model_poisson)[, 1]),
  Upper_95 = exp(confint.default(model_poisson)[, 2])
)
print(round(irr_tbl, 3))
```

### 8.4 Checking for Overdispersion

A key assumption of Poisson regression is that the mean equals the variance. We test for overdispersion by computing the ratio of residual deviance to degrees of freedom.

```{r overdispersion}
disp_ratio <- model_poisson$deviance / model_poisson$df.residual
cat(sprintf("Dispersion ratio: %.3f\n", disp_ratio))
cat("A ratio >> 1 suggests overdispersion; consider negative binomial model.\n")
```

### 8.5 AE Count Distribution by ARM

```{r ae-count-plot}
library(ggplot2)

ggplot(poisson_df, aes(x = ARM, y = N_AE, fill = ARM)) +
  geom_boxplot(alpha = 0.7, outlier.alpha = 0.4) +
  scale_fill_manual(values = c("Placebo"               = "#4393c3",
                                "Xanomeline Low Dose"   = "#f4a582",
                                "Xanomeline High Dose"  = "#d6604d")) +
  labs(title = "Distribution of AE Counts by Treatment Arm",
       x = NULL, y = "Number of Adverse Events") +
  theme_bw(base_size = 13) +
  theme(legend.position = "none",
        axis.text.x = element_text(angle = 20, hjust = 1))
```

---

## 9. Coefficient Interpretation Reference {#sec-interpretation}

| Model | Coefficient | Interpretation |
|---|---|---|
| Linear | `beta` | Change in mean outcome per 1-unit increase in predictor |
| Logistic | `exp(beta)` | Odds ratio: multiplicative change in odds of outcome |
| Poisson | `exp(beta)` | Incidence rate ratio: multiplicative change in rate |

```{r interp-summary}
# Collect all key estimates in one table
cat("=== Linear Regression: AGE coefficient ===\n")
cat(sprintf("  Beta = %.4f (per year of age, SBP in mmHg)\n",
            coef(model_multi)["AGE"]))

cat("\n=== Logistic Regression: AGE odds ratio ===\n")
cat(sprintf("  OR = %.4f (per year of age, odds of any AE)\n",
            exp(coef(model_logit)["AGE"])))

cat("\n=== Poisson Regression: AGE incidence rate ratio ===\n")
cat(sprintf("  IRR = %.4f (per year of age, AE count rate)\n",
            exp(coef(model_poisson)["AGE"])))
```

---

## Summary

This chapter demonstrated:

- **Simple and multiple linear regression** to explain SBP variation using age, sex, race and treatment arm.
- **Model diagnostics** including residual plots, Q-Q plots, and Cook's distance.
- **Model selection** via AIC and stepwise procedures.
- **Polynomial regression** to capture non-linear age effects.
- **Logistic regression** to predict binary AE occurrence, with odds ratios and a manually constructed ROC curve.
- **Poisson regression** to model count outcomes (AE frequency per subject) with incidence rate ratios.

---

## Exercises {#sec-exercises-bs05}

1. Fit a simple linear regression of diastolic blood pressure (`VSTESTCD == "DIABP"`) on AGE. Compare the slope to the SBP result.

2. Extend the multiple linear regression for SBP by including a RACE × SEX interaction term. Does the interaction improve model fit (use `anova()`)? Interpret any significant interaction.

3. Examine the Cook's distance plot for the multiple regression model. Remove the most influential observation and refit the model. How do the coefficients change?

4. Using `step()`, perform forward stepwise selection starting from an intercept-only model for SBP. Do you arrive at the same model as backward selection?

5. Fit a logistic regression predicting serious AEs (`AESER == "Y"` in the `ae` dataset) from age and ARM. Compute and interpret the odds ratio for the Xanomeline High Dose arm relative to Placebo.

6. The Poisson dispersion ratio you computed may exceed 1. Refit the AE-count model using `glm.nb()` from the `MASS` package (call it as `MASS::glm.nb()`). Compare AIC values to decide which model is preferable.

7. Add `WEIGHTBL` (baseline weight) as a covariate to the SBP multiple regression if that variable is available in `dm`. Interpret its coefficient.

8. Construct a calibration plot for the logistic regression model: bin subjects by predicted probability (e.g., deciles), compute the observed proportion of AEs in each bin, and plot observed vs predicted. Does the model appear well calibrated?
