# Advanced Biostatistical Methods in Drug Development

## Introduction

Modern clinical drug development relies on sophisticated statistical methods that go beyond traditional fixed-sample, single-endpoint trial designs. This chapter covers advanced methods essential for efficient, rigorous development programs:

- **Adaptive designs**: Flexible modification of trial characteristics based on accumulating data
- **Multiple testing**: Controlling error rates when testing many hypotheses
- **Estimand framework**: Clear specification of treatment effects for complex intercurrent events
- **Missing data**: Principled approaches to incomplete data beyond simple deletion
- **Meta-analysis**: Systematic synthesis of evidence across multiple studies
- **Biomarker-driven designs**: Enrichment and precision medicine approaches

These methods, when properly implemented, improve efficiency, reduce development time, and provide stronger evidence for regulatory decision-making.

---

## 1. Adaptive Clinical Trial Designs

### 1.1 Group Sequential Designs

A **group sequential design** allows for interim analyses, with the possibility of stopping early for efficacy, futility, or both.

#### 1.1.1 O'Brien-Fleming Boundary

More conservative early stopping for efficacy, allocates more Type I error to final analysis:

- At interim analysis 1: p-value threshold ≈ 0.005 (more stringent)
- At final analysis: p-value threshold ≈ 0.047 (less stringent)

Ensures strong early efficacy evidence before stopping, reduces false positives.


```{r}
# O'Brien-Fleming boundary for 2 interim analyses + 1 final
if (requireNamespace("gsDesign", quietly = TRUE)) {
  design_of <- gsDesign::gsDesign(k = 3,  # 3 analyses total
                                  test.type = 2,  # symmetric 2-sided (OF/Pocock require symmetric)
                                  sfu = "OF",  # O'Brien-Fleming
                                  alpha = 0.05,
                                  beta = 0.20)

  design_of$upper$bound  # Efficacy boundaries
  design_of$lower$bound  # Futility boundaries
} else {
  message("gsDesign not available; install with install.packages('gsDesign')")
}
```



#### 1.1.2 Pocock Boundary

More aggressive early stopping for efficacy, uniform p-value threshold across analyses:

- At each analysis: p-value threshold ≈ 0.016 (for 3 analyses)

Higher power to detect effect early, but slightly increased false positive risk.


```{r}
# Pocock boundary
if (requireNamespace("gsDesign", quietly = TRUE)) {
  design_poc <- gsDesign::gsDesign(k = 3,
                                  test.type = 2,  # symmetric 2-sided (Pocock requires symmetric)
                                  sfu = "Pocock",
                                  alpha = 0.05,
                                  beta = 0.20)

  design_poc$upper$bound
} else {
  message("gsDesign not available; install with install.packages('gsDesign')")
}
```



#### 1.1.3 Sample Size Inflation

Group sequential designs require **information inflation factor**: additional subjects beyond fixed-sample design to maintain power:


```{r}
# Information inflation factor
# Fixed sample size for 80% power, 2-sided test
if (requireNamespace("pwr", quietly = TRUE) &&
    requireNamespace("gsDesign", quietly = TRUE)) {
  n_fixed <- pwr::pwr.t.test(d = 0.5, sig.level = 0.05, power = 0.80)$n
  n_fixed <- ceiling(n_fixed)

  # With group sequential design (2 looks, O'Brien-Fleming)
  design_gs <- gsDesign::gsDesign(k = 2, test.type = 2,
                                 sfu = "OF", alpha = 0.05, beta = 0.20)
  inflation_factor <- design_gs$n.I[2] / n_fixed

  cat("Fixed sample size:", n_fixed, "\n")
  cat("GSD sample size:", design_gs$n.I[2], "\n")
  cat("Inflation factor:", round(inflation_factor, 3), "\n")
} else {
  message("pwr or gsDesign not available")
}
```



### 1.2 Adaptive Seamless Phase II/III Designs

Combines phase II and phase III in a single trial structure:

1. **Phase II component** (first stage): Identify promising dose(s)
2. **Seamless transition**: Drop futile doses, continue promising dose(s)
3. **Phase III component** (second stage): Confirmatory efficacy testing

Advantages:
- Faster development timeline
- Shared control group across phases
- More efficient use of subjects


```{r}
# Simplified example: two-stage adaptive dose selection
# Phase II: Three doses (5, 10, 20 mg) vs. placebo
# Phase III: Continue best-performing dose vs. placebo

# Phase II (hypothetical results)
dose_effects <- data.frame(
  Dose = c("Placebo", "5mg", "10mg", "20mg"),
  Mean_Change = c(-2, -6, -9, -8),  # Blood pressure change
  SE = c(1.5, 1.5, 1.5, 1.5),
  N = c(40, 40, 40, 40)
)

# Select dose with largest effect (10 mg)
best_dose <- dose_effects[which.max(dose_effects$Mean_Change[2:4]) + 1, ]
cat("Selected dose for Phase III:", best_dose$Dose, "\n")
```

```{r}
# Phase III: Compare selected dose to placebo with larger sample
# n = 150 per group (increased from 40 in Phase II)
```



### 1.3 Sample Size Re-estimation

Allows prospective increase in sample size based on interim analysis data:


```{r}
# Interim analysis at 50% information
# Blinded review of variability estimate
# If SD larger than anticipated, increase sample size

sd_anticipated <- 10
sd_observed_interim <- 12.5  # 25% larger

# Reestimate required sample size
n_original <- 100  # per group
variance_ratio <- (sd_observed_interim / sd_anticipated)^2
n_updated <- ceiling(n_original * variance_ratio)

cat("Original sample size per group:", n_original, "\n")
```

```{r}
cat("Updated sample size per group:", n_updated, "\n")
```

```{r}
cat("Additional subjects needed:", n_updated - n_original, "\n")
```



### 1.4 Response-Adaptive Randomization

Allocates more subjects to treatments showing promise during the trial:


```{r}
# Bandit algorithm: allocate subjects based on observed response rates
# Initially: 1:1 randomization
# As data accumulate: shift allocation toward higher-response treatment

set.seed(42)
n_total <- 200
response_rate_a <- 0.55
response_rate_b <- 0.45

# Track cumulative allocation
allocation <- numeric(n_total)
responses_a <- 0
responses_b <- 0

for (i in 1:n_total) {
  # Thompson sampling: sample from posterior distributions
  # Allocate to treatment with highest posterior probability of response

  posterior_a <- rbeta(1, responses_a + 1, (i/2 - responses_a) + 1)
  posterior_b <- rbeta(1, responses_b + 1, (i/2 - responses_b) + 1)

  if (isTRUE(posterior_a > posterior_b)) {
    allocation[i] <- 1  # Assign to A
    responses_a <- responses_a + rbinom(1, 1, response_rate_a)
  } else {
    allocation[i] <- 0  # Assign to B
    responses_b <- responses_b + rbinom(1, 1, response_rate_b)
  }
}

# Compare adaptive allocation to fixed 1:1
cat("Fixed 1:1 allocation: 100 per arm\n")
```

```{r}
cat("Adaptive allocation: ",
    sum(allocation == 1), "to A, ",
    sum(allocation == 0), "to B\n")
```



---

## 2. Multiple Testing in Clinical Trials

### 2.1 Familywise Error Rate (FWER)

**FWER** is the probability of making at least one Type I error across multiple tests:

$$FWER = P(\text{reject at least one } H_0 | \text{all } H_0 \text{ true})$$

Traditional methods control FWER at α = 0.05.

### 2.2 False Discovery Rate (FDR)

**FDR** is the proportion of false positives among all rejected hypotheses:

$$FDR = E\left[\frac{V}{R}\right]$$

where V = number of false positives, R = number of rejections.

FDR control is less stringent than FWER, useful for exploratory analyses.


```{r}
# FWER vs FDR control
# m = 10 hypotheses, true effect size for 2 of them

set.seed(42)
m <- 10
true_effects <- c(0.5, 0.5, rep(0, 8))  # True effects for first 2

# Generate p-values
p_values <- numeric(m)
for (i in 1:m) {
  if (true_effects[i] > 0) {
    # Power = 0.80 for true effects
    p_values[i] <- 1 - pnorm(abs(rnorm(1, mean = true_effects[i], sd = 1)))
  } else {
    p_values[i] <- runif(1)
  }
}

# Bonferroni (FWER control): Reject if p < α/m
bonferroni_threshold <- 0.05 / m
bonferroni_reject <- p_values < bonferroni_threshold

# FDR (Benjamini-Hochberg): Reject if p <= threshold(i)
fdr_threshold <- 0.05
sorted_p <- sort(p_values)
sorted_i <- 1:m
bh_threshold <- (sorted_i / m) * fdr_threshold
bh_reject <- sum(sorted_p <= bh_threshold)

cat("Bonferroni rejections:", sum(bonferroni_reject), "\n")
```

```{r}
cat("Benjamini-Hochberg rejections:", bh_reject, "\n")
```



### 2.3 Hierarchical Testing Procedures

#### 2.3.1 Fixed Sequence Procedure

Tests hypotheses in a pre-specified order, stops at first non-rejection:


```{r}
# Primary efficacy → Key secondary 1 → Key secondary 2
# H1: Primary endpoint (strong control of Type I error)
# H2: Key secondary 1 (only if H1 rejected)
# H3: Key secondary 2 (only if H2 rejected)

p_values_seq <- c(0.008, 0.025, 0.032)  # Primary, Key Sec 1, Key Sec 2
alpha_level <- 0.05

rejections_fixed <- numeric(3)
for (i in 1:3) {
  if (i == 1 || rejections_fixed[i-1] == 1) {
    rejections_fixed[i] <- as.numeric(p_values_seq[i] < alpha_level)
  } else {
    break
  }
}

cat("Fixed sequence rejections:\n")
```

```{r}
for (i in 1:3) {
  cat("H", i, ": ", ifelse(rejections_fixed[i] == 1, "Rejected", "Not rejected"), "\n", sep="")
}
```



#### 2.3.2 Fallback Procedure

More flexible: can test later hypotheses even if earlier ones fail, but with adjusted α:


```{r}
# Fallback procedure
# If H1 rejected: H2 tested at α = 0.05
# If H1 not rejected: H2 tested at α = 0.025 (reduced)
# If H2 rejected: H3 tested at remaining α

p_values_fb <- c(0.08, 0.02)  # Primary fails, secondary succeeds
alpha_h1 <- 0.05
alpha_h2_if_h1_rej <- 0.05
alpha_h2_if_h1_not_rej <- 0.025

h1_rejected <- p_values_fb[1] < alpha_h1
h2_threshold <- ifelse(h1_rejected, alpha_h2_if_h1_rej, alpha_h2_if_h1_not_rej)
h2_rejected <- p_values_fb[2] < h2_threshold

cat("H1 (Primary):", ifelse(h1_rejected, "Rejected", "Not rejected"), "\n")
```

```{r}
cat("H2 threshold: alpha =", h2_threshold, "\n")
```

```{r}
cat("H2 (Secondary):", ifelse(h2_rejected, "Rejected", "Not rejected"), "\n")
```



### 2.4 Gatekeeping Procedures for Complex Hierarchies

Manages hierarchical families of hypotheses with conditional testing based on parent hypothesis results:


```{r}
# Hierarchical structure:
# Primary: PK endpoint (Pharmacokinetics)
#   └─ Secondary 1a: Efficacy measure 1
#   └─ Secondary 1b: Efficacy measure 2
# Safety: Adverse events
#   └─ Secondary 2a: Specific AE rate
#   └─ Secondary 2b: Lab abnormality rate

p_values_gate <- list(
  Primary = 0.015,
  Secondary_1a = 0.032,
  Secondary_1b = 0.048,
  Safety = 0.001,
  Secondary_2a = 0.087,
  Secondary_2b = 0.034
)

# Primary family (PK) has alpha = 0.04 (before safety)
# Safety family has alpha = 0.01

alpha_pk <- 0.04
alpha_safety <- 0.01

# Primary hypothesis
pk_rejected <- p_values_gate$Primary < alpha_pk

# Secondary hypotheses in efficacy (gated by PK)
if (pk_rejected) {
  sec1a_rejected <- p_values_gate$Secondary_1a < alpha_pk/2
  sec1b_rejected <- p_values_gate$Secondary_1b < alpha_pk/2
} else {
  sec1a_rejected <- sec1b_rejected <- FALSE
}

# Safety hypotheses (separate gating)
safety_rejected <- p_values_gate$Safety < alpha_safety
if (safety_rejected) {
  sec2a_rejected <- p_values_gate$Secondary_2a < alpha_safety/2
  sec2b_rejected <- p_values_gate$Secondary_2b < alpha_safety/2
} else {
  sec2a_rejected <- sec2b_rejected <- FALSE
}

cat("PK Endpoint:", ifelse(pk_rejected, "Positive", "Negative"), "\n")
```

```{r}
cat("├─ Efficacy Measure 1:", ifelse(sec1a_rejected, "Positive", "Negative"), "\n")
```

```{r}
cat("├─ Efficacy Measure 2:", ifelse(sec1b_rejected, "Positive", "Negative"), "\n")
```

```{r}
cat("Safety Endpoint:", ifelse(safety_rejected, "Positive", "Negative"), "\n")
```

```{r}
cat("├─ Specific AE:", ifelse(sec2a_rejected, "Positive", "Negative"), "\n")
```

```{r}
cat("└─ Lab Abnormality:", ifelse(sec2b_rejected, "Positive", "Negative"), "\n")
```



### 2.5 Multiplicity Adjustment Methods

#### 2.5.1 Bonferroni

Divides α equally among m tests:

$$p\text{-value threshold} = \frac{\alpha}{m}$$

Conservative (FWER control), but powerful when few tests.

#### 2.5.2 Holm's Procedure

More powerful than Bonferroni; rejects ordered hypotheses at thresholds:
- H₁: p < α/m
- H₂: p < α/(m-1)
- H₃: p < α/(m-2), etc.

Maintains FWER control while improving power.

#### 2.5.3 Hochberg and Hommel

Even more powerful methods; use step-up approach:


```{r}
# Example: 5 hypotheses with p-values
p_values_mult <- c(0.001, 0.008, 0.035, 0.042, 0.105)
alpha <- 0.05
m <- length(p_values_mult)

# Bonferroni
bonferroni_sig <- p_values_mult < alpha / m

# Holm
sorted_p <- sort(p_values_mult)
holm_threshold <- alpha / (m:1)
holm_sig <- sorted_p < holm_threshold

# Benjamini-Hochberg (FDR)
sorted_i <- 1:m
bh_threshold <- (sorted_i / m) * alpha
bh_sig <- sorted_p < bh_threshold

results_table <- data.frame(
  Test = 1:m,
  P_value = p_values_mult,
  Bonferroni = bonferroni_sig,
  Holm = cumsum(holm_sig) > 0,
  BH_FDR = cumsum(bh_sig) > 0
)
print(results_table)
```



---

## 3. Estimand Framework (ICH E9(R1))

### 3.1 Definition and Components

An **estimand** specifies:
1. **Population**: Target population
2. **Variable**: What is measured (primary outcome)
3. **Intercurrent events**: Events that occur during trial (drop-out, medication switches, disease progression)
4. **Population-level summary**: How individual observations are combined (mean difference, probability difference)

### 3.2 Handling Intercurrent Events

Five strategies for handling intercurrent events:

#### 3.2.1 Treatment Policy Strategy
Ignores intercurrent event, analyzes all subjects as randomized (intention-to-treat).


```{r}
# Analyze all randomized subjects regardless of adherence/dropout
# This is the standard ITT approach
```



#### 3.2.2 Composite Strategy
Defines intercurrent event as part of outcome (e.g., failure if subject stops treatment).


```{r}
# Failure = either lack of efficacy OR treatment discontinuation
# More conservative, clinically relevant approach
```



#### 3.2.3 Hypothetical Strategy
Estimates what would have happened if intercurrent event had not occurred.


```{r}
# Estimate efficacy adjusting for non-adherence
# More ambitious, requires strong assumptions (MCAR)
```



#### 3.2.4 While-on-Treatment Strategy
Analyzes only the period while subject is on treatment.


```{r}
# Exclude post-discontinuation data
# Can introduce bias if discontinuation related to outcome
```



#### 3.2.5 Principal Stratum Strategy
Conditions on whether subject would experience intercurrent event.

---

## 4. Missing Data Methods

### 4.1 Missing Data Mechanisms

- **MCAR (Missing Completely at Random)**: Missingness unrelated to observed or unobserved outcomes
- **MAR (Missing at Random)**: Missingness depends on observed data but not unobserved outcomes
- **MNAR (Missing Not at Random)**: Missingness depends on unobserved outcomes

Only MCAR justifies listwise deletion. MAR and MNAR require more sophisticated methods.

### 4.2 Multiple Imputation (MICE Framework)

Multiple Imputation by Chained Equations:

1. **Imputation**: Create m complete datasets by imputing missing values
2. **Analysis**: Analyze each complete dataset separately
3. **Pooling**: Combine results using Rubin's rules


```{r}
if (requireNamespace("mice", quietly = TRUE)) {
  # Create example dataset with missing values
  set.seed(42)
  trial_data_incomplete <- data.frame(
    subject_id = 1:100,
    treatment = rep(c("Active", "Control"), 50),
    age = rnorm(100, 55, 10),
    baseline_outcome = rnorm(100, 50, 10),
    final_outcome = rnorm(100, 55, 12)
  )

  # Introduce missing data (MCAR for simplicity)
  missing_idx <- sample(1:100, 20)
  trial_data_incomplete$final_outcome[missing_idx] <- NA

  # Multiple imputation (m = 5 imputations)
  imputed_data <- mice::mice(trial_data_incomplete,
                             m = 5, maxit = 50,
                             method = "norm", print = FALSE)

  # Analyze and pool: use base-R with() for S3 dispatch on mids objects
  fits_mids <- with(imputed_data,
                    lm(final_outcome ~ treatment + baseline_outcome))
  pooled_results <- mice::pool(fits_mids)
  print(summary(pooled_results))
} else {
  message("mice not available; install with install.packages('mice')")
}
```



**Advantages**:
- Handles MAR data properly
- Preserves uncertainty from missingness
- Flexible for different imputation models

**Disadvantages**:
- Assumes MAR (not testable)
- Requires specification of imputation model

### 4.3 Pattern Mixture Models

Explicitly model the relationship between missingness and outcome:

$$E[Y | X, \text{dropout}] = \alpha_{\text{dropout}} + \beta X$$

Coefficients can differ between subjects who drop out vs. complete follow-up.


```{r}
#| label: pattern-mixture-model
# Pattern mixture model: complete vs. dropout patterns
# Create a self-contained dataset for this demonstration
set.seed(42)
trial_data_pmm <- data.frame(
  subject_id       = 1:100,
  treatment        = rep(c("Active", "Control"), 50),
  baseline_outcome = rnorm(100, 50, 10),
  final_outcome    = rnorm(100, 55, 12)
)
# Introduce missingness
missing_idx2 <- sample(1:100, 20)
trial_data_pmm$final_outcome[missing_idx2] <- NA
trial_data_pmm$pattern <- ifelse(is.na(trial_data_pmm$final_outcome),
                                  "Dropout", "Complete")

complete_data <- subset(trial_data_pmm, pattern == "Complete")
dropout_data  <- subset(trial_data_pmm, pattern == "Dropout")

cat("Complete follow-up:", nrow(complete_data), "subjects
")
cat("Early dropout:     ", nrow(dropout_data),  "subjects
")

# Analyze observed data in completers
fit_complete <- lm(final_outcome ~ treatment + baseline_outcome,
                   data = complete_data)
cat("
Pattern mixture model (completers):
")
print(coef(summary(fit_complete)))
```



### 4.4 Sensitivity and Tipping Point Analysis

**Sensitivity analysis** explores robustness to different assumptions about missingness:


```{r}
# Tipping point analysis: at what assumed outcome for Active dropouts
# does the treatment conclusion flip to zero?
# Key: vary Active dropout assumption; keep Control dropout = observed mean (50)

n_completers <- 80; n_missing <- 20; n_total <- 100
obs_active <- 55; obs_control <- 50

# Range must cross the tipping point: when Active imputed = 30, effect = 0
# (55*80 + 30*20)/100 - 50 = 50 - 50 = 0
missing_outcome_active <- seq(28, 57, by = 1)

treatment_effect <- sapply(missing_outcome_active, function(x) {
  mean_active  <- (obs_active  * n_completers + x  * n_missing) / n_total
  mean_control <- (obs_control * n_completers + obs_control * n_missing) / n_total
  mean_active - mean_control
})

# Find tipping point (treatment effect = 0) via linear interpolation
tipping_point_result <- tryCatch(
  approx(treatment_effect, missing_outcome_active, xout = 0)$y,
  error = function(e) NA
)

cat("Observed treatment effect:", obs_active - obs_control, "\n")
```

```{r}
if (!is.na(tipping_point_result)) {
  cat("Tipping point (Active dropout mean that nullifies effect):",
      round(tipping_point_result, 2), "\n")
  cat("(Effect flips sign if Active dropouts score below ~",
      round(tipping_point_result, 1), "on average)\n")
} else {
  cat("Tipping point: treatment effect does not cross zero in specified range\n")
}
```



---

## 5. Meta-Analysis

### 5.1 Fixed-Effects Model

Assumes all studies estimate the same true effect:

$$\theta_i = \theta + \epsilon_i$$

where $\epsilon_i$ is within-study variability.

Pooled estimate (inverse variance weighting):

$$\hat{\theta} = \frac{\sum w_i \hat{\theta}_i}{\sum w_i}$$

where $w_i = 1 / SE_i^2$ (inverse variance weight).


```{r}
if (!requireNamespace("meta", quietly = TRUE)) {
  message("meta package not available; skipping meta-analysis examples.")
} else {

# Meta-analysis of antihypertensive drug efficacy
# 5 studies comparing drug to placebo

studies <- data.frame(
  Study = c("Study A", "Study B", "Study C", "Study D", "Study E"),
  Mean_Diff = c(-10, -12, -8, -11, -9),  # mmHg reduction
  SE = c(2.0, 2.2, 1.8, 2.1, 1.9)
)

# Meta-analysis (fixed-effects)
meta_fe <- meta::metagen(TE = Mean_Diff,
                        seTE = SE,
                        studlab = Study,
                        data = studies,
                        fixed = TRUE,
                        random = FALSE)

summary(meta_fe)
} # end requireNamespace("meta") guard
```



### 5.2 Random-Effects Model

Assumes studies estimate different true effects, drawn from distribution:

$$\theta_i = \mu + \eta_i + \epsilon_i$$

where:
- μ = mean effect across studies
- η_i = between-study heterogeneity
- ε_i = within-study error

DerSimonian-Laird estimator for between-study variance:

$$\hat{\tau}^2 = \frac{Q - (k-1)}{\sum w_i - \frac{\sum w_i^2}{\sum w_i}}$$

where Q = heterogeneity statistic.


```{r}
if (requireNamespace("meta", quietly = TRUE) && exists("studies")) {
  # Meta-analysis (random-effects)
  meta_re <- meta::metagen(TE = Mean_Diff,
                           seTE = SE,
                           studlab = Study,
                           data = studies,
                           fixed = FALSE,
                           random = TRUE,
                           method.tau = "DL")  # DerSimonian-Laird
  summary(meta_re)
} else {
  message("meta package not available or studies dataset not found")
}
```



### 5.3 Heterogeneity Assessment

#### 5.3.1 I² Statistic

Proportion of total variance due to heterogeneity:

$$I^2 = \frac{Q - (k-1)}{Q} \times 100\%$$

- I² < 25%: Low heterogeneity
- 25% < I² < 75%: Moderate heterogeneity
- I² > 75%: High heterogeneity

#### 5.3.2 τ² (Tau-Squared)

Between-study variance in original scale:


```{r}
# Extract heterogeneity measures
if (requireNamespace("meta", quietly = TRUE) && exists("meta_re")) {
  # Note: requires meta_re object from previous random-effects meta-analysis chunk
  i2_value <- tryCatch(meta_re$I2$TE, error = function(e) NA)
  tau2_value <- tryCatch(meta_re$tau2, error = function(e) NA)
  tau_value <- tryCatch(meta_re$tau, error = function(e) NA)

  if (!is.na(i2_value)) {
    cat("I²:", round(i2_value, 1), "%\n")
  }
  if (!is.na(tau2_value)) {
    cat("τ²:", round(tau2_value, 2), "\n")
  }
  if (!is.na(tau_value)) {
    cat("τ:", round(tau_value, 2), "\n")
  }
  if (is.na(i2_value) && is.na(tau2_value) && is.na(tau_value)) {
    message("Could not extract heterogeneity measures from meta_re object")
  }
} else {
  message("meta package not available or meta_re object not found")
}
```



### 5.4 Forest Plots

Graphical representation of individual studies and pooled estimate:


```{r}
# Forest plot
if (requireNamespace("meta", quietly = TRUE) && exists("meta_re")) {
  plot(meta_re, xlim = c(-20, 0))
} else {
  message("meta package or meta_re object not found; run random-effects meta-analysis chunk first")
}
```



### 5.5 Publication Bias

#### 5.5.1 Funnel Plot

Scatterplot of effect size vs. study precision; asymmetry suggests publication bias:


```{r}
# Funnel plot
if (requireNamespace("metafor", quietly = TRUE) && requireNamespace("meta", quietly = TRUE) && exists("meta_re")) {
  tryCatch(
    metafor::funnel(meta_re, main = "Funnel Plot"),
    error = function(e) message("Could not create funnel plot: ", e$message)
  )
} else {
  message("metafor/meta package or meta_re object not found")
}
```



#### 5.5.2 Egger's Test

Tests for asymmetry in funnel plot:

$$z_i = \frac{\hat{\theta}_i}{SE_i} = \alpha + \beta \frac{1}{SE_i}$$

Significant intercept (α) suggests publication bias.


```{r}
# Egger's test for publication bias
if (requireNamespace("meta", quietly = TRUE) && exists("meta_re")) {
  tryCatch(
    {
      egger_result <- meta::metabias(meta_re, method = "Egger")
      print(egger_result)
    },
    error = function(e) message("Could not compute Egger's test: ", e$message)
  )
} else {
  message("meta package or meta_re object not found")
}
```



---

## 6. Biomarker Analysis and Enrichment

### 6.1 Predictive vs. Prognostic Biomarkers

- **Predictive**: Modifies treatment response (predicts who benefits)
- **Prognostic**: Associated with outcome regardless of treatment


```{r}
# Analyze biomarker as predictor of treatment response
set.seed(42)
biomarker_trial <- data.frame(
  subject_id = 1:200,
  treatment = rep(c("Active", "Control"), 100),
  biomarker = rnorm(200, 0, 1),
  response = rbinom(200, 1, 0.5)
)

# Logistic regression: test biomarker × treatment interaction (predictive)
fit_predictive <- glm(response ~ treatment + biomarker + treatment:biomarker,
                     family = binomial,
                     data = biomarker_trial)
summary(fit_predictive)
```

```{r}
# If treatment:biomarker significant, biomarker is predictive
# If only biomarker significant (no interaction), biomarker is prognostic
```



### 6.2 Biomarker-Stratified Design

Enrolls only patients with specific biomarker status:


```{r}
# Enrichment design: enroll only biomarker-positive patients
enriched_trial <- biomarker_trial[biomarker_trial$biomarker > 0, ]

# Compare treatment effect in enriched population
fit_enriched <- glm(response ~ treatment,
                   family = binomial,
                   data = enriched_trial)
summary(fit_enriched)
```

```{r}
# Typically larger treatment effect in enriched population
```



### 6.3 Adaptive Biomarker Thresholding

Optimize biomarker threshold using interim data:


```{r}
# Interim analysis: optimize biomarker cutoff
interim_data <- biomarker_trial[1:100, ]

# Test different cutoffs
cutoffs <- seq(-2, 2, by = 0.5)
treatment_effects <- numeric(length(cutoffs))

for (i in seq_along(cutoffs)) {
  stratified_data <- interim_data[interim_data$biomarker > cutoffs[i], ]
  if (nrow(stratified_data) > 10) {
    fit_temp <- glm(response ~ treatment,
                   family = binomial,
                   data = stratified_data)
    treatment_effects[i] <- coef(fit_temp)["treatmentControl"]
  } else {
    treatment_effects[i] <- NA
  }
}

# Select cutoff with largest treatment effect
optimal_cutoff <- cutoffs[which.max(abs(treatment_effects))]
cat("Optimal biomarker cutoff:", round(optimal_cutoff, 2), "\n")
```



---

## 7. R Implementations

### 7.1 Adaptive Designs


```{r}
# gsDesign: group sequential designs
if (requireNamespace("gsDesign", quietly = TRUE)) {
  design <- gsDesign::gsDesign(k = 2, test.type = 2, sfu = "OF",
                    alpha = 0.05, beta = 0.20)

  # Sequential spending function (test.type = 2 required for Pocock/OF boundaries)
  gsDesign::gsDesign(k = 4, test.type = 2, sfu = "Pocock",
          alpha = 0.05, beta = 0.20)
} else {
  message("gsDesign not available; install with install.packages('gsDesign')")
}
```

```{r}
# rpact: adaptive designs (guard — may not be installed)
if (requireNamespace("rpact", quietly = TRUE)) {
  design_rpact <- rpact::getDesignGroupSequential(kMax = 3,
                                                  typeOfDesign = "OF",
                                                  alpha = 0.05,
                                                  beta = 0.20)
  print(design_rpact)
} else {
  message("rpact not available; install with install.packages('rpact')")
}
```



### 7.2 Multiple Testing


```{r}
# Multiplicity adjustment (example with p_values from earlier sections)
p.adjust(p_values, method = "bonferroni")
p.adjust(p_values, method = "holm")
p.adjust(p_values, method = "BH")  # Benjamini-Hochberg

# Hierarchical testing
# Manual implementation or use gatekeeping() functions
```



### 7.3 Missing Data


```{r}
#| eval: false
# Reference: multiple imputation workflow (requires mice, Hmisc, missRanger)
require(mice)
imputed <- mice(data_with_missing, m = 5)
fits <- with(imputed, lm(outcome ~ predictor))
pooled <- pool(fits)

require(Hmisc)
Hmisc::aregImpute()  # Imputation by additive regression, bootstrapping

require(missRanger)
missRanger::missRanger()  # Random forest imputation
```



### 7.4 Meta-Analysis


```{r}
#| eval: false
# Reference: meta-analysis workflow (requires meta, metafor, netmeta)
require(meta)
result <- metagen(TE = effect, seTE = se_effect,
                 studlab = study_name,
                 fixed = FALSE, random = TRUE)

require(metafor)  # Meta-regression, subgroup analyses
require(netmeta)  # Network meta-analysis
```



### 7.5 Biomarker Analysis


```{r}
#| eval: false
# Reference: biomarker analysis (requires pROC)
# Biomarker × treatment interaction
glm(response ~ treatment + biomarker + treatment:biomarker,
   family = binomial)

# ROC curves for biomarker threshold optimization
require(pROC)
roc_obj <- roc(disease, biomarker_level)
```



---

## 8. Exercises

1. **Adaptive Designs**:
   - Design group sequential trial with O'Brien-Fleming boundary
   - Calculate sample size inflation factor
   - Determine interim stopping thresholds

2. **Multiple Testing**:
   - Apply Bonferroni, Holm, and BH corrections to multiple p-values
   - Compare number of rejections across methods
   - Design hierarchical testing procedure for complex endpoint hierarchy

3. **Estimands**:
   - Specify estimand for trial with anticipated dropouts
   - Choose appropriate intercurrent event strategy
   - Plan sensitivity analyses

4. **Missing Data**:
   - Implement multiple imputation for MAR data
   - Conduct tipping point analysis
   - Compare treatment effect under different missingness assumptions

5. **Meta-Analysis**:
   - Pool results from multiple studies using fixed-effects model
   - Compare to random-effects model
   - Assess publication bias with funnel plot and Egger's test
   - Interpret heterogeneity statistics

6. **Biomarker Analysis**:
   - Test for biomarker × treatment interaction (predictive)
   - Evaluate biomarker-stratified design
   - Optimize biomarker cutoff using interim data

7. **Integration**:
   - Design adaptive enrichment trial
   - Specify multiple testing strategy for hierarchical endpoints
   - Plan missing data handling approach
   - Conduct sensitivity analyses

---

## Summary

Advanced biostatistical methods enhance clinical trial efficiency and rigor:

- **Adaptive designs**: Early stopping, seamless phase designs, sample size re-estimation
- **Multiple testing**: FWER vs FDR control, hierarchical and gatekeeping procedures
- **Estimands**: Clear specification of treatment effects with intercurrent event handling
- **Missing data**: Multiple imputation, pattern mixture models, sensitivity analyses
- **Meta-analysis**: Fixed and random-effects models, heterogeneity assessment, publication bias
- **Biomarker designs**: Predictive biomarker identification, enrichment strategies

When properly integrated, these methods accelerate drug development while maintaining scientific rigor and regulatory acceptability.

---

## References and Further Reading

- ICH E9(R1). (2019). Addendum on estimands and sensitivity analyses in clinical trials.
- Peto, R. (1978). The design and analysis of long-term animal experiments. *Journal of the Royal Statistical Society*, 142, 151–183.
- DerSimonian, R., & Laird, N. (1986). Meta-analysis in clinical trials. *Controlled Clinical Trials*, 7(3), 177–188.
- Higgins, J.P.T., & Green, S. (Eds.). (2011). *Cochrane Handbook for Systematic Reviews of Interventions*. Wiley-Blackwell.
- Little, R.J.A., & Rubin, D.B. (2002). *Statistical Analysis with Missing Data* (2nd ed.). Wiley.
- Pocock, S.J., & Simon, R. (1975). Sequential treatment assignment with balancing for prognostic factors in the controlled clinical trial. *Biometrika*, 62(2), 375–386.
- Jennison, C., & Turnbull, B.W. (2000). *Group Sequential Methods with Applications to Clinical Trials*. Chapman & Hall/CRC.
- Moyé, L.A. (2006). *Multiple Analyses in Clinical Trials: Fundamentals for Investigators*. Springer.
