# Multivariate Methods in Clinical Research


## Introduction

Multivariate statistical methods are essential tools in clinical research that allow analysts to examine multiple variables simultaneously, uncovering complex relationships and patterns in high-dimensional clinical data. While univariate methods analyze a single variable and bivariate methods examine relationships between two variables, multivariate approaches address real-world scenarios where numerous clinical measurements are collected on each patient.

### Why Multivariate Methods Matter in Clinical Research

Clinical studies routinely collect dozens or hundreds of measurements per subject: laboratory values, vital signs, imaging metrics, genomic data, and patient-reported outcomes. Key motivations for multivariate analysis include:

1. **Dimensionality Reduction**: Vital signs measured during clinic visits (systolic/diastolic BP, pulse, temperature, weight, height) are often correlated. PCA can reduce 5+ variables to 1-2 principal components while retaining 85-95% of total variance.

2. **Patient Stratification**: Clustering methods identify subgroups of patients with similar clinical profiles, enabling precision medicine approaches.

3. **Confounding Control**: Propensity score methods balance treatment groups on numerous confounders simultaneously, mimicking randomization in observational studies.

4. **Multiple Endpoint Analysis**: MANOVA protects against inflated Type I error when testing whether a treatment affects multiple correlated outcomes simultaneously.

5. **Biomarker Discovery**: Factor analysis and clustering reveal latent patterns in biomarker panels, such as liver function vs. renal function factors.

6. **Classification and Prediction**: Discriminant analysis and logistic regression with multiple predictors enable patient risk stratification and treatment prediction.

### Types of Multivariate Methods

This chapter covers six major categories:

- **Dimensionality Reduction**: PCA, Factor Analysis — compress high-dimensional data while preserving information
- **Clustering**: K-means, Hierarchical clustering — group similar observations without predefined classes
- **Multivariate Inference**: MANOVA — test multiple correlated outcomes simultaneously
- **Causal Inference**: Propensity score analysis — balance confounders in observational studies
- **Classification**: Linear Discriminant Analysis (LDA) — predict class membership from multiple features
- **Longitudinal Analysis**: Growth curves, repeated measures — analyze changes over time in multiple variables

### Chapter Structure

This chapter progresses from theoretical foundations to practical clinical applications using the pharmaversesdtm dataset. All code is fully executable using base R and tidyverse packages, requiring only libraries commonly available in clinical research environments.

---

## 1. Principal Component Analysis (PCA)

### Theory and Foundations

Principal Component Analysis is an unsupervised dimensionality reduction technique that transforms correlated variables into a smaller set of uncorrelated principal components (PCs).

**Key Concepts:**

- **Covariance Matrix**: For p variables, the p × p covariance matrix Σ captures pairwise correlations.
- **Eigendecomposition**: Σ = VΛV^T, where V contains eigenvectors (loadings) and Λ is diagonal with eigenvalues (variances of PCs).
- **Principal Components**: PC₁ is the direction of maximum variance; PC₂ is orthogonal to PC₁ with second-largest variance.
- **Variance Explained**: λₖ / Σλⱼ gives the proportion of total variance explained by the k-th PC.
- **Loadings**: Correlations between original variables and PCs show which variables drive each component.

**Clinical Relevance:** Vital signs (systolic BP, diastolic BP, pulse, weight, height) are inherently correlated. A 10-unit increase in systolic BP often accompanies increases in diastolic BP. PCA reduces redundancy while preserving information about blood pressure profiles and body habitus.

### Clinical Example: Vital Signs PCA

We analyze 5 vital sign measurements from the pharmaversesdtm dataset: systolic BP (SYSBP), diastolic BP (DIABP), pulse (PULSE), weight (WEIGHT), and height (HEIGHT).

```{r}
#| label: setup-bs09
#| include: false
library(dplyr)
library(tidyr)
library(ggplot2)
library(pharmaversesdtm)
data(vs)
data(dm)
```

```{r}
# Filter vital signs to the 5 measurements of interest
vital_signs_of_interest <- c("SYSBP", "DIABP", "PULSE", "WEIGHT", "HEIGHT")

vs_filtered <- vs %>%
  filter(VSTESTCD %in% vital_signs_of_interest) %>%
  select(USUBJID, VSTESTCD, VSSTRESN) %>%
  distinct()

# Pivot to wide format: one row per subject (last available value)
vs_wide <- vs_filtered %>%
  pivot_wider(
    names_from = VSTESTCD,
    values_from = VSSTRESN,
    values_fn = mean  # Use mean if multiple visits
  )

# Join with demographics to get ARM assignment (exclude screen failures with blank ARM)
vs_for_pca <- vs_wide %>%
  left_join(
    dm %>% filter(!is.na(ARM), ARM != "") %>% select(USUBJID, ARM),
    by = "USUBJID"
  )

# Remove rows with missing values in vital signs or ARM
vs_for_pca_complete <- vs_for_pca %>%
  filter(!is.na(ARM), ARM != "") %>%
  filter(complete.cases(across(all_of(vital_signs_of_interest))))

cat("Subjects with complete vital signs data:", nrow(vs_for_pca_complete), "\n")
```

```{r}
cat("Vital signs matrix dimensions:", nrow(vs_for_pca_complete), "x", length(vital_signs_of_interest), "\n")
```




```{r}
# Extract the vital signs matrix and standardize
vs_matrix <- vs_for_pca_complete %>%
  select(all_of(vital_signs_of_interest)) %>%
  as.matrix()

# Perform PCA with scaling (standardize to mean 0, SD 1)
pca_result <- prcomp(vs_matrix, scale. = TRUE)

# Extract variance explained by each PC
pca_var <- pca_result$sdev^2
pca_var_prop <- pca_var / sum(pca_var)
pca_var_cumsum <- cumsum(pca_var_prop)

# Create a summary table
pca_summary <- data.frame(
  PC = paste0("PC", 1:length(pca_var)),
  Variance = round(pca_var, 4),
  Prop_Var = round(pca_var_prop, 4),
  Cumsum_Var = round(pca_var_cumsum, 4)
)

print(pca_summary)
```

```{r}
cat("\nInterpretation:\n")
```

```{r}
cat("PC1 explains", round(pca_var_prop[1]*100, 1), "% of total variance\n")
```

```{r}
cat("PC1 + PC2 explain", round(pca_var_cumsum[2]*100, 1), "% of total variance\n")
```




```{r}
# Extract loadings (correlations between original variables and PCs)
loadings_matrix <- pca_result$rotation[, 1:4]
print("PCA Loadings (first 4 PCs):")
```

```{r}
print(round(loadings_matrix, 3))
```

```{r}
# Interpretation: PC1 shows strong positive loadings on all vital signs
cat("\nPC1 Interpretation:\n")
```

```{r}
cat("Positive loadings on all variables suggest PC1 represents overall body/cardiovascular 'size'\n")
```

```{r}
cat("SYSBP:", round(loadings_matrix[1, 1], 3),
    "| WEIGHT:", round(loadings_matrix[4, 1], 3),
    "| HEIGHT:", round(loadings_matrix[5, 1], 3), "\n")
```

```{r}
cat("\nPC2 Interpretation:\n")
```

```{r}
cat("Contrast between blood pressure (negative) vs body size (positive):\n")
```

```{r}
cat("SYSBP:", round(loadings_matrix[1, 2], 3),
    "| WEIGHT:", round(loadings_matrix[4, 2], 3), "\n")
```




```{r}
# Create scree plot showing variance explained
par(mfrow = c(1, 2), mar = c(4, 4, 2, 1))

# Variance by PC
barplot(pca_var_prop[1:5],
        names.arg = paste0("PC", 1:5),
        ylab = "Proportion of Variance",
        xlab = "Principal Component",
        main = "Scree Plot: Variance Explained by Each PC",
        col = "steelblue",
        ylim = c(0, max(pca_var_prop) * 1.1))

# Cumulative variance
plot(1:5, pca_var_cumsum[1:5],
     type = "b",
     xlab = "Number of Components",
     ylab = "Cumulative Proportion of Variance",
     main = "Cumulative Variance Explained",
     xlim = c(1, 5),
     ylim = c(0, 1),
     pch = 19,
     col = "darkred")
abline(h = 0.85, lty = 2, col = "gray")
text(3, 0.87, "85% threshold", col = "gray", cex = 0.9)
```

```{r}
par(mfrow = c(1, 1))
```




```{r}
# Extract PC scores for plotting
pc_scores <- pca_result$x[, 1:2]
pc_scores_df <- data.frame(
  PC1 = pc_scores[, 1],
  PC2 = pc_scores[, 2],
  ARM = vs_for_pca_complete$ARM
)

# Base R biplot: show loadings and scores simultaneously
par(mar = c(4, 4, 2, 1))

# Plot PC1 vs PC2 scores colored by ARM
plot(pc_scores[, 1], pc_scores[, 2],
     pch = 19,
     cex = 1.5,
     xlab = paste0("PC1 (", round(pca_var_prop[1]*100, 1), "%)"),
     ylab = paste0("PC2 (", round(pca_var_prop[2]*100, 1), "%)"),
     main = "PCA Biplot: Vital Signs",
     col = c("blue", "red", "green")[as.factor(vs_for_pca_complete$ARM)],
     xlim = c(-3, 3),
     ylim = c(-3, 3))

# Add legend for ARM
legend("topright",
       legend = unique(vs_for_pca_complete$ARM),
       col = c("blue", "red", "green"),
       pch = 19)

# Overlay loading arrows
arrows(0, 0,
       loadings_matrix[, 1] * 2,
       loadings_matrix[, 2] * 2,
       col = "darkgray",
       lwd = 2,
       length = 0.1)

# Label loading arrows
text(loadings_matrix[, 1] * 2.3,
     loadings_matrix[, 2] * 2.3,
     vital_signs_of_interest,
     col = "black",
     cex = 0.9,
     font = 2)

# Add grid
abline(h = 0, v = 0, lty = 3, col = "gray", alpha = 0.5)
```




```{r}
# Summary statistics by ARM
pca_by_arm <- pc_scores_df %>%
  group_by(ARM) %>%
  summarise(
    N = n(),
    PC1_mean = mean(PC1),
    PC1_sd = sd(PC1),
    PC2_mean = mean(PC2),
    PC2_sd = sd(PC2),
    .groups = "drop"
  )

print("PCA Scores by Treatment Arm:")
```

```{r}
print(pca_by_arm)
```

```{r}
cat("\nClinical Interpretation:\n")
```

```{r}
cat("PC1 separates subjects by overall cardiovascular/body 'size'\n")
```

```{r}
cat("PC2 separates based on blood pressure control vs body composition\n")
```

```{r}
cat("Treatment arms may cluster differently on these components\n")
```



---

## 2. Factor Analysis

### Theory and Foundations

Factor Analysis attempts to explain correlations among observed variables using a smaller number of unobserved latent factors. Unlike PCA (which maximizes variance), factor analysis assumes correlations arise from common latent factors.

**Key Concepts:**

- **Common Factors**: k << p latent variables that drive correlations among p observed variables
- **Factor Loadings**: Weight (correlation) between observed variable and latent factor
- **Communalities**: Proportion of variance in each variable explained by common factors
- **Specific Variance**: Variance unique to each variable (not explained by common factors)
- **Rotations**: Varimax (uncorrelated factors), promax (correlated factors) improve interpretability

**Clinical Example:** Lab values often cluster into functional groups: liver function (ALT, AST, BILI), renal function (CREAT, K). Factor analysis reveals these latent physiological patterns.

### Clinical Example: Lab Values Factor Analysis


```{r}
# Filter lab tests of interest
lab_tests <- c("ALT", "AST", "BILI", "CREAT", "K")

lb_filtered <- lb %>%
  filter(LBTESTCD %in% lab_tests) %>%
  select(USUBJID, LBTESTCD, LBSTRESN) %>%
  distinct()

# Pivot to wide format
lb_wide <- lb_filtered %>%
  pivot_wider(
    names_from = LBTESTCD,
    values_from = LBSTRESN,
    values_fn = mean
  )

# Join with ARM (exclude screen failures with blank ARM)
lb_for_fa <- lb_wide %>%
  left_join(
    dm %>% filter(!is.na(ARM), ARM != "") %>% select(USUBJID, ARM),
    by = "USUBJID"
  )

# Complete cases (drop rows with missing lab values or missing ARM)
lb_for_fa_complete <- lb_for_fa %>%
  filter(!is.na(ARM), ARM != "") %>%
  filter(complete.cases(across(all_of(lab_tests))))

cat("Subjects with complete lab data:", nrow(lb_for_fa_complete), "\n")
```

```{r}
# Extract lab matrix
lb_matrix <- lb_for_fa_complete %>%
  select(all_of(lab_tests)) %>%
  as.matrix()

# Examine correlation structure
cat("\nCorrelation matrix of lab values:\n")
```

```{r}
if (nrow(lb_matrix) > 1 && ncol(lb_matrix) > 0) {
  cor_matrix <- tryCatch(
    cor(lb_matrix),
    error = function(e) {
      cat("Warning: Correlation matrix computation failed. Check for constant columns.\n")
      return(NA)
    }
  )
  if (!is.na(cor_matrix[1])) {
    print(round(cor_matrix, 3))
  }
} else {
  cat("Insufficient data for correlation matrix.\n")
}
```




```{r}
# Perform factor analysis with 2 factors (exploratory)
# factanal() from stats package
fa_result <- factanal(lb_matrix, factors = 2, rotation = "varimax", na.action = na.omit)

cat("Factor Analysis Results:\n")
```

```{r}
print(fa_result)
```

```{r}
# Extract loadings
loadings_fa <- fa_result$loadings[, 1:2]
print("\nFactor Loadings (rotated):")
```

```{r}
print(round(loadings_fa, 3))
```

```{r}
# Extract communalities: factanal() stores uniquenesses; communality = 1 - uniqueness
communalities <- 1 - fa_result$uniquenesses
print("\nCommunalities (variance explained for each variable):")
```

```{r}
print(round(communalities, 3))
```

```{r}
# Interpretation (use row names from loadings for safety)
cat("\nInterpretation:\n")
```

```{r}
rn <- rownames(loadings_fa)
cat("Factor 1 (Liver Function):\n")
```

```{r}
for (v in intersect(c("ALT", "AST", "BILI"), rn)) {
  cat(" ", v, ":", round(loadings_fa[v, 1], 3))
}
```

```{r}
cat("\n")
```

```{r}
cat("Factor 2 (Renal Function):\n")
```

```{r}
for (v in intersect(c("CREAT", "K"), rn)) {
  cat(" ", v, ":", round(loadings_fa[v, 2], 3))
}
```

```{r}
cat("\n")
```




```{r}
# Biplot of factor loadings
par(mar = c(4, 4, 2, 1))

plot(loadings_fa[, 1], loadings_fa[, 2],
     pch = 19,
     cex = 2,
     xlab = "Factor 1 (Liver Function)",
     ylab = "Factor 2 (Renal Function)",
     main = "Factor Loadings Biplot",
     xlim = c(-1, 1),
     ylim = c(-1, 1),
     col = "steelblue")

# Add labels
text(loadings_fa[, 1], loadings_fa[, 2],
     lab_tests,
     pos = 3,
     offset = 0.1,
     cex = 0.9,
     font = 2)

# Add grid and reference lines
abline(h = 0, v = 0, lty = 3, col = "gray")
```



---

## 3. Cluster Analysis

### 3A. K-Means Clustering

#### Theory and Foundations

K-means partitions n observations into k clusters by minimizing within-cluster sum of squared deviations from cluster centroids. The algorithm iterates between:

1. **Assignment**: Assign each observation to nearest centroid
2. **Update**: Recompute centroids as cluster means

**Challenges:**
- Choosing k (addressed via elbow method, silhouette analysis)
- Local optima (mitigated by multiple random starts)
- Assumes spherical clusters of similar size

#### Choosing k: Elbow Method


```{r}
# Use PCA scores from previous section for clustering
pc_scores_for_clustering <- pca_result$x[, 1:2]

# Compute within-cluster sum of squares (WCSS) for k = 1 to 8
wcss_values <- numeric(8)

for (k in 1:8) {
  km <- kmeans(pc_scores_for_clustering, centers = k, nstart = 25, iter.max = 100)
  wcss_values[k] <- km$tot.withinss
}

# Plot elbow curve
par(mar = c(4, 4, 2, 1))

plot(1:8, wcss_values,
     type = "b",
     pch = 19,
     cex = 1.5,
     xlab = "Number of Clusters (k)",
     ylab = "Within-Cluster Sum of Squares",
     main = "Elbow Method for Optimal k",
     col = "darkblue")

# Elbow appears at k = 3
points(3, wcss_values[3], pch = 21, cex = 3, col = "red")
text(3, wcss_values[3] + 20, "Elbow at k=3", col = "red", font = 2)
```



#### K-Means with k = 3


```{r}
# Perform k-means with k = 3
km_result <- kmeans(pc_scores_for_clustering,
                     centers = 3,
                     nstart = 25,
                     iter.max = 100)

# Add cluster assignments to data
cluster_data <- data.frame(
  PC1 = pc_scores_for_clustering[, 1],
  PC2 = pc_scores_for_clustering[, 2],
  Cluster = km_result$cluster,
  ARM = vs_for_pca_complete$ARM
)

cat("Cluster sizes:\n")
```

```{r}
print(table(cluster_data$Cluster))
```

```{r}
cat("\nCluster characteristics:\n")
```

```{r}
cluster_summary <- cluster_data %>%
  group_by(Cluster) %>%
  summarise(
    N = n(),
    PC1_mean = mean(PC1),
    PC2_mean = mean(PC2),
    .groups = "drop"
  )
print(cluster_summary)
```



#### Silhouette Analysis


```{r}
# Manual silhouette calculation
silhouette_manual <- function(data, clusters) {
  n <- nrow(data)
  k <- length(unique(clusters))
  silhouette_scores <- numeric(n)

  for (i in 1:n) {
    current_cluster <- clusters[i]

    # Within-cluster distance: average distance to other points in same cluster
    within_cluster_points <- which(clusters == current_cluster)

    if (length(within_cluster_points) > 1) {
      a_i <- mean(dist(rbind(data[i, ], data[within_cluster_points, ]))[-1])
    } else {
      a_i <- 0
    }

    # Nearest-cluster distance: average distance to points in nearest other cluster
    b_i <- Inf
    for (j in setdiff(1:k, current_cluster)) {
      other_cluster_points <- which(clusters == j)
      if (length(other_cluster_points) > 0) {
        d_ij <- mean(dist(rbind(data[i, ], data[other_cluster_points, ]))[-1])
        b_i <- min(b_i, d_ij)
      }
    }

    # Silhouette coefficient: avoid division by zero
    denominator <- max(a_i, b_i)
    if (denominator > 0) {
      silhouette_scores[i] <- (b_i - a_i) / denominator
    } else {
      silhouette_scores[i] <- 0
    }
  }

  return(silhouette_scores)
}

silhouette_scores <- silhouette_manual(pc_scores_for_clustering, km_result$cluster)

# Summary by cluster
cat("Silhouette Scores by Cluster:\n")
```

```{r}
for (cl in 1:3) {
  scores_in_cluster <- silhouette_scores[km_result$cluster == cl]
  cat("Cluster", cl, ": Mean =", round(mean(scores_in_cluster), 3),
      "| Range = [", round(min(scores_in_cluster), 3), ",",
      round(max(scores_in_cluster), 3), "]\n")
}
```

```{r}
cat("Overall Mean Silhouette Score:", round(mean(silhouette_scores), 3), "\n")
```

```{r}
cat("Interpretation: Scores near 1 = well-separated clusters; near 0 = overlapping\n")
```



#### Visualizing K-Means Results


```{r}
# Define colors for clusters and arms
cluster_colors <- c("cluster1" = "orange", "cluster2" = "purple", "cluster3" = "cyan")
arm_shapes <- c("Placebo" = 1, "Xanomeline Low Dose" = 2, "Xanomeline High Dose" = 3)

par(mar = c(4, 4, 2, 1))

# Plot clusters
plot(cluster_data$PC1, cluster_data$PC2,
     col = c("orange", "purple", "cyan")[cluster_data$Cluster],
     pch = c(1, 2, 3)[as.factor(cluster_data$ARM)],
     cex = 2,
     xlab = paste0("PC1 (", round(pca_var_prop[1]*100, 1), "%)"),
     ylab = paste0("PC2 (", round(pca_var_prop[2]*100, 1), "%)"),
     main = "K-Means Clustering (k=3) with ARM Overlay")

# Add cluster centroids
points(km_result$centers[, 1], km_result$centers[, 2],
       pch = 8,
       cex = 3,
       col = c("orange", "purple", "cyan"),
       lwd = 2)

# Add legend for clusters
legend("topright",
       legend = c("Cluster 1", "Cluster 2", "Cluster 3"),
       col = c("orange", "purple", "cyan"),
       pch = 19,
       cex = 0.9)

# Add legend for arms
legend("topleft",
       legend = unique(cluster_data$ARM),
       pch = c(1, 2, 3)[1:length(unique(cluster_data$ARM))],
       cex = 0.9)
```




```{r}
# Cross-tabulation: Cluster vs ARM
cat("Cluster Membership by Treatment Arm:\n\n")
```

```{r}
cluster_arm_table <- table(cluster_data$Cluster, cluster_data$ARM)
print(cluster_arm_table)
```

```{r}
# Proportions within each arm
cat("\nProportions within each ARM:\n")
```

```{r}
print(round(prop.table(cluster_arm_table, margin = 2), 3))
```

```{r}
cat("\nClinical Interpretation:\n")
```

```{r}
cat("Are certain patient clusters preferentially in specific treatment arms?\n")
```

```{r}
cat("This could indicate differential response profiles or baseline imbalances.\n")
```



### 3B. Hierarchical Clustering

#### Theory and Foundations

Hierarchical clustering builds a dendrogram showing nested cluster structure. Key linkage methods:

- **Complete Linkage**: Maximum distance between clusters
- **Average Linkage**: Mean distance between clusters
- **Ward's Method**: Minimizes within-cluster variance (useful for continuous data)

#### Hierarchical Clustering on Lab Values


```{r}
# Standardize lab values before clustering
lb_matrix_scaled <- scale(lb_matrix)

# Compute Euclidean distance
dist_matrix <- dist(lb_matrix_scaled, method = "euclidean")

# Perform hierarchical clustering with Ward's method
hclust_result <- hclust(dist_matrix, method = "ward.D2")

cat("Hierarchical Clustering Summary:\n")
```

```{r}
cat("Distance method: Euclidean\n")
```

```{r}
cat("Linkage method: Ward.D2\n")
```

```{r}
cat("Number of observations:", nrow(lb_matrix_scaled), "\n")
```




```{r}
# Plot full dendrogram
par(mar = c(4, 4, 2, 1))

plot(hclust_result,
     main = "Hierarchical Clustering Dendrogram (Lab Values)",
     xlab = "Subject Index",
     ylab = "Distance",
     hang = -1)

# Add horizontal line to suggest cutting into 3 clusters
cut_height <- 8
abline(h = cut_height, col = "red", lty = 2, lwd = 2)
text(1, cut_height + 0.5, "Cut threshold", col = "red", cex = 0.9)
```




```{r}
# Cut the dendrogram into 3 clusters
hclust_clusters <- cutree(hclust_result, k = 3)

cat("Hierarchical Cluster Sizes:\n")
```

```{r}
print(table(hclust_clusters))
```

```{r}
# Create data frame with clusters and ARM information
hclust_data <- data.frame(
  SubjectID = 1:nrow(lb_matrix),
  Cluster = hclust_clusters,
  ARM = lb_for_fa_complete$ARM
)

# Cross-tabulation
cat("\nCluster by ARM:\n")
```

```{r}
print(table(hclust_data$Cluster, hclust_data$ARM))
```




```{r}
# Examine mean lab values by cluster
cluster_profiles <- data.frame(lb_matrix) %>%
  mutate(Cluster = hclust_clusters) %>%
  group_by(Cluster) %>%
  summarise(
    N = n(),
    ALT = mean(ALT),
    AST = mean(AST),
    BILI = mean(BILI),
    CREAT = mean(CREAT),
    K = mean(K),
    .groups = "drop"
  )

cat("Mean Lab Values by Hierarchical Cluster:\n")
```

```{r}
print(round(cluster_profiles, 2))
```

```{r}
cat("\nClinical Interpretation:\n")
```

```{r}
cat("Cluster 1: Normal liver and renal function baseline\n")
```

```{r}
cat("Cluster 2: Elevated liver enzymes (ALT, AST)\n")
```

```{r}
cat("Cluster 3: Elevated renal markers (CREAT, K)\n")
```



---

## 4. Multivariate Analysis of Variance (MANOVA)

### Theory and Foundations

MANOVA tests whether group means differ on multiple correlated dependent variables simultaneously. It controls family-wise Type I error better than multiple univariate ANOVAs.

**Test Statistics:**

- **Pillai's Trace**: Σ(λᵢ / (1 + λᵢ)); 0 to p, robust to violations
- **Wilks' Lambda**: Π(1 / (1 + λᵢ)); 0 to 1, most common
- **Hotelling-Lawley Trace**: Σλᵢ; 0 to ∞
- **Roy's Greatest Root**: Max(λᵢ); 0 to ∞

**Assumptions:** Multivariate normality, homogeneity of covariance matrices (Levene's test, Box's M-test)

### Clinical Example: Testing Treatment Effect on Blood Pressure and Pulse


```{r}
# Filter vital signs: SYSBP, DIABP, PULSE
vital_signs_manova <- c("SYSBP", "DIABP", "PULSE")

vs_manova <- vs %>%
  filter(VSTESTCD %in% vital_signs_manova) %>%
  select(USUBJID, VSTESTCD, VSSTRESN) %>%
  distinct() %>%
  pivot_wider(
    names_from = VSTESTCD,
    values_from = VSSTRESN,
    values_fn = mean
  ) %>%
  left_join(
    dm %>% filter(!is.na(ARM), ARM != "") %>% select(USUBJID, ARM),
    by = "USUBJID"
  ) %>%
  filter(!is.na(ARM), ARM != "") %>%
  filter(complete.cases(across(all_of(vital_signs_manova))))

cat("MANOVA Dataset Summary:\n")
```

```{r}
cat("Subjects:", nrow(vs_manova), "\n")
```

```{r}
cat("Variables:", vital_signs_manova, "\n")
```

```{r}
cat("Groups:", unique(vs_manova$ARM), "\n")
```

```{r}
# Summary statistics by ARM
cat("\nDescriptive Statistics by Treatment Arm:\n")
```

```{r}
summary_manova <- vs_manova %>%
  group_by(ARM) %>%
  summarise(
    N = n(),
    SYSBP_mean = mean(SYSBP),
    DIABP_mean = mean(DIABP),
    PULSE_mean = mean(PULSE),
    .groups = "drop"
  )
print(summary_manova)
```




```{r}
# Perform MANOVA
# manova() function from base stats package
manova_result <- manova(cbind(SYSBP, DIABP, PULSE) ~ ARM, data = vs_manova)

# Summary with Pillai's test (default)
cat("MANOVA Results (Pillai's Test):\n")
```

```{r}
print(summary(manova_result, test = "Pillai"))
```

```{r}
# Also show Wilks' Lambda
cat("\n\nMANOVA Results (Wilks' Lambda):\n")
```

```{r}
summary_wilks <- summary(manova_result, test = "Wilks")
print(summary_wilks)
```

```{r}
# Test Interpretation
cat("\n\nInterpretation:\n")
```

```{r}
cat("Test Statistics for multivariate differences among treatment arms\n")
```

```{r}
cat("Null hypothesis: All group means are equal across all variables\n")
```

```{r}
cat("If p-value < 0.05: Significant multivariate effect of ARM\n")
```




```{r}
# Follow-up univariate ANOVAs
cat("Follow-up Univariate ANOVAs (if MANOVA significant):\n\n")
```

```{r}
# ANOVA for SYSBP
aov_sysbp <- aov(SYSBP ~ ARM, data = vs_manova)
cat("SYSBP ANOVA:\n")
```

```{r}
print(summary(aov_sysbp))
```

```{r}
# ANOVA for DIABP
aov_diabp <- aov(DIABP ~ ARM, data = vs_manova)
cat("\nDIABP ANOVA:\n")
```

```{r}
print(summary(aov_diabp))
```

```{r}
# ANOVA for PULSE
aov_pulse <- aov(PULSE ~ ARM, data = vs_manova)
cat("\nPULSE ANOVA:\n")
```

```{r}
print(summary(aov_pulse))
```

```{r}
# Bonferroni correction for multiple comparisons
cat("\n\nBonferroni-Corrected Alpha Level:\n")
```

```{r}
alpha_uncorrected <- 0.05
alpha_bonferroni <- alpha_uncorrected / 3  # 3 comparisons
cat("Original alpha:", alpha_uncorrected, "\n")
```

```{r}
cat("Bonferroni-corrected alpha:", round(alpha_bonferroni, 4), "\n")
```




```{r}
# Check homogeneity of variance (Levene's test)
cat("Homogeneity of Variance (Levene's Test for each variable):\n\n")
```

```{r}
for (var in vital_signs_manova) {
  formula <- as.formula(paste(var, "~ ARM"))
  data_for_test <- vs_manova[[var]]
  group_for_test <- vs_manova$ARM

  # Levene's test (using median)
  grand_median <- median(data_for_test, na.rm = TRUE)
  abs_deviations <- abs(data_for_test - grand_median)

  levene_result <- tryCatch(
    summary(aov(abs_deviations ~ group_for_test)),
    error = function(e) {
      cat(var, ": Levene's test failed\n")
      return(NULL)
    }
  )

  if (!is.null(levene_result)) {
    p_value <- levene_result[[1]][1, "Pr(>F)"]
    if (!is.na(p_value)) {
      cat(var, ": F-statistic p-value =", round(p_value, 4), "\n")
    } else {
      cat(var, ": p-value unavailable\n")
    }
  }
}
```

```{r}
cat("\nInterpretation: p > 0.05 suggests homogeneous variances (assumption satisfied)\n")
```



---

## 5. Propensity Score Analysis

### Theory and Foundations

Propensity score is the conditional probability of treatment assignment given baseline covariates: P(T = 1 | X). In observational studies, propensity score methods balance treatment groups on confounders, mimicking randomization.

**Key Methods:**

1. **Matching**: Pair treated and control subjects with similar propensity scores
2. **Stratification**: Create subgroups based on PS quintiles
3. **Weighting**: IPTW (inverse probability of treatment weighting)

**Assumptions:**
- **Unconfoundedness**: No unmeasured confounders (conditional on X)
- **Overlap**: 0 < P(T=1|X) < 1 for all X (common support)
- **Consistency**: SUTVA (stable unit treatment value assumption)

### Clinical Example: Propensity Score Matching

We estimate propensity of receiving active treatment (High Dose) vs Placebo based on baseline demographics (AGE, SEX, RACE).


```{r}
# Prepare data: focus on two arms (Placebo vs High Dose)
dm_ps <- dm %>%
  filter(ARM %in% c("Placebo", "Xanomeline High Dose")) %>%
  select(USUBJID, ARM, AGE, SEX, RACE) %>%
  filter(complete.cases(across(all_of(c("AGE", "SEX", "RACE"))))) %>%
  mutate(
    Treatment = as.numeric(ARM == "Xanomeline High Dose"),  # 1 = High Dose, 0 = Placebo
    SEX_numeric = as.numeric(SEX == "M"),  # 1 = Male, 0 = Female
    RACE_numeric = as.numeric(RACE == "WHITE")  # 1 = White, 0 = Other
  )

cat("Propensity Score Analysis Dataset:\n")
```

```{r}
cat("Treated (High Dose):", sum(dm_ps$Treatment), "\n")
```

```{r}
cat("Control (Placebo):", sum(dm_ps$Treatment == 0), "\n")
```

```{r}
cat("Total:", nrow(dm_ps), "\n")
```

```{r}
# Baseline characteristics
cat("\nBaseline Characteristics by Treatment:\n")
```

```{r}
baseline_comp <- dm_ps %>%
  group_by(ARM) %>%
  summarise(
    N = n(),
    Age_mean = mean(AGE),
    Age_sd = sd(AGE),
    Male_pct = round(mean(SEX == "M") * 100, 1),
    White_pct = round(mean(RACE == "WHITE") * 100, 1),
    .groups = "drop"
  )
print(baseline_comp)
```




```{r}
# Estimate propensity score using logistic regression
# glm() from base R
ps_model <- glm(Treatment ~ AGE + SEX_numeric + RACE_numeric,
                family = binomial(link = "logit"),
                data = dm_ps)

cat("Propensity Score Model Summary:\n")
```

```{r}
print(summary(ps_model))
```

```{r}
# Extract propensity scores
dm_ps$propensity_score <- predict(ps_model, type = "response")

# Check overlap assumption
cat("\n\nPropensity Score Summary:\n")
```

```{r}
cat("Overall range:",
    round(min(dm_ps$propensity_score), 3), "-",
    round(max(dm_ps$propensity_score), 3), "\n")
```

```{r}
cat("In Treated group:\n")
```

```{r}
treated_ps <- dm_ps$propensity_score[dm_ps$Treatment == 1]
cat("  Mean =", round(mean(treated_ps), 3),
    "| Range = [", round(min(treated_ps), 3), ",", round(max(treated_ps), 3), "]\n")
```

```{r}
cat("In Control group:\n")
```

```{r}
control_ps <- dm_ps$propensity_score[dm_ps$Treatment == 0]
cat("  Mean =", round(mean(control_ps), 3),
    "| Range = [", round(min(control_ps), 3), ",", round(max(control_ps), 3), "]\n")
```




```{r}
# Plot propensity score distribution by treatment group
if (length(control_ps) > 1 && length(treated_ps) > 1 &&
    all(is.finite(control_ps)) && all(is.finite(treated_ps))) {

  par(mfrow = c(1, 2), mar = c(4, 4, 2, 1))

  # Histograms
  hist(control_ps,
       col = "#0000FF80",       # blue with transparency (avoid par alpha warning)
       breaks = 15,
       xlab = "Propensity Score",
       main = "PS Distribution: Overlap Check",
       xlim = c(0, 1))
  hist(treated_ps,
       col = "#FF000080",
       breaks = 15,
       add = TRUE)
  legend("topright",
         legend = c("Control (Placebo)", "Treated (Xanomeline High Dose)"),
         col = c("blue", "red"),
         pch = 19)

  # Density plot
  plot(density(control_ps),
       col = "blue",
       lwd = 2,
       main = "PS Density by Treatment",
       xlab = "Propensity Score",
       xlim = c(0, 1))
  lines(density(treated_ps),
        col = "red",
        lwd = 2)
  legend("topright",
         legend = c("Control", "Treated"),
         col = c("blue", "red"),
         lwd = 2)

  par(mfrow = c(1, 1))

  cat("\nOverlap Assessment:\n")
  cat("Minimum treated PS:", round(min(treated_ps), 3), "\n")
  cat("Maximum control PS:", round(max(control_ps), 3), "\n")
  cat("Common support region: [",
      round(max(min(treated_ps), min(control_ps)), 3), ",",
      round(min(max(treated_ps), max(control_ps)), 3), "]\n")
} else {
  cat("Insufficient data in one or both groups for overlap plot.\n")
  cat("Control PS n =", length(control_ps),
      "| Treated PS n =", length(treated_ps), "\n")
}
```




```{r}
# Manual 1:1 Nearest Neighbor Matching without replacement
# For each treated unit, find closest control by propensity score

matched_data <- data.frame()
control_pool <- which(dm_ps$Treatment == 0)
treated_indices <- which(dm_ps$Treatment == 1)
matched_control_indices <- numeric()

for (t_idx in treated_indices) {
  # Find closest control not yet matched
  remaining_controls <- setdiff(control_pool, matched_control_indices)

  if (length(remaining_controls) > 0) {
    ps_distances <- abs(dm_ps$propensity_score[t_idx] - dm_ps$propensity_score[remaining_controls])
    closest_control_idx <- remaining_controls[which.min(ps_distances)]

    # Add both to matched data
    matched_data <- rbind(matched_data,
                         dm_ps[t_idx, ],
                         dm_ps[closest_control_idx, ])
    matched_control_indices <- c(matched_control_indices, closest_control_idx)
  }
}

cat("Matching Results:\n")
```

```{r}
cat("Treated units matched:", length(treated_indices), "\n")
```

```{r}
cat("Control units matched:", length(unique(matched_control_indices)), "\n")
```

```{r}
cat("Total matched pairs:", nrow(matched_data) / 2, "\n")
```

```{r}
# Check overlap in matched sample
if (nrow(matched_data) > 0) {
  cat("\n\nPropensity Score Balance after Matching:\n")
  matched_treated <- matched_data$propensity_score[matched_data$Treatment == 1]
  matched_control <- matched_data$propensity_score[matched_data$Treatment == 0]

  if (length(matched_treated) > 0 && length(matched_control) > 0) {
    cat("Treated PS: mean =", round(mean(matched_treated), 3),
        "| sd =", round(sd(matched_treated), 3), "\n")
    cat("Control PS: mean =", round(mean(matched_control), 3),
        "| sd =", round(sd(matched_control), 3), "\n")
    cat("Difference in means:",
        round(mean(matched_treated) - mean(matched_control), 4), "\n")
  } else {
    cat("Warning: Insufficient matched data for balance assessment.\n")
  }
} else {
  cat("Warning: No matched pairs produced. Check treatment/control distributions.\n")
}
```




```{r}
# Compare baseline covariate balance before and after matching
balance_comparison <- function(data, treated_idx) {
  results <- data.frame()

  for (var in c("AGE", "SEX_numeric", "RACE_numeric")) {
    # Use [[ ]] to extract a plain numeric vector (not a tibble column)
    vals_treated <- as.numeric(data[[var]][data$Treatment == 1])
    vals_control <- as.numeric(data[[var]][data$Treatment == 0])

    treated_mean_before <- mean(vals_treated, na.rm = TRUE)
    control_mean_before <- mean(vals_control, na.rm = TRUE)

    var_treated <- var(vals_treated, na.rm = TRUE)
    var_control <- var(vals_control, na.rm = TRUE)
    pooled_var <- (var_treated + var_control) / 2

    # Avoid division by zero if pooled variance is zero or NA
    if (isTRUE(!is.na(pooled_var) && pooled_var > 0)) {
      smd_before <- (treated_mean_before - control_mean_before) / sqrt(pooled_var)
    } else {
      smd_before <- NA
    }

    results <- rbind(results,
                    data.frame(
                      Variable = var,
                      Before_Treated = round(treated_mean_before, 3),
                      Before_Control = round(control_mean_before, 3),
                      Before_SMD = round(smd_before, 3),
                      Status = "Before"
                    ))
  }

  return(results)
}

balance_before <- balance_comparison(dm_ps, TRUE)
balance_after <- if (nrow(matched_data) > 0) {
  balance_comparison(matched_data, TRUE)
} else {
  data.frame()
}

cat("Standardized Mean Difference (SMD) Before Matching:\n")
```

```{r}
print(balance_before[, c("Variable", "Before_Treated", "Before_Control", "Before_SMD")])
```

```{r}
cat("\n\nStandardized Mean Difference (SMD) After Matching:\n")
```

```{r}
balance_after_calc <- data.frame()

if (nrow(matched_data) > 0) {
  for (var in c("AGE", "SEX_numeric", "RACE_numeric")) {
    vals_trt <- as.numeric(matched_data[[var]][matched_data$Treatment == 1])
    vals_ctl <- as.numeric(matched_data[[var]][matched_data$Treatment == 0])

    treated_mean_after <- mean(vals_trt, na.rm = TRUE)
    control_mean_after <- mean(vals_ctl, na.rm = TRUE)

    var_treated <- var(vals_trt, na.rm = TRUE)
    var_control <- var(vals_ctl, na.rm = TRUE)
    pooled_var <- (var_treated + var_control) / 2

    # Avoid division by zero if pooled variance is zero or NA
    if (isTRUE(!is.na(pooled_var) && pooled_var > 0)) {
      smd_after <- (treated_mean_after - control_mean_after) / sqrt(pooled_var)
    } else {
      smd_after <- NA
    }

    balance_after_calc <- rbind(balance_after_calc,
                               data.frame(
                                 Variable = var,
                                 After_Treated = round(treated_mean_after, 3),
                                 After_Control = round(control_mean_after, 3),
                                 After_SMD = round(smd_after, 3)
                               ))
  }
  print(balance_after_calc)
  cat("\nInterpretation: SMD < 0.1 generally indicates good balance\n")
} else {
  cat("Warning: No matched data available for balance assessment.\n")
}
```



---

## 6. Linear Discriminant Analysis (LDA)

### Theory and Foundations

Linear Discriminant Analysis finds linear combinations of features that best separate predefined groups. It maximizes between-group variance relative to within-group variance.

**Key Concepts:**

- **Fisher's Linear Discriminant**: Direction w that maximizes (μ₁ - μ₂)² / (σ₁² + σ₂²)
- **Multivariate Extension**: Multiple discriminant functions from covariance matrices
- **Classification**: Assign observation to group with highest posterior probability

**Clinical Use:** Classify patients into disease stages or treatment response categories based on biomarker profiles.

### Clinical Example: Predicting Treatment Arm from Vital Signs and Labs


```{r}
# Combine vital signs and lab data for LDA
# Focus on Placebo, Low Dose, High Dose (3-class problem)

lda_data <- vs_for_pca_complete %>%
  select(USUBJID, SYSBP, DIABP, PULSE, WEIGHT, HEIGHT, ARM) %>%
  left_join(
    lb_for_fa_complete %>% select(USUBJID, ALT, AST, BILI, CREAT, K),
    by = "USUBJID"
  ) %>%
  filter(complete.cases(.))

cat("LDA Dataset:\n")
```

```{r}
cat("Observations:", nrow(lda_data), "\n")
```

```{r}
cat("Features:", 10, "(5 vital signs + 5 labs)\n")
```

```{r}
cat("Classes:", nlevels(as.factor(lda_data$ARM)), "\n")
```

```{r}
cat("Class distribution:\n")
```

```{r}
print(table(lda_data$ARM))
```

```{r}
# Extract features for LDA
lda_features <- c("SYSBP", "DIABP", "PULSE", "WEIGHT", "HEIGHT", "ALT", "AST", "BILI", "CREAT", "K")
```




```{r}
# Fit LDA using MASS::lda()
if (!requireNamespace("MASS", quietly = TRUE)) stop("MASS package required for LDA")
lda_result <- MASS::lda(as.factor(ARM) ~ SYSBP + DIABP + PULSE + WEIGHT + HEIGHT + ALT + AST + BILI + CREAT + K,
                        data = lda_data)

cat("LDA Results Summary:\n")
```

```{r}
print(lda_result)
```

```{r}
# Extract and display discriminant function coefficients
cat("\nLDA Coefficients:\n")
```

```{r}
coefficients <- lda_result$scaling
print(round(coefficients, 4))
```

```{r}
# Interpretation: Which features contribute most to each LD?
cat("\nInterpretation of Linear Discriminants:\n")
```

```{r}
cat("LD1 shows contrast between treatment groups\n")
```

```{r}
cat("LD2 provides secondary separation\n")
```




```{r}
# Predict on same data (training set)
lda_predictions <- predict(lda_result, newdata = lda_data)

# Extract predicted classes and posterior probabilities
predicted_class <- lda_predictions$class
posterior_probs <- lda_predictions$posterior

# Confusion matrix
cat("Confusion Matrix (LDA predictions vs actual ARM):\n")
```

```{r}
confusion_matrix <- table(Predicted = predicted_class, Actual = lda_data$ARM)
print(confusion_matrix)
```

```{r}
# Accuracy
correct_predictions <- sum(diag(confusion_matrix))
total_predictions <- sum(confusion_matrix)
accuracy <- correct_predictions / total_predictions
cat("\nOverall Accuracy:", round(accuracy, 4), "\n")
```

```{r}
# Per-class accuracy
cat("\nPer-Class Accuracy:\n")
```

```{r}
for (cl in colnames(confusion_matrix)) {
  class_accuracy <- confusion_matrix[cl, cl] / sum(confusion_matrix[, cl])
  cat(cl, ":", round(class_accuracy, 4), "\n")
}
```




```{r}
# Extract LD scores (linear discriminant scores)
ld_scores <- lda_predictions$x

# Plot LD1 vs LD2
par(mar = c(4, 4, 2, 1))

arm_colors <- c("Placebo" = "blue", "Xanomeline Low Dose" = "red", "Xanomeline High Dose" = "green")
arm_numeric <- as.numeric(as.factor(lda_data$ARM))
arm_pch <- c(16, 17, 18)[arm_numeric]

plot(ld_scores[, 1], ld_scores[, 2],
     pch = arm_pch,
     cex = 1.5,
     col = unname(arm_colors[lda_data$ARM]),
     xlab = "LD1",
     ylab = "LD2",
     main = "LDA Score Plot by Treatment Arm")

legend("topright",
       legend = names(arm_colors),
       col = unname(arm_colors),
       pch = c(16, 17, 18),
       cex = 0.9)

# Add grid
abline(h = 0, v = 0, lty = 3, col = "gray")
```



---

## 7. Repeated Measures and Multivariate Longitudinal Data

### Theory and Foundations

Repeated measures designs collect multiple outcomes on same subjects over time. Multivariate approaches model:

- **Compound Symmetry**: All correlations equal, homogeneous variances
- **Unstructured**: Unrestricted covariance (most flexible, fewest assumptions)
- **Growth Curves**: Polynomial trends over time with individual random effects

### Clinical Example: Blood Pressure Trajectories by Treatment Arm


```{r}
# Get SYSBP measurements across visits
vs_sysbp <- vs %>%
  filter(VSTESTCD == "SYSBP") %>%
  select(USUBJID, VSSTRESN, VISITNUM) %>%
  left_join(
    dm %>% filter(!is.na(ARM), ARM != "") %>% select(USUBJID, ARM),
    by = "USUBJID"
  ) %>%
  filter(!is.na(ARM), ARM != "") %>%
  filter(complete.cases(.)) %>%
  arrange(USUBJID, VISITNUM)

# Calculate mean SYSBP by ARM and visit
sysbp_trajectory <- vs_sysbp %>%
  group_by(ARM, VISITNUM) %>%
  summarise(
    N = n(),
    Mean_SYSBP = mean(VSSTRESN),
    SD_SYSBP = sd(VSSTRESN),
    SE_SYSBP = sd(VSSTRESN) / sqrt(n()),
    .groups = "drop"
  )

cat("Blood Pressure Trajectories by Visit and ARM:\n")
```

```{r}
print(sysbp_trajectory)
```




```{r}
# Spaghetti plot: individual trajectories with mean overlay
par(mar = c(4, 4, 2, 1))

# Initialize empty plot
plot(NULL, NULL,
     xlim = c(0.5, max(vs_sysbp$VISITNUM) + 0.5),
     ylim = c(100, 160),
     xlab = "Visit Number",
     ylab = "Systolic Blood Pressure (mmHg)",
     main = "SysBP Trajectories by Treatment Arm")

# Define colors for treatment arms
arm_colors_long <- c("Placebo" = "blue", "Xanomeline Low Dose" = "red", "Xanomeline High Dose" = "green")

# Plot individual trajectories
for (arm in unique(vs_sysbp$ARM)) {
  arm_data <- vs_sysbp %>% filter(ARM == arm)

  for (subj in unique(arm_data$USUBJID)) {
    subj_data <- arm_data %>% filter(USUBJID == subj) %>% arrange(VISITNUM)
    lines(subj_data$VISITNUM, subj_data$VSSTRESN,
          col = adjustcolor(arm_colors_long[arm], alpha.f = 0.3),
          type = "l",
          lwd = 1)
  }
}

# Overlay mean trajectories (thicker lines)
for (arm in unique(sysbp_trajectory$ARM)) {
  arm_mean_data <- sysbp_trajectory %>% filter(ARM == arm)
  lines(arm_mean_data$VISITNUM, arm_mean_data$Mean_SYSBP,
        col = arm_colors_long[arm],
        type = "b",
        lwd = 3,
        pch = 19,
        cex = 1.5)
}

# Add legend
legend("topright",
       legend = names(arm_colors_long),
       col = unname(arm_colors_long),
       lwd = 2,
       cex = 0.9)
```




```{r}
# Fit polynomial growth curve model
# Model: SYSBP ~ VISITNUM + ARM + VISITNUM*ARM + VISITNUM^2 + VISITNUM^2*ARM

# Prepare data for lm()
vs_sysbp_poly <- vs_sysbp %>%
  mutate(
    Visit = VISITNUM,
    Visit_Squared = VISITNUM^2,
    Arm_LD = as.numeric(ARM == "Xanomeline Low Dose"),
    Arm_HD = as.numeric(ARM == "Xanomeline High Dose")
  )

# Fit model with error handling
growth_model <- tryCatch(
  lm(VSSTRESN ~ Visit + Arm_LD + Arm_HD + Visit:Arm_LD + Visit:Arm_HD +
     Visit_Squared + Visit_Squared:Arm_LD + Visit_Squared:Arm_HD,
    data = vs_sysbp_poly),
  error = function(e) {
    cat("Warning: Growth curve model fitting failed:", e$message, "\n")
    return(NULL)
  }
)

if (!is.null(growth_model)) {
  cat("Growth Curve Model Results:\n")
  print(summary(growth_model))

  # Extract model coefficients
  coefs <- coef(growth_model)
  cat("\nInterpretation:\n")
  cat("Intercept (Placebo at Visit 1):", round(coefs[1], 2), "\n")
  cat("Visit (linear trend for Placebo):", round(coefs[2], 4), "\n")
  if (length(coefs) >= 8) {
    cat("Visit^2 (quadratic trend for Placebo):", round(coefs[8], 4), "\n")
  }
} else {
  cat("Growth curve model not fitted due to errors.\n")
}
```




```{r}
# Generate fitted curves for visualization (only if model was fitted successfully)
if (!is.null(growth_model)) {
  visit_range <- seq(1, max(vs_sysbp$VISITNUM), length.out = 50)

  # Predict for each arm
  placebo_pred <- predict(growth_model,
                         newdata = data.frame(
                           Visit = visit_range,
                           Visit_Squared = visit_range^2,
                           Arm_LD = 0,
                           Arm_HD = 0
                         ))

  ld_pred <- predict(growth_model,
                    newdata = data.frame(
                      Visit = visit_range,
                      Visit_Squared = visit_range^2,
                      Arm_LD = 1,
                      Arm_HD = 0
                    ))

  hd_pred <- predict(growth_model,
                    newdata = data.frame(
                      Visit = visit_range,
                      Visit_Squared = visit_range^2,
                      Arm_LD = 0,
                      Arm_HD = 1
                    ))

  # Plot fitted curves
  par(mar = c(4, 4, 2, 1))

  plot(NULL, NULL,
       xlim = c(0.5, max(vs_sysbp$VISITNUM) + 0.5),
       ylim = c(100, 160),
       xlab = "Visit Number",
       ylab = "Systolic Blood Pressure (mmHg)",
       main = "Growth Curve Model: Fitted SysBP Trajectories")

  # Mean observed data
  for (arm in unique(sysbp_trajectory$ARM)) {
    arm_mean_data <- sysbp_trajectory %>% filter(ARM == arm)
    points(arm_mean_data$VISITNUM, arm_mean_data$Mean_SYSBP,
           col = arm_colors_long[arm],
           pch = 19,
           cex = 1.2)
  }

  # Fitted curves
  lines(visit_range, placebo_pred, col = arm_colors_long["Placebo"], lwd = 2.5, lty = 1)
  lines(visit_range, ld_pred, col = arm_colors_long["Xanomeline Low Dose"], lwd = 2.5, lty = 1)
  lines(visit_range, hd_pred, col = arm_colors_long["Xanomeline High Dose"], lwd = 2.5, lty = 1)

  legend("topright",
         legend = names(arm_colors_long),
         col = unname(arm_colors_long),
         lwd = 2,
         cex = 0.9)
} else {
  cat("Skipping fitted curves plot: Growth curve model not available.\n")
}
```



---

## 8. Multiple Testing in Multivariate Settings

### Background: The Multiple Comparisons Problem

When performing k statistical tests independently, the family-wise error rate (FWER) inflates:

FWER = 1 - (1 - α)^k

For 5 tests at α = 0.05: FWER = 1 - (0.95)^5 ≈ 0.226 (22.6% chance of any false positive!)

### Multiple Testing Corrections


```{r}
# Simulate p-values from a study with 20 independent tests
# Assume true null for all tests (no real effects)

n_tests <- 20
n_simulations <- 1000

# Generate random p-values under null hypothesis
pvalues_matrix <- matrix(runif(n_tests * n_simulations),
                        nrow = n_simulations,
                        ncol = n_tests)

# Apply different correction methods
uncorrected_reject <- apply(pvalues_matrix < 0.05, 1, sum)
bonferroni_reject <- apply(pvalues_matrix < 0.05/n_tests, 1, sum)
holm_rejected <- numeric(n_simulations)
fdr_rejected <- numeric(n_simulations)

for (i in 1:n_simulations) {
  # Holm-Bonferroni
  holm_corrected <- p.adjust(pvalues_matrix[i, ], method = "holm")
  holm_rejected[i] <- sum(holm_corrected < 0.05)

  # Benjamini-Hochberg (FDR)
  fdr_corrected <- p.adjust(pvalues_matrix[i, ], method = "BH")
  fdr_rejected[i] <- sum(fdr_corrected < 0.05)
}

# Summary statistics
cat("Multiple Testing Simulation Results\n")
```

```{r}
cat("(n_tests =", n_tests, ", n_simulations =", n_simulations, ")\n\n")
```

```{r}
cat("Uncorrected (α = 0.05):\n")
```

```{r}
cat("  Any rejection rate:", round(sum(uncorrected_reject > 0) / n_simulations, 4), "\n")
```

```{r}
cat("  Mean rejections per sim:", round(mean(uncorrected_reject), 2), "\n")
```

```{r}
cat("\nBonferroni (α = 0.05/k):\n")
```

```{r}
cat("  Any rejection rate:", round(sum(bonferroni_reject > 0) / n_simulations, 4), "\n")
```

```{r}
cat("  Mean rejections per sim:", round(mean(bonferroni_reject), 2), "\n")
```

```{r}
cat("\nHolm-Bonferroni:\n")
```

```{r}
cat("  Any rejection rate:", round(sum(holm_rejected > 0) / n_simulations, 4), "\n")
```

```{r}
cat("  Mean rejections per sim:", round(mean(holm_rejected), 2), "\n")
```

```{r}
cat("\nBenjamini-Hochberg (FDR = 0.05):\n")
```

```{r}
cat("  Any rejection rate:", round(sum(fdr_rejected > 0) / n_simulations, 4), "\n")
```

```{r}
cat("  Mean rejections per sim:", round(mean(fdr_rejected), 2), "\n")
```

```{r}
cat("\nInterpretation:\n")
```

```{r}
cat("Uncorrected: High false positive rate (not recommended)\n")
```

```{r}
cat("Bonferroni: Conservative, controls FWER strictly\n")
```

```{r}
cat("Holm: More powerful than Bonferroni, still controls FWER\n")
```

```{r}
cat("FDR: Controls expected proportion of false discoveries (less conservative)\n")
```




```{r}
# Real clinical example: Test if treatment differs from control on 10 variables
# Suppose 2 true effects, 8 nulls

set.seed(42)

# Simulate p-values: 2 significant, 8 null
pvalues_clinical <- c(
  0.001, 0.002,              # True positives
  runif(8, 0.1, 0.95)        # Null hypotheses
)

# Apply corrections
uncorrected_sig <- pvalues_clinical < 0.05
bonferroni_sig <- pvalues_clinical < (0.05 / 10)
holm_sig <- p.adjust(pvalues_clinical, method = "holm") < 0.05
bh_sig <- p.adjust(pvalues_clinical, method = "BH") < 0.05

# Summary table
results_summary <- data.frame(
  Test = paste0("Variable_", 1:10),
  P_value = round(pvalues_clinical, 4),
  Uncorrected = uncorrected_sig,
  Bonferroni = bonferroni_sig,
  Holm = holm_sig,
  FDR_BH = bh_sig
)

cat("Multiple Testing Correction Comparison:\n")
```

```{r}
print(results_summary)
```

```{r}
cat("\nNumber of Significant Tests (α = 0.05):\n")
```

```{r}
cat("Uncorrected:", sum(uncorrected_sig), "\n")
```

```{r}
cat("Bonferroni:", sum(bonferroni_sig), "\n")
```

```{r}
cat("Holm:", sum(holm_sig), "\n")
```

```{r}
cat("Benjamini-Hochberg (FDR):", sum(bh_sig), "\n")
```

```{r}
cat("\nRecommendation for Clinical Studies:\n")
```

```{r}
cat("- Pre-specify primary endpoints (no correction needed)\n")
```

```{r}
cat("- For secondary endpoints: Use Holm-Bonferroni or FDR control\n")
```

```{r}
cat("- Genome-wide association: Use FDR control for power\n")
```



---

## 9. Summary: Choosing the Right Multivariate Method

| **Problem** | **Method** | **Use Case** | **Key Assumption** |
|---|---|---|---|
| Reduce 10 correlated variables to 3 principal components | PCA | Exploratory data analysis, data compression | None (unsupervised) |
| Identify latent factors explaining correlations | Factor Analysis | Biomarker discovery, construct validation | Multivariate normality |
| Group subjects into homogeneous clusters | K-Means or Hierarchical | Patient stratification, phenotyping | Spherical clusters (K-means) |
| Test treatment effect on 3 correlated outcomes | MANOVA | Multiple related endpoints | Multivariate normality |
| Balance confounders in observational study | Propensity Scoring | Causal inference without RCT | Unconfoundedness, overlap |
| Classify subjects into predefined groups | LDA | Risk stratification, diagnosis prediction | Multivariate normality, equal covariances |
| Model outcome trends over multiple visits | Growth Curves / MANOVA repeated measures | Longitudinal analysis | Appropriate covariance structure |
| Control false positive rate across 20 biomarkers | Multiple testing correction (Holm, FDR) | Biomarker discovery, panel testing | Depends on method |

---

## 10. Key Takeaways

1. **Dimensionality Reduction** (PCA, Factor Analysis): Simplify high-dimensional data while preserving structure. PCA is unsupervised and variance-focused; FA explains correlations via latent factors.

2. **Clustering** (K-Means, Hierarchical): Partition subjects into similar groups. Use elbow method or silhouette analysis to choose k. Hierarchical clustering provides a dendrogram for visual exploration.

3. **Multivariate Inference** (MANOVA): Test multiple correlated endpoints simultaneously, controlling family-wise Type I error better than separate ANOVAs.

4. **Causal Inference** (Propensity Scores): Balance treatment groups on confounders in observational studies. Check overlap assumption and assess covariate balance before/after matching.

5. **Classification** (LDA): Assign new observations to predefined classes based on multiple features. Provides interpretable linear discriminant scores.

6. **Longitudinal Analysis** (Growth Curves): Model trends over time by treatment group. Use polynomial models or more complex mixed-effects approaches.

7. **Multiple Testing**: Account for inflated false positive rate when testing many hypotheses. Use Bonferroni for strict FWER control or FDR (Benjamini-Hochberg) for exploratory studies.

---

## 11. Exercises

### Exercise 1: PCA on Lab Values

Using the lb dataset from pharmaversesdtm, perform PCA on the five lab tests (ALT, AST, BILI, CREAT, K).

a) How much variance do the first two principal components explain?
b) Create a biplot showing subjects colored by ARM.
c) Interpret the loadings for PC1 and PC2. Which lab values drive each component?
d) Are there visual differences between treatment arms on the PC1-PC2 plane?

### Exercise 2: Optimal Clustering for Vital Signs

Perform both K-means and hierarchical clustering on the vital signs data (SYSBP, DIABP, PULSE, WEIGHT, HEIGHT).

a) Use the elbow method to determine the optimal number of clusters.
b) Compute silhouette scores for k = 2, 3, 4 and recommend the best k.
c) Create a contingency table showing cluster membership vs treatment ARM.
d) Are certain clusters preferentially represented in specific treatment arms?

### Exercise 3: Propensity Score Analysis for Two-Arm Comparison

Extend the propensity score analysis to include a third arm (Low Dose vs Placebo) or modify the confounders to include AGE, SEX, RACE, and baseline WEIGHT.

a) Estimate propensity scores using logistic regression.
b) Check the overlap assumption visually.
c) Perform 1:1 nearest-neighbor matching.
d) Compare standardized mean differences before and after matching for all confounders.
e) What is your conclusion about balance?

### Exercise 4: MANOVA with Lab Values

Test whether treatment ARM affects three liver function markers (ALT, AST, BILI) simultaneously using MANOVA.

a) Fit the MANOVA model and report Wilks' Lambda test statistic and p-value.
b) If MANOVA is significant, perform follow-up univariate ANOVAs with Bonferroni correction.
c) Check assumptions: multivariate normality, homogeneity of covariance matrices.
d) Interpret clinical significance of any differences found.

---

## References and Further Reading

1. **Principal Component Analysis**: Jolliffe, I. T. (2002). Principal Component Analysis (2nd ed.). Springer.

2. **Cluster Analysis**: Everitt, B. S., Landau, S., Leese, M., & Stahl, D. (2011). Cluster Analysis (5th ed.). Wiley.

3. **Propensity Scores**: Rosenbaum, P. R., & Rubin, D. B. (1983). The central role of the propensity score in observational studies. Biometrika, 70(1), 41-55.

4. **MANOVA**: Johnson, R. A., & Wichern, D. W. (2007). Applied Multivariate Statistical Analysis (6th ed.). Pearson.

5. **Clinical Trial Analysis**: Rosenberger, W. F., & Lachin, J. M. (2015). Randomization in Clinical Trials: Theory and Practice (2nd ed.). Wiley.

6. **Multiple Testing**: Benjamini, Y., & Hochberg, Y. (1995). Controlling the false discovery rate. Journal of the Royal Statistical Society: Series B, 57(1), 289-300.

---

**End of Chapter BS9**

---

## Session Information


```{r}
cat("R Version:", as.character(getRversion()), "\n")
```

```{r}
cat("Session Date:", format(Sys.Date(), "%Y-%m-%d"), "\n")
```

```{r}
cat("\nLoaded Packages:\n")
```

```{r}
cat("  - dplyr:", as.character(packageVersion("dplyr")), "\n")
```

```{r}
cat("  - tidyr:", as.character(packageVersion("tidyr")), "\n")
```

```{r}
cat("  - ggplot2:", as.character(packageVersion("ggplot2")), "\n")
```

```{r}
cat("  - MASS:", as.character(packageVersion("MASS")), "\n")
```

```{r}
cat("  - pharmaversesdtm:", as.character(packageVersion("pharmaversesdtm")), "\n")
```


